### 2021-11-18 manuscript/main.tex
Added manuscript

\documentclass{article}
\usepackage{pdfpages}    % To import the PDF document
\usepackage{xcolor}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTBLUELINK}{HTML}{3366BB}
\definecolor{PURPLELINK}{HTML}{663366}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false ]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
linkcolor=DARKBLUELINK,
anchorcolor=DARKBLUELINK,
citecolor=DARKBLUELINK,
filecolor=DARKBLUELINK,
menucolor=DARKBLUELINK,
urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\definecolor{RED}{HTML}{EB6231}
\definecolor{YELLOW}{HTML}{E29D26}
\definecolor{BLUE}{HTML}{5D80B4}
\definecolor{LIGHTGREEN}{HTML}{6ABD9B}
\definecolor{GREEN}{HTML}{8FB03E}
\definecolor{PURPLE}{HTML}{BE1E2D}
\definecolor{BROWN}{HTML}{A97C50}
\definecolor{PINK}{HTML}{DA1C5C}

\usepackage[margin=60pt]{geometry}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{blkarray}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{verbatim}
\usepackage{float}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\lstset{%
	backgroundcolor=\color{gray!25},
	basicstyle=\ttfamily,
	breaklines=true,
	columns=fullflexible
}
\usepackage[export]{adjustbox}
\usepackage{tabu}
\tabulinesep=0.6mm
\newcommand\cellwidth{\TX@col@width}
\usepackage{hhline}
\setlength{\arrayrulewidth}{1.2pt}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\AtBeginEnvironment{tabu}{\footnotesize}
\usepackage{booktabs}

\usepackage{graphicx}
\graphicspath{{artworks/}}
\makeatletter
\def\input@path{{artworks/}}
\makeatother
\pdfstringdefDisableCommands{%
\renewcommand*{\bm}[1]{#1}%
% any other necessary redefinitions
}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{natbib}
\pdfinclusioncopyfonts=1

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\angstrom}{\text{\normalfont\AA}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Pfix}{2 \Ne \proba_{\text{fix}}}

\newcommand{\VarPhenotype}{V_{P}}
\newcommand{\VarGenetic}{V_{G}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{M}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}
\newcommand{\MutationRate}{\mu}
\newcommand{\NbrLoci}{L}
\newcommand{\Trait}{X}

\begin{document}

\part*{Evolutionary quantitative genomics}
\tableofcontents

\section{Quantitative genetics theory}

In this section, we seek to derive phenotypic and genetic variance as a function of mutation, selection and drift.
First, the genetic variance ($\VarGenetic$) is related to the observed phenotypic variance ($\VarPhenotype$) through heredity:
\begin{gather}
    \VarGenetic =  h^2 \VarPhenotype \label{eq-heredity}.
\end{gather}

\subsection{Mutation-drift equilibrium}

Mutation supply new variants in the population, while genetic random drift depletes standing variation in the population. At equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the genetic variance is thus a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$):
\begin{gather}
    \VarGenetic =  2 \Ne \VarMutation \label{eq-var-genetic}.
\end{gather}

The mutational variance ($\VarMutation$) is the rate at which new mutations contributes to new mutations. As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ depends on the mutation rate per loci per generation ($\MutationRate$), the number of loci encoding the trait ($\NbrLoci$) and finally the random effect on the phenotype of a mutation ($a$):
\begin{gather}
    \VarMutation =  2 \NbrLoci \MutationRate \E \left[ a^2 \right] \label{eq-var-mutation}.
\end{gather}

\subsection{Mutation-selection-drift equilibrium}

Selection, whether positive or negative, depletes standing variation in the population and reduces to genetic variance. Hence, at equilibrium between mutation, selection and drift, the observed genetic variance is lower than the theoretical neutral variance:
\begin{align}
    \VarGenetic & < 2 \Ne \VarMutation \\
    \iff \VarGenetic & < 4 \Ne \NbrLoci \MutationRate \E \left[ a^2 \right]
\end{align}

Under the House of cards model of stabilizing selection~\citep{kingman_simple_1978}, the trait's fitness is normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$. Moreover, the trait is encoded by a few loci with large effects and the genetic variance at equilibrium between mutation, selection and drift is: 
\begin{gather}
    \VarGenetic = 2 \NbrLoci \MutationRate \VarSelection_S \label{eq-mutsel-HC}.
\end{gather}

Alternatively, under the Gaussian model of stabilizing selection~\citep{lande_natural_1976}, the trait's fitness is also normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$. However, the trait is encoded by a large number of loci traits with small effect and the genetic variance at equilibrium is: 
\begin{gather}
    \VarGenetic = \sqrt{2 \NbrLoci \VarMutation \VarSelection_S} \label{eq-mutsel-G}.
\end{gather}

The boundary between the two models is given by \citet{turelli_heritable_1984} as:
\begin{align}
    20 \MutationRate \VarSelection_S &< \E \left[ a^2 \right] \text{ for the House of card model}, \\
    20 \MutationRate \VarSelection_S &> \E \left[ a^2 \right] \text{ for the geometric model}.
\end{align}

\section{Quantitative traits}

By definition, quantitative trait is an observable can be measured for an organisms, for example genome size~\citep{alfsnes_genome_2017}.

\subsection{Life history traits}

Life history traits are a subset of quantitative traits related to the age and stage-specific patterns and timing of events that make up an organism's life such as such as age at maturity, fecundity, life expectancy, height, weight, and so on. Life history traits are often suspected to be under selection and encoded by many loci.

\subsection{Genetic architecture of quantitative traits}
Due to recent advances in Genome Wide Association Studies (GWAS), the availability of  datasets in human (\textit{e.g.} UK Biobank) allows to dissect the genetic underpinning of quantitative traits such as Body Mass Index (BMI) or disease related susceptibility.
For example by relating observable summary statistics (phenotypic distribution, allelic frequencies, variance per site) to parameters of the genetic architecture (pleiotropy, distribution of selection coefficients, effect size), as in \citet{simons_population_2018}.
Overall, as reviewed from GWAS evidences in \citet{sella_thinking_2019}, most quantitative traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection in humans.

\subsection{Gene expression}

In transcriptomic dataset, the expression level of each gene can be seen a quantitative trait.
First, $\VarGenetic$ can be estimated from the standing variation inside natural populations. 
Second, $\VarMutation$ can be  estimated from mutation accumulation experiments in laboratory clonal lines.
Third, $\NbrLoci$ can be  estimated from effect of gene knockout of expression level.
Forth, $\MutationRate$ can be  estimated from the frequency of mutation altering expression level.
Fifth, $\Ne$ can be estimated from standing genetic polymorphism ($\pi = 4 \Ne u$), by correcting for the mutation rate ($u$).
Altogether, the evolutionary regime (neutral or selection) of gene expression can be tested across all genes, as suggested in \citet{fay_evaluating_2008}.

In the yeast \textit{Saccharomyces cerevisiae}, the fruit fly \textit{Drosophila melanogaster} and the nematode \textit{Caenorhabditis elegans},  \citet{hodgins-davis_gene_2015} found that gene expression is under stabilizing selection, such that neutral evolution is ruled out for most genes.
Moreover, the House of card model is preferred against the Gaussian model of stabilizing selection, also suggesting that gene expression is encoded by a few loci with large effects.

Finally, mutation accumulation experiments with different population size have shown that $\VarMutation$ decreases with $\Ne$, suggesting that gene expression is under strong stabilizing selection~\citep{deiss_global_2021}.

\section{Phylogenetic trait evolution}

\subsection{Relationship between micro and macro evolution. }

We denote $\Trait_i$ the quantitative trait $\Trait$ for species $i$.
\citet{hansen_translating_1996} derived a relation between trait variation between species and the genetic variance of this trait as 
\begin{gather}
    \Cov \left( \Trait_i, \Trait_j \right) = \Cov \left( \E \left[ \Trait_i | \Trait_z \right ], \E \left[ \Trait_j | \Trait_z \right ] \right),  \label{eq-cov-traits}
\end{gather}
where $z$ is the ancestor between species $i$ an $j$.

If the trait is neutral, evolving as a Brownian process, then $\E \left[ \Trait_i | \Trait_z \right ] = \Trait_z$ and equation \ref{eq-cov-traits} simplifies to:
\begin{align}
\Cov \left( \Trait_i, \Trait_j \right) &= \Var \left( \Trait_z \right),\\
    &= \frac{\VarGenetic}{\Ne} t_{z}, \\
    &=  2 t_{z} \VarMutation, \label{eq-cov-traits-neutral} \text{ from equation } \ref{eq-var-genetic},
\end{align}
where $t_{z}$ is the time between species $z$ an both $i$ an $j$.

The relationship between trait variation between species and the mutational variance can be used as method to detect whether traits are evolving under a neutral model of evolution, similar to the use of synonymous ($\ds$) and non-synonymous ($\dn$) substitution rates in molecular evolution, as noted in \citet{fay_evaluating_2008}.

\begin{table}[H]
	\centering
	\noindent\adjustbox{max width=\textwidth}{%
		\begin{tabu}{|c||c|c|}
			\hline
			 & Coding sequences & Quantitative traits \\ \hline \hline
			Adaptation or moving optimum & $\dnds > 1$ & $\Cov \left( \Trait_i, \Trait_j \right) > 2 t_{z} \VarMutation $ \\ \hline
			Neutral regime of evolution & $\dnds = 1$ & $\Cov \left( \Trait_i, \Trait_j \right) = 2 t_{z} \VarMutation $ \\ \hline
			Purifying or directional selection & $\dnds < 1$ & $\Cov \left( \Trait_i, \Trait_j \right) < 2 t_{z} \VarMutation $ \\ \hline
		\end{tabu}}
\end{table}

If the trait is under directional selection, but the strength and direction of selection varies randomly from one generation to the next (variance $\VarSelection_D$), then the trait still follows a Brownian process model and equation \ref{eq-cov-traits} reduces to:
\begin{align}
\Cov \left( \Trait_i, \Trait_j \right) &= \Var \left( \Trait_z \right), \\
    &= \left(\frac{\VarGenetic}{\Ne} + \VarGenetic \VarSelection_D \VarGenetic \right)  t_{z}. \label{eq-selection-vary-direct}
\end{align}

If the trait is under stabilizing selection, but the strength and direction of the optimum varies randomly from one generation to the next (variance $\VarSelection_{\theta}$), then the trait still follows a Brownian process model and equation \ref{eq-cov-traits} reduces to:
\begin{align}
\Cov \left( \Trait_i, \Trait_j \right) &= \VarSelection_{\theta} t_{z}. \label{eq-selection-vary-stabi}
\end{align}

Altogether, traits that are found to be under a long term Brownian process are not necessarily neutral.

To read : \citet{arnold_adaptive_2001}
\subsection{Trait evolution under a Brownian or Ornstein-Uhlenbeck process}

Modelling $\Trait$ as a Brownian process allows theoretically to test for neutrality.
In \citet{catalan_drift_2019}, gene expression is modelled either as a Brownian (supposedly neutral regime), or as a Brownian with a shift in the mean for one branch (supposedly directional selection), or as an Ornstein-Uhlenbeck process (supposedly stabilizing selection) or finally as an Ornstein-Uhlenbeck process with a shift in the optimal for one branch (supposedly directional selection). 
The analysis on the transcriptome across different individuals and species concluded that 81\% of genes are under a neutral regime and 16\% under directional selection. 
However, the alternative model of a varying directional (equation~ \ref{eq-selection-vary-direct}) or stabilizing selection (equation~ \ref{eq-selection-vary-stabi}) was not implemented, and thus can not be ruled out by this analysis.

\subsection{Variance evolution}

The multidimensional generalization of $\VarGenetic$ to several traits is the of the genetic covariance matrix $\MatrixGenetic$.
Stability of $\MatrixGenetic$, can be reformulated as whether $\MatrixGenetic (t) = \MatrixGenetic$, an assumption discussed in \citet{arnold_understanding_2008} and \citet{hohenlohe_mipod_2008}. Having a time dependent matrix is implemented in JIVE, a Bayesian hierarchical model developed by \citet{kostikova_bridging_2016} and implemented in BEAST by \citet{gaboriau_multi-platform_2020}.

\section{Discussion \& perspectives}
At the population level, gene expression is shown to be empirically under stabilizing selection. However at the phylogenetic scale, gene expression is under a neutral regime. Can we reconcile these observations?

We can find the genes under adaptive regime (fluctuating optimal gene expression) at the phylogenetic scale and compare that to the genes under adaptation using $\dnds$ based methods. Do we find the same genes?
\section{Appendix}

\subsection{Definitions}

We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $i$.

\begin{align}
    \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
\end{align}

\bibliographystyle{natbib}%%%%Bibliography style file
\bibliography{references}

\end{document}

### 2021-11-18 copy_artworks.sh

@@ -1,0 +1,5 @@
+#!/usr/bin/env bash
+\cp -rf ./Contrasts/Analysis/* ./manuscript/artworks/
+\cp -f ./Contrasts/scatterplot_gene.pdf ./manuscript/artworks/
+\cp -f ./Contrasts/scatterplot_site.pdf ./manuscript/artworks/
+\cp -f ./Contrasts/ontology/*.core.tex ./manuscript/artworks/

### 2021-11-18 data_simulated/simulated_experiment.py

@@ -6,7 +6,7 @@ from subprocess import run
 
 def create_experiment(config, screen, sbatch, nbr_cpu):
     experiment = config.replace(".yaml", "")
-    exp_path = os.getcwd() + '/Experiments/' + experiment
+    exp_path = os.getcwd() + '/experiments/' + experiment
 
     os.makedirs(exp_path, exist_ok=True)
     os.system('cp {0} {1}/config.yaml'.format(config, exp_path))

### 2021-11-18 scripts/plot_scatter.py

@@ -37,16 +37,14 @@ def get_node_names(root, target1, target2):
     return nodes
 
 
-def scatter_plot(x, y, x_label, y_label, output, scale, nbr_genes, title="", histy_log=False, loglog=False):
+def scatter_plot(x, y, x_label, y_label, output, nbr_genes, title="", histy_log=False, loglog=False):
     color_models = colors(x)
     fig = plt.figure(figsize=(1920 / my_dpi, 880 / my_dpi), dpi=my_dpi)
     set_nb_points = set([len(nb) for nb in x.values()])
     assert len(set_nb_points) == 1
     nb_points = set_nb_points.pop()
-    if scale == "pairs":
-        fig.suptitle("{0} genes for each pairs of species ({1} points)".format(nbr_genes, nb_points) + title)
-    elif scale == "species":
-        fig.suptitle("{0} genes for each species ({1} points)".format(nbr_genes, nb_points) + title)
+    assert nb_points == nbr_genes
+    fig.suptitle("{0} genes".format(nbr_genes) + title)
 
     gs = fig.add_gridspec(2, 2, width_ratios=(7, 2), height_ratios=(2, 7),
                           left=0.1, right=0.9, bottom=0.1, top=0.9,
@@ -73,8 +71,8 @@ def scatter_plot(x, y, x_label, y_label, output, scale, nbr_genes, title="", his
         reg = '{0} - slope of {1:.2g} ($r^2$={2:.2g})'.format(m.replace("_", " ").capitalize(), a, results.rsquared)
         ax.plot(idf, linear, '-', linestyle="--", label=reg, color=color_models[id_m])
 
-    bins_x = np.geomspace(max(1e-6, min_x), max_x, 150) if loglog else "auto"
-    bins_y = np.geomspace(max(1e-6, min_y), max_y, 150) if loglog else "auto"
+    bins_x = np.geomspace(max(1e-6, min_x), max_x, 100) if loglog else 100
+    bins_y = np.geomspace(max(1e-6, min_y), max_y, 100) if loglog else 100
     ax_histx.hist(x.values(), bins=bins_x, color=color_models, **hist_filled)
     ax_histx.hist(x.values(), bins=bins_x, color=color_models, **hist_step)
[…]

### 2021-11-18 scripts/plot_trajectory.py

@@ -7,7 +7,7 @@ import argparse
 
 mpl.use('Agg')
 
-my_dpi = 128
+my_dpi = 256
 fontsize = 14
 fontsize_legend = 12
 
@@ -26,36 +26,29 @@ if __name__ == '__main__':
     args = parser.parse_args()
 
     groups = [v for k, v in pd.read_csv(args.input, sep="\t").groupby("Lineage")]
-    fig, ((ax1, ax2, ax3, ax4), (ax5, ax6, ax7, ax8)) = plt.subplots(2, 4, sharex='all', linewidth=0.5,
-                                                                     figsize=(1920 / my_dpi, 880 / my_dpi), dpi=my_dpi)
-    fig.suptitle(basename(args.input).replace(".tsv", '').replace("_", ' '))
-
+    fig, ax1 = plt.subplots(1, 1, sharex='all', linewidth=0.5, figsize=(1920 / my_dpi, 1440 / my_dpi), dpi=my_dpi)
     for ddf in groups:
         plot_trajectory(ax1, ddf["Generation"], ddf["Phenotype mean"])
-        plot_trajectory(ax2, ddf["Generation"], ddf["Population size"])
-        plot_trajectory(ax3, ddf["Generation"], ddf["Fitness mean"])
-        plot_trajectory(ax4, ddf["Generation"], ddf["Mutational var"], True)
-        plot_trajectory(ax5, ddf["Generation"], ddf["Genotype var"], True)
-        plot_trajectory(ax6, ddf["Generation"], ddf["Phenotype var"], True)
-        plot_trajectory(ax7, ddf["Generation"], ddf["Heritability"], True)
-        plot_trajectory(ax8, ddf["Generation"], ddf["Genotype var"] / ddf["Phenotype var"], True)
-
     ax1.set_ylabel("Mean phenotype ($\\bar{X}$)", fontsize=fontsize)
-    ax2.set_ylabel("Population size ($N_e$)", fontsize=fontsize)
-    ax3.set_ylabel("Fitness mean ($\\bar{w}$)", fontsize=fontsize)
-    ax4.set_ylabel("Mutational variance ($V_M$)", fontsize=fontsize)
-    ax5.set_ylabel("Genotype variance ($V_G$)", fontsize=fontsize)
-    ax6.set_ylabel("Phenotype variance ($V_P$)", fontsize=fontsize)
[…]

### 2023-01-13 manuscript/main.tex
Draft of the manuscript for JEB.

@@
+%! BibTeX Compiler = biber
+%TC:ignore
 \documentclass{article}
-\usepackage{pdfpages}    % To import the PDF document
-\usepackage{xcolor}
+
+\usepackage{xcolor, colortbl}
 \definecolor{BLUELINK}{HTML}{0645AD}
 \definecolor{DARKBLUELINK}{HTML}{0B0080}
-\definecolor{LIGHTBLUELINK}{HTML}{3366BB}
-\definecolor{PURPLELINK}{HTML}{663366}
 \PassOptionsToPackage{hyphens}{url}
-\usepackage[colorlinks=false ]{hyperref}
+\usepackage[colorlinks=false]{hyperref}
 % for linking between references, figures, TOC, etc in the pdf document
@@
 
-\definecolor{RED}{HTML}{EB6231}
-\definecolor{YELLOW}{HTML}{E29D26}
-\definecolor{BLUE}{HTML}{5D80B4}
-\definecolor{LIGHTGREEN}{HTML}{6ABD9B}
-\definecolor{GREEN}{HTML}{8FB03E}
-\definecolor{PURPLE}{HTML}{BE1E2D}
-\definecolor{BROWN}{HTML}{A97C50}
-\definecolor{PINK}{HTML}{DA1C5C}
+\usepackage{biorxiv}
+\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
+\addbibresource{references.bib}
 
-\usepackage[margin=60pt]{geometry}
+\usepackage{url}
 \usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
-\usepackage{blkarray}
 \usepackage{lmodern}
-\usepackage{bm,bbold}
-\usepackage{verbatim}
-\usepackage{float}
+\usepackage{xfrac, nicefrac}
 \usepackage{bm}
 \usepackage{listings, enumerate, enumitem}
-\lstset{%
-	backgroundcolor=\color{gray!25},
-	basicstyle=\ttfamily,
-	breaklines=true,
-	columns=fullflexible
+\usepackage[export]{adjustbox}
+\usepackage{graphicx}
+\usepackage{bbold}
+\usepackage{pdfpages}
+\pdfinclusioncopyfonts=1
+\usepackage{lineno}
+\renewcommand{\baselinestretch}{1.25}
+\renewcommand{\arraystretch}{1.6}
+
+\input{notations.tex}
+
+\title{A neutrality index for quantitative traits}
+
+\author{
+\large
+\textbf{T. {Latrille}$^{1}$, T. {Gaboriau}$^{1}$, N. {Salamin}$^{1}$}\\
+\normalsize
+$^{1}$Université de Lausanne, Lausanne, Switzerland\\
+\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
 }
-\usepackage[export]{adjustbox}
-\usepackage{tabu}
-\tabulinesep=0.6mm
-\newcommand\cellwidth{\TX@col@width}
-\usepackage{hhline}
-\setlength{\arrayrulewidth}{1.2pt}
-\usepackage{multicol,multirow,array}
-\usepackage{etoolbox}
-\AtBeginEnvironment{tabu}{\footnotesize}
-\usepackage{booktabs}
-
-\usepackage{graphicx}
-\graphicspath{{artworks/}}
-\makeatletter
-\def\input@path{{artworks/}}
-\makeatother
-\pdfstringdefDisableCommands{%
-\renewcommand*{\bm}[1]{#1}%
-% any other necessary redefinitions
-}
-\usepackage{xfrac, nicefrac}
-\usepackage{lmodern}
-\usepackage{natbib}
-\pdfinclusioncopyfonts=1
-
-\newcommand{\UniDimArray}[1]{\bm{#1}}
-\newcommand{\BiDimArray}[1]{\bm{#1}}
-\DeclareMathOperator{\E}{\mathbb{E}}
-\DeclareMathOperator{\Var}{\text{Var}}
-\DeclareMathOperator{\Cov}{\text{Cov}}
-\newcommand{\der}{\mathrm{d}}
-\newcommand{\angstrom}{\text{\normalfont\AA}}
-\newcommand{\e}{\text{e}}
-\newcommand{\Ne}{N_{\text{e}}}
-\newcommand{\dn}{d_N}
-\newcommand{\ds}{d_S}
-\newcommand{\dnds}{\dn / \ds}
-\newcommand{\pn}{\pi_N}
-\newcommand{\ps}{\pi_S}
-\newcommand{\pnps}{\pn / \ps}
-\newcommand{\proba}{\mathbb{P}}
-\newcommand{\pfix}{\proba_{\text{fix}}}
-\newcommand{\Pfix}{2 \Ne \proba_{\text{fix}}}
-
-\newcommand{\VarPhenotype}{V_{P}}
-\newcommand{\VarGenetic}{V_{G}}
-\newcommand{\MatrixGenetic}{\BiDimArray{G}}
-\newcommand{\VarMutation}{V_{M}}
-\newcommand{\StdSelection}{\sigma}
-\newcommand{\VarSelection}{\StdSelection^2}
-\newcommand{\MutationRate}{\mu}
-\newcommand{\NbrLoci}{L}
-\newcommand{\Trait}{X}
 
@@
 
-\part*{Evolutionary quantitative genomics}
-\tableofcontents
+\maketitle
 
-\section{Quantitative genetics theory}
+% Abstract (≤ 250 words)
+%TC:endignore
+\begin{abstract}
+    Current methods at the population scale do not integrate the changes over long evolutionary time, while at the same time the methods require laboratory clonal lines to determine the neutral expectation.
+    Conversely, at the phylogenetic scale, current methods discard the variance in traits between individuals, leaving out valuable information.
+    The main goal of this project is to combine within-species and between-species variation of quantitative traits.
+    To this aim, we define a neutrality index for of a quantitative trait, devoted to determine its evolutionary regime, leveraging both data at the population and phylogenetic scale.
+\end{abstract}
 
-In this section, we seek to derive phenotypic and genetic variance as a function of mutation, selection and drift.
-First, the genetic variance ($\VarGenetic$) is related to the observed phenotypic variance ($\VarPhenotype$) through heredity:
+\keywords{Quantitative genetics \and Phylogenetics \and Population-genetics \and Selection }
+
+% Research Article (7500 words)
+\section{Introduction}\label{sec:introduction}
+A quantitative trait is by definition an observable can be measured for an organisms, for example genome size\cite{alfsnes_genome_2017}.
+Life history traits (LHTs) are such quantitative traits, related to the age and stage-specific patterns and timing of events that make up an organism's life such as age at maturity, fecundity, life expectancy, height, weight, and so on.
+LHTs are often suspected to be under selection and encoded by many loci.
+Appart from LHTs, the genetic underpinning of other quantitative traits such as Body Mass Index (BMI) or disease related susceptibility can now be dissected due to recent advances in Genome Wide Association Studies (GWAS) and the availability of datasets in human (\textit{e.g.} UK Biobank)\cite{simons_population_2018}.
+Overall, GWAS have revealed that traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection in humans\cite{sella_thinking_2019}.
+At the genomic level, the expression level of each gene can also be seen as a quantitative trait.
+Surveys of gene expression reveal extensive variability both within and between a wide range of species\cite{fay_evaluating_2008}.
+
+From the standing variation both within and between a wide range of species, distinguishing adaptive changes driven by positive selection from neutral evolution is critical for understanding the evolution of trait evolution.
+At the population level, the variance between individuals of the same population can be estimated for any quantitative trait.
+This observed variance can be compared to the expected variance that would be obtained under the assumption that the trait is not under selection, which can be estimated from mutation accumulation experiments in laboratory clonal lines.
+At the phylogenetic scale, between different populations, the evolutionary regime for a quantitative trait can also be tested.
+Typically, the trait mean (across individuals of the same population) is modelled as a Brownian process, allowing theoretically to test for neutrality.
+Alternatively, a Brownian with a shift in the mean for one branch is supposedly directional selection, an Ornstein-Uhlenbeck process is supposedly stabilizing selection or finally an Ornstein-Uhlenbeck process with a shift in the optimal for one branch is supposedly directional selection\cite{catalan_drift_2019}.
+
+At the frontier between phylogenetics and population genetics, studies comparing between species variation have acknowledged the importance of within species variation\cite{whitehead_variation_2006}.
+However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is a promising line of research\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Altogether, there is a push toward integrating within and between species variation into a unified framework.
+This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
+However, it has not yet been generalized for any number of species related by their phylogenetic tree.
+Finally, studies on quantitative trait evolution
+In this study, we propose a neutrality index for a quantitative trait, leveraging data within and between species, normalizing the trait evolution by its genetic polymorphism and divergence.
+
+\section{Materials and Methods}\label{sec:materials-and-methods}
+\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
+\subsubsection{Mutation-drift equilibrium for a trait}
+
+For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect on the random of a mutation ($a$).
+New mutations are thus generating variance for such a trait.
+We define the expected effect of a mutation on the variance of a trait as $\RateMut$, given as:
 \begin{gather}
-    \VarGenetic =  h^2 \VarPhenotype \label{eq-heredity}.
+    \RateMut \defEqual \NbrLoci \Multiply \E \left[ a^2 \right]. \label{eq:rate-mut}
 \end{gather}
 
-\subsection{Mutation-drift equilibrium}
-
-Mutation supply new variants in the population, while genetic random drift depletes standing variation in the population.
-At equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the genetic variance is thus a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$):
+At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contributes to the trait variance per generation.
+As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of $\RateMut$ and the mutation rate per loci per generation ($\MutationRate$), as:
 \begin{gather}
-    \VarGenetic =  2 \Ne \VarMutation \label{eq-var-genetic}.
+    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
 \end{gather}
 
-The mutational variance ($\VarMutation$) is the rate at which new mutations contributes to new mutations.
-As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ depends on the mutation rate per loci per generation ($\MutationRate$), the number of loci encoding the trait ($\NbrLoci$) and finally the random effect on the phenotype of a mutation ($a$):
+At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
+As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in the population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$) as:
+\begin{align}
+    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
+    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
+\end{align}
+
+Moreover, for any neutral genomic region of interest, the genetic diversity, called $\pi$, is measured as the number of mutation segregating in the population divided by the size of the region.
+Any of the segregating mutation will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
+As shown in \textcite{tajima_statistical_1989}, $\pi$ is thus a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$), given as:
 \begin{gather}
-    \VarMutation =  2 \NbrLoci \MutationRate \E \left[ a^2 \right] \label{eq-var-mutation}.
+    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
 \end{gather}
 
-\subsection{Mutation-selection-drift equilibrium}
-
-Selection, whether positive or negative, depletes standing variation in the population and reduces to genetic variance.
-Hence, at equilibrium between mutation, selection and drift, the observed genetic variance is lower than the theoretical neutral variance:
+Altogether, at the population level, we define define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$), which identify with $\RateMut$:
 \begin{align}
-    \VarGenetic & < 2 \Ne \VarMutation \\
-    \iff \VarGenetic & < 4 \Ne \NbrLoci \MutationRate \E \left[ a^2 \right]
+    \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
+    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and \ref{eq:genetic-diversity}}, \\
+    & = \RateMut. \label{eq:rate-pop}
 \end{align}
 
-Under the House of cards model of stabilizing selection~\citep{kingman_simple_1978}, the trait's fitness is normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$.
-Moreover, the trait is encoded by a few loci with large effects and the genetic variance at equilibrium between mutation, selection and drift is:
+Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($h^2$), giving also $\RatePop$ as:
+\begin{align}
+    \RatePop & = \frac{h^2 \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
+\end{align}
+
+\subsubsection{Phylogenetic evolution of a trait}
+
+Along a lineage, we denote $\Trait$ as the mean value of the trait over the individuals in the population.
+If the trait is neutral and encoded by many loci, $\Trait$ evolves as Brownian process\cite{hansen_translating_1996}.
+After the population has evolved for $\Time$ generations, the variance of $\Trait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
+\begin{align}
+    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
+    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+\end{align}
+
+Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated in the populations, some will we eventually reach fixation in the population due to genetic drift, resulting in a substitution.
+For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
+As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
+\begin{align}
+    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
+    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
+    & = \MutationRate. \label{eq:substitution-rate}
+\end{align}
+That is, if mutations are neutral, the rate with which a new neutral allele is fixed in the population in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.
+
+Thus, after having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
+\begin{align}
+    d & = \Time \Multiply \SubRate \\
+    & = \Time \Multiply \MutationRate \text{ from eq.\ref{eq:substitution-rate}}. \label{eq:distance}
+\end{align}
+Altogether, the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$) also identify with $\RateMut$:
+\begin{align}
+    \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
+    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and \ref{eq:distance}}, \\
+    & = \RateMut \label{eq:rate-phy}.
+\end{align}
+
+\begin{figure*}[!ht]
+    \centering
+    \includegraphics[width=\textwidth, page=1] {artworks/fig-summary}
+    \caption{
+        In red, at the phylogenetic scale, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
+        In blue, at the population scale, trait variance within species ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
+        Under neutral evolution $\RatePhy$ should equal $\RatePop$.
+        Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
+    }
+    \label{fig:methods}
+\end{figure*}
+
+\subsubsection{Neutrality index}
+
+ The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
+ At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
+Combined as a ratio, we obtain $\RatePop$ at the population level, which is a proxy for $\RateMut$, namely the expected effect of a mutation on the variance of a trait.
+At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allow to compute $d$.
+Similarly, combined as a ratio, we obtain $\RatePhy$ which is another proxy for $\RateMut$.
+Altogether, we thus have:
+\begin{align}
+    \RateMut & = \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and \ref{eq:rate-phy}} , \\
+    \Rightarrow \frac{\RatePhy}{\RatePop} & = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
+\end{align}
+Altogether, $\RatePhy / \RatePop$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
+Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between populations, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.
+
+\subsection{Estimates}
+
+\subsubsection{Maximum likelihood estimate}
+For each population of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences thus normalizing .
+Averaged across all populations we obtain the mean as:
 \begin{gather}
-    \VarGenetic = 2 \NbrLoci \MutationRate \VarSelection_S \label{eq-mutsel-HC}.
+    \EstRatePop = \left< \frac{\VarGenetic}{\pi} \right> = \left< \frac{h^2 \Multiply \VarPhenotype }{\pi } \right> \label{eq:estimated-rate-pop}
 \end{gather}
 
-Alternatively, under the Gaussian model of stabilizing selection~\citep{lande_natural_1976}, the trait's fitness is also normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$.
-However, the trait is encoded by a large number of loci traits with small effect and the genetic variance at equilibrium is:
+At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
+The contrast matrix is symmetric, along the diagonal are the total distances ($d_{i}$) from the root of the tree to each taxon ($i$).
+The off-diagonal elements $C_{\Spi,\Spj}$ are the total branch lengths shared by particular pairs of taxa $\Spi$ and $\Spj$, computed from the distance between the pair ($d_{\Spi,\Spj}$) as:
 \begin{gather}
-    \VarGenetic = \sqrt{2 \NbrLoci \VarMutation \VarSelection_S} \label{eq-mutsel-G}.
+    C_{\Spi,\Spj} = d_{\mathcal{A}(\Spi, \Spj)} = \frac{d_{\Spi} + d_{\Spj} - d_{\Spi,\Spj}}{2}
 \end{gather}
+Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrSpecies \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
+\begin{gather}
+    \RootTrait = \left( \VecOne\tr \Cr\inv \VecOne \right)\inv \left( \VecOne\tr \Cr\inv \VecTrait \right), \label{eq:estimated-root-trait}
+\end{gather}
+where $\VecOne$ is an $\NbrSpecies \times 1$ column vector of ones.
 
-The boundary between the two models is given by \citet{turelli_heritable_1984} as:
-\begin{align}
-    20 \MutationRate \VarSelection_S &< \E \left[ a^2 \right] \text{ for the House of card model}, \\
-    20 \MutationRate \VarSelection_S &> \E \left[ a^2 \right] \text{ for the geometric model}.
-\end{align}
+Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
+\begin{gather}
+    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \VecOne \right)\tr \Cr\inv \left( \VecTrait -  \RootTrait  \VecOne  \right)}{\NbrSpecies - 1}. \label{eq:estimated-rate-phy}
+\end{gather}
+Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
 
-\section{Quantitative traits}
+\subsection{Simulations}\label{subsec:simulations}
+The evolutionary dynamics was formalized as a Wright-Fisher model with mutation, selection and drift.
+The population is assumed to be panmictic, with {effective population size} $\Ne$ and with non-overlapping generations.
+We simulate explicitly each generation along the phylogeny under a Wright-Fisher population, consisting of three steps: mutation, selection and {genetic drift} of currently segregating alleles.
+Mutations are drawn based on a user-defined nucleotide matrix, where our simulations used a symmetric time-reversible mutation matrix.
+Drift is induced by the multinomial resampling of the currently segregating alleles.
+Our simulator use a geometric Brownian multivariate process to model the changes in the mutation rate per generation, the generation time and $\Ne$ along the lineages.
+Moreover, the instant value of log-$\Ne$ can also be modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
+The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
+In our simulations, the tree is composed of 77 species (see supplementary), the tree root is $150$ million years old, the initial mutation rate is $\smash{10^{-8}}$ per site per generation and the initial generation time is $10$ years.
+The simulator starts from an initial sequence at equilibrium, at each node, the process is split, and finally stopped at the leaves of the tree.
 
-By definition, quantitative trait is an observable can be measured for an organisms, for example genome size~\citep{alfsnes_genome_2017}.
+We introduced selection as a one dimensional Fisher's geometric landscape~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+The phenotype is a vector in a multidimensional space, where the number of dimensions is often termed complexity.
+From a {phenotype}, the fitness is a monotonously decreasing function of the {phenotype} distance to $0$.
+The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength ($\alpha$) and epistasis ($\beta$).
+If the phenotype-fitness map is explicit, the genotype-phenotype map is more pervasive.
+Mutations are seen has displacement of the {phenotype} in the multidimensional space.
+Beneficial mutations are moving the {phenotype} closer to $0$, whereas deleterious mutations are moving the {phenotype} further away.
+In such original context, the distribution of mutational effects is not dependent on the current genotype, but this can be relaxed using a genotype-phenotype map.
 
-\subsection{Life history traits}
+\begin{figure*}[!ht]
+    \centering
+    \includegraphics[width=\textwidth, page=1] {artworks/fig-simulator}
+    \caption{
+        Wright-Fisher simulations with mutation, selection and drift.
+    }
+    \label{fig:simulator}
+\end{figure*}
 
-Life history traits are a subset of quantitative traits related to the age and stage-specific patterns and timing of events that make up an organism's life such as such as age at maturity, fecundity, life expectancy, height, weight, and so on.
-Life history traits are often suspected to be under selection and encoded by many loci.
+\section{Results}
 
-\subsection{Genetic architecture of quantitative traits}
-Due to recent advances in Genome Wide Association Studies (GWAS), the availability of  datasets in human (\textit{e.g.} UK Biobank) allows to dissect the genetic underpinning of quantitative traits such as Body Mass Index (BMI) or disease related susceptibility.
-For example by relating observable summary statistics (phenotypic distribution, allelic frequencies, variance per site) to parameters of the genetic architecture (pleiotropy, distribution of selection coefficients, effect size), as in \citet{simons_population_2018}.
-Overall, as reviewed from GWAS evidences in \citet{sella_thinking_2019}, most quantitative traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection in humans.
+To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at the population and phylogenetic scale.
+Quantitative genetics formalism has related both within and between species variation to the underlying mutational input for a neutral trait.
+More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
+At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
+This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
+Altogether, we define the neutrality index as $\RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
+Under a depleted variation between species, meaning $\RatePhy < \RatePop$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
+In contrast, under an inflated variation between species, meaning $\RatePhy > \RatePop$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.
 
-\subsection{Gene expression}
-
-In transcriptomic dataset, the expression level of each gene can be seen a quantitative trait.
-First, $\VarGenetic$ can be estimated from the standing variation inside natural populations. 
-Second, $\VarMutation$ can be  estimated from mutation accumulation experiments in laboratory clonal lines.
-Third, $\NbrLoci$ can be  estimated from effect of gene knockout of expression level.
-Forth, $\MutationRate$ can be  estimated from the frequency of mutation altering expression level.
-Fifth, $\Ne$ can be estimated from standing genetic polymorphism ($\pi = 4 \Ne u$), by correcting for the mutation rate ($u$).
-Altogether, the evolutionary regime (neutral or selection) of gene expression can be tested across all genes, as suggested in \citet{fay_evaluating_2008}.
-
-In the yeast \textit{Saccharomyces cerevisiae}, the fruit fly \textit{Drosophila melanogaster} and the nematode \textit{Caenorhabditis elegans},  \citet{hodgins-davis_gene_2015} found that gene expression is under stabilizing selection, such that neutral evolution is ruled out for most genes.
-Moreover, the House of card model is preferred against the Gaussian model of stabilizing selection, also suggesting that gene expression is encoded by a few loci with large effects.
-
-Finally, mutation accumulation experiments with different population size have shown that $\VarMutation$ decreases with $\Ne$, suggesting that gene expression is under strong stabilizing selection~\citep{deiss_global_2021}.
-
-\section{Phylogenetic trait evolution}
-
-\subsection{Relationship between micro and macro evolution. }
-
-We denote $\Trait_i$ the quantitative trait $\Trait$ for species $i$.
-\citet{hansen_translating_1996} derived a relation between trait variation between species and the genetic variance of this trait as 
-\begin{gather}
-    \Cov \left( \Trait_i, \Trait_j \right) = \Cov \left( \E \left[ \Trait_i | \Trait_z \right ], \E \left[ \Trait_j | \Trait_z \right ] \right),  \label{eq-cov-traits}
-\end{gather}
-where $z$ is the ancestor between species $i$ an $j$.
-
-If the trait is neutral, evolving as a Brownian process, then $\E \left[ \Trait_i | \Trait_z \right ] = \Trait_z$ and equation \ref{eq-cov-traits} simplifies to:
-\begin{align}
-\Cov \left( \Trait_i, \Trait_j \right) &= \Var \left( \Trait_z \right),\\
-    &= \frac{\VarGenetic}{\Ne} t_{z}, \\
-    &=  2 t_{z} \VarMutation, \label{eq-cov-traits-neutral} \text{ from equation } \ref{eq-var-genetic},
-\end{align}
-where $t_{z}$ is the time between species $z$ an both $i$ an $j$.
-
-The relationship between trait variation between species and the mutational variance can be used as method to detect whether traits are evolving under a neutral model of evolution, similar to the use of synonymous ($\ds$) and non-synonymous ($\dn$) substitution rates in molecular evolution, as noted in \citet{fay_evaluating_2008}.
-
-\begin{table}[H]
+\begin{table*}[tb]
     \centering
-	\noindent\adjustbox{max width=\textwidth}{%
-		\begin{tabu}{|c||c|c|}
+    \begin{adjustbox}{width = 0.75\textwidth}
+        \begin{tabular}{|c||c|c|}
             \hline
             & Coding sequences                                  & Quantitative traits                                               \\ \hline \hline
-			Adaptation or moving optimum & $\dnds > 1$ & $\Cov \left( \Trait_i, \Trait_j \right) > 2 t_{z} \VarMutation $ \\ \hline
-			Neutral regime of evolution & $\dnds = 1$ & $\Cov \left( \Trait_i, \Trait_j \right) = 2 t_{z} \VarMutation $ \\ \hline
-			Purifying or directional selection & $\dnds < 1$ & $\Cov \left( \Trait_i, \Trait_j \right) < 2 t_{z} \VarMutation $ \\ \hline
-		\end{tabu}}
-\end{table}
+            Adaptation or moving optimum       & $\frac{\dn \Multiply \ps}{\ds \Multiply \pn} > 1$ & $ \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d } > 1 $ \\ \hline
+            Neutral regime of evolution        & $\frac{\dn \Multiply \ps}{\ds \Multiply \pn} = 1$ & $ \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d } = 1 $ \\ \hline
+            Purifying or directional selection & $\frac{\dn \Multiply \ps}{\ds \Multiply \pn} < 1$ & $\frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d } < 1 $ \\ \hline
+        \end{tabular}
+    \end{adjustbox}
+    \caption{
+        The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
+        Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
+    }
+    \label{table:unfolded-MK}
+\end{table*}
 
-If the trait is under directional selection, but the strength and direction of selection varies randomly from one generation to the next (variance $\VarSelection_D$), then the trait still follows a Brownian process model and equation \ref{eq-cov-traits} reduces to:
-\begin{align}
-\Cov \left( \Trait_i, \Trait_j \right) &= \Var \left( \Trait_z \right), \\
-    &= \left(\frac{\VarGenetic}{\Ne} + \VarGenetic \VarSelection_D \VarGenetic \right)  t_{z}. \label{eq-selection-vary-direct}
-\end{align}
+Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
+Analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
+Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
+This statistic can be obtained as a maximum likelihood estimate, from equations \ref{eq:estimated-rate-pop} and \ref{eq:estimated-rate-phy}.
 
-If the trait is under stabilizing selection, but the strength and direction of the optimum varies randomly from one generation to the next (variance $\VarSelection_{\theta}$), then the trait still follows a Brownian process model and equation \ref{eq-cov-traits} reduces to:
-\begin{align}
-\Cov \left( \Trait_i, \Trait_j \right) &= \VarSelection_{\theta} t_{z}. \label{eq-selection-vary-stabi}
-\end{align}
+\subsection{Results against simulations}
 
-Altogether, traits that are found to be under a long term Brownian process are not necessarily neutral.
+The inference framework was first tested on independently simulated datasets.
+With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant empirical regime.
+Thus, the tree topology and the branch lengths were chosen based on a tree estimated on the mammalian dataset further considered below.
 
-To read : \citet{arnold_adaptive_2001}
-\subsection{Trait evolution under a Brownian or Ornstein-Uhlenbeck process}
+\begin{figure*}[!ht]
+    \centering
+    \includegraphics[width=\textwidth, page=1] {artworks/constant_pop_size_phy_pop.hist}
+    \caption{
+        $\RatePhy$ estimated at the phylogenetic scale as a function of $\RatePop$ estimated at the population scale, for $30.000$ genes simulated under different evolutionary regimes.
+        $\NIx < 1$ for traits simulated under selection (stabilizing selection in yellow).
+        $\NIx = 1$ for traits simulated under a neutral evolution (in blue).
+        $\NIx > 1$ for genes simulated under a moving optimum (diversifying selection in red).
+    }
+    \label{fig:constant_pop_size_phy_pop}
+\end{figure*}
 
-Modelling $\Trait$ as a Brownian process allows theoretically to test for neutrality.
-In \citet{catalan_drift_2019}, gene expression is modelled either as a Brownian (supposedly neutral regime), or as a Brownian with a shift in the mean for one branch (supposedly directional selection), or as an Ornstein-Uhlenbeck process (supposedly stabilizing selection) or finally as an Ornstein-Uhlenbeck process with a shift in the optimal for one branch (supposedly directional selection).
-The analysis on the transcriptome across different individuals and species concluded that 81\% of genes are under a neutral regime and 16\% under directional selection. 
-However, the alternative model of a varying directional (equation~ \ref{eq-selection-vary-direct}) or stabilizing selection (equation~ \ref{eq-selection-vary-stabi}) was not implemented, and thus can not be ruled out by this analysis.
+\section{Discussion}\label{sec:discussion}
+% Consequences of our neutrality index
+% Use case for gene expression level
+ Distinguishing adaptive changes driven by positive selection from neutral divergence resulting from mutation and genetic drift is critical for understanding the evolution of a quantitative trait.
+ At the population scale, other tests of neutrality for a quantitative trait have been developed, by leveraging mutation accumulation experiments.
+ In the yeast \textit{Saccharomyces cerevisiae}, the fruit fly \textit{Drosophila melanogaster} and in the nematode \textit{Caenorhabditis elegans}, gene expression level has been shown to under stabilizing selection, such that neutral evolution of expression level is ruled out for most genes\cite{hodgins-davis_gene_2015}.
+Finally, mutation accumulation experiments with different population size have also shown that gene expression is under strong stabilizing selection\cite{deiss_global_2021}.
+Altogether, at the population level, gene expression has been shown to be empirically under stabilizing selection.
 
-\subsection{Variance evolution}
+At the phylogenetic scale, other tests of neutrality for a quantitative trait have been developed, modelling the underlying process generating the change in mean trait value across different species at the phylogenetic scale\cite{fay_evaluating_2008}.
+These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
+However, deviation from a Brownian process at the phylogenetic scale, does not mean that the
+trait is not under selection.
+Indeed, a trait under selection for which the optimal trait is evolving as a Brownian process will not deviate from a Brownian process\cite{hansen_translating_1996}.
+For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection\cite{catalan_drift_2019}.
+However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
+We argue that our diversity index, by leveraging trait variation at different scale and normalizing by divergence and polymorphism, we can set a threshold for neutrality and thus is able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
+To do so, our neutrality index leverage sequence divergence and polymorphism, as well as the heritability of traits.
+We argue that this is the main novelty of this study, being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
+We showed, that by doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny.
+Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
+We argue that the main drawback of our method is that the heritability ($h^2$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
 
-The multidimensional generalization of $\VarGenetic$ to several traits is the of the genetic covariance matrix $\MatrixGenetic$.
-Stability of $\MatrixGenetic$, can be reformulated as whether $\MatrixGenetic (t) = \MatrixGenetic$, an assumption discussed in \citet{arnold_understanding_2008} and \citet{hohenlohe_mipod_2008}.
-Having a time dependent matrix is implemented in JIVE, a Bayesian hierarchical model developed by \citet{kostikova_bridging_2016} and implemented in BEAST by \citet{gaboriau_multi-platform_2020}.
+% Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
+More theoretically, our development is based on several assumption that could be relaxed such as constant genetic architecture and independence of traits.
+First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
+Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Moreover, even though the neutrality index can be applied for has many traits as available,  each trait is so far assumed to be independent.
+Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework\cite{huelsenbeck_detecting_2003,  lartillot_phylogenetic_2011,lartillot_joint_2012}.
 
-\section{Discussion \& perspectives}
-At the population level, gene expression is shown to be empirically under stabilizing selection.
-However at the phylogenetic scale, gene expression is under a neutral regime.
-Can we reconcile these observations?
+%TC:ignore
+\section*{Acknowledgements}
+\label{sec:acknowledgment}
+\textbf{Funding:}
+Université de Lausanne.
+\textbf{Author contributions:}
+Original idea: T.L.\ and N.S.;
+Model conception: T.L.\ and N.S.;
+Code: T.L.;
+Data analyses: T.L.;
+Interpretation: T.L., T.G\ and N.S.;
+First draft: T.L.;
+Editing and revisions: T.L., T.G\ and N.S.;
+Project management and funding: N.S\@.
+\textbf{Competing interests:}
+The authors declare no conflicts of interest.
+\textbf{Data and materials availability:}
+Snakemake pipeline, analysis scripts and documentation are available at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}.
 
-We can find the genes under adaptive regime (fluctuating optimal gene expression) at the phylogenetic scale and compare that to the genes under adaptation using $\dnds$ based methods.
-Do we find the same genes?
-\section{Appendix}
-
-\subsection{Definitions}
-
-We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $i$.
-
-\begin{align}
-    \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
-\end{align}
-
-\bibliographystyle{natbib}%%%%Bibliography style file
-\bibliography{references}
+\printbibliography
 
 \end{document}
+%TC:endignore

### 2023-01-13 manuscript/supp-mat.tex
Draft of the manuscript for JEB.

%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
}
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
\addbibresource{references.bib}

\usepackage[margin=50pt]{geometry}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{verbatim}
\usepackage{float}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{tabu}
\usepackage{longtable}
\tabulinesep=0.6mm
\newcommand\cellwidth{\TX@col@width}
\usepackage{hhline}
\setlength{\arrayrulewidth}{1.2pt}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\AtBeginEnvironment{tabu}{\footnotesize}
\usepackage{booktabs}
\usepackage{graphicx}
\pdfinclusioncopyfonts=1

\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

\input{notations.tex}

\title{Detection of selection for a quantitative traits through the scaled mutational variance at the phylogenetic and population-genetic level}

\author{
    \large
    \textbf{T. {Latrille}$^{1}$, T. {Gaboriau}$^{1}$, N. {Salamin}$^{1}$}\\
    \normalsize
    $^{1}$Université de Lausanne, Lausanne, Switzerland\\
    \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

\date{}

\begin{document}
    \maketitle
    \part*{Supplementary materials}
    \tableofcontents
    \clearpage

    \section{Theoretical derivation}
    \subsection{Definitions}

    We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $\species$.

    \begin{align}
        \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
    \end{align}

    \subsection{Mutation-selection-drift equilibrium}

    Selection, whether positive or negative, depletes standing variation in the population and reduces to genetic variance.
    Hence, at equilibrium between mutation, selection and drift, the observed genetic variance is lower than the theoretical neutral variance:
    \begin{align}
        \VarGenetic & < 2 \Ne \VarMutation \\
        \iff \VarGenetic & < 4 \Ne \NbrLoci \MutationRate \E \left[ a^2 \right]
    \end{align}

    Under the House of cards model of stabilizing selection\cite{kingman_simple_1978}, the trait's fitness is normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$.
    Moreover, the trait is encoded by a few loci with large effects and the genetic variance at equilibrium between mutation, selection and drift is:
    \begin{gather}
        \VarGenetic = 2 \NbrLoci \MutationRate \VarSelection_S \label{eq-mutsel-HC}.
    \end{gather}

    Alternatively, under the Gaussian model of stabilizing selection\cite{lande_natural_1976}, the trait's fitness is also normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$.
    However, the trait is encoded by a large number of loci traits with small effect and the genetic variance at equilibrium is:
    \begin{gather}
        \VarGenetic = \sqrt{2 \NbrLoci \VarMutation \VarSelection_S} \label{eq-mutsel-G}.
    \end{gather}

    The boundary between the two models is given by \textcite{turelli_heritable_1984} as:
    \begin{align}
        20 \MutationRate \VarSelection_S &< \E \left[ a^2 \right] \text{ for the House of card model}, \\
        20 \MutationRate \VarSelection_S &> \E \left[ a^2 \right] \text{ for the geometric model}.
    \end{align}

    \subsection{Selection at the phylogenetic scale}
    If the trait is under directional selection, but the strength and direction of selection varies randomly from one generation to the next (variance $\VarSelection_D$), then the trait still follows a Brownian process:
    \begin{align}
        \VarPhy &= \Var \left( \Trait_z \right), \\
        &= \left(\frac{\VarGenetic}{\Ne} + \VarGenetic \VarSelection_D \VarGenetic \right) t. \label{eq-selection-vary-direct}
    \end{align}

    If the trait is under stabilizing selection, but the strength and direction of the optimum varies randomly from one generation to the next (variance $\VarSelection_{\theta}$), then the trait still follows a Brownian process:
    \begin{align}
        \VarPhy &= \VarSelection_{\theta} t. \label{eq-selection-vary-stabi}
    \end{align}

    Altogether, traits that are found to be under a long term Brownian process are not necessarily neutral.

    \printbibliography

\end{document}

### 2023-01-13 scripts/libraries.py

@@ -121,6 +121,7 @@ def scatter_plot(x, y, x_label, y_label, output, nbr_genes, title="", histy_log=
         ax.set_xlim((0.95 * min_x, max_x * 1.05))
         ax.set_ylim((0.95 * min_y, max_y * 1.05))
     ax.legend(fontsize=fontsize_legend)
+    plt.tight_layout()
     plt.savefig(output, format="pdf")
     plt.savefig(output.replace(".pdf", ".png"), format="png")
     plt.close('all')
@@ -143,10 +144,8 @@ def hist_plot(x, x_label, output, nbr_genes, title=""):
 
     for id_m, m in enumerate(x):
         x_mean = np.mean(x[m])
-        x_median = np.median(x[m])
         ax.plot((x_mean, x_mean), (0, max_y), linewidth=3, color=color_models[id_m],
-                label=(m.replace("_", " ").capitalize() + ' (mean {0:.2g}; median {1:.2g})'.format(x_mean, x_median)))
-        ax.plot((x_median, x_median), (0, max_y), linewidth=2, linestyle='--', color=color_models[id_m])
+                label=(m.replace("_", " ").capitalize() + ' (mean {0:.2g})'.format(x_mean)))
 
     ax.set_xlabel(x_label, fontsize=fontsize)
     ax.set_xlim((0.95 * min_x, 1.05 * max_x))
@@ -154,6 +153,7 @@ def hist_plot(x, x_label, output, nbr_genes, title=""):
     ax.set_ylim((0, max_y))
     ax.set_ylabel("Density", fontsize=fontsize)
     ax.legend(fontsize=fontsize_legend)
+    plt.tight_layout()
     plt.savefig(output, format="pdf")
     plt.savefig(output.replace(".pdf", ".png"), format="png")
     plt.clf()

### 2023-01-13 scripts/plot_scatter_stats.py

@@ -47,9 +47,13 @@ def main(folder, output):
     # to break out the lists into columns
     df.to_csv(output, sep="\t", index=False)
     nb_genes = set([len(v) for v in var_dict["between"].values()]).pop()
+    x_str = r"Neutrality index $\left( \frac{\sigma^2_{Phy}}{\sigma^2_{Pop}} \right)$"
+    ratio_scaled = {m: np.array(var_dict["between"][m]) / np.array(var_dict["within_sampled"][m]) for m in models}
+    hist_plot(ratio_scaled, x_str, output.replace(".tsv.gz", ".hist.pdf"), nb_genes)
+
     mut_str = r"Variance mutational $\left( \frac{V_M}{2u} \right)$"
     within_str = r"Variance within $\left( \frac{V_G}{\theta} \right)$"
-    between_str = r"Variance between $\left( \frac{V_B}{4d} \right)$"
+    between_str = r"Variance between $\left( \frac{Var[\overline{X}_t]}{4d} \right)$"
     scatter_plot(var_dict["within_sampled"], var_dict["between"], within_str, between_str,
                  output.replace(".tsv.gz", "-H.pdf"), nb_genes, title=' - Sample theta', histy_log=True)
     scatter_plot(var_dict["within_tajima"], var_dict["between"], within_str, between_str,

### 2023-01-16 manuscript/main.tex
Updated manuscript

@@
 \begin{abstract}
-    Current methods at the population scale do not integrate the changes over long evolutionary time, while at the same time the methods require laboratory clonal lines to determine the neutral expectation.
+    To determine whether a trait is neutrally evolving or under selection, methods have been developed both at the population and at the phylogenetic scale.
+    At the population scale, methods do not integrate the changes over long evolutionary time.
     Conversely, at the phylogenetic scale, current methods discard the variance in traits between individuals, leaving out valuable information.
-    The main goal of this project is to combine within-species and between-species variation of quantitative traits.
-    To this aim, we define a neutrality index for of a quantitative trait, devoted to determine its evolutionary regime, leveraging both data at the population and phylogenetic scale.
+    The main goal of this project is to combine within-species and between-species variation of quantitative traits to determine the evolutionary regime of a trait.
+    We define a neutrality index for a quantitative trait, leveraging both data at the population and phylogenetic scale.
+    Moreover, this index leverages sequence divergence and polymorphism to normalize the variation of trait between and within species.
+    Our neutrality index equals to one for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
+    If our neutrality index is lower than one for traits putatively under stabilizing selection, and conversely greater than one fors trait putatively under diversifying selection.
+    Finally, we show that our test is not sensitive to the assumption that population sizes and mutation rates are constant across the phylogeny, and automatically adjust for it.
 \end{abstract}
@@
 \section{Introduction}\label{sec:introduction}
-A quantitative trait is by definition an observable can be measured for an organisms, for example genome size\cite{alfsnes_genome_2017}.
-Life history traits (LHTs) are such quantitative traits, related to the age and stage-specific patterns and timing of events that make up an organism's life such as age at maturity, fecundity, life expectancy, height, weight, and so on.
-LHTs are often suspected to be under selection and encoded by many loci.
-Appart from LHTs, the genetic underpinning of other quantitative traits such as Body Mass Index (BMI) or disease related susceptibility can now be dissected due to recent advances in Genome Wide Association Studies (GWAS) and the availability of datasets in human (\textit{e.g.} UK Biobank)\cite{simons_population_2018}.
-Overall, GWAS have revealed that traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection in humans\cite{sella_thinking_2019}.
-At the genomic level, the expression level of each gene can also be seen as a quantitative trait.
-Surveys of gene expression reveal extensive variability both within and between a wide range of species\cite{fay_evaluating_2008}.
 
 From the standing variation both within and between a wide range of species, distinguishing adaptive changes driven by positive selection from neutral evolution is critical for understanding the evolution of trait evolution.
-At the population level, the variance between individuals of the same population can be estimated for any quantitative trait.
-This observed variance can be compared to the expected variance that would be obtained under the assumption that the trait is not under selection, which can be estimated from mutation accumulation experiments in laboratory clonal lines.
+At the population level, Genome Wide Association Studies (GWAS) in humans have revealed that traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
 At the phylogenetic scale, between different populations, the evolutionary regime for a quantitative trait can also be tested.
 Typically, the trait mean (across individuals of the same population) is modelled as a Brownian process, allowing theoretically to test for neutrality.
-Alternatively, a Brownian with a shift in the mean for one branch is supposedly directional selection, an Ornstein-Uhlenbeck process is supposedly stabilizing selection or finally an Ornstein-Uhlenbeck process with a shift in the optimal for one branch is supposedly directional selection\cite{catalan_drift_2019}.
+Alternatively, a deviation from the Brownian process is interpreted as selection\cite{catalan_drift_2019}.
 
-At the frontier between phylogenetics and population genetics, studies comparing between species variation have acknowledged the importance of within species variation\cite{whitehead_variation_2006}.
+At the frontier between phylogenetics and population genetics, studies comparing between species variation have acknowledged the importance of within species variation\cite{whitehead_variation_2006, fay_evaluating_2008}.
 However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
@@
 Altogether, there is a push toward integrating within and between species variation into a unified framework.
-This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
-However, it has not yet been generalized for any number of species related by their phylogenetic tree.
-Finally, studies on quantitative trait evolution
 In this study, we propose a neutrality index for a quantitative trait, leveraging data within and between species, normalizing the trait evolution by its genetic polymorphism and divergence.
+% This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
+% However, it has not yet been generalized for any number of species related by their phylogenetic tree.
 
@@
     \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
-    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and \ref{eq:genetic-diversity}}, \\
+    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
     & = \RateMut. \label{eq:rate-pop}
@@
     d & = \Time \Multiply \SubRate \\
-    & = \Time \Multiply \MutationRate \text{ from eq.\ref{eq:substitution-rate}}. \label{eq:distance}
+    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
 \end{align}
@@
     \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
-    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and \ref{eq:distance}}, \\
+    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
     & = \RateMut \label{eq:rate-phy}.
@@
 \begin{align}
-    \RateMut & = \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and \ref{eq:rate-phy}} , \\
+    \RateMut & = \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}} , \\
     \Rightarrow \frac{\RatePhy}{\RatePop} & = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
@@
 
-\subsection{Estimates}
+\subsection{Estimate}
 
 \subsubsection{Maximum likelihood estimate}
-For each population of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences thus normalizing .
+For each population of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences.
 Averaged across all populations we obtain the mean as:
@@
 At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
-The contrast matrix is symmetric, along the diagonal are the total distances ($d_{i}$) from the root of the tree to each taxon ($i$).
+The contrast matrix is symmetric, along the diagonal are the total distances ($d_{\Spi}$) from the root of the tree to each taxon ($\Spi$).
 The off-diagonal elements $C_{\Spi,\Spj}$ are the total branch lengths shared by particular pairs of taxa $\Spi$ and $\Spj$, computed from the distance between the pair ($d_{\Spi,\Spj}$) as:
@@
 
-\subsection{Simulations}\label{subsec:simulations}
+\subsection{Simulation}\label{subsec:simulations}
 The evolutionary dynamics was formalized as a Wright-Fisher model with mutation, selection and drift.
@@
 Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
-This statistic can be obtained as a maximum likelihood estimate, from equations \ref{eq:estimated-rate-pop} and \ref{eq:estimated-rate-phy}.
+This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
 
@@
     \includegraphics[width=\textwidth, page=1] {artworks/constant_pop_size_phy_pop.hist}
+    \includegraphics[width=\textwidth, page=1] {artworks/fluctuating_pop_size_phy_pop.hist}
     \caption{
@@
         $\NIx > 1$ for genes simulated under a moving optimum (diversifying selection in red).
+        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top panel), or fluctuating as a Brownian process along the phylogeny (panel B).
     }

### 2023-01-16 scripts/libraries.py

@@ -131,7 +131,7 @@ def scatter_plot(x, y, x_label, y_label, output, nbr_genes, title="", histy_log=
 def hist_plot(x, x_label, output, nbr_genes, title=""):
     output = output.replace("scatter_plot", "histogram")
     color_models = colors(x)
-    fig = plt.figure(figsize=(1920 / my_dpi, 880 / my_dpi), dpi=my_dpi)
+    fig = plt.figure(figsize=(1920 / my_dpi, 480 / my_dpi), dpi=my_dpi)
     fig.suptitle("{0} genes".format(nbr_genes) + title)
     ax = fig.add_subplot(1, 1, 1)
     x = {k: [i for i in v if np.isfinite(i)] for k, v in x.items()}

### 2023-02-01 manuscript/main.tex
Updated section introduction.

@@
 
-\title{A neutrality index for quantitative traits}
+\title{A neutrality index for quantitative traits from within and between variations leveraging sequences evolution}
 
@@
     At the population scale, methods do not integrate the changes over long evolutionary time.
-    Conversely, at the phylogenetic scale, current methods discard the variance in traits between individuals, leaving out valuable information.
+    Conversely, at the phylogenetic scale, current methods usually discards the variance in traits between individuals, leaving out valuable information.
     The main goal of this project is to combine within-species and between-species variation of quantitative traits to determine the evolutionary regime of a trait.
@@
 
-From the standing variation both within and between a wide range of species, distinguishing adaptive changes driven by positive selection from neutral evolution is critical for understanding the evolution of trait evolution.
-At the population level, Genome Wide Association Studies (GWAS) in humans have revealed that traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
-At the phylogenetic scale, between different populations, the evolutionary regime for a quantitative trait can also be tested.
-Typically, the trait mean (across individuals of the same population) is modelled as a Brownian process, allowing theoretically to test for neutrality.
-Alternatively, a deviation from the Brownian process is interpreted as selection\cite{catalan_drift_2019}.
-
-At the frontier between phylogenetics and population genetics, studies comparing between species variation have acknowledged the importance of within species variation\cite{whitehead_variation_2006, fay_evaluating_2008}.
-However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is a promising line of research\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Altogether, there is a push toward integrating within and between species variation into a unified framework.
-In this study, we propose a neutrality index for a quantitative trait, leveraging data within and between species, normalizing the trait evolution by its genetic polymorphism and divergence.
+Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
+Fundamentally, distinguishing neutral evolution from selection requires to determine which model of evolution is supported from the observed variation of traits and sequences.
+This variation of traits can be observed at different scales, at the individual level across different development stages, at the population level across different individuals, at the species level across different populations, and finally at the phylogenetic level across different species.
+All these scales requires different assumptions and methodology, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods and developments across different fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
+At the population level across individuals, Genome Wide Association Studies (GWAS) in humans have revealed that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
+At the species level, by contrasting variation within and between different populations, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes\cite{whitehead_neutral_2006, gilad_natural_2006, gilad_expression_2006}.
+However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated\cite{signor_evolution_2018, price_detecting_2022}.
+For a given population, by contrasting to the divergence to a close species, studies have confirms that changes in gene expression accumulate linearly differences with time from divergence\cite{khaitovich_neutral_2004}.
 % This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
 % However, it has not yet been generalized for any number of species related by their phylogenetic tree.
+% However, this divergence in the level of expression stabilizes over a long distance: the difference between human and chicken (having diverged 300 million years) is similar to that between human and opossum (160 million years).
+% This observation suggests that the conservation of essential functions of organs could define a limit for the divergence of transcriptomes.
+
+% Phylogenetics (mean trait evolution)
+At the phylogenetic scale, by determining how much the mean trait value (taken as average across different individual for a given species) is changing across different species, the mode of selection for a quantitative trait can also be tested while accounting for the underlying relationships between species\cite{felsenstein_phylogenies_1985}.
+Typically, the mean trait value is modelled as a Brownian process, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Alternatively, a deviation from the Brownian process, typically an Ornstein-Uhlenbeck process is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
+However, studies have shown that comparative approaches are subject to different biases\cite{harmon_phylogenetic_2018}.
+First, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
+Indeed, a trait under selection for which the optimal trait is evolving as a Brownian process will not deviate from a Brownian process\cite{hansen_translating_1996}.
+Secondly and contrarily, even under a neutral trait the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling   artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+
+% At the frontier Phylogenetics/Population-genetics
+At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+Altogether, there is a push toward integrating within and between species variation into a unified framework.
+Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
+In this study, we propose a neutrality index for a quantitative trait, leveraging data within and between species, normalizing the trait evolution with both genetic polymorphism within species and divergence between species.
+In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
+Instead, our procedure can be applied to a single trait, without prior knowledge, estimating a statistic called $\RatePhy / \RatePop$.
+Technically, $\RatePhy / \RatePop$ is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
 
@@
  Distinguishing adaptive changes driven by positive selection from neutral divergence resulting from mutation and genetic drift is critical for understanding the evolution of a quantitative trait.
- At the population scale, other tests of neutrality for a quantitative trait have been developed, by leveraging mutation accumulation experiments.
- In the yeast \textit{Saccharomyces cerevisiae}, the fruit fly \textit{Drosophila melanogaster} and in the nematode \textit{Caenorhabditis elegans}, gene expression level has been shown to under stabilizing selection, such that neutral evolution of expression level is ruled out for most genes\cite{hodgins-davis_gene_2015}.
-Finally, mutation accumulation experiments with different population size have also shown that gene expression is under strong stabilizing selection\cite{deiss_global_2021}.
-Altogether, at the population level, gene expression has been shown to be empirically under stabilizing selection.
-
-At the phylogenetic scale, other tests of neutrality for a quantitative trait have been developed, modelling the underlying process generating the change in mean trait value across different species at the phylogenetic scale\cite{fay_evaluating_2008}.
+At the phylogenetic scale, tests of neutrality for a quantitative trait have been developed, modelling the underlying process generating the change in mean trait value across different species at the phylogenetic scale\cite{fay_evaluating_2008}.
 These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
-However, deviation from a Brownian process at the phylogenetic scale, does not mean that the
-trait is not under selection.
-Indeed, a trait under selection for which the optimal trait is evolving as a Brownian process will not deviate from a Brownian process\cite{hansen_translating_1996}.
 For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection\cite{catalan_drift_2019}.
 However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
-We argue that our diversity index, by leveraging trait variation at different scale and normalizing by divergence and polymorphism, we can set a threshold for neutrality and thus is able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
+Our diversity index, by leveraging trait variation at different scale and normalizing by divergence and polymorphism, we can set a threshold for neutrality and thus is able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
 To do so, our neutrality index leverage sequence divergence and polymorphism, as well as the heritability of traits.
-We argue that this is the main novelty of this study, being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
-We showed, that by doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny.
+We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
+By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
 Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.

### 2023-02-03 manuscript/main.tex
Intro/discussion

@@
 Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
-Fundamentally, distinguishing neutral evolution from selection requires to determine which model of evolution is supported from the observed variation of traits and sequences.
-This variation of traits can be observed at different scales, at the individual level across different development stages, at the population level across different individuals, at the species level across different populations, and finally at the phylogenetic level across different species.
+Fundamentally, distinguishing neutral evolution from selection requires to determine which mode of selection is supported from the observed variation of traits and sequences.
+This variation of traits can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
 All these scales requires different assumptions and methodology, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods and developments across different fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
@@
 For a given population, by contrasting to the divergence to a close species, studies have confirms that changes in gene expression accumulate linearly differences with time from divergence\cite{khaitovich_neutral_2004}.
+However, studies including variation across individuals and population typically include species divergence to a close species in a pairwise comparison, and do not integrate the changes across the whole phylogenetic tree.
 % This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
@@
 % Phylogenetics (mean trait evolution)
-At the phylogenetic scale, by determining how much the mean trait value (taken as average across different individual for a given species) is changing across different species, the mode of selection for a quantitative trait can also be tested while accounting for the underlying relationships between species\cite{felsenstein_phylogenies_1985}.
-Typically, the mean trait value is modelled as a Brownian process, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Contrarily, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale\cite{felsenstein_phylogenies_1985}.
+Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a variance proportional to the length of the branch\cite{hansen_translating_1996}.
+As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
 Alternatively, a deviation from the Brownian process, typically an Ornstein-Uhlenbeck process is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
@@
 First, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
-Indeed, a trait under selection for which the optimal trait is evolving as a Brownian process will not deviate from a Brownian process\cite{hansen_translating_1996}.
-Secondly and contrarily, even under a neutral trait the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling   artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Indeed, a trait under selection for which the optimal trait value is evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral\cite{hansen_translating_1996}.
+Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling   artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Indeed, in such framework, the variation within species is discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
+Altogether, the classical comparative framework allows tests for neutrality at the phylogenetic scale but discards the variance in traits between individuals, leaving out valuable information.
 
@@
 However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-Altogether, there is a push toward integrating within and between species variation into a unified framework.
 Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
-In this study, we propose a neutrality index for a quantitative trait, leveraging data within and between species, normalizing the trait evolution with both genetic polymorphism within species and divergence between species.
-In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
-Instead, our procedure can be applied to a single trait, without prior knowledge, estimating a statistic called $\RatePhy / \RatePop$.
-Technically, $\RatePhy / \RatePop$ is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
+In this study, we propose a neutrality index for a quantitative trait articulating trait variation at both the phylogenetic and population scale.
+Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales.
 
@@
 To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at the population and phylogenetic scale.
-Quantitative genetics formalism has related both within and between species variation to the underlying mutational input for a neutral trait.
+The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
 More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
@@
 \section{Discussion}\label{sec:discussion}
-% Consequences of our neutrality index
-% Use case for gene expression level
- Distinguishing adaptive changes driven by positive selection from neutral divergence resulting from mutation and genetic drift is critical for understanding the evolution of a quantitative trait.
-At the phylogenetic scale, tests of neutrality for a quantitative trait have been developed, modelling the underlying process generating the change in mean trait value across different species at the phylogenetic scale\cite{fay_evaluating_2008}.
+
+In this study, we proposed a neutrality index for a quantitative trait.
+At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
+Similarly, at the population scale, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
+Our neutrality index for a trait is the ratio of between species variation to within species variation.
+Such neutrality index can be compared to 1 to seek a statistical deviation from a neutral trait.
+In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
+Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
+Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
+Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
+We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
+By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
+Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
+Contrarily, we argue that the main drawback of our method is that the heritability ($h^2$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
+
+At the phylogenetic scale, the change in mean trait value across different species is typically used to test for neutrality for a quantitative trait.
 These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
@@
 However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
-Our diversity index, by leveraging trait variation at different scale and normalizing by divergence and polymorphism, we can set a threshold for neutrality and thus is able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
-To do so, our neutrality index leverage sequence divergence and polymorphism, as well as the heritability of traits.
-We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
-By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
-Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
-We argue that the main drawback of our method is that the heritability ($h^2$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
+With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
+Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$), as well as estimate for heritabilities ($h^2$).
 
 % Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
-More theoretically, our development is based on several assumption that could be relaxed such as constant genetic architecture and independence of traits.
+More theoretically, our development is based on several assumption that could be relaxed.
 First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
@@
 Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework\cite{huelsenbeck_detecting_2003,  lartillot_phylogenetic_2011,lartillot_joint_2012}.
+Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within populations in a unified framework.
 

### 2023-05-15 manuscript/main.tex
Added data from the Zoonomia project published in the Science SI.

@@
 
-\title{A neutrality index for quantitative traits from within and between variations leveraging sequences evolution}
+\title{Testing for neutrality of a trait from within and between species variations}
 
@@
     Our neutrality index equals to one for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
-    If our neutrality index is lower than one for traits putatively under stabilizing selection, and conversely greater than one fors trait putatively under diversifying selection.
+    If our neutrality index is lower than one for traits putatively under stabilizing selection, and conversely greater than one for trait putatively under diversifying selection.
     Finally, we show that our test is not sensitive to the assumption that population sizes and mutation rates are constant across the phylogeny, and automatically adjust for it.
@@
 
-For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect on the random of a mutation ($a$).
+For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
 New mutations are thus generating variance for such a trait.
-We define the expected effect of a mutation on the variance of a trait as $\RateMut$, given as:
+We define the expected effect of mutations on the variance of a trait as $\RateMut$, given as:
 \begin{gather}

### 2023-05-15 scripts/pre_processed_traits_mammals.py

@@ -10,7 +10,10 @@ from neutrality_index import open_tree, prune_tree
 def rename_tree(tree: Tree) -> Tree:
     for node in tree.traverse("levelorder"):
         if node.is_leaf():
-            node.name = "_".join(node.name.replace("_E", "").split("_")[:-1])
+            if node.name.endswith("_E"):
+                node.name = "_".join(node.name.replace("_E", "").split("_")[:-1])
+            else:
+                node.name = node.name.replace(" ", "_")
     return tree
 
 
@@ -38,26 +41,32 @@ def main(path_input_traits, path_input_pS, path_input_dS, path_output_tree, path
     tree = rename_tree(tree)
     set_taxa_names = set(tree.get_leaf_names())
     print(f"The tree has {len(set_taxa_names)} leaves")
+
+    # Open the pS/hererozygosity file
     read_pS = open(path_input_pS, "r").readlines()
     if read_pS[0].strip() == "#TRAITS":
         columns = read_pS[1].strip().split(" ")
         n_taxa = int(columns[0])
         n_traits = int(columns[1])
-        columns = ["specie"] + columns[2:]
+        columns = ["species"] + columns[2:]
         assert len(columns) == n_traits + 1
         pS_df = pd.read_csv(path_input_pS, sep=" ", skiprows=2, names=columns)
         # Replace -1 by NaN
         pS_df = pS_df.replace(-1, np.nan)
         # Remove "_E" from the specie name
-        pS_df["specie"] = pS_df["specie"].apply(lambda x: x.replace("_E", ""))
+        pS_df["species"] = pS_df["species"].apply(lambda x: x.replace("_E", ""))
         assert len(pS_df) == n_taxa
     else:
         pS_df = pd.read_csv(path_input_pS, sep=",")
+
+    # Remove all spaces in the specie name
+    pS_df["species"] = pS_df["species"].apply(lambda x: x.replace(" ", "_"))
     print(f"The pS dataframe has {len(pS_df)} rows before filtering.")
-    pS_df = pS_df[pS_df["specie"].isin(set_taxa_names)]
-    pS_col = "pS" if "pS" in pS_df.columns else "pS_uniq"
+    pS_df = pS_df[pS_df["species"].isin(set_taxa_names)]
+    pS_col = "pS" if "pS" in pS_df.columns else "heterozygosity"
     pS_df = pS_df[np.isfinite(pS_df[pS_col]) & (pS_df[pS_col] > 0)]
[…]

### 2023-05-23 data_empirical/scripts/merge_results.py
Added trees obtained on neutral markers from https://doi.org/10.1126/science.abl8189

@@ -1,34 +1,0 @@
-import os
-import argparse
-import pandas as pd
-
-
-def replace_last(s: str, old: str, new: str) -> str:
-    li = s.rsplit(old, 1)
-    return new.join(li)
-
-
-def main(tsv_ML_list: str, tsv_Bayes_list: str, output: str):
-    os.makedirs(os.path.dirname(output), exist_ok=True)
-    list_df = []
-    for method, tsv_list in [("ML", tsv_ML_list), ("Bayesian", tsv_Bayes_list)]:
-        for path in tsv_list:
-            name_split = os.path.basename(path).replace(".tsv", "").split("_")
-            df = pd.read_csv(path, sep='\t')
-            df["method"] = method
-            for i, p in enumerate(["dataset", "sex", "logT"]):
-                df[p] = name_split[i]
-            list_df.append(df)
-    df_out = pd.concat(list_df)
-    # Sort by trait, dataset, sex, logT and then method
-    df_out = df_out.sort_values(by=["trait", "dataset", "sex", "logT", "method"])
-    df_out.to_csv(output, sep="\t", index=False)
-
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-    parser.add_argument("--tsv_ML", nargs="+", help="Input tsv ML file", required=True)
-    parser.add_argument("--tsv_Bayes", nargs="+", help="Input tsv BayesCode file", required=True)
-    parser.add_argument("--output", help="Output tsv file", required=True)
-    args = parser.parse_args()
-    main(args.tsv_ML, args.tsv_Bayes, args.output)

### 2023-05-23 data_empirical/scripts/neutrality_index.py

@@ -1,158 +1,0 @@
-import os
-import argparse
-import itertools
-from gzip import open as gzopen
-from collections import defaultdict
-import pandas as pd
-import numpy as np
-from ete3 import Tree
-
-
-def open_tree(tree_path: str, format_ete3: int = 1) -> Tree:
-    if tree_path.endswith(".gz"):
-        newick = gzopen(tree_path).read().decode()
-        return Tree(newick, format=format_ete3)
-    else:
-        return Tree(tree_path, format=format_ete3)
-
-
-def prune_tree(input_tree: Tree, list_taxa: list) -> Tree:
-    # Prune tree
-    tree = input_tree.copy()
-    tree.prune(list_taxa)
-    # Add polytomies if branch length are 0
-    remove_nodes = set([n for n in tree.traverse() if (n.dist == 0.0 and not n.is_root())])
-    for n in remove_nodes:
-        n.delete()
-    assert len(set([n for n in tree.traverse()]).intersection(remove_nodes)) == 0
-    for n in tree.traverse():
-        if not n.is_root():
-            assert n.dist > 0.0
-    return tree
-
-
-def brownian_fitting(tree: Tree, trait: str) -> (float, float):
-    leaves = tree.get_leaves()
-    n = len(leaves)
-    C = np.zeros((n, n))
-    x = np.zeros(n)
-    for i, leaf in enumerate(leaves):
-        x[i] = float(getattr(leaf, trait))
-
-    for node in tree.traverse("levelorder"):
-        if node.is_root():
-            node.t = 0
-        else:
-            # Don't forget the 4 factor (sigma^2_phy = Var(x)/4t)
-            node.t = 4 * node.dist + node.up.t
-
-    for i, j in itertools.product(range(n), range(n)):
-        if i == j:
-            C[i, j] = leaves[i].t
-        elif i < j:
-            ancestor = leaves[i].get_common_ancestor(leaves[j])
-            C[i, j] = ancestor.t
-            C[j, i] = ancestor.t
-
-    invC = np.linalg.inv(C)
-    ones = np.ones(n)
-    v = np.dot(ones.T, invC)
-    anc_z = float(np.dot(v, x)) / float(np.dot(v, ones))
-    assert np.isfinite(anc_z)
-
-    d = (x - anc_z * ones)
-    var = np.dot(d.T, np.dot(invC, d)) / (n - 1)
-    return anc_z, var
-
-
[…]

### 2023-05-23 data_empirical/scripts/pre_processed_traits.py

@@ -1,165 +1,0 @@
-import os
-import argparse
-from gzip import open as gzopen
-from collections import defaultdict
-import pandas as pd
-import numpy as np
-from ete3 import Tree
-
-
-def open_tree(tree_path: str, format_ete3: int = 1) -> Tree:
-    if tree_path.endswith(".gz"):
-        newick = gzopen(tree_path).read().decode()
-        t = Tree(newick, format=format_ete3)
-    else:
-        t = Tree(tree_path, format=format_ete3)
-    for node in t.traverse("levelorder"):
-        if node.is_leaf():
-            node.name = "_".join(node.name.replace("_E", "").split("_")[:-1])
-    return t
-
-
-def main(path_input_traits, path_input_pS, path_input_dS, path_output_tree, path_output_traits,
-         path_output_variance_pop, log_transform, sex):
-    for path in [path_input_traits, path_input_pS, path_input_dS]:
-        assert os.path.exists(path), f"Path {path} does not exist"
-    for path in [path_output_tree, path_output_traits, path_output_variance_pop]:
-        os.makedirs(os.path.dirname(path), exist_ok=True)
-
-    dS_tree = open_tree(path_input_dS, format_ete3=1)
-    print(f"The tree has {len(dS_tree.get_leaves())} leaves")
-    read_pS = open(path_input_pS, "r").readlines()
-    if read_pS[0].strip() == "#TRAITS":
-        columns = read_pS[1].strip().split(" ")
-        n_taxa = int(columns[0])
-        n_traits = int(columns[1])
-        columns = ["specie"] + columns[2:]
-        assert len(columns) == n_traits + 1
-        pS_df = pd.read_csv(path_input_pS, sep=" ", skiprows=2, names=columns)
-        # Replace -1 by NaN
-        pS_df = pS_df.replace(-1, np.nan)
-        # Remove "_E" from the specie name
-        pS_df["specie"] = pS_df["specie"].apply(lambda x: x.replace("_E", ""))
-        assert len(pS_df) == n_taxa
-    else:
-        pS_df = pd.read_csv(path_input_pS, sep=",")
-    print(f"The pS dataframe has {len(pS_df)} rows before filtering.")
-    pS_df = pS_df[pS_df["specie"].isin([n.name for n in dS_tree.get_leaves()])]
-    pS_col = "pS" if "pS" in pS_df.columns else "pS_uniq"
-    pS_df = pS_df[np.isfinite(pS_df[pS_col]) & (pS_df[pS_col] > 0)]
[…]

### 2023-05-23 data_empirical/scripts/read_trace.py

@@ -1,72 +1,0 @@
-import os
-import argparse
-import numpy as np
-import pandas as pd
-from collections import defaultdict
-
-
-def open_trace_file(trace_file: str, burn_in: int) -> dict[str: np.ndarray]:
-    df = pd.read_csv(trace_file, sep="\t")
-    out_dict = {}
-    for col in df.columns:
-        if col.startswith("Var_"):
-            assert col.endswith("_mean")
-            out_dict[replace_last(col[4:], "_mean", "")] = df[col].values[burn_in:]
-    n_taxa = int(df["n_taxa"][0])
-    return out_dict, n_taxa
-
-
-def replace_last(s: str, old: str, new: str) -> str:
-    li = s.rsplit(old, 1)
-    return new.join(li)
-
-
-def main(inference_path: str, variance_pop_path: str, burn_in: int, output_path: str):
-    os.makedirs(os.path.dirname(output_path), exist_ok=True)
-
-    traces_dict, n_taxa = open_trace_file(inference_path, burn_in=burn_in)
-    var_pop_df = pd.read_csv(variance_pop_path, sep="\t")
-
-    output_dict = defaultdict(list)
-    for trait, trait_trace in traces_dict.items():
-        variance_phy = np.mean(trait_trace) / 4
-        assert variance_phy > 0
-
-        assert f"{trait}_variance" in var_pop_df.columns
-        notna = np.isfinite(var_pop_df[f"{trait}_variance"])
-        if f"{trait}_heritability" not in var_pop_df.columns:
-            print(f"Warning: column {f'{trait}_heritability'} not found in {variance_pop_path}.")
-            print("Assuming heritability = 1.0.")
-            var_pop_df[f"{trait}_heritability"] = 1.0
-        # Computing the genetic variance (geno = h² * pheno)
-        genetic_variance_array = var_pop_df[f"{trait}_heritability"][notna] * var_pop_df[f"{trait}_variance"][notna]
-        variance_pop_array = genetic_variance_array / (var_pop_df["pS"][notna])
-        print(f'Found {len(genetic_variance_array)} species with a variance for {trait}.')
-        variance_pop = np.mean(variance_pop_array)
-        assert variance_pop > 0
-
-        ratio = variance_phy / variance_pop
-        ratio_pv = np.mean(trait_trace > variance_pop)
-        print(f"variance_phy = {variance_phy}")
-        print(f"variance_pop = {variance_pop}.")
-        print(f"ratio = {ratio}.")
-        output_dict["trait"].append(trait)
[…]

### 2023-05-23 scripts/libraries.py

@@ -7,8 +7,8 @@ mpl.use('Agg')
 hist_filled = {'alpha': 0.3, 'histtype': 'stepfilled'}
 hist_step = {'histtype': 'step'}
 my_dpi = 128
-fontsize = 14
-fontsize_legend = 12
+fontsize = 24
+fontsize_legend = 18
 
 
 def colors(x):
@@ -85,7 +85,7 @@ def hist_plot(x, x_label, output, nbr_genes, title="", xscale="log"):
         bins = np.geomspace(min_x, max_x, 100)
     else:
         min_x = np.min([min(i) for i in x.values()])
-        bins = np.linspace(min_x, max_x, 1000)
+        bins = np.linspace(min_x, max_x, 100)
     hist, _, _ = ax.hist(x.values(), bins=bins, color=color_models, **hist_filled)
     hist, _, _ = ax.hist(x.values(), bins=bins, color=color_models, **hist_step)
     max_y = 1.2 * (max([max(h) for h in hist]) if len(x) > 1 else max(hist))
@@ -103,6 +103,11 @@ def hist_plot(x, x_label, output, nbr_genes, title="", xscale="log"):
     else:
         ax.set_xlim((-0.01, 1.01))
         ax.set_yscale("log")
+    # Change x and y ticks font size
+    for tick in ax.xaxis.get_major_ticks():
+        tick.label.set_fontsize(fontsize)
+    for tick in ax.yaxis.get_major_ticks():
+        tick.label.set_fontsize(fontsize)
     ax.set_ylim((0, max_y))
     ax.set_ylabel("Density", fontsize=fontsize)
     ax.legend(fontsize=fontsize_legend)

### 2023-05-23 scripts/plot_simulations_bayesian.py

@@ -105,18 +105,18 @@ def main(folder, tsv_path, burn_in, output):
     rename = lambda x: output.replace(".tsv.gz", x)
     hist_plot(var_dict["phy_pop"], "$\\frac{\\sigma_{B}^2}{\\sigma_{W}^2}$", rename(".pdf"), nb_genes,
               'Var inter divided by Var intra')
-
     hist_plot(var_dict["phy_pop_pv"], "$P ( \\sigma_{B}^2 > \\sigma_{W}^2 )$", rename(".pvalues.pdf"), nb_genes,
               'Var inter divided by Var intra', xscale="linear")
 
-    scatter_plot(var_dict["mut"], var_dict["phy"], "mut", "phy", rename(".mutphy.pdf"), nb_genes,
-                 title=' - Vm versus Vd', histy_log=True)
-
-    scatter_plot(var_dict["mut"], var_dict["pop"], "mut", "pop", rename(".mutpop.pdf"), nb_genes,
-                 title=' - Vm versus Vg', histy_log=True)
-
-    scatter_plot(var_dict["pop"], var_dict["phy"], "pop", "phy", rename(".popphy.pdf"), nb_genes,
-                 title=' - Vg versus Vd', histy_log=True)
+    mut_str = r"Variance mutational $\left( \sigma^2_{M} \right)$"
+    within_str = r"Variance within $\left( \sigma^2_{W} \right)$"
+    between_str = r"Variance between $\left( \sigma^2_{B} \right)$"
+    scatter_plot(var_dict["mut"], var_dict["phy"], mut_str, between_str, rename(".mutphy.pdf"), nb_genes,
+                 title='', histy_log=True)
+    scatter_plot(var_dict["mut"], var_dict["pop"], mut_str, within_str, rename(".mutpop.pdf"), nb_genes,
+                 title='', histy_log=True)
+    scatter_plot(var_dict["pop"], var_dict["phy"], within_str, between_str, rename(".popphy.pdf"), nb_genes,
+                 title='', histy_log=True)
 
 
 if __name__ == '__main__':

### 2023-05-23 scripts/pre_processed_traits_mammals.py

@@ -63,8 +63,12 @@ def main(path_input_traits, path_input_pS, path_input_dS, path_output_tree, path
     pS_df["species"] = pS_df["species"].apply(lambda x: x.replace(" ", "_"))
     print(f"The pS dataframe has {len(pS_df)} rows before filtering.")
     pS_df = pS_df[pS_df["species"].isin(set_taxa_names)]
-    pS_col = "pS" if "pS" in pS_df.columns else "heterozygosity"
-    pS_df = pS_df[np.isfinite(pS_df[pS_col]) & (pS_df[pS_col] > 0)]
+    pS_col_list = ["heterozygosity", "pS", 'Hzoo']
+    for pS_col in pS_col_list:
+        if pS_col in pS_df.columns:
+            print(f"Keeping only species with available '{pS_col}'.")
+            pS_df = pS_df[np.isfinite(pS_df[pS_col]) & (pS_df[pS_col] > 0)]
+            break
     print(f"The pS dataframe has {len(pS_df)} rows after filtering taxon name and available pS.")
     assert len(pS_df) >= 5, "Not enough species with pS. Exiting."
 

### 2023-05-23 manuscript/main.tex
Updated manuscript

@@
 \definecolor{DARKBLUELINK}{HTML}{0B0080}
+\definecolor{LIGHTGREY}{gray}{0.9}
 \PassOptionsToPackage{hyphens}{url}
@@
 \usepackage{lineno}
-\renewcommand{\baselinestretch}{1.25}
-\renewcommand{\arraystretch}{1.6}
+\usepackage{tabu}
+\usepackage{hhline}
+\usepackage{multicol,multirow,array}
+\usepackage{etoolbox}
+\usepackage{booktabs}
+\usepackage{makecell}
+\usepackage{orcidlink}
 
@@
 
-\title{Testing for neutrality of a trait from within and between species variations}
+\renewcommand{\baselinestretch}{1.5}
+\renewcommand{\arraystretch}{1.2}
+\linenumbers
+\frenchspacing
+
+\title{A test of diversifying selection for a trait from within and between species variations}
+\rhead{\scshape Test of diversifying selection}
 
@@
 \large
-\textbf{T. {Latrille}$^{1}$, T. {Gaboriau}$^{1}$, N. {Salamin}$^{1}$}\\
+\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M. {Bastian}$^{2}$, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
 \normalsize
-$^{1}$Université de Lausanne, Lausanne, Switzerland\\
+$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
+$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
 \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
@@
 \begin{abstract}
-    To determine whether a trait is neutrally evolving or under selection, methods have been developed both at the population and at the phylogenetic scale.
-    At the population scale, methods do not integrate the changes over long evolutionary time.
-    Conversely, at the phylogenetic scale, current methods usually discards the variance in traits between individuals, leaving out valuable information.
-    The main goal of this project is to combine within-species and between-species variation of quantitative traits to determine the evolutionary regime of a trait.
-    We define a neutrality index for a quantitative trait, leveraging both data at the population and phylogenetic scale.
-    Moreover, this index leverages sequence divergence and polymorphism to normalize the variation of trait between and within species.
-    Our neutrality index equals to one for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
-    If our neutrality index is lower than one for traits putatively under stabilizing selection, and conversely greater than one for trait putatively under diversifying selection.
-    Finally, we show that our test is not sensitive to the assumption that population sizes and mutation rates are constant across the phylogeny, and automatically adjust for it.
+    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
+    However, using within species variation, methods do not integrate the changes over long evolutionary time.
+    Conversely, using between species variation, current methods usually discards the variance in traits between individuals, leaving out valuable information.
+    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
+    This neutrality index this does not require estimation of population size nor of time of speciation for normalization since it also leverages sequence polymorphism and divergence.
+    Our index can be used to seek deviation from the null model of neutral evolution, and is greater than one for a trait under diversifying selection. 
+    As an example, we show that brain mass at the mammalian scale is under diversifying selection. 
+    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
 \end{abstract}
@@
 All these scales requires different assumptions and methodology, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods and developments across different fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
+
 At the population level across individuals, Genome Wide Association Studies (GWAS) in humans have revealed that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
-At the species level, by contrasting variation within and between different populations, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes\cite{whitehead_neutral_2006, gilad_natural_2006, gilad_expression_2006}.
+At the species level, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait\cite{martin_multivariate_2008, leinonen_qst_2013}
+For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations\cite{whitehead_neutral_2006, gilad_natural_2006}.
 However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated\cite{signor_evolution_2018, price_detecting_2022}.
-For a given population, by contrasting to the divergence to a close species, studies have confirms that changes in gene expression accumulate linearly differences with time from divergence\cite{khaitovich_neutral_2004}.
-However, studies including variation across individuals and population typically include species divergence to a close species in a pairwise comparison, and do not integrate the changes across the whole phylogenetic tree.
-% This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
-% However, it has not yet been generalized for any number of species related by their phylogenetic tree.
-% However, this divergence in the level of expression stabilizes over a long distance: the difference between human and chicken (having diverged 300 million years) is similar to that between human and opossum (160 million years).
-% This observation suggests that the conservation of essential functions of organs could define a limit for the divergence of transcriptomes.
+Theoretically, change in traits accumulate linearly with time from divergence for a pair of species, and proportionally to the trait variance\cite{lande_genetic_1980, turelli_heritable_1984}.
+For example, gene expression accumulate linearly with time from divergence, and accumulates faster for trait with large variance\cite{khaitovich_neutral_2004}.
+Such theoretical connection between trait variance and change in mean value can be used to test for a selection of trait using for a pair of species\cite{walsh_evolution_2018}.
+However, to our knowledge, a neutral theoretical expectation for trait changes across several species related by their phylogenetic tree, as a function of within species trait variation has not yet been formulated and tested.
 
@@
 Contrarily, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale\cite{felsenstein_phylogenies_1985}.
-Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a variance proportional to the length of the branch\cite{hansen_translating_1996}.
+Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a standard deviation proportional to the squared root of the divergence time\cite{hansen_translating_1996}.
 As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
-Alternatively, a deviation from the Brownian process, typically an Ornstein-Uhlenbeck process is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
-However, studies have shown that comparative approaches are subject to different biases\cite{harmon_phylogenetic_2018}.
-First, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
-Indeed, a trait under selection for which the optimal trait value is evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral\cite{hansen_translating_1996}.
+A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
+Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
+However, studies have shown that such comparative approaches are subject to different biases\cite{harmon_phylogenetic_2018}.
+First, and unfortunately, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
+A trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral\cite{hansen_translating_1996}.
 Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Indeed, in such framework, the variation within species is discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
-Altogether, the classical comparative framework allows tests for neutrality at the phylogenetic scale but discards the variance in traits between individuals, leaving out valuable information.
+In such comparative framework, the variation within species is usually discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
+Altogether, methods at the phylogenetic scale easily mis-classify selection and also discards the variance in traits between individuals, leaving out valuable information.
 
@@
 However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
-In this study, we propose a neutrality index for a quantitative trait articulating trait variation at both the phylogenetic and population scale.
-Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales.
+Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
+In this study, we propose a neutrality index for a quantitative trait articulating trait and sequence variation within and between species.
+Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
+From a population-genetic perspective, our study can be seen as the generalization of Q$_\text{ST}$--F$_\text{ST}$ methods at the phylogenetic scale to account relationships between species.
+From a phylogenetic perspective, our study can be seen as a extension of the EVE model\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} where we set a clear threshold for neutral evolution by leveraging species' sequence polymorphism and divergence.
+% Basically, we are testing whether the variance of means is greater than the man of variances
 
@@
 New mutations are thus generating variance for such a trait.
-We define the expected effect of mutations on the variance of a trait as $\RateMut$, given as:
+At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contributes to the trait variance per generation.
+As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$), as:
 \begin{gather}
-    \RateMut \defEqual \NbrLoci \Multiply \E \left[ a^2 \right]. \label{eq:rate-mut}
-\end{gather}
-
-At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contributes to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of $\RateMut$ and the mutation rate per loci per generation ($\MutationRate$), as:
-\begin{gather}
-    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
+    \VarMutation = 2 \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:var-mutation}.
 \end{gather}
@@
 At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
-As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in the population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$) as:
+As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$) as:
 \begin{align}
     \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
-    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
+    & = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
 \end{align}
@@
 
-Altogether, at the population level, we define define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$), which identify with $\RateMut$:
+Altogether, we define define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$).
+This ratio allows to the discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePop$ as:
 \begin{align}
     \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
-    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
-    & = \RateMut. \label{eq:rate-pop}
+    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right]}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
+    & = \NbrLoci \Multiply \E \left[ a^2 \right]. \label{eq:rate-pop}
 \end{align}
@@
 
-Along a lineage, we denote $\Trait$ as the mean value of the trait over the individuals in the population.
+Along a lineage, we denote $\Trait$ as the mean value of the trait over the individuals in the species.
 If the trait is neutral and encoded by many loci, $\Trait$ evolves as Brownian process\cite{hansen_translating_1996}.
-After the population has evolved for $\Time$ generations, the variance of $\Trait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
+After the species has evolved for $\Time$ generations, the variance of $\Trait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
 \begin{align}
     \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
-    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+    & = 4 \Time \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
 \end{align}
 
-Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated in the populations, some will we eventually reach fixation in the population due to genetic drift, resulting in a substitution.
+Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will we eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
 For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
@@
 \end{align}
-That is, if mutations are neutral, the rate with which a new neutral allele is fixed in the population in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.
+That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.
 
@@
 \end{align}
-Altogether, the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$) also identify with $\RateMut$:
+Altogether, we defined $\RatePhy$ as the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$).
+This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePhy$ as:
 \begin{align}
     \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
-    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
-    & = \RateMut \label{eq:rate-phy}.
+    & = \frac{4 \Time \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right]}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
+    & = \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:rate-phy}.
 \end{align}
@@
         In red, at the phylogenetic scale, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
-        In blue, at the population scale, trait variance within species ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
+        In blue, trait variance within species ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
         Under neutral evolution $\RatePhy$ should equal $\RatePop$.
@@
  At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
-Combined as a ratio, we obtain $\RatePop$ at the population level, which is a proxy for $\RateMut$, namely the expected effect of a mutation on the variance of a trait.
-At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allow to compute $d$.
-Similarly, combined as a ratio, we obtain $\RatePhy$ which is another proxy for $\RateMut$.
+Combined as a ratio, we obtain $\RatePop$ at the population level, which depends solely on the genetic architecture of the trait (as $\NbrLoci \Multiply \E \left[ a^2 \right]$).
+At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
+Similarly, combined as a ratio, we obtain $\RatePhy$ at the phylogenetic scale, which also depends solely on the genetic architecture of the trait.
 Altogether, we thus have:
-\begin{align}
-    \RateMut & = \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}} , \\
-    \Rightarrow \frac{\RatePhy}{\RatePop} & = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
-\end{align}
+\begin{gather}
+    \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}} , \\
+    \Rightarrow \frac{\RatePhy}{\RatePop} = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
+\end{gather}
 Altogether, $\RatePhy / \RatePop$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
-Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between populations, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.
+Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.
 
@@
 \subsubsection{Maximum likelihood estimate}
-For each population of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences.
-Averaged across all populations we obtain the mean as:
+For each species of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences.
+Averaged across all species we obtain the mean as:
 \begin{gather}
@@
 \subsection{Simulation}\label{subsec:simulations}
-The evolutionary dynamics was formalized as a Wright-Fisher model with mutation, selection and drift.
-The population is assumed to be panmictic, with {effective population size} $\Ne$ and with non-overlapping generations.
-We simulate explicitly each generation along the phylogeny under a Wright-Fisher population, consisting of three steps: mutation, selection and {genetic drift} of currently segregating alleles.
-Mutations are drawn based on a user-defined nucleotide matrix, where our simulations used a symmetric time-reversible mutation matrix.
-Drift is induced by the multinomial resampling of the currently segregating alleles.
-Our simulator use a geometric Brownian multivariate process to model the changes in the mutation rate per generation, the generation time and $\Ne$ along the lineages.
-Moreover, the instant value of log-$\Ne$ can also be modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
+
+We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing ly to the genotypic value, these effect are Gamma distributed with mean $a$ (Fig.~\ref{fig:simulator}).
+We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
+Assuming a panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
+Mutations will be based on a user-defined nucleotide matrix, while drift will be modelled by the multinomial resampling of the currently segregating alleles.
+The changes in the mutation rate per generation, the generation time and $\Ne$ along the lineages will be modelled by a geometric Brownian multivariate and an Ornstein-Uhlenbeck process to test different processes of evolution of these parameters.
+The instant value of log-$\Ne$ can also be modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
 The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
-In our simulations, the tree is composed of 77 species (see supplementary), the tree root is $150$ million years old, the initial mutation rate is $\smash{10^{-8}}$ per site per generation and the initial generation time is $10$ years.
-The simulator starts from an initial sequence at equilibrium, at each node, the process is split, and finally stopped at the leaves of the tree.
-
-We introduced selection as a one dimensional Fisher's geometric landscape~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
-The phenotype is a vector in a multidimensional space, where the number of dimensions is often termed complexity.
-From a {phenotype}, the fitness is a monotonously decreasing function of the {phenotype} distance to $0$.
-The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength ($\alpha$) and epistasis ($\beta$).
-If the phenotype-fitness map is explicit, the genotype-phenotype map is more pervasive.
-Mutations are seen has displacement of the {phenotype} in the multidimensional space.
-Beneficial mutations are moving the {phenotype} closer to $0$, whereas deleterious mutations are moving the {phenotype} further away.
-In such original context, the distribution of mutational effects is not dependent on the current genotype, but this can be relaxed using a genotype-phenotype map.
+The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
+We will introduce selection as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype \cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
+Mutations are seen has displacement of the phenotype in the multidimensional space.
+Beneficial mutations will move the phenotype closer to the optimum, while deleterious mutations will move it further away.
+Stabilizing selection will be implemented by means of a fixed optimum phenotype, while diversifying selection will be implemented by means of a moving optimum phenotype, changing along the phylogeny.
 
@@
 
-To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at the population and phylogenetic scale.
+To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at different scales.
 The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
@@
 
-\begin{table*}[tb]
+\begin{table*}[t!]
     \centering
-    \begin{adjustbox}{width = 0.75\textwidth}
+    \begin{adjustbox}{width = 0.5\textwidth}
     \begin{tabular}{|c||c|c|}
@@
             & Coding sequences & Quantitative traits \\ \hline \hline
-            Adaptation or moving optimum       & $\frac{\dn \Multiply \ps}{\ds \Multiply \pn} > 1$ & $ \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d } > 1 $ \\ \hline
-            Neutral regime of evolution        & $\frac{\dn \Multiply \ps}{\ds \Multiply \pn} = 1$ & $ \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d } = 1 $ \\ \hline
-            Purifying or directional selection & $\frac{\dn \Multiply \ps}{\ds \Multiply \pn} < 1$ & $\frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d } < 1 $ \\ \hline
+            Adaptation / Diversifying selection & $\dnds > \pnps$ & $ \RatePhy > \RatePop  $ \\ \hline
+            Neutral regime of evolution & $\dnds = \pnps$ & $ \RatePhy = \RatePop  $ \\ \hline
+            Purifying / Stabilizing selection & $\dnds < \pnps$ & $ \RatePhy < \RatePop  $ \\ \hline
     \end{tabular}
@@
     \caption{
-        $\RatePhy$ estimated at the phylogenetic scale as a function of $\RatePop$ estimated at the population scale, for $30.000$ genes simulated under different evolutionary regimes.
+        $\RatePhy / \RatePop$ for $30.000$ genes simulated under different evolutionary regimes.
         $\NIx < 1$ for traits simulated under selection (stabilizing selection in yellow).
@@
 At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
-Similarly, at the population scale, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
+Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
 Our neutrality index for a trait is the ratio of between species variation to within species variation.
@@
 
+% What about the other methods? What are we testing for?
 At the phylogenetic scale, the change in mean trait value across different species is typically used to test for neutrality for a quantitative trait.
@@
 
+% What do we expect for biological traits
+% Problem for slow evolving diversifying selection
+% Modelling the mechantistic expectation Stabilizting selection with a moving optimum. 
+% How to include Vm?
+
 % Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
 More theoretically, our development is based on several assumption that could be relaxed.
-First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
-Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+First, $\NbrLoci \Multiply \E \left[ a^2 \right]$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
+Having a constant $\NbrLoci \Multiply \E \left[ a^2 \right]$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\NbrLoci \Multiply \E \left[ a^2 \right](t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 Moreover, even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
-Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework\cite{huelsenbeck_detecting_2003,  lartillot_phylogenetic_2011,lartillot_joint_2012}.
-Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within populations in a unified framework.
+Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012}.
+Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within species in a unified framework.
 
@@
 Code: T.L.;
-Data analyses: T.L.;
+Data analyses: T.L.\ and M.B.;
 Interpretation: T.L., T.G\ and N.S.;
 First draft: T.L.;
-Editing and revisions: T.L., T.G\ and N.S.;
+Editing and revisions: T.L., M.B, T.G\ and N.S.;
 Project management and funding: N.S\@.

### 2023-06-06 manuscript/main.tex
Updated manuscript

@@
 Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
-Fundamentally, distinguishing neutral evolution from selection requires to determine which mode of selection is supported from the observed variation of traits and sequences.
-This variation of traits can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these scales requires different assumptions and methodology, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods and developments across different fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
+Fundamentally, distinguishing neutral evolution from selection requires determining which mode of selection is supported by the observed variation of traits and sequences.
+Trait variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
+All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
 
@@
 New mutations are thus generating variance for such a trait.
-At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contributes to the trait variance per generation.
+At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
 As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$), as:
 \begin{gather}
-    \VarMutation = 2 \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:var-mutation}.
+    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
 \end{gather}
@@
 At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
-As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$) as:
+As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
 \begin{align}
     \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
-    & = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
+    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
 \end{align}
@@
 Moreover, for any neutral genomic region of interest, the genetic diversity, called $\pi$, is measured as the number of mutation segregating in the population divided by the size of the region.
-Any of the segregating mutation will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
+Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
 As shown in \textcite{tajima_statistical_1989}, $\pi$ is thus a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$), given as:
@@
 
-Altogether, we define define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$).
-This ratio allows to the discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePop$ as:
+Altogether, we define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$).
+This ratio allows discarding the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePop$ as:
 \begin{align}
     \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
-    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right]}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
-    & = \NbrLoci \Multiply \E \left[ a^2 \right]. \label{eq:rate-pop}
+    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
+    & = \RateMut. \label{eq:rate-pop}
 \end{align}
 
-Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($h^2$), giving also $\RatePop$ as:
+Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($\Heredity$), giving also $\RatePop$ as:
 \begin{align}
-    \RatePop & = \frac{h^2 \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
+    \RatePop & = \frac{\Heredity \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
 \end{align}
@@
 
-Along a lineage, we denote $\Trait$ as the mean value of the trait over the individuals in the species.
-If the trait is neutral and encoded by many loci, $\Trait$ evolves as Brownian process\cite{hansen_translating_1996}.
-After the species has evolved for $\Time$ generations, the variance of $\Trait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
+Along a lineage, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
+If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process\cite{hansen_translating_1996}.
+After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
 \begin{align}
     \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
-    & = 4 \Time \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
 \end{align}
 
-Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will we eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
+Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
 For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
@@
     \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
-    & = \frac{4 \Time \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right]}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
-    & = \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:rate-phy}.
+    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
+    & = \RateMut \label{eq:rate-phy}.
 \end{align}
@@
  At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
-Combined as a ratio, we obtain $\RatePop$ at the population level, which depends solely on the genetic architecture of the trait (as $\NbrLoci \Multiply \E \left[ a^2 \right]$).
+Combined as a ratio, we obtain $\RatePop$ at the population level, which depends solely on the genetic architecture of the trait (as $\RateMut$).
 At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
@@
 \subsubsection{Maximum likelihood estimate}
-For each species of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences.
-Averaged across all species we obtain the mean as:
-\begin{gather}
-    \EstRatePop = \left< \frac{\VarGenetic}{\pi} \right> = \left< \frac{h^2 \Multiply \VarPhenotype }{\pi } \right> \label{eq:estimated-rate-pop}
-\end{gather}
-
 At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
@@
 \begin{gather}
-    \RootTrait = \left( \VecOne\tr \Cr\inv \VecOne \right)\inv \left( \VecOne\tr \Cr\inv \VecTrait \right), \label{eq:estimated-root-trait}
+    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
 \end{gather}
@@
 \begin{gather}
-    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \VecOne \right)\tr \Cr\inv \left( \VecTrait -  \RootTrait  \VecOne  \right)}{\NbrSpecies - 1}. \label{eq:estimated-rate-phy}
+    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrSpecies - 1}.
+\label{eq:estimated-rate-phy}
 \end{gather}
+
+For a species $i$ of interest with inter-individual data available, $\RatePopSpi$ can be estimated as the ratio of genetic variance of a trait ($\VarGeneticSpi = \VarPhenotypeSpi \Multiply \HereditySpi$) over genetic diversity of neutrally evolving sequences ($\pi_{\Spi}$).
+Averaged across all species, we obtain the mean as:
+\begin{gather}
+    \EstRatePop = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies}\RatePopSpi = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies} \frac{  V_{\Trait, i} \Multiply \Heredity_{i}}{ \pi_{i}}.
+\label{eq:estimated-rate-pop}
+\end{gather}
+
 Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
+
+\subsubsection{Bayesian estimate}
+We used the software Bayescode
 
@@
 
-We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing ly to the genotypic value, these effect are Gamma distributed with mean $a$ (Fig.~\ref{fig:simulator}).
+We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator}).
 We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
-Assuming a panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
-Mutations will be based on a user-defined nucleotide matrix, while drift will be modelled by the multinomial resampling of the currently segregating alleles.
-The changes in the mutation rate per generation, the generation time and $\Ne$ along the lineages will be modelled by a geometric Brownian multivariate and an Ornstein-Uhlenbeck process to test different processes of evolution of these parameters.
-The instant value of log-$\Ne$ can also be modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
+Assuming a diploid panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
+Genetic drift is modelled by the multinomial resampling of the currently segregating alleles.
+The number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $.
+The changes in the mutation rate per generation ($\MutationRate$) and $\Ne$ along the lineages is modelled by a geometric Brownian process.
+Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
 The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
 The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
-We will introduce selection as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype \cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
+Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype \cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
 The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
 Mutations are seen has displacement of the phenotype in the multidimensional space.
-Beneficial mutations will move the phenotype closer to the optimum, while deleterious mutations will move it further away.
-Stabilizing selection will be implemented by means of a fixed optimum phenotype, while diversifying selection will be implemented by means of a moving optimum phenotype, changing along the phylogeny.
+Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
+Stabilizing selection is implemented by means of a fixed optimum phenotype.
+Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.
 
@@
     \centering
-    \begin{adjustbox}{width = 0.5\textwidth}
-    \begin{tabular}{|c||c|c|}
+    \begin{adjustbox}{width = 1.0\textwidth}
+    \begin{tabular}{||l|l||c|c||c|c||}
             \hline
-            & Coding sequences & Quantitative traits \\ \hline \hline
-            Adaptation / Diversifying selection & $\dnds > \pnps$ & $ \RatePhy > \RatePop  $ \\ \hline
-            Neutral regime of evolution & $\dnds = \pnps$ & $ \RatePhy = \RatePop  $ \\ \hline
-            Purifying / Stabilizing selection & $\dnds < \pnps$ & $ \RatePhy < \RatePop  $ \\ \hline
+            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NIx > 1$  & Test $\NIx < 1$ \\ \hline \hline
+            Changing $\MutationRate$ & Scaling with $\pi$, $d$, $\VarPhenotype$ and $\VarPhy$ & Adjusted & Adjusted & -- & --  \\ \hline
+            Changing $\Ne$ & Scaling with $\pi$ and $\VarPhenotype$ & Adjusted & -- & -- & --  \\ \hline
+            Unknown heredity & $\Heredity = 1$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
+            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
+            Polymorphism are selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
+            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
+            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & False positives  \\ \hline
+            Divergence are selected & $d$ is under-estimated & -- & Over-estimated & False positives & Conservative  \\ \hline
+            Structured population & unknown & unknown & unknown & unknown & unknown \\ \hline
+            Gene flow between species & unknown & unknown & unknown & unknown & unknown \\ \hline
     \end{tabular}
     \end{adjustbox}
-    \caption{
-        The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
-        Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
-    }
-    \label{table:unfolded-MK}
+    \caption{Assumptions and their consequences on deviation from neutraliity.}
+    \label{table:assumptions}
 \end{table*}
@@
 Analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
+The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
+Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
 Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
@@
 Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
-Contrarily, we argue that the main drawback of our method is that the heritability ($h^2$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
+Contrarily, we argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
 
@@
 With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
-Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$), as well as estimate for heritabilities ($h^2$).
+Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$), as well as estimate for heritabilities ($\Heredity$).
 
@@
 More theoretically, our development is based on several assumption that could be relaxed.
-First, $\NbrLoci \Multiply \E \left[ a^2 \right]$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
-Having a constant $\NbrLoci \Multiply \E \left[ a^2 \right]$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\NbrLoci \Multiply \E \left[ a^2 \right](t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
+Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 Moreover, even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
@@
 
+\part*{Supplementary materials}
+
+\section{Theoretical derivation}
+\subsection{Definitions}
+
+We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $i$.
+
+\begin{align}
+    \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
+\end{align}
+
+\subsection{Simulator}
+\begin{itemize}
+    \setlength\itemsep{-0.2em}
+    \item $\NbrLoci$ is the number of loci encoding the trait.
+    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
+    \item $\Ne$ is the effective number of individual in the population.
+    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotype at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
+    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the influence of genotype on the trait for individual $i$.
+\end{itemize}
+The mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
+\begin{equation}
+    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad}	\VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2
+\end{equation}
+\begin{itemize}
+    \setlength\itemsep{-0.2em}
+    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
+    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
+\end{itemize}
+The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
+\begin{equation}
+    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2
+\end{equation}
+Heritability ($h^2$) is defined as:
+\begin{equation}
+    h^2 = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}
+\end{equation}
+\begin{gather}
+    G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} \\
+    \Trait_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} + \xi_i
+\end{gather}
+
+\newpage
+
+\begin{itemize}
+    \item $W_i$ is the weight for individual $i$ to be chosen for the next generation.
+    \item Neutral model with no fitness function: $W_i = 1$.
+    \item Stabilizing selection with fitness $W_i = \e^{(\Trait_i - \theta)^2/ \sigma}$, parameters $\theta=0$ and $\sigma=0.02$.
+    \item Moving optimum with fitness $W_i = \e^{(\Trait_i - \theta_t)^2/ \sigma}$, parameters $\theta_t \sim \mathcal{B}_t$ and $\sigma=0.02$.
+\end{itemize}
+\begin{itemize}
+    \item Trait encoded by $5,000$ loci, $1000$ replicates.
+    \item $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
+\end{itemize}
+
+\begin{table*}[t!]
+    \centering
+    \begin{adjustbox}{width = 0.5\textwidth}
+        \begin{tabular}{|c||c|c|}
+            \hline
+            & Coding sequences & Quantitative traits \\ \hline \hline
+            Adaptation / Diversifying selection & $\dnds > \pnps$ & $ \RatePhy > \RatePop  $ \\ \hline
+            Neutral regime of evolution & $\dnds = \pnps$ & $ \RatePhy = \RatePop  $ \\ \hline
+            Purifying / Stabilizing selection & $\dnds < \pnps$ & $ \RatePhy < \RatePop  $ \\ \hline
+        \end{tabular}
+    \end{adjustbox}
+    \caption{
+        The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
+        Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
+    }
+    \label{table:dnds}
+\end{table*}
+
 \end{document}

### 2023-06-06 scripts/pre_processed_traits_mammals.py
Added primate dataset

@@ -7,12 +7,14 @@ from ete3 import Tree
 from neutrality_index import open_tree, prune_tree
 
 
-
 def rename_tree(tree: Tree) -> Tree:
     for node in tree.traverse("levelorder"):
         if node.is_leaf():
             if node.name.endswith("_E"):
                 node.name = "_".join(node.name.replace("_E", "").split("_")[:-1])
+            elif "PD_" in node.name[:3]:
+                node.name = "_".join(node.name.split("_")[2:])
+                assert node.name.count("_") == 1
             else:
                 node.name = node.name.replace(" ", "_")
     return tree
@@ -59,12 +61,14 @@ def main(path_input_traits, path_input_pS, path_input_dS, path_output_tree, path
         assert len(pS_df) == n_taxa
     else:
         pS_df = pd.read_csv(path_input_pS, sep=",")
+        if "SPECIES_BINOMIAL" in pS_df.columns:
+            pS_df["species"] = pS_df["SPECIES_BINOMIAL"]
 
     # Remove all spaces in the specie name
     pS_df["species"] = pS_df["species"].apply(lambda x: x.replace(" ", "_"))
     print(f"The pS dataframe has {len(pS_df)} rows before filtering.")
     pS_df = pS_df[pS_df["species"].isin(set_taxa_names)]
-    pS_col = [p for p in ["heterozygosity", "pS", 'Hzoo'] if p in pS_df.columns][0]
+    pS_col = [p for p in ["MEDIAN_HETEROZYGOSITY", "heterozygosity", "pS", 'Hzoo'] if p in pS_df.columns][0]
     print(f"Keeping only species with available '{pS_col}'.")
     pS_df = pS_df[np.isfinite(pS_df[pS_col]) & (pS_df[pS_col] > 0)]
     print(f"The pS dataframe has {len(pS_df)} rows after filtering taxon name and available pS.")
@@ -124,29 +128,21 @@ def main(path_input_traits, path_input_pS, path_input_dS, path_output_tree, path
 
         # Filter out the species with a unique row in the trait dataframe
         var_df = trait_df.copy()
-        var_df = var_df[var_df["Sample size (Brain)"] == 1]
+        var_df = var_df[var_df[ss_brain] == 1]
         print(f"The trait dataframe has {len(var_df)} rows after filtering for Nsize != 1.")
         var_grouped = {k: v for k, v in var_df.groupby("Genus_Species") if len(v) > 1}
[…]

### 2023-06-06 scripts/scale_tree.py

@@ -1,0 +1,50 @@
+import argparse
+from neutrality_index import open_tree, prune_tree
+from pre_processed_traits_mammals import rename_tree
+
+
+def main(primate_tree, mammal_neutral_tree, output_tree):
+    """
+    Because the primate tree is not based on neutral markers, we scaled it so that the total branch length of the
+    primate tree and the mammal tree (based on neutral markers) are the same when both trees are pruned to the same
+    intersection of taxa.
+    :param primate_tree:
+    :param mammal_neutral_tree:
+    :param output_tree:
+    :return:
+    """
+    t_primate = rename_tree(open_tree(primate_tree, format_ete3=1))
+    discard_list = ["Galeopterus_variegatus", "Tupaia_belangeri", "Oryctolagus_cuniculus", "Mus_musculus"]
+    keep_species = list(set(t_primate.get_leaf_names()).difference(set(discard_list)))
+    t_primate = prune_tree(t_primate, keep_species)
+    t_mammal = rename_tree(open_tree(mammal_neutral_tree, format_ete3=1))
+    intersection = list(set(t_primate.get_leaf_names()).intersection(set(t_mammal.get_leaf_names())))
+    print(f"Intersection between mammals and primates has {len(intersection)} leaves")
+    pruned_primate = prune_tree(t_primate, intersection)
+    t_mammal = prune_tree(t_mammal, intersection)
+    assert set(pruned_primate.get_leaf_names()) == set(t_mammal.get_leaf_names())
+    total_mammal_distance = sum([n.dist for n in t_mammal.traverse()])
+    total_primate_distance = sum([n.dist for n in pruned_primate.traverse()])
+    ratio = total_mammal_distance / total_primate_distance
+    print(f"Total mammal distance: {total_mammal_distance}")
+    print(f"Total primate distance: {total_primate_distance}")
+    print(f"Ratio: {ratio}")
+    # Scale the primate tree
+    for n in t_primate.traverse():
+        n.dist = n.dist * ratio
+    # Scale the pruned primate tree
+    for n in pruned_primate.traverse():
+        n.dist = n.dist * ratio
+
+    distance_scaled = sum([n.dist for n in pruned_primate.traverse()])
+    assert abs(distance_scaled - total_mammal_distance) < 1e-6, f"{distance_scaled} != {total_mammal_distance}"
+    t_primate.write(outfile=output_tree, format=1)
+
+
+if __name__ == '__main__':
[…]

### 2023-06-07 scripts/libraries.py
Scale the sigma for the Brownian process

@@ -16,15 +16,14 @@ def colors(x):
     return [cs[idx % len(cs)] for idx in range(len(x))]
 
 
-def scatter_plot(x, y, x_label, y_label, output, nbr_genes, title="", histy_log=False, loglog=False):
+def scatter_plot(x_input, y_input, x_label, y_label, output, histy_log=False, loglog=False):
+    x = {k: np.array(v) for k, v in x_input.items()}
+    y = {k: np.array(v) for k, v in y_input.items()}
+    xy_filtered = {k: (np.isfinite(x[k]) & np.isfinite(y[k]) & (x[k] > 0) & (y[k] > 0)) for k in x.keys()}
+    x = {k: x[k][xy_filtered[k]] for k in x.keys()}
+    y = {k: y[k][xy_filtered[k]] for k in y.keys()}
     color_models = colors(x)
-    fig = plt.figure(figsize=(1920 / my_dpi, 880 / my_dpi), dpi=my_dpi)
-    set_nb_points = set([len(nb) for nb in x.values()])
-    assert len(set_nb_points) == 1
-    nb_points = set_nb_points.pop()
-    assert nb_points == nbr_genes
-    fig.suptitle("{0} genes".format(nbr_genes) + title)
-
+    fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
     gs = fig.add_gridspec(2, 2, width_ratios=(7, 2), height_ratios=(2, 7),
                           left=0.1, right=0.9, bottom=0.1, top=0.9,
                           wspace=0.05, hspace=0.05)
@@ -65,30 +64,30 @@ def scatter_plot(x, y, x_label, y_label, output, nbr_genes, title="", histy_log=
         ax.set_xlim((0.95 * min_x, max_x * 1.05))
         ax.set_ylim((0.95 * min_y, max_y * 1.05))
     ax.legend(fontsize=fontsize_legend)
-    plt.tight_layout()
     plt.savefig(output, format="pdf")
     plt.savefig(output.replace(".pdf", ".png"), format="png")
     plt.close('all')
     print(output)
 
 
-def hist_plot(x, x_label, output, nbr_genes, title="", xscale="log"):
+def hist_plot(x_input, x_label, output, xscale="log"):
+    x = {k: np.array(v) for k, v in x_input.items()}
+    xy_filtered = {k: (np.isfinite(x[k]) & (x[k] >= 0)) for k in x.keys()}
[…]

### 2023-06-07 scripts/plot_simulations_ML.py

@@ -36,7 +36,8 @@ def main(folder, neutral_tree_path, output):
 
             leaves_value = defaultdict(list)
             for n, n_neutral in zip(tree.get_leaves(), neutral_tree.get_leaves()):
-                vg = float(getattr(n, "Phenotype_var")) * float(getattr(n, "Heritability"))
+                # vg = float(getattr(n, "Phenotype_var")) * float(getattr(n, "Heritability"))
+                vg = float(getattr(n, "Genotype_var"))
                 leaves_value['within_tajima'].append(vg / float(getattr(n_neutral, "Theta_Tajima")))
                 leaves_value['within_watterson'].append(vg / float(getattr(n_neutral, "Theta_Watterson")))
                 leaves_value["within_fay_wu"].append(vg / float(getattr(n_neutral, "Theta_Fay_Wu")))
@@ -56,26 +57,25 @@ def main(folder, neutral_tree_path, output):
     df = pd.concat(dict_of_df, axis=1)
     # to break out the lists into columns
     df.to_csv(output, sep="\t", index=False)
-    nb_genes = set([len(v) for v in var_dict["between"].values()]).pop()
-    x_str = r"Neutrality index $\left( \frac{\sigma^2_{B}}{\sigma^2_{W}} \right)$"
+    x_str = r"Neutrality index $\left( \rho \right)$"
     ratio_scaled = {m: np.array(var_dict["between"][m]) / np.array(var_dict["within_sampled"][m]) for m in models}
-    hist_plot(ratio_scaled, x_str, output.replace(".tsv.gz", ".hist.pdf"), nb_genes)
+    hist_plot(ratio_scaled, x_str, output.replace(".tsv.gz", ".hist.pdf"))
 
     mut_str = r"Variance mutational $\left( \sigma^2_{M} \right)$"
     within_str = r"Variance within $\left( \sigma^2_{W} \right)$"
     between_str = r"Variance between $\left( \sigma^2_{B} \right)$"
     scatter_plot(var_dict["within_sampled"], var_dict["between"], within_str, between_str,
-                 output.replace(".tsv.gz", "-H.pdf"), nb_genes, title=' - Sample theta', histy_log=True)
+                 output.replace(".tsv.gz", "-H.pdf"), histy_log=True)
     scatter_plot(var_dict["within_tajima"], var_dict["between"], within_str, between_str,
[…]

### 2023-06-07 scripts/plot_simulations_bayesian.py

@@ -87,10 +87,11 @@ def main(folder, tsv_path, burn_in, output):
             rep = os.path.basename(filepath).replace(".trace.gz", "").split("_")[-1]
             var_phy_trace = open_trace_file(filepath, burn_in=burn_in)
             var_phy = np.mean(var_phy_trace)
-            var_pop = abs(dict_tree[("within_sampled", m, rep)])
+            var_pop = dict_tree[("within_sampled", m, rep)]
+            if var_pop <= 0:
+                continue
             var_mut = dict_tree[("mutational", m, rep)]
             assert var_phy > 0
-            assert var_pop > 0
             assert var_mut > 0
             var_dict["phy"][m].append(var_phy)
             var_dict["pop"][m].append(var_pop)
@@ -101,22 +102,16 @@ def main(folder, tsv_path, burn_in, output):
 
     df = pd.DataFrame(var_dict)
     df.to_csv(output, sep="\t", index=False)
-    nb_genes = set([len(v) for v in var_dict["phy_pop"].values()]).pop()
     rename = lambda x: output.replace(".tsv.gz", x)
-    hist_plot(var_dict["phy_pop"], "$\\frac{\\sigma_{B}^2}{\\sigma_{W}^2}$", rename(".pdf"), nb_genes,
-              'Var inter divided by Var intra')
-    hist_plot(var_dict["phy_pop_pv"], "$P ( \\sigma_{B}^2 > \\sigma_{W}^2 )$", rename(".pvalues.pdf"), nb_genes,
-              'Var inter divided by Var intra', xscale="linear")
+    hist_plot(var_dict["phy_pop"], "$\\rho$", rename(".pdf"))
+    hist_plot(var_dict["phy_pop_pv"], "$P ( \\rho > 1 )$", rename(".pvalues.pdf"), xscale="linear")
 
     mut_str = r"Variance mutational $\left( \sigma^2_{M} \right)$"
     within_str = r"Variance within $\left( \sigma^2_{W} \right)$"
     between_str = r"Variance between $\left( \sigma^2_{B} \right)$"
-    scatter_plot(var_dict["mut"], var_dict["phy"], mut_str, between_str, rename(".mutphy.pdf"), nb_genes,
-                 title='', histy_log=True)
-    scatter_plot(var_dict["mut"], var_dict["pop"], mut_str, within_str, rename(".mutpop.pdf"), nb_genes,
-                 title='', histy_log=True)
[…]

### 2023-06-07 simulator/src/lib/wright_fisher.hpp

@@ -45,15 +45,22 @@ class MutationRateProcess {
           mutation_rate_max{mutation_rate_max},
           brownian_sigma{brownian_sigma} {
         brownian = log(mutation_rate);
-        brownian_min = log(mutation_rate_min);
-        brownian_max = log(mutation_rate_max);
+        if (brownian_sigma == 0.0) {
+            brownian_min = brownian;
+            brownian_max = brownian;
+        } else {
+            brownian_min = log(mutation_rate_min);
+            brownian_max = log(mutation_rate_max);
+        }
         normal_distrib = normal_distribution<double>(0.0, 1.0);
     }
 
     void update() {
-        brownian += brownian_sigma * normal_distrib(generator_mut_rate);
-        if (brownian < brownian_min) { brownian = 2 * brownian_min - brownian; }
-        if (brownian > brownian_max) { brownian = 2 * brownian_max - brownian; }
+        if (brownian_sigma != 0.0) {
+            brownian += brownian_sigma * normal_distrib(generator_mut_rate);
+            if (brownian < brownian_min) { brownian = 2 * brownian_min - brownian; }
+            if (brownian > brownian_max) { brownian = 2 * brownian_max - brownian; }
+        }
         mutation_rate = exp(brownian);
     }
     double get_mutation_rate() const {
@@ -74,13 +81,17 @@ class MutationRateArgParse {
         "mutation_rate_per_loci_per_generation", "Mutation rate per locus per generation", false,
         1e-5, "double", cmd};
     TCLAP::ValueArg<double> mutation_rate_min{
-        "", "mutation_rate_min", "Minimum mutation_rate", false, 1e-7, "double", cmd};
+        "", "mutation_rate_min", "Minimum mutation_rate", false, 0.0, "double", cmd};
     TCLAP::ValueArg<double> mutation_rate_max{
-        "", "mutation_rate_max", "Maximum mutation_rate", false, 1e-3, "double", cmd};
+        "", "mutation_rate_max", "Maximum mutation_rate", false, 1.0, "double", cmd};
     TCLAP::ValueArg<double> mutation_brownian_sigma{"", "mutation_brownian_sigma",
-        "The Brownian sigma (0<sigma) applied to μ at each generation", false, 0.05, "double", cmd};
+        "The Brownian sigma (0<sigma) applied to μ at each generation", false, 0.0, "double", cmd};
 
     MutationRateProcess get_model() {
+        assert(mutation_rate_min.getValue() <= mutation_rate_per_loci_per_generation.getValue());
[…]

### 2023-06-07 simulator/src/neutral.cpp

@@ -24,7 +24,7 @@ int main(int argc, char *argv[]) {
     tree.set_root_age(root_age);
 
     Trace trace(args.ouput_tsv.getValue());
-    population.run(args.burn_in.getValue(), trace);
+    population.run(args.burn_in.getValue(), trace, true);
     Process process(tree, population, trace);
     trace.write_tsv(args.output_path.getValue());
     tree.write_newick(args.output_path.getValue());

### 2023-06-07 simulator/src/selection.cpp

@@ -25,7 +25,7 @@ int main(int argc, char *argv[]) {
     tree.set_root_age(root_age);
 
     Trace trace(args.ouput_tsv.getValue());
-    population.run(args.burn_in.getValue(), trace);
+    population.run(args.burn_in.getValue(), trace, true);
     Process process(tree, population, trace);
     trace.write_tsv(args.output_path.getValue());
     tree.write_newick(args.output_path.getValue());

### 2023-06-07 simulator/src/lib/wright_fisher.hpp
Changed option 'mutation_brownian_sigma' to 'mutation_rate_brownian_sigma' in simulator

@@ -84,17 +84,17 @@ class MutationRateArgParse {
         "", "mutation_rate_min", "Minimum mutation_rate", false, 0.0, "double", cmd};
     TCLAP::ValueArg<double> mutation_rate_max{
         "", "mutation_rate_max", "Maximum mutation_rate", false, 1.0, "double", cmd};
-    TCLAP::ValueArg<double> mutation_brownian_sigma{"", "mutation_brownian_sigma",
+    TCLAP::ValueArg<double> mutation_rate_brownian_sigma{"", "mutation_rate_brownian_sigma",
         "The Brownian sigma (0<sigma) applied to μ at each generation", false, 0.0, "double", cmd};
 
     MutationRateProcess get_model() {
         assert(mutation_rate_min.getValue() <= mutation_rate_per_loci_per_generation.getValue());
         assert(mutation_rate_per_loci_per_generation.getValue() <= mutation_rate_max.getValue());
         assert(mutation_rate_per_loci_per_generation.getValue() >= 0.0);
-        assert(mutation_brownian_sigma.getValue() >= 0.0);
+        assert(mutation_rate_brownian_sigma.getValue() >= 0.0);
         return MutationRateProcess(mutation_rate_per_loci_per_generation.getValue(),
             mutation_rate_min.getValue(), mutation_rate_max.getValue(),
-            mutation_brownian_sigma.getValue());
+            mutation_rate_brownian_sigma.getValue());
     }
 };
 

### 2023-06-07 manuscript/main.tex
Updated manuscript

@@
 \documentclass{article}
-
+\usepackage{caption}
 \usepackage{xcolor, colortbl}
@@
 
-\input{notations.tex}
+\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
+\newcommand{\Multiply}{\cdot}
+\newcommand{\MultiplyMatrix}{\times}
+\newcommand{\UniDimArray}[1]{\bm{#1}}
+\newcommand{\BiDimArray}[1]{\bm{#1}}
+\newcommand{\tr}{^{\intercal}}
+\newcommand{\inv}{^{-1}}
+\DeclareMathOperator{\E}{\mathbb{E}}
+\DeclareMathOperator{\Var}{\text{Var}}
+\DeclareMathOperator{\Cov}{\text{Cov}}
+\newcommand{\der}{\mathrm{d}}
+\newcommand{\e}{\text{e}}
+\newcommand{\Ne}{N_{\text{e}}}
+\newcommand{\dn}{d_N}
+\newcommand{\ds}{d_S}
+\newcommand{\dnds}{\dn / \ds}
+\newcommand{\pn}{\pi_N}
+\newcommand{\ps}{\pi_S}
+\newcommand{\pnps}{\pn / \ps}
+\newcommand{\proba}{\mathbb{P}}
+\newcommand{\pfix}{\proba_{\text{fix}}}
+\newcommand{\Spi}{i}
+\newcommand{\Spj}{j}
+\newcommand{\NbrTaxa}{n}
+\newcommand{\Time}{t}
+\newcommand{\Trait}{P}
+\newcommand{\Heredity}{h^2}
+\newcommand{\HereditySpi}{\Heredity_{\Spi}}
+\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
+\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
+\newcommand{\Root}{0}
+\newcommand{\RootTrait}{\widehat{\theta}}
+\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
+\newcommand{\VecOne}{\UniDimArray{1}}
+\newcommand{\Distance}{\BiDimArray{D}}
+\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
+\newcommand{\MutationRate}{\mu}
+\newcommand{\SubRate}{q}
+\newcommand{\NbrLoci}{L}
+\newcommand{\VarPhenotype}{V_{\Trait}}
+\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
+\newcommand{\VarGenetic}{V_{\mathrm{G}}}
+\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
+\newcommand{\VarEnv}{V_{\mathrm{E}}}
+\newcommand{\MatrixGenetic}{\BiDimArray{G}}
+\newcommand{\VarMutation}{V_{\mathrm{M}}}
+\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
+\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
+\newcommand{\RatePhy}{\sigma^2_{\mathrm{B}}}
+\newcommand{\RatePop}{\sigma^2_{\mathrm{W}}}
+\newcommand{\RatePopSpi}{\sigma^2_{\mathrm{W}, \Spi}}
+\newcommand{\VecRatePop}{\UniDimArray{\RatePop}}
+\newcommand{\EstRatePhy}{\widehat{\RatePhy}}
+\newcommand{\EstRatePop}{\widehat{\RatePop}}
+\newcommand{\NI}{\rho}
+\newcommand{\StdSelection}{\sigma}
+\newcommand{\VarSelection}{\StdSelection^2}
+
+% Tree
+\newcommand{\branch}{\text{b}}
+\newcommand{\branchexp}{^{(\branch)}}
+\newcommand{\Nbranch}{2 \NbrTaxa - 2}
+\newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
+\newcommand{\Setbranch}{\branch \in \{1, \hdots, \Nbranch \}}
+\newcommand{\nodeUp}{\branch^{\uparrow}}
+\newcommand{\nodeDown}{\branch^{\downarrow}}
+\newcommand{\Ntrait}{K}
+\newcommand{\contrast}{\UniDimArray{C}}
+\newcommand{\Covariancematrix}{\Sigma}
+\newcommand{\CovarianceMatrix}{\BiDimArray{\Covariancematrix}}
+\newcommand{\Precisionmatrix}{\Omega}
+\newcommand{\PrecisionMatrix}{\BiDimArray{\Precisionmatrix}}
+\newcommand{\Identitymatrix}{\BiDimArray{I}}
+\newcommand{\brownian}{\mathcal{B}}
+\newcommand{\Brownian}{\UniDimArray{\brownian}}
+\newcommand{\Scattermatrix}{\BiDimArray{A}}
+\newcommand{\Multivariate}{\UniDimArray{Z}}
+\newcommand{\indice}{a}
+\newcommand{\indiceexp}{^{(\indice)}}
+\newcommand{\vecZero}{\UniDimArray{0}}
+\newcommand{\vecOne}{\UniDimArray{1}}
 
@@
     \centering
-    \includegraphics[width=\textwidth, page=1] {artworks/fig-summary}
+    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
     \caption{
-        In red, at the phylogenetic scale, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
-        In blue, trait variance within species ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
+        Between species, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
+        Within species, genetic variance ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
         Under neutral evolution $\RatePhy$ should equal $\RatePop$.
@@
 \subsubsection{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
-The contrast matrix is symmetric, along the diagonal are the total distances ($d_{\Spi}$) from the root of the tree to each taxon ($\Spi$).
-The off-diagonal elements $C_{\Spi,\Spj}$ are the total branch lengths shared by particular pairs of taxa $\Spi$ and $\Spj$, computed from the distance between the pair ($d_{\Spi,\Spj}$) as:
+At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
+Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
+The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be root, in which case the distance is 0).
+Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
 \begin{gather}
-    C_{\Spi,\Spj} = d_{\mathcal{A}(\Spi, \Spj)} = \frac{d_{\Spi} + d_{\Spj} - d_{\Spi,\Spj}}{2}
+    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
 \end{gather}
-Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrSpecies \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
-\begin{gather}
-    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
-\end{gather}
-where $\VecOne$ is an $\NbrSpecies \times 1$ column vector of ones.
+where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.
 
@@
 \begin{gather}
-    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrSpecies - 1}.
+    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
 \label{eq:estimated-rate-phy}
@@
 
-For a species $i$ of interest with inter-individual data available, $\RatePopSpi$ can be estimated as the ratio of genetic variance of a trait ($\VarGeneticSpi = \VarPhenotypeSpi \Multiply \HereditySpi$) over genetic diversity of neutrally evolving sequences ($\pi_{\Spi}$).
-Averaged across all species, we obtain the mean as:
+For a given species $\Spi$ with inter-individual data available, the ratio of genetic variance of a trait ($\VarGeneticSpi$) over genetic diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RatePop$.
+Averaged across all species, we obtain the estimate $\EstRatePop$ as:
 \begin{gather}
-    \EstRatePop = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies}\RatePopSpi = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies} \frac{  V_{\Trait, i} \Multiply \Heredity_{i}}{ \pi_{i}}.
+    \EstRatePop = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heredity_{i}}{ \pi_{i}}.
 \label{eq:estimated-rate-pop}
@@
 \subsubsection{Bayesian estimate}
-We used the software Bayescode
+
+Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
+Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
+The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
+We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution.
 
@@
             \hline
-            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NIx > 1$  & Test $\NIx < 1$ \\ \hline \hline
+            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
             Changing $\MutationRate$ & Scaling with $\pi$, $d$, $\VarPhenotype$ and $\VarPhy$ & Adjusted & Adjusted & -- & --  \\ \hline
@@
     \centering
-    \includegraphics[width=\textwidth, page=1] {artworks/constant_pop_size_phy_pop.hist}
-    \includegraphics[width=\textwidth, page=1] {artworks/fluctuating_pop_size_phy_pop.hist}
+    \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
+    \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
     \caption{
-        $\RatePhy / \RatePop$ for $30.000$ genes simulated under different evolutionary regimes.
-        $\NIx < 1$ for traits simulated under selection (stabilizing selection in yellow).
-        $\NIx = 1$ for traits simulated under a neutral evolution (in blue).
-        $\NIx > 1$ for genes simulated under a moving optimum (diversifying selection in red).
+        $\RatePhy / \RatePop$ for $1,000$ replicates under different evolutionary regimes.
+        Traits are encoded by $5,000$ loci with $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
+        $\NI < 1$ for traits simulated under selection (stabilizing selection in yellow).
+        $\NI = 1$ for traits simulated under a neutral evolution (in blue).
+        $\NI > 1$ for genes simulated under a moving optimum (diversifying selection in red).
         Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top panel), or fluctuating as a Brownian process along the phylogeny (panel B).
@@
 
+\newpage
+
 \part*{Supplementary materials}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
 
-\section{Theoretical derivation}
-\subsection{Definitions}
+\section{Simulator}
 
-We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $i$.
+\subsection{Genotype-phenotype map}
 
-\begin{align}
-    \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
-\end{align}
-
-\subsection{Simulator}
 \begin{itemize}
@@
     \item $\Ne$ is the effective number of individual in the population.
-    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotype at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
-    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the influence of genotype on the trait for individual $i$.
+    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
+    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the genotypic value for individual $i$.
 \end{itemize}
-The mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
+
+\begin{center}
+    \captionof{figure}{Summary of the simulator.}
+    \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
+    \label{fig:simulator-summary}
+\end{center}
+
+\subsection{Heritability, genotypic and phenotypic variance}
+At the population level, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
 \begin{equation}
@@
 
-\newpage
+\subsection{Phenotype-fitness map}
 
@@
 \end{itemize}
-\begin{itemize}
-    \item Trait encoded by $5,000$ loci, $1000$ replicates.
-    \item $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
-\end{itemize}
 
-\begin{table*}[t!]
-    \centering
+\section{Analogy with molecular evolution}
+
+\begin{center}
+    \captionof{table}{Analogy with molecular evolution}
+    \scriptsize
     \begin{adjustbox}{width = 0.5\textwidth}
@@
     \end{adjustbox}
-    \caption{
+\end{center}
 The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
 Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
-    }
-    \label{table:dnds}
-\end{table*}
 
+\section{Bayesian estimate}
+
+\subsection{Multivariate Brownian process}
+Here we generalize to $\Ntrait$  traits evolving along the phylogeny and are correlated between them.
+Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
+The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
+For a branch $\branch$ with length $d\branchexp$, the Brownian process start at node $\nodeUp$ with value $\Brownian(\nodeUp)$, at ends at node $\nodeDown$ with value $\Brownian(\nodeDown)$.
+The independent contrast $\contrast\branchexp$ defined as change in trait along the branch normalized by $\sqrt {d}$ is a multivariate Gaussian:
+\begin{equation}
+    \label{eq:DistribBrownian}
+    \contrast\branchexp = \frac{\Brownian^{(\nodeDown)} - \Brownian^{(\nodeUp)} }{\sqrt {d}} \sim \mathcal{N}\left(\vecZero, \CovarianceMatrix \right).
+\end{equation}
+We defined the {prior} on the covariance matrix as an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
+\begin{equation}
+    \label{eq:Distribcovariance}
+    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
+\end{equation}
+
+\subsection{Sampling the covariance matrix}
+From the independent contrast at each branch of the tree ($\contrast\branchexp$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$ as:
+\begin{equation}
+    \Scattermatrix = \sum\limits_{b=1}^{\Nbranch} \contrast\branchexp \MultiplyMatrix \left[\contrast\branchexp\right]\tr
+\end{equation}
+By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
+\begin{equation}
+    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)
+\end{equation}
+This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate\indiceexp$ defined by
+\begin{equation}
+    \Multivariate\indiceexp \sim \mathcal{N} \left( \vecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).
+\end{equation}
+And from these multivariate samples, $\CovarianceMatrix$ is Gibbs sampled as:
+\begin{equation}
+    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate\indiceexp \MultiplyMatrix  \left[\Multivariate\indiceexp \right] \tr \right)\inv
+\end{equation}
+
+\subsection{Implementation}
+\label{sec:site-specific-mutation-selection-codon-models}
+In \textit{BayesCode} (\url{https://github.com/ThibaultLatrille/bayescode}), we ran the Brownian multivariate model \textit{nodetraits} for 2000 points of MCMC with the options:
+\begin{scriptsize}
+    \begin{verbatim}
+ nodetraits --tree my_tree.newick --traits mean_traits.tsv -u 2000 run_name_chain1
+    \end{verbatim}
+\end{scriptsize}
+The covariance matrix ($\CovarianceMatrix$) is also obtained by running \textit{readnodetraits}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
+\begin{scriptsize}
+    \begin{verbatim}
+ readnodetraits --every 1 --until 2000 --burnin 1000 --cov run_name_chain1
+    \end{verbatim}
+\end{scriptsize}
 \end{document}

### 2023-07-11 data_simulated/update_snake.sh
Use of 'readnodetraits' to compute posterior probabilities.

@@ -4,6 +4,6 @@ for EXPERIMENT in $(pwd)/experiments/*; do
   rm -rf "${EXPERIMENT}"/Snakefile
   ln $(pwd)/Snakefile "${EXPERIMENT}"/Snakefile
   cd "${EXPERIMENT}"
-  snakemake --touch -j 8 --rerun-incomplete
+  snakemake -j 8 --rerun-incomplete
   cd "${FOLDER}"
 done

### 2023-07-11 scripts/libraries.py

@@ -84,7 +84,7 @@ def hist_plot(x_input, x_label, output, xscale="log"):
         min_x = max(1e-6, min_x)
         bins = np.geomspace(min_x, max_x, 100)
     else:
-        bins = np.linspace(min_x, max_x, 100)
+        bins = np.linspace(min_x, max_x, 50)
     hist, _, _ = ax.hist(x.values(), bins=bins, color=color_models, **hist_filled)
     hist, _, _ = ax.hist(x.values(), bins=bins, color=color_models, **hist_step)
     max_y = 1.2 * (np.max([np.max(h) for h in hist]) if len(x) > 1 else np.max(hist))

### 2023-07-11 scripts/merge_results_mammals.py

@@ -13,7 +13,7 @@ def main(tsv_ML_list: str, tsv_Bayes_list: str, output: str):
     list_df = []
     for method, tsv_list in [("ML", tsv_ML_list), ("Bayesian", tsv_Bayes_list)]:
         for path in tsv_list:
-            name_split = os.path.basename(path).replace(".tsv", "").split("_")
+            name_split = os.path.basename(path).replace(".ratio.tsv", "").split("_")
             df = pd.read_csv(path, sep='\t')
             df["method"] = method
             for i, p in enumerate(["dataset", "sex", "logT"]):
@@ -22,6 +22,8 @@ def main(tsv_ML_list: str, tsv_Bayes_list: str, output: str):
     df_out = pd.concat(list_df)
     # Sort by trait, dataset, sex, logT and then method
     df_out = df_out.sort_values(by=["trait", "dataset", "sex", "logT", "method"])
+    # Fill missing values with NaN
+    df_out = df_out.fillna("NaN")
     df_out.to_csv(output, sep="\t", index=False, float_format="%.3f")
 
 

### 2023-07-11 scripts/neutrality_index.py

@@ -1,3 +1,4 @@
+#!/usr/bin/env python3
 import os
 import argparse
 import itertools
@@ -28,7 +29,8 @@ def prune_tree(input_tree: Tree, list_taxa: list = None) -> Tree:
     remove_nodes = set([n for n in tree.traverse() if (n.dist == 0.0 and not n.is_root())])
     for n in remove_nodes:
         n.delete()
-    assert len(set([n for n in tree.traverse()]).intersection(remove_nodes)) == 0, "Some polytomies could not be removed"
+    assert len(
+        set([n for n in tree.traverse()]).intersection(remove_nodes)) == 0, "Some polytomies could not be removed"
     for n in tree.traverse():
         if not n.is_root():
             assert n.dist > 0.0, f"Branch length is 0.0 for node {n.name}"
@@ -102,12 +104,13 @@ def main(input_traits: str, input_tree: str, input_var_within: str, output_tsv: 
     # The column names should be in the format "*_variance" and "*_heritability" (where * is the trait name)
     # The trait names should be the same as in the mean trait file
     # The first column should be "TaxonName"
-    # The second column should be "pS", and not contain any missing value
+    # The second column should be "Nucleotide_diversity", and not contain any missing value
     # "*_variance" is mandatory, but can contain missing values (ideally not)
     # "*_heritability" is optional, if not found, it will be set to 1.0
     var_pop_df = pd.read_csv(input_var_within, sep="\t")
     assert "TaxonName" == var_pop_df.columns[0], f"Column 'TaxonName' is not the first column in {input_var_within}."
-    assert "pS" == var_pop_df.columns[1], f"Column 'pS' is not the second column in {input_var_within}."
+    assert "Nucleotide_diversity" == var_pop_df.columns[
+        1], f"Column 'Nucleotide_diversity' is not the second column in {input_var_within}."
     for trait in traits:
         assert f"{trait}_variance" in var_pop_df.columns, f"Column {trait}_variance not found in {input_var_within}."
     for taxa in var_pop_df["TaxonName"]:
@@ -118,14 +121,22 @@ def main(input_traits: str, input_tree: str, input_var_within: str, output_tsv: 
     output_dict = defaultdict(list)
     for trait in traits:
         print(f"\nProcessing phenotype {trait}.")
-        if f"{trait}_heritability" not in var_pop_df.columns:
[…]

### 2023-07-11 scripts/plot_simulations_ML.py

@@ -12,7 +12,7 @@ from libraries import hist_plot, scatter_plot
 
 def main(folder, neutral_tree_path, output):
     os.makedirs(os.path.dirname(output), exist_ok=True)
-    
+
     model_prefs = {"moving_optimum": 0, "directional": 1, "stabilizing": 2, "neutral": 3}
     models_path = {basename(p): p for p in glob(f"{folder}/*") if isdir(p) and basename(p) in model_prefs}
     models = list(sorted(models_path, key=lambda x: model_prefs[x] if x in model_prefs else -1))
@@ -36,12 +36,15 @@ def main(folder, neutral_tree_path, output):
 
             leaves_value = defaultdict(list)
             for n, n_neutral in zip(tree.get_leaves(), neutral_tree.get_leaves()):
-                # vg = float(getattr(n, "Phenotype_var")) * float(getattr(n, "Heritability"))
-                vg = float(getattr(n, "Genotype_var"))
-                leaves_value['within_tajima'].append(vg / float(getattr(n_neutral, "Theta_Tajima")))
-                leaves_value['within_watterson'].append(vg / float(getattr(n_neutral, "Theta_Watterson")))
-                leaves_value["within_fay_wu"].append(vg / float(getattr(n_neutral, "Theta_Fay_Wu")))
-                leaves_value["within_sampled"].append(vg / float(getattr(n_neutral, "Theta")))
+                var_pheno = float(getattr(n, "Phenotype_var")) * float(getattr(n, "BranchMean_Heritability"))
+                leaves_value["within"].append(var_pheno / float(getattr(n_neutral, "Theta")))
+                leaves_value['within_tajima'].append(var_pheno / float(getattr(n_neutral, "Theta_Tajima")))
+                leaves_value['within_watterson'].append(var_pheno / float(getattr(n_neutral, "Theta_Watterson")))
+                leaves_value["within_fay_wu"].append(var_pheno / float(getattr(n_neutral, "Theta_Fay_Wu")))
+
+                var_geno = float(getattr(n, "Genotype_var"))
+                leaves_value["within_geno"].append(var_geno / float(getattr(n_neutral, "Theta")))
+
                 r = float(getattr(n, "Expected_Mutational_var")) / (2 * float(getattr(n_neutral, "Mutational_rate")))
                 leaves_value["mutational"].append(r)
 
[…]

### 2023-07-11 scripts/plot_simulations_bayesian.py

@@ -2,123 +2,49 @@ import os
 import argparse
 from glob import glob
 from natsort import natsorted
-import numpy as np
 import pandas as pd
 from collections import defaultdict
 from os.path import basename, isdir
 from libraries import hist_plot, scatter_plot
 
 
-def open_covar_file(covar_file):
-    """
-    The file has the following format:
-    entries are in the following order:
-    Phenotype_mean
-    Genotype_mean
-
-    covariances
-
-    62.266  61.58
-    61.58 61.114
-
-    correlation coefficients
-
-    1  0.998
-    0.998      1
-
-    posterior probs
-
-    -      1
-    1      -
-    """
-    with open(covar_file) as f:
-        f.readline()
-        header = []
-        for line in f:
-            if line.strip() == "":
-                break
-            header.append(line.strip())
-        matrix = [[0. for _ in range(len(header))] for _ in range(len(header))]
-        assert f.readline().strip() == "covariances"
-        assert f.readline().strip() == ""
-        row = 0
-        for line in f:
-            if line.strip() == "":
-                break
-            matrix[row] = [float(x.strip()) for x in line.strip().split(" ") if x.strip() != ""]
-            if matrix[row][row] < 0:
-                matrix[row][row] = 0
-                exit(1)
-            assert len(matrix[row]) == len(header)
-            row += 1
-    return header, matrix
-
-
-def open_trace_file(trace_file, burn_in):
-    df = pd.read_csv(trace_file, sep="\t")
-    return df["Var_Phenotype_mean"].values[burn_in:] / 4.0
-
-
-def main(folder, tsv_path, burn_in, output):
+def main(folder, output):
     os.makedirs(os.path.dirname(output), exist_ok=True)
-
-    # Read the tsv file with multiindex columns
-    df = pd.read_csv(tsv_path, sep="\t", header=None)
-    df.columns = pd.MultiIndex.from_arrays([df.iloc[0], df.iloc[1]])
-    df = df.iloc[2:]  # trim off the first 2 rows after we put them in the MultiIndex
-    dict_tree = dict()
-    for name, model in df.columns:
-        if "replicate" == name:
-            continue
-        reps = df[('replicate', model)].values
-        for rep, val in zip(reps, df[(name, model)].values):
[…]

### 2023-07-11 scripts/pre_processed_traits_mammals.py

@@ -33,46 +33,46 @@ def name_internal_nodes(tree: Tree) -> Tree:
     return tree
 
 
-def main(path_input_traits, path_input_pS, path_input_dS, path_output_tree, path_output_traits,
+def main(path_input_traits, path_input_nuc_div, path_input_dS, path_output_tree, path_output_traits,
          path_output_var_within, log_transform, sex):
-    for path in [path_input_traits, path_input_pS, path_input_dS]:
+    for path in [path_input_traits, path_input_nuc_div, path_input_dS]:
         assert os.path.exists(path), f"Path {path} does not exist"
     for path in [path_output_tree, path_output_traits, path_output_var_within]:
         os.makedirs(os.path.dirname(path), exist_ok=True)
 
-    tree = open_tree(path_input_dS, format_ete3=1)
-    tree = rename_tree(tree)
+    t = open_tree(path_input_dS, format_ete3=1)
+    tree = name_internal_nodes(rename_tree(t))
     set_taxa_names = set(tree.get_leaf_names())
     print(f"The tree has {len(set_taxa_names)} leaves")
 
-    # Open the pS/hererozygosity file
-    read_pS = open(path_input_pS, "r").readlines()
-    if read_pS[0].strip() == "#TRAITS":
-        columns = read_pS[1].strip().split(" ")
+    # Open the nucleotide diversity file
+    read_nuc_div = open(path_input_nuc_div, "r").readlines()
+    if read_nuc_div[0].strip() == "#TRAITS":
+        columns = read_nuc_div[1].strip().split(" ")
         n_taxa = int(columns[0])
         n_traits = int(columns[1])
         columns = ["species"] + columns[2:]
         assert len(columns) == n_traits + 1
-        pS_df = pd.read_csv(path_input_pS, sep=" ", skiprows=2, names=columns)
+        nuc_div_df = pd.read_csv(path_input_nuc_div, sep=" ", skiprows=2, names=columns)
         # Replace -1 by NaN
-        pS_df = pS_df.replace(-1, np.nan)
+        nuc_div_df = nuc_div_df.replace(-1, np.nan)
         # Remove "_E" from the specie name
-        pS_df["species"] = pS_df["species"].apply(lambda x: x.replace("_E", ""))
-        assert len(pS_df) == n_taxa
[…]

### 2023-07-11 scripts/pre_processed_traits_simulations.py

@@ -4,7 +4,7 @@ import pandas as pd
 from neutrality_index import open_tree, prune_tree
 
 
-def main(input_path, input_neutral_path, tree_path, traits_path):
+def main(input_path, input_neutral_path, tree_path, traits_path, var_within_path):
     tree = open_tree(input_path, format_ete3=1)
 
     dico_traits = defaultdict(list)
@@ -27,12 +27,25 @@ def main(input_path, input_neutral_path, tree_path, traits_path):
     # Write the topology to a newick file
     neutral_tree.write(outfile=tree_path, format=3)
 
+    var_within = defaultdict(list)
+    for n, n_neutral in zip(tree.get_leaves(), neutral_tree.get_leaves()):
+        assert n.name == n_neutral.name
+        var_within["TaxonName"].append(n.name)
+        var_within["Nucleotide_diversity"].append(float(getattr(n_neutral, "Theta")))
+        var_within["Phenotype_variance"].append(float(getattr(n, "Phenotype_var")))
+        var_within["Phenotype_heritability"].append(float(getattr(n, "BranchMean_Heritability")))
+        var_within["Genotype_variance"].append(float(getattr(n, "Genotype_var")))
+        var_within["Genotype_heritability"].append(1.0)
+    df = pd.DataFrame(var_within)
+    df.to_csv(var_within_path, sep="\t", index=False)
+
 
 if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
     parser.add_argument("--input", help="Input tree file", required=True)
     parser.add_argument("--neutral_tree", help="Input neutral tree file", required=True)
     parser.add_argument("--tree", help="Output tree file", required=True)
-    parser.add_argument("--traits", help="Output traits file", required=True)
+    parser.add_argument("--traitsfile", help="Output traits file", required=True)
+    parser.add_argument("--var_within", help="Output var_within file", required=True)
     args = parser.parse_args()
-    main(args.input, args.neutral_tree, args.tree, args.traits)
+    main(args.input, args.neutral_tree, args.tree, args.traitsfile, args.var_within)

### 2023-07-11 scripts/ratio_pvalue.py

@@ -1,72 +1,0 @@
-import os
-import argparse
-import numpy as np
-import pandas as pd
-from collections import defaultdict
-
-
-def open_trace_file(trace_file: str, burn_in: int) -> dict[str: np.ndarray]:
-    df = pd.read_csv(trace_file, sep="\t")
-    var_phy_dict = {}
-    for col in df.columns:
-        if col.startswith("Var_"):
-            assert col.endswith("_mean"), f"Column {col} does not end with _mean"
-            assert len(df[col].values) > burn_in, f"Column {col} has less than {burn_in} values"
-            var_phy_dict[replace_last(col[4:], "_mean", "")] = df[col].values[burn_in:] / 4.0
-    n_taxa = int(df["n_taxa"][0])
-    return var_phy_dict, n_taxa
-
-
-def replace_last(s: str, old: str, new: str) -> str:
-    li = s.rsplit(old, 1)
-    return new.join(li)
-
-
-def main(inference_path: str, var_within_path: str, burn_in: int, output_path: str):
-    os.makedirs(os.path.dirname(output_path), exist_ok=True)
-
-    traces_var_phy_dict, n_taxa = open_trace_file(inference_path, burn_in=burn_in)
-    var_pop_df = pd.read_csv(var_within_path, sep="\t")
-
-    output_dict = defaultdict(list)
-    for trait, var_phy_trace in traces_var_phy_dict.items():
-        var_between = np.mean(var_phy_trace)
-
-        assert f"{trait}_variance" in var_pop_df.columns
-        notna = np.isfinite(var_pop_df[f"{trait}_variance"])
-        if f"{trait}_heritability" not in var_pop_df.columns:
-            print(f"Warning: column {f'{trait}_heritability'} not found in {var_within_path}.")
-            print("Assuming heritability = 1.0.")
-            var_pop_df[f"{trait}_heritability"] = 1.0
-        # Computing the genetic variance (geno = h² * pheno)
-        genetic_variance_array = var_pop_df[f"{trait}_heritability"][notna] * var_pop_df[f"{trait}_variance"][notna]
-        var_within_array = genetic_variance_array / (var_pop_df["pS"][notna])
-        print(f'Found {len(genetic_variance_array)} species with a variance for {trait}.')
-        var_within = np.mean(var_within_array)
-        assert var_within > 0
-
-        ratio = var_between / var_within
-        ratio_pv = np.mean(var_phy_trace > var_within)
-        print(f"var_between = {var_between}")
-        print(f"var_within = {var_within}.")
[…]

### 2023-07-11 manuscript/main.tex
Updated manuscript

@@
 \usepackage{orcidlink}
+\usepackage{marvosym}
 
@@
 \newcommand{\NI}{\rho}
+\newcommand{\EstNI}{\widehat{\rho}}
 \newcommand{\StdSelection}{\sigma}
@@
 New mutations are thus generating variance for such a trait.
-At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$), as:
+At the individual level, the mutational variance ($\VarMutation = \NbrLoci \Multiply  \E [a^2]$) is the rate at which new mutations contribute to the trait variance per generation.
+As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is also a function of the mutation rate per loci per generation ($\MutationRate$), as:
 \begin{gather}
@@
 
-Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($\Heredity$), giving also $\RatePop$ as:
+Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heredity$), giving also $\RatePop$ as:
 \begin{align}
@@
 
+Altogether, the neutrality index is estimated as:
+\begin{gather}
+    \EstNI = \frac{\EstRatePhy}{\EstRatePop}. \label{eq:estimated-NI}
+\end{gather}
+
 \subsubsection{Bayesian estimate}
 
-Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
+Even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
+Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
+Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021} .
 Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
+The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
 We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution.
+The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba{\NI > 1 }$ to test for evidence of diversifying selection and $\proba{\NI < 1 }$ to test for evidence of stabilizing selection.
 
@@
 From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
-Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype \cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
 The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
@@
 Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.
+Genetic diversity is measured as the heterozygosity of the neutral markers, and the genetic variance of the trait is measured as the variance of the trait genotypic value.
+Genetic divergence is measured as the number of substitutions per site along branches of the phylogeny.
+Heritability is estimated from the slope of the regression of offspring trait values on parental trait values\cite{lynch_genetics_1998}, using the information from the last two generations of the simulation.
 
@@
 
+\subsection{Empirical dataset}
+
+We first focused on a mammalian dataset, with the body size and brain mass as quantitative traits.
+For the genetic variation, the mammalian phylogeny is extracted from the Zoonomia project, obtained on a set of neutral markers in \textcite{foley_genomic_2023}, with genetic diversity measured as heterozygosity in \textcite{wilder_contribution_2023}
+For quantitative trait variation, the log-transformed body size and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
+We removed individuals not marked as adult, and split the data into males and females.
+We discarded species with no intra-specific variability.
+
+We also focused on a dataset of Primates species, with the genetic variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
+Unfortunately, the primate phylogeny was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
+
 \section{Results}
@@
 This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
-Altogether, we define the neutrality index as $\RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
-Under a depleted variation between species, meaning $\RatePhy < \RatePop$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
-In contrast, under an inflated variation between species, meaning $\RatePhy > \RatePop$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.
+Altogether, we define the neutrality index as $\NI = \RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
+Under a depleted variation between species, meaning $\NI < 1$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
+In contrast, under an inflated variation between species, meaning $\NI > 1$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.
 
@@
             Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
-            Changing $\MutationRate$ & Scaling with $\pi$, $d$, $\VarPhenotype$ and $\VarPhy$ & Adjusted & Adjusted & -- & --  \\ \hline
-            Changing $\Ne$ & Scaling with $\pi$ and $\VarPhenotype$ & Adjusted & -- & -- & --  \\ \hline
-            Unknown heredity & $\Heredity = 1$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
-            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
-            Polymorphism are selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
-            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
-            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & False positives  \\ \hline
-            Divergence are selected & $d$ is under-estimated & -- & Over-estimated & False positives & Conservative  \\ \hline
-            Structured population & unknown & unknown & unknown & unknown & unknown \\ \hline
-            Gene flow between species & unknown & unknown & unknown & unknown & unknown \\ \hline
+            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & Invalid  \\ \hline
+            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
+            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are negatively selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
+            Markers for divergence are positively selected & $d$ is over-estimated & -- & Under-estimated & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are positively selected & $\pi$ is over-estimated  & Under-estimated & -- & Invalid & Conservative  \\ \hline
+            Markers for divergence are negatively selected & $d$ is under-estimated & -- & Over-estimated & Invalid & Conservative  \\ \hline
     \end{tabular}
     \end{adjustbox}
-    \caption{Assumptions and their consequences on deviation from neutraliity.}
+    \caption{Assumptions and their consequences on the estimation of $\EstRatePop$ and $\EstRatePhy$ and on the ratio $\NI = \EstRatePhy/\EstRatePop$.
+    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
+    $\pi$ is the genetic diversity, $d$ is the genetic divergence, $\VarPhenotype$ is the phenotypic variance, $\VarPhy$ is the phylogenetic variance.
+    }
     \label{table:assumptions}
@@
 Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
-Analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
 The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
-Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
 Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
 This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
+We also devised a Bayesian estimated allowing to obtain the posterior distribution of the neutrality index, and to test for evidence of diversifying selection as $\proba [\EstNI > 1]$, and evidence for stabilizing selection as $\proba [\EstNI < 1]$.
 
@@
     \centering
+    \begin{minipage}{0.49\textwidth}
         \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
+    \end{minipage}
+    \begin{minipage}{0.49\textwidth}
         \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
+    \end{minipage}
+    \begin{minipage}{0.49\textwidth}
+        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.rho}
+    \end{minipage}
+    \begin{minipage}{0.49\textwidth}
+        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.pvalues}
+    \end{minipage}
     \caption{
@@
         Traits are encoded by $5,000$ loci with $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
-        $\NI < 1$ for traits simulated under selection (stabilizing selection in yellow).
-        $\NI = 1$ for traits simulated under a neutral evolution (in blue).
-        $\NI > 1$ for genes simulated under a moving optimum (diversifying selection in red).
-        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top panel), or fluctuating as a Brownian process along the phylogeny (panel B).
+        Traits simulated under stabilizing selection in yellow, under a neutral evolution in blue, and under a diver optimum (in red).
+        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
     }
@@
 \end{figure*}
+
+\subsection{Results on empirical data}\label{subsec:results-on-empirical-data}
+For mammalian body size, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
+The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.346$ ($[0.183,~0.677]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.278$ ($[0.136,~0.639]$ for $95\%$ credible interval).
+For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.375$ ($[0.844,~2.212]$ for $95\%$ credible interval, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.762$ ($[1.021,~3.010]$ for $95\%$ credible interval, $\NbrTaxa=26$).
+In both males and females, brain mass is under diversifying selection, with p-values of $0.116$ and $0.022$ respectively.
+
+For primates body size, we obtained $\EstNI_{\text{\Male}} = 0.557$ ($[0.406,~0.767]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.393$ ($[0.277,~0.561]$ for $95\%$ credible interval, $\NbrTaxa=65$).
+For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.920$ ($[1.397,~2.616]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.974$ ($[1.393,~2.798]$ for $95\%$ credible interval, $\NbrTaxa=65$).
+In both males and females, brain mass is under diversifying selection, with p-values of $0.00$ and $0.00$ respectively.
+However, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).
 
@@
 Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
-Our neutrality index for a trait is the ratio of between species variation to within species variation.
-Such neutrality index can be compared to 1 to seek a statistical deviation from a neutral trait.
+Our neutrality inde for a trait, called $\NI$, is the ratio of between species variation to within species variation.
+$\NI$ can be compared to 1 to seek a statistical deviation from a neutral trait.
 In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
 Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
-Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
+Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
+$ \proba [ \NI > 1 ]$).
 Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
@@
 Contrarily, we argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
+If heritability is not known, the test for diversifying selection can still be performed (although under-powered) by setting the heritability to 1.0, but the test for stabilizing selection is invalid.
+Because our test is based on several assumptions that might not hold on empirical data, we provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
+As example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$ (see fig.~\ref{fig:empirical_results}).
+However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for genetic divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
+
+% Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
+More theoretically, our development is based on several assumption that could be relaxed.
+First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
+Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of polymorphism and divergence.
+Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within species in a unified framework.
+
+Our test bears analogy with others tests for neutrality developed, providing insight into its behavior.
+First, analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
+Second, because $\NI$ is compared to 1, our test is similar to the codon-based test of selection of \textcite{goldman_codonbased_1994, muse_likelihood_1994}, where the ratio of non-synonymous to synonymous substitutions ($\omega$) is compared to 1.
+As $\omega < 1$ is interpreted as purifying selection, $\NI < 1$ is interpreted as stabilizing selection.
+Also, as $\omega > 1$ is interpreted as diversifying selection, $\NI > 1$ is interpreted as adaptation.
+With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\cite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
+First, not rejecting the hypothesis that $\omega = 1$ does not mean that the trait is effectively neutral, since a combination of adaptation and purifying selection can also lead to $\omega = 1$\cite{nielsen_molecular_2005}.
+Analogously, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$.
+Secondly, $\omega < 1$ is not evidence for the absence of adaptation, but rather that purifying selection is strong enough to overcome the effect of adaptation.
+For example, by explicitly modelling purifying selection it is possible to detect genes and sites under adaptation even though they display $\omega < 1$ ~\cite{latrille_genes_2023}.
+Similarly, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection.
+We thus argue that by explicitly modelling stabilizing selection as a moving optimum, it would theoretically allow to tease-out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
 
@@
 
-% What do we expect for biological traits
-% Problem for slow evolving diversifying selection
-% Modelling the mechantistic expectation Stabilizting selection with a moving optimum. 
-% How to include Vm?
-
-% Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
-More theoretically, our development is based on several assumption that could be relaxed.
-First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
-Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Moreover, even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
-Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012}.
-Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within species in a unified framework.
+It tells us about the phenotype fitness map
 
@@
     \begin{verbatim}
- nodetraits --tree my_tree.newick --traits mean_traits.tsv -u 2000 run_name_chain1
+ nodetraits --tree my_tree.newick --traitsfile mean_traits.tsv -u 2000 run_name_chain1
     \end{verbatim}

### 2023-07-13 data_simulated/run.sh
Use the mammal phylogeny in simulations derived from empirical data

@@ -1,4 +1,4 @@
-CPU=6
-for EXPERIMENT in ./constant_pop_size_L*.yaml; do
+CPU=8
+for EXPERIMENT in ./constant_pop_size*.yaml; do
 python3 simulated_experiment.py -c $(basename "${EXPERIMENT}") -j ${CPU}
 done

### 2023-07-13 simulator/trees/compute_params.py

@@ -1,0 +1,46 @@
+import numpy as np
+import pandas as pd
+from ete3 import Tree
+
+
+def total_tree_length(t: Tree):
+    return sum([n.dist for n in t.traverse() if not n.is_root()])
+
+
+def distances_from_root_to_leaves(t: Tree):
+    assert t.is_root()
+    return [n.get_distance(t) for n in t.get_leaves()]
+
+
+if __name__ == '__main__':
+    var_within_path = "MamGenTimescale.m.var_within.tsv"
+    dS_tree_path = "MamGenTimescale.m.tree"
+    time_tree_path = "MamGenTimescale.m.species.nwk"
+
+    var_within = pd.read_csv(var_within_path, sep="\t")
+    pS = var_within["Nucleotide_diversity"].mean()
+    print(f"pS = {pS}")
+    dS_tree = Tree(dS_tree_path, format=1)
+    dS_total_tree_length = total_tree_length(dS_tree)
+    print(f"Tree length (dS) = {dS_total_tree_length}")
+    time_tree = Tree(time_tree_path, format=1)
+    time_total_tree_length = total_tree_length(time_tree)
+    print(f"Tree length (My) = {time_total_tree_length}")
+    ages_list = distances_from_root_to_leaves(time_tree)
+    root_age = np.mean(ages_list)
+    print(f"Root age (My) = {root_age}")
+    root_dS = dS_total_tree_length * root_age / time_total_tree_length
+    print(f"Root age (dS) = {root_dS}")
+
+    pop_size = 50
+    nbr_loci = 10000
+    nbr_sites_var = nbr_loci * pS
+    print(f"nbr_sites_var = {nbr_sites_var}")
+    # pS = 4 * N * u
+    # u = pS / (4 * N)
+    u = pS / (4 * pop_size)
+    print(f"u = {u}")
+
+    nbr_generations = root_dS / u
+    print(f"nbr_generations = {nbr_generations}")
+

### 2023-07-13 manuscript/main.tex
Updated manuscript

@@
 \usepackage{marvosym}
+
+% -- Defining colors:
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
+\lstdefinestyle{mystyle}{
+    backgroundcolor=\color{backcolour},
+    basicstyle=\ttfamily\scriptsize\bfseries,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=t,
+    keepspaces=true,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2
+}% -- Setting up the custom style:
+\lstset{style=mystyle}
 
@@
 
-\title{A test of diversifying selection for a trait from within and between species variations}
+\title{A test for diversifying selection for a trait from within and between species variations while integrating for genomic diversity}
 \rhead{\scshape Test of diversifying selection}
@@
     To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
-    However, using within species variation, methods do not integrate the changes over long evolutionary time.
-    Conversely, using between species variation, current methods usually discards the variance in traits between individuals, leaving out valuable information.
+    Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
+    Conversely, using between species variation, current methods usually discards within species variation, thus not accounting for processes at the micro-evolutionary scale.
     The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
-    This neutrality index this does not require estimation of population size nor of time of speciation for normalization since it also leverages sequence polymorphism and divergence.
-    Our index can be used to seek deviation from the null model of neutral evolution, and is greater than one for a trait under diversifying selection.
-    As an example, we show that brain mass at the mammalian scale is under diversifying selection.
+    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
+    As such, it does not require estimation of population size nor of time of speciation for normalization.
+    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
+    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
     Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
@@
 % Research Article (7500 words)
-\section{Introduction}\label{sec:introduction}
+\section*{Introduction}\label{sec:introduction}
 
@@
 Trait variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
+All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.
 
-At the population level across individuals, Genome Wide Association Studies (GWAS) in humans have revealed that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
-At the species level, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait\cite{martin_multivariate_2008, leinonen_qst_2013}
-For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations\cite{whitehead_neutral_2006, gilad_natural_2006}.
-However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated\cite{signor_evolution_2018, price_detecting_2022}.
-Theoretically, change in traits accumulate linearly with time from divergence for a pair of species, and proportionally to the trait variance\cite{lande_genetic_1980, turelli_heritable_1984}.
-For example, gene expression accumulate linearly with time from divergence, and accumulates faster for trait with large variance\cite{khaitovich_neutral_2004}.
-Such theoretical connection between trait variance and change in mean value can be used to test for a selection of trait using for a pair of species\cite{walsh_evolution_2018}.
+First, at the micro-evolutionary scale, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
+At the species level, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
+For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
+However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
+Theoretically, change in traits accumulate linearly with time from divergence for a pair of species, and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
+For example, gene expression accumulate linearly with time from divergence, and accumulates faster for trait with large variance~\cite{khaitovich_neutral_2004}.
+Such theoretical connection between trait variance and change in mean value can be used to test for a selection of trait using for a pair of species~\cite{walsh_evolution_2018}.
 However, to our knowledge, a neutral theoretical expectation for trait changes across several species related by their phylogenetic tree, as a function of within species trait variation has not yet been formulated and tested.
@@
 % Phylogenetics (mean trait evolution)
-Contrarily, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale\cite{felsenstein_phylogenies_1985}.
-Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a standard deviation proportional to the squared root of the divergence time\cite{hansen_translating_1996}.
-As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
-A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
+Contrarily, at the macro-evolutionary scale, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
+Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a standard deviation proportional to the squared root of the divergence time~\cite{hansen_translating_1996}.
+As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
+A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
 Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
-However, studies have shown that such comparative approaches are subject to different biases\cite{harmon_phylogenetic_2018}.
+However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
 First, and unfortunately, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
-A trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral\cite{hansen_translating_1996}.
-Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+A trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
+Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 In such comparative framework, the variation within species is usually discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
@@
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
 Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
@@
 Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
-From a population-genetic perspective, our study can be seen as the generalization of Q$_\text{ST}$--F$_\text{ST}$ methods at the phylogenetic scale to account relationships between species.
-From a phylogenetic perspective, our study can be seen as a extension of the EVE model\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} where we set a clear threshold for neutral evolution by leveraging species' sequence polymorphism and divergence.
+From a population-geneticist perspective, our study can be seen as the generalization at the macro scale of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
+From a phylogeneticist perspective, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} where we set a clear threshold for neutral evolution by leveraging species' sequence polymorphism and divergence.
 % Basically, we are testing whether the variance of means is greater than the man of variances
 
-\section{Materials and Methods}\label{sec:materials-and-methods}
-\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
-\subsubsection{Mutation-drift equilibrium for a trait}
+\section*{Materials and Methods}\label{sec:materials-and-methods}
+\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
+\subsubsection*{Mutation-drift equilibrium for a trait}
 
@@
 
-\subsubsection{Phylogenetic evolution of a trait}
+\subsubsection*{Phylogenetic evolution of a trait}
 
 Along a lineage, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
-If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process\cite{hansen_translating_1996}.
+If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
 After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
@@
 \end{align}
-That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.
+That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.
 
@@
 
-\subsubsection{Neutrality index}
+\subsubsection*{Neutrality index}
 
@@
 
-\subsection{Estimate}
+\subsection*{Estimate}\label{subsec:estimate}
 
-\subsubsection{Maximum likelihood estimate}
+\subsubsection*{Maximum likelihood estimate}
 At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
@@
 The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be root, in which case the distance is 0).
-Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
+Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate~\cite{omeara_testing_2006}, given as:
 \begin{gather}
@@
 
-Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
+Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated~\cite{omeara_testing_2006} as:
 \begin{gather}
@@
 
-\subsubsection{Bayesian estimate}
+\subsubsection*{Bayesian estimate}
 
@@
 Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
-Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021} .
-Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
+Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
+Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
 The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
-We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution.
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba{\NI > 1 }$ to test for evidence of diversifying selection and $\proba{\NI < 1 }$ to test for evidence of stabilizing selection.
+We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
+The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\NI > 1 ]$ to test for evidence of diversifying selection and $\proba [\NI < 1 ]$ to test for evidence of stabilizing selection.
 
-\subsection{Simulation}\label{subsec:simulations}
+\subsection*{Simulation}\label{subsec:simulations}
 
-We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator}).
+We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator} and section~S\ref{subsec:genotype-phenotype-map}).
 We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
@@
 Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
-The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
+The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype (see section~S\ref{subsec:phenotype-fitness-map}).
 Mutations are seen has displacement of the phenotype in the multidimensional space.
@@
 Genetic divergence is measured as the number of substitutions per site along branches of the phylogeny.
-Heritability is estimated from the slope of the regression of offspring trait values on parental trait values\cite{lynch_genetics_1998}, using the information from the last two generations of the simulation.
+Heritability is estimated from the slope of the regression of offspring trait values on parental trait values~\cite{lynch_genetics_1998}, using the information from the last two generations of the simulation.
 
@@
         Wright-Fisher simulations with mutation, selection and drift.
+        Panel A: the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
+        The trait genotypic value is then transformed into a phenotypic value by adding a Gaussian noise with standard deviation $\VarEnv$.
+        Parents are selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
+        Mutations are drawn from a Poisson distribution, with each loci having a probability $\MutationRate$ to mutate.
+        Drift is modelled by the resampling of parents in a finite population.
+        Panel B: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (grey area).
     }
@@
 
-\subsection{Empirical dataset}
+\subsection*{Empirical dataset}\label{subsec:empirical-dataset}
 
@@
 
-\section{Results}
+\section*{Results}\label{sec:results}
 
@@
 The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
-More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
+More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$), denoted $\RatePop$, is a proxy for the mutational variance ($\RateMut$).
 At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
-This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
+This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), denoted $\RatePhy$, allowing us to obtain another proxy for the mutational variance ($\RateMut$).
 Altogether, we define the neutrality index as $\NI = \RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
@@
             Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
-            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & Invalid  \\ \hline
-            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
-            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are negatively selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
-            Markers for divergence are positively selected & $d$ is over-estimated & -- & Under-estimated & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are positively selected & $\pi$ is over-estimated  & Under-estimated & -- & Invalid & Conservative  \\ \hline
-            Markers for divergence are negatively selected & $d$ is under-estimated & -- & Over-estimated & Invalid & Conservative  \\ \hline
+            Trait encoded by few loci & Between species trait variation is under-estimated  & -- & Under-estimated & Conservative & Invalid  \\ \hline
+            Sexual dimorphism &  Within species trait variation is over-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
+            Inbreeding & Nucleotide diversity ($\pi$) is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
+            Markers for divergence are positively selected & Nucleotide divergence ($d$) is over-estimated & -- & Under-estimated & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is over-estimated  & Under-estimated & -- & Invalid & Conservative  \\ \hline
+            Markers for divergence are negatively selected & Nucleotide divergence ($d$) is under-estimated & -- & Over-estimated & Invalid & Conservative  \\ \hline
     \end{tabular}
     \end{adjustbox}
-    \caption{Assumptions and their consequences on the estimation of $\EstRatePop$ and $\EstRatePhy$ and on the ratio $\NI = \EstRatePhy/\EstRatePop$.
+    \caption{Assumptions and their consequences on the estimation of within species variation ($\EstRatePop$), between species variation ($\EstRatePhy$), and on the neutrality index $\NI = \EstRatePhy/\EstRatePop$.
     The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
-    $\pi$ is the genetic diversity, $d$ is the genetic divergence, $\VarPhenotype$ is the phenotypic variance, $\VarPhy$ is the phylogenetic variance.
     }
@@
 
-\subsection{Results against simulations}
+\subsection*{Results against simulations}\label{subsec:results-against-simulations}
 
@@
 
-\subsection{Results on empirical data}\label{subsec:results-on-empirical-data}
+\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
 For mammalian body size, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
@@
 
-\section{Discussion}\label{sec:discussion}
+\section*{Discussion}\label{sec:discussion}
 
@@
 Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
-Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
+Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation (see section~S\ref{sec:implementation}) also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
 $ \proba [ \NI > 1 ]$).
-Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
+Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits ($\Heredity$).
 We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
-By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
-Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
-Contrarily, we argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
-If heritability is not known, the test for diversifying selection can still be performed (although under-powered) by setting the heritability to 1.0, but the test for stabilizing selection is invalid.
-Because our test is based on several assumptions that might not hold on empirical data, we provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
-As example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$ (see fig.~\ref{fig:empirical_results}).
+By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time~\cite{litsios_effects_2012}.
+Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes~\cite{seo_estimating_2004}.
+
+We argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically harder to measure for any trait.
+Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is under-powered.
+Indeed, because $\Heredity \leq 1$ by definition, if we set $\Heredity \leq 1$, we overestimate the within species variation and thus underestimate $ \NI$ and can test for diversifying selection.
+In other words, having $\NI > 1$ while setting $\Heredity = 1$ means that knowing the real heritability would only increase the evidence for diversifying selection.
+Contrarily, the test for stabilizing selection is invalid if heritability is not known.
+Because our test is also based on several assumptions that might not hold on empirical data, we provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
+As example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$.
 However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for genetic divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
@@
 First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
-Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component~\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny~\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of polymorphism and divergence.
 Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within species in a unified framework.
+Also, we don't know how our test would behave in the context of gene flows and introgression, which could be investigated in future studies.
 
-Our test bears analogy with others tests for neutrality developed, providing insight into its behavior.
-First, analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
-Second, because $\NI$ is compared to 1, our test is similar to the codon-based test of selection of \textcite{goldman_codonbased_1994, muse_likelihood_1994}, where the ratio of non-synonymous to synonymous substitutions ($\omega$) is compared to 1.
+Theoretically, our test bears analogy with others tests for neutrality developed, providing insight into its behavior.
+First, taking inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
+Second, because $\NI$ is compared to 1, our test bears analogy to the codon-based test of selection of \textcite{goldman_codonbased_1994, muse_likelihood_1994}, where the ratio of non-synonymous to synonymous substitutions ($\omega$) is compared to 1.
 As $\omega < 1$ is interpreted as purifying selection, $\NI < 1$ is interpreted as stabilizing selection.
@@
 With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\cite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the hypothesis that $\omega = 1$ does not mean that the trait is effectively neutral, since a combination of adaptation and purifying selection can also lead to $\omega = 1$\cite{nielsen_molecular_2005}.
+First, not rejecting the hypothesis that $\omega = 1$ does imply a neutral evolution, since a combination of adaptation and purifying selection can also lead to $\omega = 1$~\cite{nielsen_molecular_2005}.
 Analogously, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$.
@@
 % What about the other methods? What are we testing for?
+From an empirical point of view, our test is also related to other methods that have been developed to test for selection on quantitative traits.
 At the phylogenetic scale, the change in mean trait value across different species is typically used to test for neutrality for a quantitative trait.
-These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
-For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection\cite{catalan_drift_2019}.
+These models are seeking deviation from a Brownian process~\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation~\cite{rohlfs_phylogenetic_2015}.
+For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\cite{catalan_drift_2019}.
 However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
 With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
-Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$), as well as estimate for heritabilities ($\Heredity$).
-
-It tells us about the phenotype fitness map
+Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$).
+Altogether, our study provides a statistical framework to test for neutrality of a quantitative trait while integrating the trove of genomic data available both within and between species.
+, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
 
@@
 \setcounter{table}{0}
+\setcounter{section}{0}
 
-\section{Simulator}
+\tableofcontents
 
-\subsection{Genotype-phenotype map}
+\section{Genetic architecture of the trait}\label{sec:simulator}
+
+\subsection{Genotype-phenotype map}\label{subsec:genotype-phenotype-map}
 
@@
     \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the genotypic value for individual $i$.
+    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
+    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
 \end{itemize}
@@
 \begin{center}
-    \captionof{figure}{Summary of the simulator.}
+    \captionof{figure}{Summary of trait's genetic architecture.}
     \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
@@
 
-\subsection{Heritability, genotypic and phenotypic variance}
 At the population level, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
 \begin{equation}
-    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad}	\VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2
+    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad}	\VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2\label{eq:simu-genotype}
 \end{equation}
-\begin{itemize}
-    \setlength\itemsep{-0.2em}
-    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
-    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
-\end{itemize}
+
 The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
 \begin{equation}
-    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2
+    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2 \label{eq:simu-between}
 \end{equation}
+
 Heritability ($h^2$) is defined as:
 \begin{equation}
-    h^2 = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}
+    h^2 = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
 \end{equation}
-\begin{gather}
-    G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} \\
-    \Trait_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} + \xi_i
-\end{gather}
+Heritability can also be estimated from the slope of the linear regression of the phenotype on the genotype.
 
-\subsection{Phenotype-fitness map}
+\subsection{Phenotype-fitness map}\label{subsec:phenotype-fitness-map}
 
@@
 
-\section{Analogy with molecular evolution}
+\section{Bayesian estimate}\label{sec:bayesian-estimate}
 
-\begin{center}
-    \captionof{table}{Analogy with molecular evolution}
-    \scriptsize
-    \begin{adjustbox}{width = 0.5\textwidth}
-        \begin{tabular}{|c||c|c|}
-            \hline
-            & Coding sequences & Quantitative traits \\ \hline \hline
-            Adaptation / Diversifying selection & $\dnds > \pnps$ & $ \RatePhy > \RatePop  $ \\ \hline
-            Neutral regime of evolution & $\dnds = \pnps$ & $ \RatePhy = \RatePop  $ \\ \hline
-            Purifying / Stabilizing selection & $\dnds < \pnps$ & $ \RatePhy < \RatePop  $ \\ \hline
-        \end{tabular}
-    \end{adjustbox}
-\end{center}
-The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
-Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
-
-\section{Bayesian estimate}
-
-\subsection{Multivariate Brownian process}
+\subsection{Multivariate Brownian process}\label{subsec:multivariate-brownian-process}
 Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
@@
 
-\subsection{Sampling the covariance matrix}
+\subsection{Sampling the covariance matrix}\label{subsec:sampling-the-covariance-matrix}
 From the independent contrast at each branch of the tree ($\contrast\branchexp$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$ as:
 \begin{equation}
-    \Scattermatrix = \sum\limits_{b=1}^{\Nbranch} \contrast\branchexp \MultiplyMatrix \left[\contrast\branchexp\right]\tr
+    \Scattermatrix = \sum\limits_{b=1}^{\Nbranch} \contrast\branchexp \MultiplyMatrix \left[\contrast\branchexp\right]\tr\label{eq:bayes-scatter}
 \end{equation}
@@
 \begin{equation}
-    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)
+    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)\label{eq:bayes-posterior}
 \end{equation}
@@
 \begin{equation}
-    \Multivariate\indiceexp \sim \mathcal{N} \left( \vecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).
+    \Multivariate\indiceexp \sim \mathcal{N} \left( \vecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
 \end{equation}
@@
 \begin{equation}
-    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate\indiceexp \MultiplyMatrix  \left[\Multivariate\indiceexp \right] \tr \right)\inv
+    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate\indiceexp \MultiplyMatrix  \left[\Multivariate\indiceexp \right] \tr \right)\inv \label{eq:bayes-gibbs}
 \end{equation}
 
-\subsection{Implementation}
-\label{sec:site-specific-mutation-selection-codon-models}
-In \textit{BayesCode} (\url{https://github.com/ThibaultLatrille/bayescode}), we ran the Brownian multivariate model \textit{nodetraits} for 2000 points of MCMC with the options:
-\begin{scriptsize}
-    \begin{verbatim}
- nodetraits --tree my_tree.newick --traitsfile mean_traits.tsv -u 2000 run_name_chain1
-    \end{verbatim}
-\end{scriptsize}
-The covariance matrix ($\CovarianceMatrix$) is also obtained by running \textit{readnodetraits}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
-\begin{scriptsize}
-    \begin{verbatim}
- readnodetraits --every 1 --until 2000 --burnin 1000 --cov run_name_chain1
-    \end{verbatim}
-\end{scriptsize}
+\section{Bayesian and Maximum-likelihood implementation}\label{sec:implementation}
+
+Implementation is included within the \textit{BayesCode} software, available at \url{https://github.com/ThibaultLatrille/bayescode}.
+
+\subsection{Data formatting}\label{subsec:data-formatting}
+
+Running the analysis on your dataset and compute posterior probabilities requires three files:
+
+\begin{enumerate}
+    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers)
+    \item A file containing the mean trait values for each species.
+    \item A file containing the variation within species for each trait and the genetic variation within species (neutral markers).
+\end{enumerate}
+
+\subsubsection{Phylogenetic tree}\label{subsubsec:phylogenetic-tree}
+
+The phylogenetic tree must be in newick format, with branch lengths in number of substitutions per site (neutral markers).
+
+\subsubsection{Mean trait for each species}\label{subsubsec:mean-trait-for-each-species}
+
+The file containing mean trait values for each species must be in a tab-delimited file with the following format:
+
+\begin{table}[h]
+    \centering
+    \begin{tabular}{|l|c|c|}
+        \hline
+        TaxonName & Body\_mass & Brain\_mass \\
+        \hline
+        Panthera\_tigris & 12.26 & 5.676 \\
+        Pithecia\_pithecia & 7.256 & 3.436 \\
+        Colobus\_angolensis & 9.176 & 4.284 \\
+        Saimiri\_boliviensis & 6.845 & 3.279 \\
+        $\hdots$ & $\hdots$ & $\hdots$ \\
+        \hline
+    \end{tabular}\label{tab:trait-mean}
+\end{table}
+
+The columns are:
+
+\begin{itemize}
+    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
+    \item As many columns as traits, without spaces or special characters in the trait.
+    \item The values can be \texttt{NaN} to indicate that the trait is not available for that taxon.
+\end{itemize}
+
+\subsubsection{Trait variation for each species}\label{subsubsec:trait-variation-for-each-species}
+
+The file containing trait variation for each species must be in a tab-delimited file with the following format:
+
+\begin{table}[h]
+    \centering
+    \begin{adjustbox}{width = 1.0\textwidth}
+    \begin{tabular}{|l|c|c|c|c|c|c|}
+        \hline
+        TaxonName & Nucleotide\_diversity & Body\_mass\_variance & Body\_mass\_heritability & Brain\_mass\_variance & Brain\_mass\_heritability \\
+        \hline
+        Pithecia\_pithecia & 0.0016 & 0.22871 & 0.2 & 0.00737 & 0.2 \\
+        Colobus\_angolensis & 0.0017 & 0.00393 & 0.2 & 0.00416 & 0.2 \\
+        Saimiri\_boliviensis & 0.0013 & 0.00022 & 0.2 & 0.00045 & 0.2 \\
+        Pygathrix\_nemaeus & 0.0016 & 0.00347 & 0.2 & 0.00097 & 0.2 \\
+        $\hdots$ & $\hdots$ & $\hdots$ & $\hdots$ & $\hdots$ & $\hdots$ \\
+        \hline
+    \end{tabular}
+    \end{adjustbox}\label{tab:trait-variance}
+\end{table}
+
+\begin{itemize}
+    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
+    \item \emph{Nucleotide\_diversity}: the nucleotide diversity within species (neutral markers), cannot be \texttt{NaN}.
+    \item As many columns as traits, without spaces or special characters in the trait.
+    \item \emph{TraitName\_variance}: the phenotypic variance of the trait within species, can be \texttt{NaN} to indicate that the trait variance is not available for that taxon.
+    \item \emph{TraitName\_heritability} (optional): the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
+    \item The columns with the suffix \texttt{\_variance} and \texttt{\_heritability} are repeated for each trait.
+    \item \emph{TraitName\_heritability\_lower} (optional): the lower bound of the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
+    \item \emph{TraitName\_heritability\_upper} (optional): the upper bound of the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
+    \item If the columns with the suffix \texttt{\_heritability\_lower} and \texttt{\_heritability\_upper} are present, the heritability is randomly drawn from a uniform distribution between the lower and upper bounds.
+    \item If the columns with the suffix \texttt{\_heritability} is present, it is taken as is.
+    \item If the genetic variance (instead of phenotypic variance) is available for a trait, the heritability can be omitted and will automatically be set to 1.0.
+\end{itemize}
+
+\subsection{Bayesian estimation}\label{subsec:running-nodetraitsand-readnodetraits}
+
+The executable \texttt{nodetraits} from \textit{BayesCode} is used to run the Bayesian estimation of the model, and the executable \texttt{readnodetraits} is used to read the results.
+
+Assuming that the file \texttt{data/body\_size/mammals.male.tsv} contains the mean trait values for each species, the file \texttt{data/body\_size/mammals.male.var\_trait.tsv} contains the variation within species for each trait and the genetic variation within species (neutral markers), and the file \texttt{data/body\_size/mammals.male.tree} contains the phylogenetic tree, the following commands are used to run the model and read the results.
+
+\subsubsection{Running the model}\label{subsubsec:running-nodetraits}
+\texttt{nodetraits} is run with the following command:
+\begin{lstlisting}[language = sh,label={lst:nodetraits-run}]
+nodetraits  --until 2000
+            --tree data/body_size/mammals.male.tree
+            --traitsfile data/body_size/mammals.male.tsv
+            run_mammals_male
+\end{lstlisting}
+
+\subsubsection{Reading the results}\label{subsubsec:reading-the-results}
+Once the model has ran, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between species variation over within species variation with \texttt{readnodetraits}:
+\begin{lstlisting}[language = sh,label={lst:readnodetraits-rho}]
+readnodetraits --burnin 1000
+               --var_within data/body_size/mammals.male.var_trait.tsv
+               --output results_mammals_male.tsv
+               run_mammals_male
+\end{lstlisting}
+The file \texttt{data\_empirical/chain\_name.ratio.tsv} then contains the posterior mean of the ratio of between species variation over within species variation, the 95\% and 99\% credible interval, and the posterior probability that the ratio is greater than 1.
+
+\subsection{Maximum likelihood estimation}\label{subsec:maximum-likelihood-estimation}
+
+To obtain the ratio (without the posterior credible interval and probability) using maximum likelihood computation, the following python script can be used:
+\begin{lstlisting}[language = sh, label={lst:neutrality_index}]
+python3 utils/neutrality_index.py --tree data/body_size/mammals.male.tree
+                                  --traitsfile data/body_size/mammals.male.tsv
+                                  --var_within data/body_size/mammals.male.var_trait.tsv
+                                  --output results_ML_mammals_male.tsv
+\end{lstlisting}
 \end{document}

### 2023-07-17 manuscript/main.tex
Updated manuscript

@@
 \newcommand{\Trait}{P}
-\newcommand{\Heredity}{h^2}
-\newcommand{\HereditySpi}{\Heredity_{\Spi}}
+\newcommand{\Heritability}{h^2}
+\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
 \newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
@@
 \newcommand{\Root}{0}
-\newcommand{\RootTrait}{\widehat{\theta}}
+\newcommand{\RootTrait}{\Trait_{\Root}}
 \newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
+\newcommand{\VecZero}{\UniDimArray{0}}
 \newcommand{\VecOne}{\UniDimArray{1}}
@@
 \newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
-\newcommand{\RatePhy}{\sigma^2_{\mathrm{B}}}
-\newcommand{\RatePop}{\sigma^2_{\mathrm{W}}}
-\newcommand{\RatePopSpi}{\sigma^2_{\mathrm{W}, \Spi}}
-\newcommand{\VecRatePop}{\UniDimArray{\RatePop}}
-\newcommand{\EstRatePhy}{\widehat{\RatePhy}}
-\newcommand{\EstRatePop}{\widehat{\RatePop}}
+\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
+\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
+\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
+\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
+\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
+\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
 \newcommand{\NI}{\rho}
@@
 % Tree
-\newcommand{\branch}{\text{b}}
-\newcommand{\branchexp}{^{(\branch)}}
 \newcommand{\Nbranch}{2 \NbrTaxa - 2}
 \newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
-\newcommand{\Setbranch}{\branch \in \{1, \hdots, \Nbranch \}}
-\newcommand{\nodeUp}{\branch^{\uparrow}}
-\newcommand{\nodeDown}{\branch^{\downarrow}}
 \newcommand{\Ntrait}{K}
@@
 \newcommand{\Multivariate}{\UniDimArray{Z}}
-\newcommand{\indice}{a}
-\newcommand{\indiceexp}{^{(\indice)}}
-\newcommand{\vecZero}{\UniDimArray{0}}
-\newcommand{\vecOne}{\UniDimArray{1}}
 
@@
 
-\title{A test for diversifying selection for a trait from within and between species variations while integrating for genomic diversity}
-\rhead{\scshape Test of diversifying selection}
+\title{Trait' selection from within and between species variations taking in genomics}
+\rhead{\scshape Test of trait selection}
 
@@
 
-\keywords{Quantitative genetics \and Phylogenetics \and Population-genetics \and Selection }
+\keywords{Quantitative genetics \and Trait evolution \and Selection }
 
@@
 Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
-Fundamentally, distinguishing neutral evolution from selection requires determining which mode of selection is supported by the observed variation of traits and sequences.
-Trait variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.
+Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits and sequences.
+Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
+All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.
 
-First, at the micro-evolutionary scale, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
-At the species level, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
+First, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
+Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
 For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
-However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
-Theoretically, change in traits accumulate linearly with time from divergence for a pair of species, and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
-For example, gene expression accumulate linearly with time from divergence, and accumulates faster for trait with large variance~\cite{khaitovich_neutral_2004}.
-Such theoretical connection between trait variance and change in mean value can be used to test for a selection of trait using for a pair of species~\cite{walsh_evolution_2018}.
-However, to our knowledge, a neutral theoretical expectation for trait changes across several species related by their phylogenetic tree, as a function of within species trait variation has not yet been formulated and tested.
+However, dominant selective regime acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
+For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
+Gene expression exhibit such accumulation, and divergence in expression accumulates faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
+Altogether, both the trait variance and the evolution in mean value can be used to test for a selection of trait in a pair of species~\cite{walsh_evolution_2018}.
 
 % Phylogenetics (mean trait evolution)
-Contrarily, at the macro-evolutionary scale, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
-Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a standard deviation proportional to the squared root of the divergence time~\cite{hansen_translating_1996}.
+Alternatively, at the macro-evolutionary scale, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
+Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\cite{hansen_translating_1996}.
 As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
@@
 However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
-First, and unfortunately, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
-A trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
+First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
+In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
 Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-In such comparative framework, the variation within species is usually discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
-Altogether, methods at the phylogenetic scale easily mis-classify selection and also discards the variance in traits between individuals, leaving out valuable information.
+Altogether, methods at the phylogenetic scale easily mis-classify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.
 
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+At the frontier between micro and macro evolution, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
 Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
-In this study, we propose a neutrality index for a quantitative trait articulating trait and sequence variation within and between species.
-Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
-From a population-geneticist perspective, our study can be seen as the generalization at the macro scale of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
-From a phylogeneticist perspective, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} where we set a clear threshold for neutral evolution by leveraging species' sequence polymorphism and divergence.
+In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
+Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
+For population-geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
+For a phylogeneticist, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
 % Basically, we are testing whether the variance of means is greater than the man of variances
@@
 For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
-New mutations are thus generating variance for such a trait.
-At the individual level, the mutational variance ($\VarMutation = \NbrLoci \Multiply  \E [a^2]$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is also a function of the mutation rate per loci per generation ($\MutationRate$), as:
+New mutations are thus generating variance for such a trait, where the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
+At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
+As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$ as:
 \begin{gather}
@@
 
-At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
-As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
+Mutations supply new variants, while genetic random drift depletes standing variation.
+As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
 \begin{align}
@@
 
-Moreover, for any neutral genomic region of interest, the genetic diversity, called $\pi$, is measured as the number of mutation segregating in the population divided by the size of the region.
+For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutation segregating divided by the size of the region.
 Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
-As shown in \textcite{tajima_statistical_1989}, $\pi$ is thus a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$), given as:
+As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
 \begin{gather}
@@
 
-Altogether, we define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$).
-This ratio allows discarding the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePop$ as:
+Altogether, we define $\RateWhithin$ as the ratio of genetic variance of the trait ($\VarGenetic$) over nucleotide diversity of any neutral genomic region of interest ($\pi$).
+This ratio allows to discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
 \begin{align}
-    \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
+    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
     & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
@@
 
-Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heredity$), giving also $\RatePop$ as:
+Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), giving also $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
 \begin{align}
-    \RatePop & = \frac{\Heredity \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
+    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
 \end{align}
@@
 
-Along a lineage, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
+At generation $\Time$, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
 If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
-After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
+After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
 \begin{align}
@@
 
-Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
+Moreover, for any neutral genomic region of interest, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
 For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
@@
 
-Thus, after having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
+After having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
 \begin{align}
@@
 \end{align}
-Altogether, we defined $\RatePhy$ as the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$).
-This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePhy$ as:
+Altogether, we define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region of interest ($d$).
+This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateBetween$ as another proxy of $\RateMut$:
 \begin{align}
-    \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
+    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
     & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
@@
     \caption{
-        Between species, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
-        Within species, genetic variance ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
-        Under neutral evolution $\RatePhy$ should equal $\RatePop$.
-        Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
+        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalised by nucleotide divergence.
+        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalised by nucleotide diversity.
+        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
+        Under neutral evolution, $\EstNI$ is expected to be equal to one.
+        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
+        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
+        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the counfunding effect on diversity of mutation rate, population size and divergence time.
     }
@@
 The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
- At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
-Combined as a ratio, we obtain $\RatePop$ at the population level, which depends solely on the genetic architecture of the trait (as $\RateMut$).
+Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
+Combined as a ratio, $\RateWhithin$ depends solely on the genetic architecture of the trait as $\RateMut$.
 At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
-Similarly, combined as a ratio, we obtain $\RatePhy$ at the phylogenetic scale, which also depends solely on the genetic architecture of the trait.
+Similarly, combined as a ratio, $\RateBetween$ depends solely on $\RateMut$
 Altogether, we thus have:
 \begin{gather}
-    \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}} , \\
-    \Rightarrow \frac{\RatePhy}{\RatePop} = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
+    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
+    \Rightarrow \frac{\RateBetween}{\RateWhithin} = 1.
 \end{gather}
-Altogether, $\RatePhy / \RatePop$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
-Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.
+Altogether, $\RateBetween / \RateWhithin$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
+Importantly, $\RateBetween$ and $\RateWhithin$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
+Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
 
@@
 \subsubsection*{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
+At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in unit of substitution per site) and the topology of the phylogenetic tree.
 Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
@@
 
-Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated~\cite{omeara_testing_2006} as:
+Finally, within species variation $\EstRateBetween$ is estimated as~\cite{omeara_testing_2006}:
 \begin{gather}
-    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
+    \EstRateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
 \label{eq:estimated-rate-phy}
@@
 
-For a given species $\Spi$ with inter-individual data available, the ratio of genetic variance of a trait ($\VarGeneticSpi$) over genetic diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RatePop$.
-Averaged across all species, we obtain the estimate $\EstRatePop$ as:
+For a given species $\Spi$ with inter-individual data available, the ratio of genetic variance of a trait ($\VarGeneticSpi$) over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
+Averaged across all species, we obtain the estimate $\EstRateWhithin$ as:
 \begin{gather}
-    \EstRatePop = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heredity_{i}}{ \pi_{i}}.
+    \EstRateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}.
 \label{eq:estimated-rate-pop}
@@
 
-Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
-
 Altogether, the neutrality index is estimated as:
 \begin{gather}
-    \EstNI = \frac{\EstRatePhy}{\EstRatePop}. \label{eq:estimated-NI}
+    \EstNI = \frac{\EstRateBetween}{\EstRateWhithin}. \label{eq:estimated-NI}
 \end{gather}
@@
 
-Even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
+Even though the neutrality index can be applied for any trait, we so far assumed that several traits would to be independent of one another.
 Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
@@
 The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
-We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\NI > 1 ]$ to test for evidence of diversifying selection and $\proba [\NI < 1 ]$ to test for evidence of stabilizing selection.
+With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
+$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
+The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
 
@@
 
-We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator} and section~S\ref{subsec:genotype-phenotype-map}).
-We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
-Assuming a diploid panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
-Genetic drift is modelled by the multinomial resampling of the currently segregating alleles.
-The number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $.
-The changes in the mutation rate per generation ($\MutationRate$) and $\Ne$ along the lineages is modelled by a geometric Brownian process.
-Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
+% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
+% pS = 0.0027577478571428568
+% Tree length (dS) = 2.340579
+% Tree length (My) = 1259.4490200000002
+% Root age (My) = 98.9489413888889
+% Root age (dS) = 0.18388820079995308
+% nbr_sites_var = 27.577478571428568
+% u = 1.3788739285714283e-05
+% nbr_generations = 13336.114128321291
+We assumed that the trait genotypic value is encoded by $\NbrLoci=5,000$ loci, with each locus contributing additively to the genotypic value.
+Each effect is Normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
+The phenotypic value is the sum of phenotypic value and an environmental effect, the environmental effect is normally distributed with variance $\VarEnv$.
+We assumed a trait with an heritability of $\Heritability =0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
+We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
+Assuming a diploid panmictic population of size $\Ne=50$ at the root of tree, and with non-overlapping generations, we can simulate explicitly each generation along an ultrametric phylogenetic tree.
+For each offspring, the number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
+From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
+To scale parameters with a mammalian regime, we used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38e^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38e^{-5} = 13,500$ generations from root to leaves, and the number of generation along each branch is proportional to the branch length.
+
+The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages are both modelled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
+Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
 The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
-The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
+The simulations starts from an initial sequence at equilibrium at the root of the tree and, at each node, the process is split until it finally reaches the leaves of the tree.
 From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
+
+Genetic drift is modelled by the resampling of individuals at each generation, with each parent probability of being sampled being proportional to its fitness ($W$).
 Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
-The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype (see section~S\ref{subsec:phenotype-fitness-map}).
+More specifically, the fitness of an individual is given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ is the trait value of the individual, $\lambda=0.0$ is the optimal trait value, and $\alpha=0.02$ is the strength of selection.
 Mutations are seen has displacement of the phenotype in the multidimensional space.
 Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
-Stabilizing selection is implemented by means of a fixed optimum phenotype.
-Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.
-Genetic diversity is measured as the heterozygosity of the neutral markers, and the genetic variance of the trait is measured as the variance of the trait genotypic value.
-Genetic divergence is measured as the number of substitutions per site along branches of the phylogeny.
-Heritability is estimated from the slope of the regression of offspring trait values on parental trait values~\cite{lynch_genetics_1998}, using the information from the last two generations of the simulation.
+Stabilizing selection is implemented by means of a fixed optimum phenotype ($\lambda=0.0$).
+Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process ($\brownian \left(0, \sigma_{\lambda}=1.0\right)$).
+Neutral evolution is implemented with a fixed fitness landscape ($W=1$), where each individual has the same probability of being sampled at each generation.
+
+Nucleotide diversity ($\pi$) is measured as the heterozygosity of the neutral markers.
+Nucleotide divergence ($d$) is measured as the number of substitutions per site along branches of the phylogeny.
+The genetic variance is measured as phenotypic variance multiplied by heritability.
+Heritability is estimated from the slope of the regression of offspring's phenotypic trait values on parental's phenotypic trait values~\cite{lynch_genetics_1998} using the last generations of the simulations.
+As such heritability is not a given parameter of the simulations, but rather measured as it would be in empirical data.
 
@@
         Mutations are drawn from a Poisson distribution, with each loci having a probability $\MutationRate$ to mutate.
-        Drift is modelled by the resampling of parents in a finite population.
+        Drift is modelled by the resampling of parents.
         Panel B: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (grey area).
@@
 
-We first focused on a mammalian dataset, with the body size and brain mass as quantitative traits.
-For the genetic variation, the mammalian phylogeny is extracted from the Zoonomia project, obtained on a set of neutral markers in \textcite{foley_genomic_2023}, with genetic diversity measured as heterozygosity in \textcite{wilder_contribution_2023}
-For quantitative trait variation, the log-transformed body size and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
-We removed individuals not marked as adult, and split the data into males and females.
+We analyzed a mammalian dataset, with the body size and brain mass as quantitative traits.
+The log-transformed values of body size and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
+We removed individuals not marked as adult, and split the data into males and females due to sexual dimorphism.
 We discarded species with no intra-specific variability.
+The mammalian nucleotide diversity is obtained from the Zoonomia project, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
 
-We also focused on a dataset of Primates species, with the genetic variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
-Unfortunately, the primate phylogeny was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
+We also analyzed on a dataset of Primates species, with the nucleotide variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
+However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
 
@@
 The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
-More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$), denoted $\RatePop$, is a proxy for the mutational variance ($\RateMut$).
-At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
-This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), denoted $\RatePhy$, allowing us to obtain another proxy for the mutational variance ($\RateMut$).
-Altogether, we define the neutrality index as $\NI = \RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
+More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by nucleotide diversity of any neutral genomic region ($\pi$), denoted $\RateWhithin$, is a proxy for the mutational variance $\RateMut$.
+At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a $\Time$ generations is $\VarPhy$.
+This variance can also be normalized by the nucleotide divergence of any neutral genomic region ($d$), denoted $\RateBetween$, allowing us to obtain another proxy for $\RateMut$.
+Altogether, we defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
 Under a depleted variation between species, meaning $\NI < 1$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
@@
             \hline
-            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
+            Assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
             Trait encoded by few loci                        & Between species trait variation is under-estimated & --              & Under-estimated & Conservative & Invalid  \\ \hline
@@
     \end{adjustbox}
-    \caption{Assumptions and their consequences on the estimation of within species variation ($\EstRatePop$), between species variation ($\EstRatePhy$), and on the neutrality index $\NI = \EstRatePhy/\EstRatePop$.
+    \caption{Assumptions and their consequences on the estimation of within species variation ($\EstRateWhithin$), between species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
     The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
@@
 
-Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
+Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate, population size and divergence time.
 The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
@@
 The inference framework was first tested on independently simulated datasets.
-With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant empirical regime.
-Thus, the tree topology and the branch lengths were chosen based on a tree estimated on the mammalian dataset further considered below.
+With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant mammalian empirical regime (see Materials and Methods).
+First, we assumed a constant population size ($\Ne$) and a constant mutation rate ($\MutationRate$) across the phylogeny (fig.~\ref{fig:results-simulations}A\&B)
+Simulations performed under stabilizing selection where detected as such ($\proba [\EstNI < 1] > 0.95$), no false negative was observed (blue in fig.~\ref{fig:results-simulations}A\&B).
+Simulations performed under diversifying selection were detected as such ($\proba [\EstNI > 1] > 0.975$), no false negative was observed (red in fig.~\ref{fig:results-simulations}A\&B).
+Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 79\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B), the remaining cases were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) in 21\% of the simulations.
+Second, for simulations under a fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}C\&D), simulations of diversifying selection was detected as such and stabilizing selection was detected as such in 100\% of the cases.
+For simulations under neutral evolution, 46\% of the simulations were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), 54\% of the simulations were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
 
@@
     \caption{
-        $\RatePhy / \RatePop$ for $1,000$ replicates under different evolutionary regimes.
-        Traits are encoded by $5,000$ loci with $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
-        Traits simulated under stabilizing selection in yellow, under a neutral evolution in blue, and under a diver optimum (in red).
-        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
+        $1,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
+        Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
+        Histogram of the ratio of between species trait variation ($\RateBetween$) over within species trait variation $\RateWhithin$, as $\NI = \RateBetween / \RateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\NI$ is greater than $1$ (right).
+        Effective population size ($\Ne$) and mutation rate ($\MutationRate$) are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
     }
-    \label{fig:constant_pop_size_phy_pop}
+    \label{fig:results-simulations}
 \end{figure*}
@@
 
+% Summary of the method and results
 In this study, we proposed a neutrality index for a quantitative trait.
@@
 Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
-Our neutrality inde for a trait, called $\NI$, is the ratio of between species variation to within species variation.
+Our neutrality index for a trait, $\NI$, is the ratio of between species variation to within species variation.
 $\NI$ can be compared to 1 to seek a statistical deviation from a neutral trait.
@@
 $ \proba [ \NI > 1 ]$).
-Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits ($\Heredity$).
-We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
-By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time~\cite{litsios_effects_2012}.
-Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes~\cite{seo_estimating_2004}.
+Importantly, our neutrality index leverages nucleotide divergence and polymorphism, as well as the heritability of traits ($\Heritability$).
+We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize the variation of trait between and within species.
+By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, while not requiring to estimate divergence time~\cite{litsios_effects_2012}.
+Instead, the normalization by nucleotide divergence and polymorphism automatically absorbs these changes~\cite{seo_estimating_2004}.
 
-We argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically harder to measure for any trait.
+% Pitfalls of the method
+We argue that the main drawback of our method is that the heritability ($\Heritability$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically harder to measure for any trait.
 Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is under-powered.
-Indeed, because $\Heredity \leq 1$ by definition, if we set $\Heredity \leq 1$, we overestimate the within species variation and thus underestimate $ \NI$ and can test for diversifying selection.
-In other words, having $\NI > 1$ while setting $\Heredity = 1$ means that knowing the real heritability would only increase the evidence for diversifying selection.
+Indeed, because $\Heritability \leq 1$ by definition, if we set $\Heritability \leq 1$, we overestimate the within species variation and thus underestimate $ \NI$ and can test for diversifying selection.
+In other words, testing for $\NI > 1$ while setting $\Heritability = 1$ means that knowing the real heritability would only increase the evidence for diversifying selection.
 Contrarily, the test for stabilizing selection is invalid if heritability is not known.
@@
 As example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$.
-However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for genetic divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
+However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for nucleotide divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
 
-% Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
-More theoretically, our development is based on several assumption that could be relaxed.
+% What can and should be improved
+Our development is based on several assumption that could be relaxed.
 First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
@@
 
-Theoretically, our test bears analogy with others tests for neutrality developed, providing insight into its behavior.
-First, taking inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
-Second, because $\NI$ is compared to 1, our test bears analogy to the codon-based test of selection of \textcite{goldman_codonbased_1994, muse_likelihood_1994}, where the ratio of non-synonymous to synonymous substitutions ($\omega$) is compared to 1.
-As $\omega < 1$ is interpreted as purifying selection, $\NI < 1$ is interpreted as stabilizing selection.
-Also, as $\omega > 1$ is interpreted as diversifying selection, $\NI > 1$ is interpreted as adaptation.
+% How is our test related to other tests of selection
+Our test bears analogy with others tests for neutrality developed, providing insight into its behavior.
+First, our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
+Second, because $\NI$ is compared to 1, our test bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\cite{goldman_codonbased_1994, muse_likelihood_1994}.
+As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
+Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
 With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\cite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the hypothesis that $\omega = 1$ does imply a neutral evolution, since a combination of adaptation and purifying selection can also lead to $\omega = 1$~\cite{nielsen_molecular_2005}.
-Analogously, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$.
-Secondly, $\omega < 1$ is not evidence for the absence of adaptation, but rather that purifying selection is strong enough to overcome the effect of adaptation.
-For example, by explicitly modelling purifying selection it is possible to detect genes and sites under adaptation even though they display $\omega < 1$ ~\cite{latrille_genes_2023}.
-Similarly, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection.
-We thus argue that by explicitly modelling stabilizing selection as a moving optimum, it would theoretically allow to tease-out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
+First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\cite{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\cite{latrille_genes_2023}.
+By explicitly modelling stabilizing selection as a moving optimum, it would theoretically allow to tease-out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
 
-% What about the other methods? What are we testing for?
-From an empirical point of view, our test is also related to other methods that have been developed to test for selection on quantitative traits.
-At the phylogenetic scale, the change in mean trait value across different species is typically used to test for neutrality for a quantitative trait.
-These models are seeking deviation from a Brownian process~\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation~\cite{rohlfs_phylogenetic_2015}.
-For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\cite{catalan_drift_2019}.
-However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
+% What can our test bring to the field
+Finally, from an empirical point of view, our test is also related to other methods that have been developed to test for selection on quantitative traits.
+At the phylogenetic scale, test for neutrality of a quantitative trait are seeking deviation from a Brownian process~\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation~\cite{rohlfs_phylogenetic_2015}.
+Ahe analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\cite{catalan_drift_2019}.
+However, the alternative model of a moving optimum was not implemented, and thus can not be ruled out by this analysis.
 With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
-Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$).
-Altogether, our study provides a statistical framework to test for neutrality of a quantitative trait while integrating the trove of genomic data available both within and between species.
-, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
+Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence ($d$) and polymorphism ($\pi$).
+Altogether, our study provides a statistical framework to test for neutrality of a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
 
@@
 
+\renewcommand{\baselinestretch}{1.0}\normalsize
 \tableofcontents
+\renewcommand{\baselinestretch}{1.5}\normalsize
 
+\newpage
 \section{Genetic architecture of the trait}\label{sec:simulator}
@@
 \begin{itemize}
-    \setlength\itemsep{-0.2em}
     \item $\NbrLoci$ is the number of loci encoding the trait.
     \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
-    \item $\Ne$ is the effective number of individual in the population.
+    \item $\Ne$ is the effective number of individual.
     \item $g_{i,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
@@
 
-At the population level, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
+Within species, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
 \begin{equation}
     \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad} \VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2\label{eq:simu-genotype}
+\end{equation}
+The theoretical additive genetic variance ($\VarGenetic$) is a function of the number of loci ($\NbrLoci$) and the effect of a mutation ($a$) as:
+\begin{equation}
+    \VarGenetic = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \label{eq:simu-var-genetic}
 \end{equation}
@@
 \end{equation}
-Heritability can also be estimated from the slope of the linear regression of the phenotype on the genotype.
+Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($h^2$) as:
+\begin{equation}
+    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{h^2} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{h^2} - 1 \right) \label{eq:simu-var-env}
+\end{equation}
 
-\subsection{Phenotype-fitness map}\label{subsec:phenotype-fitness-map}
-
-\begin{itemize}
-    \item $W_i$ is the weight for individual $i$ to be chosen for the next generation.
-    \item Neutral model with no fitness function: $W_i = 1$.
-    \item Stabilizing selection with fitness $W_i = \e^{(\Trait_i - \theta)^2/ \sigma}$, parameters $\theta=0$ and $\sigma=0.02$.
-    \item Moving optimum with fitness $W_i = \e^{(\Trait_i - \theta_t)^2/ \sigma}$, parameters $\theta_t \sim \mathcal{B}_t$ and $\sigma=0.02$.
-\end{itemize}
-
+\newpage
 \section{Bayesian estimate}\label{sec:bayesian-estimate}
@@
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-For a branch $\branch$ with length $d\branchexp$, the Brownian process start at node $\nodeUp$ with value $\Brownian(\nodeUp)$, at ends at node $\nodeDown$ with value $\Brownian(\nodeDown)$.
-The independent contrast $\contrast\branchexp$ defined as change in trait along the branch normalized by $\sqrt {d}$ is a multivariate Gaussian:
+Along branch $j$ with length $d_{j}$, the Brownian process start at the ancestral node $\mathcal{A}(j)$ with value $\Brownian(\mathcal{A}(j))$, and ends at node $\mathcal{R}(j)$  with value $\Brownian(\mathcal{R}(j))$.
+The independent contrast $\contrast_{j}$ defined as change in trait along the branch normalized by $\sqrt {d_{j}}$ is a multivariate Gaussian:
 \begin{equation}
     \label{eq:DistribBrownian}
-    \contrast\branchexp = \frac{\Brownian^{(\nodeDown)} - \Brownian^{(\nodeUp)} }{\sqrt {d}} \sim \mathcal{N}\left(\vecZero, \CovarianceMatrix \right).
+    \contrast_{j} = \frac{\Brownian_{\mathcal{R}(j)} - \Brownian_{\mathcal{A}(j)} }{\sqrt {d_{j}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
 \end{equation}
-We defined the {prior} on the covariance matrix as an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
+The {prior} on the covariance matrix as an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
 \begin{equation}
@@
 \subsection{Sampling the covariance matrix}\label{subsec:sampling-the-covariance-matrix}
-From the independent contrast at each branch of the tree ($\contrast\branchexp$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$ as:
+From the independent contrast at each branch of the tree ($\contrast_{j}$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$, as:
 \begin{equation}
-    \Scattermatrix = \sum\limits_{b=1}^{\Nbranch} \contrast\branchexp \MultiplyMatrix \left[\contrast\branchexp\right]\tr\label{eq:bayes-scatter}
+    \Scattermatrix = \sum\limits_{j=1}^{\Nbranch} \contrast_{j} \MultiplyMatrix \left[\contrast_{j}\right]\tr\label{eq:bayes-scatter},
 \end{equation}
-By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
+where $\Nbranch$ is the number of branches in the tree and $\NbrTaxa$ the number of taxa.
+
+By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\Brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
 \begin{equation}
@@
 \end{equation}
-This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate\indiceexp$ defined by
+This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate_{k}$ defined by
 \begin{equation}
-    \Multivariate\indiceexp \sim \mathcal{N} \left( \vecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
+    \Multivariate_{k} \sim \mathcal{N} \left( \VecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
 \end{equation}
@@
 \begin{equation}
-    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate\indiceexp \MultiplyMatrix  \left[\Multivariate\indiceexp \right] \tr \right)\inv \label{eq:bayes-gibbs}
+    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate_{k} \MultiplyMatrix  \left[\Multivariate_{k} \right] \tr \right)\inv \label{eq:bayes-gibbs}
 \end{equation}
 
+\newpage
 \section{Bayesian and Maximum-likelihood implementation}\label{sec:implementation}
@@
 Running the analysis on your dataset and compute posterior probabilities requires three files:
-
 \begin{enumerate}
@@
 The file containing mean trait values for each species must be in a tab-delimited file with the following format:
-
-\begin{table}[h]
-    \centering
+\begin{center}
+    \begin{adjustbox}{width = 0.35\textwidth}
         \begin{tabular}{|l|c|c|}
@@
             Saimiri\_boliviensis & 6.845      & 3.279       \\
-        $\hdots$ & $\hdots$ & $\hdots$ \\
+            $\vdots$             & $\vdots$   & $\vdots$    \\
             \hline
         \end{tabular}\label{tab:trait-mean}
-\end{table}
+    \end{adjustbox}
+\end{center}
 
 The columns are:
-
 \begin{itemize}
@@
 
+\newpage
 \subsubsection{Trait variation for each species}\label{subsubsec:trait-variation-for-each-species}
@@
 The file containing trait variation for each species must be in a tab-delimited file with the following format:
-
-\begin{table}[h]
-    \centering
+\begin{center}
     \begin{adjustbox}{width = 1.0\textwidth}
@@
             Pygathrix\_nemaeus   & 0.0016                & 0.00347              & 0.2                      & 0.00097               & 0.2                       \\
-        $\hdots$ & $\hdots$ & $\hdots$ & $\hdots$ & $\hdots$ & $\hdots$ \\
+            $\vdots$             & $\vdots$              & $\vdots$             & $\vdots$                 & $\vdots$              & $\vdots$                  \\
             \hline
         \end{tabular}
-    \end{adjustbox}\label{tab:trait-variance}
-\end{table}
+        \label{tab:trait-variance}
+    \end{adjustbox}
+\end{center}
 
@@
 
+\newpage
 \subsection{Bayesian estimation}\label{subsec:running-nodetraitsand-readnodetraits}

### 2023-07-18 manuscript/main.tex
Updated manuscript. Ensure simulated variances are >0

@@
 
-\title{Trait' selection from within and between species variations taking in genomics}
-\rhead{\scshape Test of trait selection}
+\title{Trait selection from within and between species variations while integrating genomics}
+\rhead{\scshape Trait selection from within and between species variations}
 
@@
 $\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
+For each trait and each species, heritability ($\Heritability$) is set as a uniform distribution with user-defined boundaries.
 The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
@@
 
-We analyzed a mammalian dataset, with the body size and brain mass as quantitative traits.
-The log-transformed values of body size and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
+We analyzed a mammalian dataset, with the body mass and brain mass as quantitative traits.
+The log-transformed values of body mass and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
 We removed individuals not marked as adult, and split the data into males and females due to sexual dimorphism.
@@
 Simulations performed under diversifying selection were detected as such ($\proba [\EstNI > 1] > 0.975$), no false negative was observed (red in fig.~\ref{fig:results-simulations}A\&B).
-Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 79\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B), the remaining cases were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) in 21\% of the simulations.
+Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 77\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B),while detected as stabilizing selection in 21\% of the cases, and detected as diversifying selection in 2\% of cases.
 Second, for simulations under a fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}C\&D), simulations of diversifying selection was detected as such and stabilizing selection was detected as such in 100\% of the cases.
-For simulations under neutral evolution, 46\% of the simulations were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), 54\% of the simulations were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
+For simulations under neutral evolution, 51\% of the simulations were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), 49\% of the simulations were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
 
@@
 \subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
-For mammalian body size, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
-The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.346$ ($[0.183,~0.677]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.278$ ($[0.136,~0.639]$ for $95\%$ credible interval).
-For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.375$ ($[0.844,~2.212]$ for $95\%$ credible interval, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.762$ ($[1.021,~3.010]$ for $95\%$ credible interval, $\NbrTaxa=26$).
-In both males and females, brain mass is under diversifying selection, with p-values of $0.116$ and $0.022$ respectively.
+For mammalian body mass, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
+The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.34$ ($[0.22--0.52]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.28$ ($[0.16--0.49]$ for $95\%$ CI).
+In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.0$ and $0.0$ respectively.
+Moreover, assuming that heritability ($\Heritability$) of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection ($ \proba [\EstNI > 1]$) becomes $0.635$ in males and $0.914$ in females.
+For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.35$ ($[0.85--2.17]$ for $95\%$ CI, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.72$ ($[1.00--2.94]$ for $95\%$ CI, $\NbrTaxa=26$).
+In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.877$ and $0.972$ respectively.
+Moreover, assuming that heritability of brain mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in both males and females.
 
-For primates body size, we obtained $\EstNI_{\text{\Male}} = 0.557$ ($[0.406,~0.767]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.393$ ($[0.277,~0.561]$ for $95\%$ credible interval, $\NbrTaxa=65$).
-For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.920$ ($[1.397,~2.616]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.974$ ($[1.393,~2.798]$ for $95\%$ credible interval, $\NbrTaxa=65$).
-In both males and females, brain mass is under diversifying selection, with p-values of $0.00$ and $0.00$ respectively.
-However, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).
+For primates body mass, we obtained $\EstNI_{\text{\Male}} = 0.56$ ($[0.40,~0.78]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.39$ ($[0.28,~0.55]$ for $95\%$ CI, $\NbrTaxa=65$).
+Assuming that heritability of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in males and $0.914$ in females.
+For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.95$ ($[1.40,~2.79]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.95$ ($[1.39,~2.79]$ for $95\%$ CI, $\NbrTaxa=65$).
+In both males and females, brain mass is under diversifying selection, with associated posterior probabilities of $1.0$.
+Importantly, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).
 
@@
 
-Heritability ($h^2$) is defined as:
+Heritability ($\Heritability$) is defined as:
 \begin{equation}
-    h^2 = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
+    \Heritability = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
 \end{equation}
-Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($h^2$) as:
+Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($\Heritability$) as:
 \begin{equation}
-    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{h^2} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{h^2} - 1 \right) \label{eq:simu-var-env}
+    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
 \end{equation}

### 2023-07-18 data_simulated/update_snake.sh

@@ -4,6 +4,6 @@ for EXPERIMENT in $(pwd)/experiments/*; do
   rm -rf "${EXPERIMENT}"/Snakefile
   ln $(pwd)/Snakefile "${EXPERIMENT}"/Snakefile
   cd "${EXPERIMENT}"
-  snakemake -j 8 --rerun-incomplete
+  snakemake -j 8 --touch
   cd "${FOLDER}"
 done

### 2023-07-18 scripts/pre_processed_traits_simulations.py

@@ -31,10 +31,10 @@ def main(input_path, input_neutral_path, tree_path, traits_path, var_within_path
     for n, n_neutral in zip(tree.get_leaves(), neutral_tree.get_leaves()):
         assert n.name == n_neutral.name
         var_within["TaxonName"].append(n.name)
-        var_within["Nucleotide_diversity"].append(float(getattr(n_neutral, "Theta")))
-        var_within["Phenotype_variance"].append(float(getattr(n, "Phenotype_var")))
-        var_within["Phenotype_heritability"].append(float(getattr(n, "BranchMean_Heritability")))
-        var_within["Genotype_variance"].append(float(getattr(n, "Genotype_var")))
+        var_within["Nucleotide_diversity"].append(max([0, float(getattr(n_neutral, "Theta"))]))
+        var_within["Phenotype_variance"].append(max([0, float(getattr(n, "Phenotype_var"))]))
+        var_within["Phenotype_heritability"].append(max([0, float(getattr(n, "BranchMean_Heritability"))]))
+        var_within["Genotype_variance"].append(max([0, float(getattr(n, "Genotype_var"))]))
         var_within["Genotype_heritability"].append(1.0)
     df = pd.DataFrame(var_within)
     df.to_csv(var_within_path, sep="\t", index=False)

### 2023-07-25 manuscript/main.tex
Updated manuscript

@@
 
-\title{Trait selection from within and between species variations while integrating genomics}
+% Including genomics, Inclusion of genomics, Linking genomics,
+\title{Leveraging genomics to detect trait selection from within and between species variations}
 \rhead{\scshape Trait selection from within and between species variations}
@@
     Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
-    Conversely, using between species variation, current methods usually discards within species variation, thus not accounting for processes at the micro-evolutionary scale.
+    Conversely, using between species variation, current methods usually discard within species variation, thus not accounting for processes at the micro-evolutionary scale.
     The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
@@
 Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.
+All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.
 
-First, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
+First, leveraging individual variations within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given locus), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
 Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
@@
 For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
-Gene expression exhibit such accumulation, and divergence in expression accumulates faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for a selection of trait in a pair of species~\cite{walsh_evolution_2018}.
+Gene expression exhibit such accumulation and divergence in expression accumulate faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\cite{walsh_evolution_2018}.
 
@@
 In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
-Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Altogether, methods at the phylogenetic scale easily mis-classify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.
+Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Altogether, methods at the phylogenetic scale easily misclassify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.
 
@@
 In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
-Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
-For population-geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
-For a phylogeneticist, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
+Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimate of population size (within species) nor of time of speciation (between species).
+For population geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
+For a phylogeneticist, our study can be seen as an extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
 % Basically, we are testing whether the variance of means is greater than the man of variances
@@
 
-For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutation segregating divided by the size of the region.
-Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
+For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating divided by the size of the region.
+Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the number of segregating mutations is also a balance between mutations and drift.
 As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
@@
 
-Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), giving also $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
+Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), also giving $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
 \begin{align}
@@
 \end{align}
-That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.
+That is, if mutations are neutral, the rate with which new neutral alleles are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.
 
@@
 Altogether, we define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region of interest ($d$).
-This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateBetween$ as another proxy of $\RateMut$:
+This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
 \begin{align}
@@
     \caption{
-        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalised by nucleotide divergence.
-        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalised by nucleotide diversity.
+        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
+        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
         $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
@@
         Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
-        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the counfunding effect on diversity of mutation rate, population size and divergence time.
+        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the confounding effect on diversity of mutation rate, population size and divergence time.
     }
@@
 The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
-Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
+Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allows to compute $\pi$.
 Combined as a ratio, $\RateWhithin$ depends solely on the genetic architecture of the trait as $\RateMut$.
@@
 \subsubsection*{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in unit of substitution per site) and the topology of the phylogenetic tree.
+At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitution per site) and the topology of the phylogenetic tree.
 Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
-The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be root, in which case the distance is 0).
+The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be the root, in which case the distance is 0).
 Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate~\cite{omeara_testing_2006}, given as:
@@
 Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
-Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
+Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
@@
 With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
-$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
+$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
 For each trait and each species, heritability ($\Heritability$) is set as a uniform distribution with user-defined boundaries.
@@
 The phenotypic value is the sum of phenotypic value and an environmental effect, the environmental effect is normally distributed with variance $\VarEnv$.
-We assumed a trait with an heritability of $\Heritability =0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
-We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
-Assuming a diploid panmictic population of size $\Ne=50$ at the root of tree, and with non-overlapping generations, we can simulate explicitly each generation along an ultrametric phylogenetic tree.
+We assumed a trait with a heritability of $\Heritability =0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
+We will follow a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
+Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we can simulate explicitly each generation along an ultrametric phylogenetic tree.
 For each offspring, the number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
 From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
-To scale parameters with a mammalian regime, we used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38e^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38e^{-5} = 13,500$ generations from root to leaves, and the number of generation along each branch is proportional to the branch length.
+To scale parameters with a mammalian regime, we used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38e^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38e^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch is proportional to the branch length.
 
-The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages are both modelled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
-Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
+The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages are both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
+Additionally, the instant value of log-$\Ne$ is modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
 The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
-The simulations starts from an initial sequence at equilibrium at the root of the tree and, at each node, the process is split until it finally reaches the leaves of the tree.
-From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
+The simulation starts from an initial sequence at equilibrium at the root of the tree and, at each node, the process is split until it finally reaches the leaves of the tree.
+From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.
 
-Genetic drift is modelled by the resampling of individuals at each generation, with each parent probability of being sampled being proportional to its fitness ($W$).
-Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+Genetic drift is modeled by the resampling of individuals at each generation, with each parent probability of being sampled being proportional to its fitness ($W$).
+Selection is modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
 More specifically, the fitness of an individual is given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ is the trait value of the individual, $\lambda=0.0$ is the optimal trait value, and $\alpha=0.02$ is the strength of selection.
 Mutations are seen has displacement of the phenotype in the multidimensional space.
-Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
+Beneficial mutations move the phenotype closer to the optimum, while deleterious mutations move it further away.
 Stabilizing selection is implemented by means of a fixed optimum phenotype ($\lambda=0.0$).
@@
 The genetic variance is measured as phenotypic variance multiplied by heritability.
-Heritability is estimated from the slope of the regression of offspring's phenotypic trait values on parental's phenotypic trait values~\cite{lynch_genetics_1998} using the last generations of the simulations.
+Heritability is estimated from the slope of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\cite{lynch_genetics_1998} using the last generations of the simulations.
 As such heritability is not a given parameter of the simulations, but rather measured as it would be in empirical data.
@@
         Parents are selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
-        Mutations are drawn from a Poisson distribution, with each loci having a probability $\MutationRate$ to mutate.
-        Drift is modelled by the resampling of parents.
-        Panel B: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (grey area).
+        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRate$ to mutate.
+        Drift is modeled by the resampling of parents.
+        Panel B: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
     }
@@
 The log-transformed values of body mass and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
-We removed individuals not marked as adult, and split the data into males and females due to sexual dimorphism.
-We discarded species with no intra-specific variability.
+We removed individuals not marked as an adult, and split the data into males and females due to sexual dimorphism.
+We discarded species with only one representative.
 The mammalian nucleotide diversity is obtained from the Zoonomia project, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
@@
             Assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
-            Trait encoded by few loci                        & Between species trait variation is under-estimated & --              & Under-estimated & Conservative & Invalid  \\ \hline
-            Sexual dimorphism                                & Within species trait variation is over-estimated   & Over-estimated & -- & Conservative & Invalid  \\ \hline
-            Inbreeding                                       & Nucleotide diversity ($\pi$) is under-estimated    & Over-estimated  & --              & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
-            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is over-estimated & -- & Under-estimated & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is over-estimated  & Under-estimated & -- & Invalid & Conservative  \\ \hline
-            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is under-estimated & -- & Over-estimated & Invalid & Conservative  \\ \hline
+            Trait encoded by few loci                        & Between species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
+            Sexual dimorphism                                & Within species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
+            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
         \end{tabular}
@@
 Simulations performed under diversifying selection were detected as such ($\proba [\EstNI > 1] > 0.975$), no false negative was observed (red in fig.~\ref{fig:results-simulations}A\&B).
-Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 77\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B),while detected as stabilizing selection in 21\% of the cases, and detected as diversifying selection in 2\% of cases.
+Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 77\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B), while detected as stabilizing selection in 21\% of the cases, and detected as diversifying selection in 2\% of cases.
 Second, for simulations under a fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}C\&D), simulations of diversifying selection was detected as such and stabilizing selection was detected as such in 100\% of the cases.
@@
 Because our test is also based on several assumptions that might not hold on empirical data, we provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
-As example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$.
+As an example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$.
 However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for nucleotide divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
@@
 % What can and should be improved
-Our development is based on several assumption that could be relaxed.
+Our development is based on several assumptions that could be relaxed.
 First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
-Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component~\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny~\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically be relaxed by adding a time component~\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny~\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of polymorphism and divergence.
@@
 % How is our test related to other tests of selection
-Our test bears analogy with others tests for neutrality developed, providing insight into its behavior.
+Our test bears analogy with other tests for neutrality developed, providing insight into its behavior.
 First, our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
@@
 Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\cite{latrille_genes_2023}.
-By explicitly modelling stabilizing selection as a moving optimum, it would theoretically allow to tease-out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
+By explicitly modeling stabilizing selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
 
@@
 Ahe analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\cite{catalan_drift_2019}.
-However, the alternative model of a moving optimum was not implemented, and thus can not be ruled out by this analysis.
+However, the alternative model of a moving optimum was not implemented, and thus cannot be ruled out by this analysis.
 With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
@@
     \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
-    \item $\Ne$ is the effective number of individual.
+    \item $\Ne$ is the effective number of individuals.
     \item $g_{i,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
@@
 Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
-Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
+Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).

### 2023-09-13 manuscript/main.tex
M. Bastian, T. Gaboriau and N. Salamin edits

@@
 \usepackage{caption}
+\usepackage{censor}
 \usepackage{xcolor, colortbl}
@@
 \usepackage{biorxiv}
-\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
+\usepackage[backend=biber,eprint=false,isbn=false,url=false,giveninits=true, style=apa,date=year]{biblatex}
+\DeclareDelimFormat{nameyeardelim}{\addcomma\space}
 \addbibresource{references.bib}
@@
 \lstset{style=mystyle}
+\captionsetup[table]{hypcap=false}
+\captionsetup[figure]{hypcap=false}
 
@@
 \newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
-\newcommand{\VarGenetic}{V_{\mathrm{G}}}
-\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
+\newcommand{\VarGenetic}{V_{\mathrm{A}}}
+\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
 \newcommand{\VarEnv}{V_{\mathrm{E}}}
-\newcommand{\MatrixGenetic}{\BiDimArray{G}}
 \newcommand{\VarMutation}{V_{\mathrm{M}}}
@@
 % Including genomics, Inclusion of genomics, Linking genomics,
-\title{Leveraging genomics to detect trait selection from within and between species variations}
-\rhead{\scshape Trait selection from within and between species variations}
+\title{Leveraging genomics to detect trait selection from within and between-species variations}
+\rhead{\scshape Trait selection from within and between-species variations}
 
@@
 }
-
+\author{~}
 \begin{document}
@@
 \begin{abstract}
-    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
-    Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
-    Conversely, using between species variation, current methods usually discard within species variation, thus not accounting for processes at the micro-evolutionary scale.
+    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between-species variation.
+    However, methods using within-species variation do not integrate the changes at the macro-evolutionary scale.
+    Conversely, current methods using between-species variation usually discard within-species variation, thus not accounting for processes at the micro-evolutionary scale.
     The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
@@
 
-\keywords{Quantitative genetics \and Trait evolution \and Selection }
+\keywords{Quantitative genetics \and Trait evolution \and Selection \and Phylogenetics \and Population genetics }
 
@@
 
-Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
-Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits and sequences.
-Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.
+Determining whether a trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
+Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
+The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
+All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.
 
-First, leveraging individual variations within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given locus), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
-Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
-For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
-However, dominant selective regime acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
-For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
-Gene expression exhibit such accumulation and divergence in expression accumulate faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\cite{walsh_evolution_2018}.
+[not sure of this paragraph...]
+
+Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
+Additionally, contrasting both trait and genetic differentiation across several populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
+For example, studies have shown that stabilizing selection is largely dominant in the evolution of gene expression in different populations~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
+[why only example from gene expression?]
+However, dominant selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
+
+To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
+For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
+Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for gene with large within-species variation~\parencite{khaitovich_neutral_2004}.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
 
 % Phylogenetics (mean trait evolution)
-Alternatively, at the macro-evolutionary scale, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\cite{hansen_translating_1996}.
-As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
-A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
-Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
-However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
-In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
-Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Altogether, methods at the phylogenetic scale easily misclassify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.
+Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
+Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
+As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
+Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
+Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
+However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
+First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
+In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
+Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.
 
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macro evolution, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
-In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
-Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimate of population size (within species) nor of time of speciation (between species).
-For population geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
-For a phylogeneticist, our study can be seen as an extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
-% Basically, we are testing whether the variance of means is greater than the man of variances
+At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value~\parencite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
+However, in the context of detecting selection acting on a trait, within-species variation is mostly used to describe technical errors.
+One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
+We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
+Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
+For population geneticists, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
+For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
+% Basically, we are testing whether the variance of means is greater than the mean of variances
 
@@
 \subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
-\subsubsection*{Mutation-drift equilibrium for a trait}
 
-For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
-New mutations are thus generating variance for such a trait, where the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
+Prior to developing our neutrality index, we review theoretical expectations for variations of quantitative traits and genomic sequences under neutral evolution for both within- and between-species variation.
+
+\subsubsection*{Within-species trait variations}
+
+For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
+New mutations are generating trait variance and the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
 At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$ as:
+As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$:
 \begin{gather}
@@
 
-Mutations supply new variants, while genetic random drift depletes standing variation.
-As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
+While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
+For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
 \begin{align}
@@
 
-For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating divided by the size of the region.
-Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the number of segregating mutations is also a balance between mutations and drift.
-As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
+For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating in the population divided by the length of the region.
+Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
+As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$):
 \begin{gather}
@@
 
-Altogether, we define $\RateWhithin$ as the ratio of genetic variance of the trait ($\VarGenetic$) over nucleotide diversity of any neutral genomic region of interest ($\pi$).
-This ratio allows to discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
+We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
+This ratio allows to remove the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
 \begin{align}
@@
 
-Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), also giving $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
+The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ beinf a function of $\VarPhenotype$ and $\Heritability$:
 \begin{align}
@@
 
-\subsubsection*{Phylogenetic evolution of a trait}
+\subsubsection*{Between-species trait variations}
 
-At generation $\Time$, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
-If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
-After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
+For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
+If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
+The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given by \parencite{hansen_translating_1996}:
 \begin{align}
@@
 
-Moreover, for any neutral genomic region of interest, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
-For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
-As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
+Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
+The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
+We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
 \begin{align}
@@
 \end{align}
-That is, if mutations are neutral, the rate with which new neutral alleles are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.
+That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate with which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.
 
-After having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
+After $\Time$ generations and assuming that no multiple substitutions occurred at the same site, the nucleotide divergence $d$, which is the fraction of the genomic region that generated a substitution, will be $\Time$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
 \begin{align}
@@
 \end{align}
-Altogether, we define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region of interest ($d$).
-This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
+We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
+This ratio allows to remove the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
 \begin{align}
@@
     \caption{
-        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
-        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
+        Between species, the mean phenotypic trait value changes along the phylogeny, allowing estimating of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
+        Within-species, the genetic variance allows estimating of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
         $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
@@
         Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
-        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the confounding effect on diversity of mutation rate, population size and divergence time.
+        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
     }
@@
 
-The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
-Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allows to compute $\pi$.
-Combined as a ratio, $\RateWhithin$ depends solely on the genetic architecture of the trait as $\RateMut$.
-At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
-Similarly, combined as a ratio, $\RateBetween$ depends solely on $\RateMut$
-Altogether, we thus have:
+The variability between either individuals or species can be obtained for both quantitative traits and genomic sequences.
+At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$, which equals $\RateMut$ if the trait is neutrally evolving (see above).
+At the phylogenetic level, the variability of the mean trait value between species can be combined with the nucleotide divergence of any neutrally evolving genomic region to obtain $\RateBetween$.
+Similarly, $\RateBetween=\RateMut$ if the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogeny.
+We thus have, for a neutrally evolving trait:
 \begin{gather}
     \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
-    \Rightarrow \frac{\RateBetween}{\RateWhithin} = 1.
+    \Rightarrow \NI \defEqual \frac{\RateBetween}{\RateWhithin} = 1.
 \end{gather}
-Altogether, $\RateBetween / \RateWhithin$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
-Importantly, $\RateBetween$ and $\RateWhithin$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
+We define a neutrality index $\NI = \RateBetween / \RateWhithin$ that will equal $1$ for a trait evolving neutrally.
+Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
 Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
@@
 
+Based on the comparative framework accounting for phylogenetic inertia~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a Maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
+
 \subsubsection*{Maximum likelihood estimate}
 At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitution per site) and the topology of the phylogenetic tree.
-Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
-The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be the root, in which case the distance is 0).
-Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate~\cite{omeara_testing_2006}, given as:
+The diagonal $\Distance_{\Spi,\Spi}$ represents the total distances from the root of the tree to each taxon ($\Spi$).
+The off-diagonal elements ($\Distance_{\Spi,\Spj} = \Distance_{\Spj,\Spi}$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
 \begin{gather}
@@
 
-Finally, within species variation $\EstRateBetween$ is estimated as~\cite{omeara_testing_2006}:
+Finally, within-species variation $\EstRateBetween$ is estimated as~\parencite{omeara_testing_2006}:
 \begin{gather}
@@
 
-For a given species $\Spi$ with inter-individual data available, the ratio of genetic variance of a trait ($\VarGeneticSpi$) over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
+For a given species $\Spi$ with inter-individual data available, additive genetic variance of a trait ($\VarGeneticSpi$) is the product of heritability ($\Heritability_{i}$) and phenotypic variance ($V_{\Trait, i}$).
+The ratio of $\VarGeneticSpi$ over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
 Averaged across all species, we obtain the estimate $\EstRateWhithin$ as:
@@
 
-Altogether, the neutrality index is estimated as:
+As depicted in fig.~\ref{fig:methods}, the neutrality index is estimated as:
 \begin{gather}
@@
 \subsubsection*{Bayesian estimate}
-
-Even though the neutrality index can be applied for any trait, we so far assumed that several traits would to be independent of one another.
-Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
-Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
+[NS: the start of this paragraph is not so clear -> to discuss]
+Even though the neutrality index can be applied to any trait, we assume here that the variation between several traits is independent.
+Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
+Here we generalize to $\Ntrait$ non-independent traits evolving along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
 Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
@@
 With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
-$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
-For each trait and each species, heritability ($\Heritability$) is set as a uniform distribution with user-defined boundaries.
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
+$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is Gibbs sampled.
+For each trait and each species, prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
+Heritability and phenotypic variance for each trait is combined with nucleotide diversity to compute $\EstRateWhithin$ for each species, and is averaged out across species (as in eq.~\ref{eq:estimated-rate-pop}).
+From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
+The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
 
@@
 % nbr_generations = 13336.114128321291
-We assumed that the trait genotypic value is encoded by $\NbrLoci=5,000$ loci, with each locus contributing additively to the genotypic value.
-Each effect is Normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
-The phenotypic value is the sum of phenotypic value and an environmental effect, the environmental effect is normally distributed with variance $\VarEnv$.
-We assumed a trait with a heritability of $\Heritability =0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
-We will follow a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
-Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we can simulate explicitly each generation along an ultrametric phylogenetic tree.
-For each offspring, the number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
+We tested the xxx [NS: explain a bit what we test here -> NI]
+We simulated phenotypic values as the sum of genotypic value and an environmental effect.
+The environmental effect was normally distributed with variance $\VarEnv$.
+We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
+We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
+We followed a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
+Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
+For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
 From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
-To scale parameters with a mammalian regime, we used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38e^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38e^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch is proportional to the branch length.
+We scaled parameters to fit plausible values for mammals.
+We used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
 
-The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages are both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
-Additionally, the instant value of log-$\Ne$ is modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
-The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
-The simulation starts from an initial sequence at equilibrium at the root of the tree and, at each node, the process is split until it finally reaches the leaves of the tree.
+The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
+Additionally, the instant value of log-$\Ne$ was modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
+The geometric Brownian motion accounted for long-term fluctuations, while the Ornstein-Uhlenbeck introduced short-term fluctuations. [NS: not clear how you combined both here]
+The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
 From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.
 
-Genetic drift is modeled by the resampling of individuals at each generation, with each parent probability of being sampled being proportional to its fitness ($W$).
-Selection is modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
-More specifically, the fitness of an individual is given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ is the trait value of the individual, $\lambda=0.0$ is the optimal trait value, and $\alpha=0.02$ is the strength of selection.
-Mutations are seen has displacement of the phenotype in the multidimensional space.
-Beneficial mutations move the phenotype closer to the optimum, while deleterious mutations move it further away.
-Stabilizing selection is implemented by means of a fixed optimum phenotype ($\lambda=0.0$).
-Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process ($\brownian \left(0, \sigma_{\lambda}=1.0\right)$).
-Neutral evolution is implemented with a fixed fitness landscape ($W=1$), where each individual has the same probability of being sampled at each generation.
+Random genetic drift was introduced by re-sampling individuals at each generation, with each parent having a probability of being sampled that was proportional to its fitness ($W$).
+Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
+Mutations were considered as a displacement of the phenotype in the multidimensional space.
+Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
+Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
+Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\parencite{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
+Neutral evolution was implemented by fixing the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.
 
-Nucleotide diversity ($\pi$) is measured as the heterozygosity of the neutral markers.
-Nucleotide divergence ($d$) is measured as the number of substitutions per site along branches of the phylogeny.
-The genetic variance is measured as phenotypic variance multiplied by heritability.
-Heritability is estimated from the slope of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\cite{lynch_genetics_1998} using the last generations of the simulations.
-As such heritability is not a given parameter of the simulations, but rather measured as it would be in empirical data.
+Nucleotide diversity ($\pi$) was measured as the heterozygosity of neutral markers (not linked to the trait).
+Nucleotide divergence ($d$) was measured as the number of substitutions per site of neutral markers (not linked to the trait) along branches of the phylogenetic tree.
+The additive genetic variance was measured as phenotypic variance multiplied by heritability.
+Heritability was estimated from the slopes of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\parencite{lynch_genetics_1998} averaged over the last 10 simulated generations.
+Heritability was thus not a given parameter of the simulations, but rather measured as it would be in empirical data.
 
@@
         Wright-Fisher simulations with mutation, selection and drift.
-        Panel A: the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
-        The trait genotypic value is then transformed into a phenotypic value by adding a Gaussian noise with standard deviation $\VarEnv$.
+        Left panel: For a given individual, the trait phenotypic value is the sum of genotypic value and a environmental effect (standard deviation $\VarEnv$).
+        The trait's genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
         Parents are selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
@@
         Drift is modeled by the resampling of parents.
-        Panel B: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
+        Right panel: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
     }
@@
 
-We analyzed a mammalian dataset, with the body mass and brain mass as quantitative traits.
-The log-transformed values of body mass and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
-We removed individuals not marked as an adult, and split the data into males and females due to sexual dimorphism.
-We discarded species with only one representative.
-The mammalian nucleotide diversity is obtained from the Zoonomia project, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
+We analyzed a dataset of body and brain masses in mammals.
+The log-transformed values of body and brain masses were taken from \textcite{tsuboi_breakdown_2018}.
+We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
+We discarded species with only one representative sample.
+The mammalian nucleotide diversity was obtained from the Zoonomia project~\parencite{genereux_comparative_2020}, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
 
-We also analyzed on a dataset of Primates species, with the nucleotide variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
+We also analyzed a dataset of primate species, with the nucleotide variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
 However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
@@
 
-To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at different scales.
-The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
-More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by nucleotide diversity of any neutral genomic region ($\pi$), denoted $\RateWhithin$, is a proxy for the mutational variance $\RateMut$.
-At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a $\Time$ generations is $\VarPhy$.
-This variance can also be normalized by the nucleotide divergence of any neutral genomic region ($d$), denoted $\RateBetween$, allowing us to obtain another proxy for $\RateMut$.
-Altogether, we defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
-Under a depleted variation between species, meaning $\NI < 1$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
-In contrast, under an inflated variation between species, meaning $\NI > 1$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.
+\subsection*{Neutrality index}
+
+For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
+At the species level, the additive genetic variance of the neutral trait ($\VarGenetic$) is normalized by nucleotide diversity of any neutral genomic region ($\pi$) to obtain $\RateWhithin$.
+At the scale of between-species variation, assuming a neutral trait evolving along the phylogeny, the mean trait value changes as a Brownian process.
+This variance in mean trait value can also be normalized by the nucleotide divergence of any neutral genomic region ($d$) to obtain $\RateBetween$.
+[NS: -> this is repeating the mat/met, is it really needed??]
+We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see methods), suggesting that traits for which this relationship is not verified are putatively under selection.
+Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
+In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
+
+[NS: this is not a result but an assumption that we have. We also already say it above. I would move it to discussion if needed]
+Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate, population size and divergence time.
+The relationship between trait variation between and within species can be used as a method to detect whether traits are evolving under a neutral model of evolution.
+Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between-species (see fig.~\ref{fig:methods}).
+This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
 
@@
             \hline
-            Assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
-            Trait encoded by few loci                        & Between species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
-            Sexual dimorphism                                & Within species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
+            Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
+            Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
             Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
             Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
             Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
+            Markers for polymorphism under balanced selection & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
             Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
+            Multiple nucleotide substitutions at the same locus & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\\hline \hline
         \end{tabular}
     \end{adjustbox}
-    \caption{Assumptions and their consequences on the estimation of within species variation ($\EstRateWhithin$), between species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
+    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
     The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
@@
 
-Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate, population size and divergence time.
-The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
-Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
-This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
-We also devised a Bayesian estimated allowing to obtain the posterior distribution of the neutrality index, and to test for evidence of diversifying selection as $\proba [\EstNI > 1]$, and evidence for stabilizing selection as $\proba [\EstNI < 1]$.
+\subsection*{Test of selection}
+
+[NS: here again, these are not results, but a discussion]
+We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
+
+Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
+Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
+Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
+Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
+It is however possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
+Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in under-estimated $\EstNI$ since $\Heritability \leq H^2$.
+Contrarily, the test for stabilizing selection is invalid if $\EstNI$ is under-estimated.
+Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
 
@@
 With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant mammalian empirical regime (see Materials and Methods).
-First, we assumed a constant population size ($\Ne$) and a constant mutation rate ($\MutationRate$) across the phylogeny (fig.~\ref{fig:results-simulations}A\&B)
-Simulations performed under stabilizing selection where detected as such ($\proba [\EstNI < 1] > 0.95$), no false negative was observed (blue in fig.~\ref{fig:results-simulations}A\&B).
-Simulations performed under diversifying selection were detected as such ($\proba [\EstNI > 1] > 0.975$), no false negative was observed (red in fig.~\ref{fig:results-simulations}A\&B).
-Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 77\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B), while detected as stabilizing selection in 21\% of the cases, and detected as diversifying selection in 2\% of cases.
-Second, for simulations under a fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}C\&D), simulations of diversifying selection was detected as such and stabilizing selection was detected as such in 100\% of the cases.
-For simulations under neutral evolution, 51\% of the simulations were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), 49\% of the simulations were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
+[NS: I would remove this sentence]
+Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
+For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
+Once we introduced fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
+For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
 
@@
     \caption{
-        $1,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
+        $10,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
         Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-        Histogram of the ratio of between species trait variation ($\RateBetween$) over within species trait variation $\RateWhithin$, as $\NI = \RateBetween / \RateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\NI$ is greater than $1$ (right).
+        Histogram of estimated the ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$, as $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\EstNI$ is greater than $1$ (right).
         Effective population size ($\Ne$) and mutation rate ($\MutationRate$) are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
@@
 
+\begin{table*}[t!]
+    \centering
+    \begin{adjustbox}{width = 0.75\textwidth}
+        \begin{tabular}{||l|l|l|c|c|c|c|c||}
+        \toprule
+        Dataset & Trait & $\Heritability$ & Sex & $\NbrTaxa$ & $\EstNI$ & $95\%$ CI for $\EstNI$ & $\proba [\EstNI > 1 ]$ \\ \hline
+        \midrule
+        Mammals & Body mass & 1.0 & \Male & 36 & 0.340 & 0.217-0.523 & 0.000 \\ \hline
+        Mammals & Body mass & 1.0 & \Female & 26 & 0.277 & 0.160-0.490 & 0.000 \\ \hline
+        Mammals & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 36 & 1.124 & 0.721-1.754 & 0.635 \\ \hline
+        Mammals & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 26 & 0.936 & 0.523-1.715 & 0.324 \\ \hline \hline
+        Mammals & Brain mass & 1.0 & \Male & 36 & 1.351 & 0.851-2.173 & 0.877 \\ \hline
+        Mammals & Brain mass & 1.0 & \Female & 26 & 1.727 & 0.991-2.938 & 0.972 \\ \hline
+        Mammals & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 36 & 4.527 & 2.831-7.091 & 1.000 \\ \hline
+        Mammals & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 26 & 6.001 & 3.288-10.941 & 1.000 \\ \hline \hline
+        Primates & Body mass & 1.0 & \Male & 71 & 0.558 & 0.401-0.784 & 0.000 \\ \hline
+        Primates & Body mass & 1.0 & \Female & 65 & 0.389 & 0.278-0.547 & 0.000 \\ \hline
+        Primates & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 71 & 1.875 & 1.288-2.695 & 1.000 \\ \hline
+        Primates & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 65 & 1.296 & 0.899-1.821 & 0.914 \\ \hline \hline
+        Primates & Brain mass & 1.0 & \Male & 71 & 1.929 & 1.395-2.616 & 1.000 \\ \hline
+        Primates & Brain mass & 1.0 & \Female & 65 & 1.950 & 1.399-2.790 & 1.000 \\ \hline
+        Primates & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 71 & 6.479 & 4.658-8.944 & 1.000 \\ \hline
+        Primates & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 65 & 6.522 & 4.664-9.294 & 1.000 \\
+        \bottomrule
+        \end{tabular}
+    \end{adjustbox}
+    \caption{
+    Test of diversifying selection on a mammal and a primate dataset, by splitting males (\Male) and females (\Female).
+    Traits considered are body mass or brain mass (log-transformed).
+    Heritability ($\Heritability$) is either assumed complete ($\Heritability=1.0$) or uniformly distributed between 20\% and 40\%  ($\Heritability \sim \mathcal{U}(0.2, 0.4)$).
+    $\NbrTaxa$ is the number of species in the dataset.
+    $\EstNI$ is the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\EstNI$ also computed.
+    $\proba [\EstNI > 1 ]$ is the estimated posterior probability of diversifying selection.
+    }
+    \label{table:empirical}
+\end{table*}
+
 \subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
-For mammalian body mass, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
-The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.34$ ($[0.22--0.52]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.28$ ($[0.16--0.49]$ for $95\%$ CI).
-In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.0$ and $0.0$ respectively.
-Moreover, assuming that heritability ($\Heritability$) of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection ($ \proba [\EstNI > 1]$) becomes $0.635$ in males and $0.914$ in females.
-For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.35$ ($[0.85--2.17]$ for $95\%$ CI, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.72$ ($[1.00--2.94]$ for $95\%$ CI, $\NbrTaxa=26$).
-In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.877$ and $0.972$ respectively.
-Moreover, assuming that heritability of brain mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in both males and females.
 
-For primates body mass, we obtained $\EstNI_{\text{\Male}} = 0.56$ ($[0.40,~0.78]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.39$ ($[0.28,~0.55]$ for $95\%$ CI, $\NbrTaxa=65$).
-Assuming that heritability of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in males and $0.914$ in females.
-For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.95$ ($[1.40,~2.79]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.95$ ($[1.39,~2.79]$ for $95\%$ CI, $\NbrTaxa=65$).
-In both males and females, brain mass is under diversifying selection, with associated posterior probabilities of $1.0$.
-Importantly, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).
+For mammalian body and brain mass, we obtained male (\Male) and female (\Female) trait variation.
+Combined with nucleotide diversity and divergence, we estimated $\EstNI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in table~\ref{table:empirical}.
+Assuming complete heritability, brain mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
+If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
+Mammalian brain mass was found to be under diversifying selection with posterior probabilities of $0.877$ for males and $0.972$ for females when complete heritability was assumed.
+Assuming a uniform distribution between 20\% and 40\% for heritability led to posterior probabilities of diversifying selection of $1.0$ for both males and females.
+
+We also analyzed a similar dataset for body mass focussing this time only at Primates (see table~\ref{table:empirical}).
+For primate body mass, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%,
+Fixing the heritabity of body mass to $1.0$ did not change the posterior probability for males, but increased the one for female to $1.0$.
+Evidence for diversifying selection on body mass was therefore more pronounced in Primates than in Mammals.
+However, the genetic markers used to normalize trait variance with divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\EstNI$ (see table~\ref{table:assumptions} and methods).
 
@@
 % Summary of the method and results
-In this study, we proposed a neutrality index for a quantitative trait.
+In this study, we proposed a neutrality index for a quantitative trait that can be used within a statistical framework to test for selection.
+Our neutrality index for a trait, $\NI$, is the ratio of the normalized between- to within-species variation and it allows the identification of the evolutionary regime of a quantitative trait.
 At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
 Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
-Our neutrality index for a trait, $\NI$, is the ratio of between species variation to within species variation.
-$\NI$ can be compared to 1 to seek a statistical deviation from a neutral trait.
-In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
-Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
-Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation (see section~S\ref{sec:implementation}) also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
-$ \proba [ \NI > 1 ]$).
+Estimated $\EstNI$ can be compared against the null hypothesis of neutrality by testing for deviation from the value of $1.0$.
+In other words, our statistical test seeks to find traits with extremal values in comparison to a pool of traits.
+Instead, our procedure can be applied to a single trait [NS: this is not clear as the previous sentence seems to contradict this one], estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
+Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S\ref{sec:implementation}) also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
+$ \proba [ \EstNI > 1 ]$).
+We tested our statistical procedure against simulated data and showed that for simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$) our test is able to correctly detect those regimes of selection.
+However, for simulations of a neutral trait, our test detects frequently a spurious signal of stabilizing selection ($\EstNI < 1$).
+Thus, we argue that our test should be used to detect diversifying selection, but has low accuracy to detect stabilizing selection due to many false positives.
 Importantly, our neutrality index leverages nucleotide divergence and polymorphism, as well as the heritability of traits ($\Heritability$).
-We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize the variation of trait between and within species.
-By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, while not requiring to estimate divergence time~\cite{litsios_effects_2012}.
-Instead, the normalization by nucleotide divergence and polymorphism automatically absorbs these changes~\cite{seo_estimating_2004}.
+We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize trait variation between and within species.
+By doing so, we also show using simulated data that our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, while not requiring to estimate them nor to estimate divergence time~\parencite{litsios_effects_2012}.
+Instead, the normalization by nucleotide divergence and polymorphism automatically absorbs these changes.
+[NS: we need to develop these arguments and show that the normalization works because it is similar to the dN/dS tests, etc....
+I would thus move the paragraph below on this here.
+Otherwise, this paragraph is a bit hollow]
 
 % Pitfalls of the method
-We argue that the main drawback of our method is that the heritability ($\Heritability$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically harder to measure for any trait.
-Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is under-powered.
-Indeed, because $\Heritability \leq 1$ by definition, if we set $\Heritability \leq 1$, we overestimate the within species variation and thus underestimate $ \NI$ and can test for diversifying selection.
-In other words, testing for $\NI > 1$ while setting $\Heritability = 1$ means that knowing the real heritability would only increase the evidence for diversifying selection.
-Contrarily, the test for stabilizing selection is invalid if heritability is not known.
-Because our test is also based on several assumptions that might not hold on empirical data, we provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
-As an example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$.
-However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for nucleotide divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
+In the context of detecting diversifying selection, we argue that the main drawback of our method is that the narrow-sense heritability ($\Heritability$) of a trait is required if one has access only to the phenotypic variance and not the additive genetic variance.
+[NS: a bit circular.
+If we have V_G, we know h^2 as V_P can be measured]
+If phenotypic variance is used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ is ultimately under-estimated.
+Similarly, using broad-sense heritability instead of narrow-sense heritability results in under-estimated $\EstNI$.
+In such context, the test of stabilizing selection ($\EstNI < 1]$) is statistically invalid.
+However, the test of diversifying selection ($\EstNI > 1$) is underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
+Because our test is also based on several assumptions that might not hold on empirical data, we also provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
+[NS: I would discuss the elements of this table and link that with some literature if possible.]
+As an example, even though we assumed complete heritability for brain mass, we can assert that brain mass is evolving under diversifying selection in mammals since $\EstNI > 1$.
+However, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass is evolving under diversifying selection, since the markers used for nucleotide divergence are not neutral which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
 
 % What can and should be improved
-Our development is based on several assumptions that could be relaxed.
-First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
-Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically be relaxed by adding a time component~\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny~\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of polymorphism and divergence.
-Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within species in a unified framework.
-Also, we don't know how our test would behave in the context of gene flows and introgression, which could be investigated in future studies.
+The development of our neutrality index is also based on several assumptions that could be relaxed in future studies.
+First, we cannot predict the behaviour of our test in the context of population structures, gene flow and introgression.
+These factors should be thoroughly investigated using simulations.
+Second, loci are assumed to contribute additively to the phenotype~\parencite{hill_data_2008, crow_epistasis_2010}, disregarding effects of dominance and epistasis.
+[NS: here a few words on the importance of dominance and epistasis in empirical quant genet studies would be interesting.
+And what about long-term effects at the phylogenetic level?]
+Third, the genetic architecture of the trait is assumed to be constant across the phylogeny, which might actually be variable among individuals and species.
+[NS: any literature on this (or is it the following ones)?
+I'm not aware of that right now, but we should have a look]
+Such assumption can theoretically be relaxed and changes in genetic architecture along the phylogeny could jointly be estimated ~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
 
 % How is our test related to other tests of selection
-Our test bears analogy with other tests for neutrality developed, providing insight into its behavior.
-First, our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
-Second, because $\NI$ is compared to 1, our test bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\cite{goldman_codonbased_1994, muse_likelihood_1994}.
+Even though our test is developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provide insight into its behavior.
+First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations are used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
+Second, because $\NI$ is compared to 1, our test ultimately bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
 As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
 Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\cite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\cite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\cite{latrille_genes_2023}.
+With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
+First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
 By explicitly modeling stabilizing selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
@@
 % What can our test bring to the field
+[NS: should we not start with this idea?]
 Finally, from an empirical point of view, our test is also related to other methods that have been developed to test for selection on quantitative traits.
-At the phylogenetic scale, test for neutrality of a quantitative trait are seeking deviation from a Brownian process~\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation~\cite{rohlfs_phylogenetic_2015}.
-Ahe analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\cite{catalan_drift_2019}.
+At the phylogenetic scale, tests for neutrality of a quantitative trait are seeking deviation from a Brownian process~\parencite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation~\parencite{rohlfs_phylogenetic_2015}.
+The analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\parencite{catalan_drift_2019}.
 However, the alternative model of a moving optimum was not implemented, and thus cannot be ruled out by this analysis.
 With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
-Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence ($d$) and polymorphism ($\pi$).
-Altogether, our study provides a statistical framework to test for neutrality of a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
+Our diversity index can thus allow us to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence ($d$) and polymorphism ($\pi$).
+Even though our method requires integrating genomic and trait variation, we also showed that our method can be applied to the empirical dataset with the illustrative example of mammals' brain and body mass.
+Altogether, our study provides a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
 
 %TC:ignore
-\section*{Acknowledgements}
-\label{sec:acknowledgment}
-\textbf{Funding:}
-Université de Lausanne.
-\textbf{Author contributions:}
-Original idea: T.L.\ and N.S.;
-Model conception: T.L.\ and N.S.;
-Code: T.L.;
-Data analyses: T.L.\ and M.B.;
-Interpretation: T.L., T.G\ and N.S.;
-First draft: T.L.;
-Editing and revisions: T.L., M.B, T.G\ and N.S.;
-Project management and funding: N.S\@.
-\textbf{Competing interests:}
-The authors declare no conflicts of interest.
-\textbf{Data and materials availability:}
-Snakemake pipeline, analysis scripts and documentation are available at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}.
-
 \printbibliography
@@
 \begin{center}
-    \captionof{figure}{Summary of trait's genetic architecture.}
+    \captionof{figure}{summary of trait's genetic architecture.}
     \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
@@
 
-Within species, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
+within-species, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
 \begin{equation}
@@
 
-Implementation is included within the \textit{BayesCode} software, available at \url{https://github.com/ThibaultLatrille/bayescode}.
+Implementation is included within the \textit{BayesCode} software, available at \censor{\url{https://github.com/ThibaultLatrille/bayescode}}.
 
@@
     \item A file containing the mean trait values for each species.
-    \item A file containing the variation within species for each trait and the genetic variation within species (neutral markers).
+    \item A file containing the variation within-species for each trait and the genetic variation within-species (neutral markers).
 \end{enumerate}
@@
     \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
-    \item \emph{Nucleotide\_diversity}: the nucleotide diversity within species (neutral markers), cannot be \texttt{NaN}.
+    \item \emph{Nucleotide\_diversity}: the nucleotide diversity within-species (neutral markers), cannot be \texttt{NaN}.
     \item As many columns as traits, without spaces or special characters in the trait.
-    \item \emph{TraitName\_variance}: the phenotypic variance of the trait within species, can be \texttt{NaN} to indicate that the trait variance is not available for that taxon.
-    \item \emph{TraitName\_heritability} (optional): the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
+    \item \emph{TraitName\_variance}: the phenotypic variance of the trait within-species, can be \texttt{NaN} to indicate that the trait variance is not available for that taxon.
+    \item \emph{TraitName\_heritability} (optional): the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
     \item The columns with the suffix \texttt{\_variance} and \texttt{\_heritability} are repeated for each trait.
-    \item \emph{TraitName\_heritability\_lower} (optional): the lower bound of the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
-    \item \emph{TraitName\_heritability\_upper} (optional): the upper bound of the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
+    \item \emph{TraitName\_heritability\_lower} (optional): the lower bound of the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
+    \item \emph{TraitName\_heritability\_upper} (optional): the upper bound of the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
     \item If the columns with the suffix \texttt{\_heritability\_lower} and \texttt{\_heritability\_upper} are present, the heritability is randomly drawn from a uniform distribution between the lower and upper bounds.
     \item If the columns with the suffix \texttt{\_heritability} is present, it is taken as is.
-    \item If the genetic variance (instead of phenotypic variance) is available for a trait, the heritability can be omitted and will automatically be set to 1.0.
+    \item If the additive genetic variance (instead of phenotypic variance) is available for a trait, the heritability can be omitted and will automatically be set to 1.0.
 \end{itemize}
@@
 
-Assuming that the file \texttt{data/body\_size/mammals.male.tsv} contains the mean trait values for each species, the file \texttt{data/body\_size/mammals.male.var\_trait.tsv} contains the variation within species for each trait and the genetic variation within species (neutral markers), and the file \texttt{data/body\_size/mammals.male.tree} contains the phylogenetic tree, the following commands are used to run the model and read the results.
+Assuming that the file \texttt{data/body\_size/mammals.male.tsv} contains the mean trait values for each species, the file \texttt{data/body\_size/mammals.male.var\_trait.tsv} contains the variation within-species for each trait and the genetic variation within-species (neutral markers), and the file \texttt{data/body\_size/mammals.male.tree} contains the phylogenetic tree, the following commands are used to run the model and read the results.
 
@@
 \subsubsection{Reading the results}\label{subsubsec:reading-the-results}
-Once the model has ran, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between species variation over within species variation with \texttt{readnodetraits}:
+Once the model has ran, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between-species variation over within-species variation with \texttt{readnodetraits}:
 \begin{lstlisting}[language = sh,label={lst:readnodetraits-rho}]
@@
 \end{lstlisting}
-The file \texttt{data\_empirical/chain\_name.ratio.tsv} then contains the posterior mean of the ratio of between species variation over within species variation, the 95\% and 99\% credible interval, and the posterior probability that the ratio is greater than 1.
+The file \texttt{data\_empirical/chain\_name.ratio.tsv} then contains the posterior mean of the ratio of between-species variation over within-species variation, the 95\% and 99\% credible interval, and the posterior probability that the ratio is greater than 1.
 

### 2023-09-13 manuscript/title_page.tex
M. Bastian, T. Gaboriau and N. Salamin edits

--- 2023-07-25 manuscript/main.tex
@@
 
-\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
-\newcommand{\Multiply}{\cdot}
-\newcommand{\MultiplyMatrix}{\times}
-\newcommand{\UniDimArray}[1]{\bm{#1}}
-\newcommand{\BiDimArray}[1]{\bm{#1}}
-\newcommand{\tr}{^{\intercal}}
-\newcommand{\inv}{^{-1}}
-\DeclareMathOperator{\E}{\mathbb{E}}
-\DeclareMathOperator{\Var}{\text{Var}}
-\DeclareMathOperator{\Cov}{\text{Cov}}
-\newcommand{\der}{\mathrm{d}}
-\newcommand{\e}{\text{e}}
-\newcommand{\Ne}{N_{\text{e}}}
-\newcommand{\dn}{d_N}
-\newcommand{\ds}{d_S}
-\newcommand{\dnds}{\dn / \ds}
-\newcommand{\pn}{\pi_N}
-\newcommand{\ps}{\pi_S}
-\newcommand{\pnps}{\pn / \ps}
-\newcommand{\proba}{\mathbb{P}}
-\newcommand{\pfix}{\proba_{\text{fix}}}
-\newcommand{\Spi}{i}
-\newcommand{\Spj}{j}
-\newcommand{\NbrTaxa}{n}
-\newcommand{\Time}{t}
-\newcommand{\Trait}{P}
-\newcommand{\Heritability}{h^2}
-\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
-\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
-\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
-\newcommand{\Root}{0}
-\newcommand{\RootTrait}{\Trait_{\Root}}
-\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
-\newcommand{\VecZero}{\UniDimArray{0}}
-\newcommand{\VecOne}{\UniDimArray{1}}
-\newcommand{\Distance}{\BiDimArray{D}}
-\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
-\newcommand{\MutationRate}{\mu}
-\newcommand{\SubRate}{q}
-\newcommand{\NbrLoci}{L}
-\newcommand{\VarPhenotype}{V_{\Trait}}
-\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
-\newcommand{\VarGenetic}{V_{\mathrm{G}}}
-\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
-\newcommand{\VarEnv}{V_{\mathrm{E}}}
-\newcommand{\MatrixGenetic}{\BiDimArray{G}}
-\newcommand{\VarMutation}{V_{\mathrm{M}}}
-\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
-\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
-\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
-\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
-\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
-\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
-\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
-\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
-\newcommand{\NI}{\rho}
-\newcommand{\EstNI}{\widehat{\rho}}
-\newcommand{\StdSelection}{\sigma}
-\newcommand{\VarSelection}{\StdSelection^2}
-
-% Tree
-\newcommand{\Nbranch}{2 \NbrTaxa - 2}
-\newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
-\newcommand{\Ntrait}{K}
-\newcommand{\contrast}{\UniDimArray{C}}
-\newcommand{\Covariancematrix}{\Sigma}
-\newcommand{\CovarianceMatrix}{\BiDimArray{\Covariancematrix}}
-\newcommand{\Precisionmatrix}{\Omega}
-\newcommand{\PrecisionMatrix}{\BiDimArray{\Precisionmatrix}}
-\newcommand{\Identitymatrix}{\BiDimArray{I}}
-\newcommand{\brownian}{\mathcal{B}}
-\newcommand{\Brownian}{\UniDimArray{\brownian}}
-\newcommand{\Scattermatrix}{\BiDimArray{A}}
-\newcommand{\Multivariate}{\UniDimArray{Z}}
-
 \renewcommand{\baselinestretch}{1.5}
@@
 }
-
 \begin{document}
@@
 
-% Abstract (≤ 250 words)
-%TC:endignore
-\begin{abstract}
-    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
-    Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
-    Conversely, using between species variation, current methods usually discard within species variation, thus not accounting for processes at the micro-evolutionary scale.
-    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
-    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
-    As such, it does not require estimation of population size nor of time of speciation for normalization.
-    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
-    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
-    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
-\end{abstract}
+\subsection*{Acknowledgements}
+ We gratefully acknowledge the help of Nicolas Lartillot, Philippe Veber, Isabela Jeronimo do Ó, Anna Marcionetti, Julien Clavel, and Daniele Silvestro for their insightful discussions and Julien Joseph for his advice and reviews concerning this manuscript.
 
-\keywords{Quantitative genetics \and Trait evolution \and Selection }
+\subsection*{Competing interests:}
+The authors declare no conflicts of interest.
+\subsection*{Data and materials availability:}
+The data that support the findings of this study are openly available in GitHub at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}.
+Snakemake pipeline, analysis scripts and documentation are available in the repository to replicate the study.
 
-% Research Article (7500 words)
-\section*{Introduction}\label{sec:introduction}
-
-Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
-Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits and sequences.
-Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.
-
-First, leveraging individual variations within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given locus), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
-Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
-For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
-However, dominant selective regime acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
-For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
-Gene expression exhibit such accumulation and divergence in expression accumulate faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\cite{walsh_evolution_2018}.
-
-% Phylogenetics (mean trait evolution)
-Alternatively, at the macro-evolutionary scale, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\cite{hansen_translating_1996}.
-As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
-A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
-Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
-However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
-In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
-Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Altogether, methods at the phylogenetic scale easily misclassify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.
-
-% At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macro evolution, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
-In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
-Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimate of population size (within species) nor of time of speciation (between species).
-For population geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
-For a phylogeneticist, our study can be seen as an extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
-% Basically, we are testing whether the variance of means is greater than the man of variances
-
-\section*{Materials and Methods}\label{sec:materials-and-methods}
-\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
-\subsubsection*{Mutation-drift equilibrium for a trait}
-
-For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
-New mutations are thus generating variance for such a trait, where the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
-At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$ as:
-\begin{gather}
-    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
-\end{gather}
-
-Mutations supply new variants, while genetic random drift depletes standing variation.
-As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
-\begin{align}
-    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
-    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
-\end{align}
-
-For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating divided by the size of the region.
-Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the number of segregating mutations is also a balance between mutations and drift.
-As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
-\begin{gather}
-    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
-\end{gather}
-
-Altogether, we define $\RateWhithin$ as the ratio of genetic variance of the trait ($\VarGenetic$) over nucleotide diversity of any neutral genomic region of interest ($\pi$).
-This ratio allows to discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
-\begin{align}
-    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
-    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
-    & = \RateMut. \label{eq:rate-pop}
-\end{align}
-
-Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), also giving $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
-\begin{align}
-    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
-\end{align}
-
-\subsubsection*{Phylogenetic evolution of a trait}
-
-At generation $\Time$, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
-If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
-After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
-\begin{align}
-    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
-    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
-\end{align}
-
-Moreover, for any neutral genomic region of interest, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
-For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
-As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
-\begin{align}
-    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
-    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
-    & = \MutationRate. \label{eq:substitution-rate}
-\end{align}
-That is, if mutations are neutral, the rate with which new neutral alleles are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.
-
-After having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
-\begin{align}
-    d & = \Time \Multiply \SubRate \\
-    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
-\end{align}
-Altogether, we define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region of interest ($d$).
-This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
-\begin{align}
-    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
-    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
-    & = \RateMut \label{eq:rate-phy}.
-\end{align}
-
-\begin{figure*}[!ht]
-    \centering
-    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
-    \caption{
-        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
-        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
-        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
-        Under neutral evolution, $\EstNI$ is expected to be equal to one.
-        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
-        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
-        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the confounding effect on diversity of mutation rate, population size and divergence time.
-    }
-    \label{fig:methods}
-\end{figure*}
-
-\subsubsection*{Neutrality index}
-
-The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
-Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allows to compute $\pi$.
-Combined as a ratio, $\RateWhithin$ depends solely on the genetic architecture of the trait as $\RateMut$.
-At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
-Similarly, combined as a ratio, $\RateBetween$ depends solely on $\RateMut$
-Altogether, we thus have:
-\begin{gather}
-    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
-    \Rightarrow \frac{\RateBetween}{\RateWhithin} = 1.
-\end{gather}
-Altogether, $\RateBetween / \RateWhithin$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
-Importantly, $\RateBetween$ and $\RateWhithin$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
-Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
-
-\subsection*{Estimate}\label{subsec:estimate}
-
-\subsubsection*{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitution per site) and the topology of the phylogenetic tree.
-Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
-The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be the root, in which case the distance is 0).
-Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate~\cite{omeara_testing_2006}, given as:
-\begin{gather}
-    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
-\end{gather}
-where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.
-
-Finally, within species variation $\EstRateBetween$ is estimated as~\cite{omeara_testing_2006}:
-\begin{gather}
-    \EstRateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
-\label{eq:estimated-rate-phy}
-\end{gather}
-
-For a given species $\Spi$ with inter-individual data available, the ratio of genetic variance of a trait ($\VarGeneticSpi$) over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
-Averaged across all species, we obtain the estimate $\EstRateWhithin$ as:
-\begin{gather}
-    \EstRateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}.
-\label{eq:estimated-rate-pop}
-\end{gather}
-
-Altogether, the neutrality index is estimated as:
-\begin{gather}
-    \EstNI = \frac{\EstRateBetween}{\EstRateWhithin}. \label{eq:estimated-NI}
-\end{gather}
-
-\subsubsection*{Bayesian estimate}
-
-Even though the neutrality index can be applied for any trait, we so far assumed that several traits would to be independent of one another.
-Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
-Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
-Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
-The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
-With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
-$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
-For each trait and each species, heritability ($\Heritability$) is set as a uniform distribution with user-defined boundaries.
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
-
-\subsection*{Simulation}\label{subsec:simulations}
-
-% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
-% pS = 0.0027577478571428568
-% Tree length (dS) = 2.340579
-% Tree length (My) = 1259.4490200000002
-% Root age (My) = 98.9489413888889
-% Root age (dS) = 0.18388820079995308
-% nbr_sites_var = 27.577478571428568
-% u = 1.3788739285714283e-05
-% nbr_generations = 13336.114128321291
-We assumed that the trait genotypic value is encoded by $\NbrLoci=5,000$ loci, with each locus contributing additively to the genotypic value.
-Each effect is Normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
-The phenotypic value is the sum of phenotypic value and an environmental effect, the environmental effect is normally distributed with variance $\VarEnv$.
-We assumed a trait with a heritability of $\Heritability =0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
-We will follow a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
-Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we can simulate explicitly each generation along an ultrametric phylogenetic tree.
-For each offspring, the number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
-From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
-To scale parameters with a mammalian regime, we used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38e^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38e^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch is proportional to the branch length.
-
-The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages are both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
-Additionally, the instant value of log-$\Ne$ is modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
-The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
-The simulation starts from an initial sequence at equilibrium at the root of the tree and, at each node, the process is split until it finally reaches the leaves of the tree.
-From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.
-
-Genetic drift is modeled by the resampling of individuals at each generation, with each parent probability of being sampled being proportional to its fitness ($W$).
-Selection is modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
-More specifically, the fitness of an individual is given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ is the trait value of the individual, $\lambda=0.0$ is the optimal trait value, and $\alpha=0.02$ is the strength of selection.
-Mutations are seen has displacement of the phenotype in the multidimensional space.
-Beneficial mutations move the phenotype closer to the optimum, while deleterious mutations move it further away.
-Stabilizing selection is implemented by means of a fixed optimum phenotype ($\lambda=0.0$).
-Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process ($\brownian \left(0, \sigma_{\lambda}=1.0\right)$).
-Neutral evolution is implemented with a fixed fitness landscape ($W=1$), where each individual has the same probability of being sampled at each generation.
-
-Nucleotide diversity ($\pi$) is measured as the heterozygosity of the neutral markers.
-Nucleotide divergence ($d$) is measured as the number of substitutions per site along branches of the phylogeny.
-The genetic variance is measured as phenotypic variance multiplied by heritability.
-Heritability is estimated from the slope of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\cite{lynch_genetics_1998} using the last generations of the simulations.
-As such heritability is not a given parameter of the simulations, but rather measured as it would be in empirical data.
-
-\begin{figure*}[!ht]
-    \centering
-    \includegraphics[width=\textwidth, page=1] {artworks/fig-simulator}
-    \caption{
-        Wright-Fisher simulations with mutation, selection and drift.
-        Panel A: the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
-        The trait genotypic value is then transformed into a phenotypic value by adding a Gaussian noise with standard deviation $\VarEnv$.
-        Parents are selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
-        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRate$ to mutate.
-        Drift is modeled by the resampling of parents.
-        Panel B: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
-    }
-    \label{fig:simulator}
-\end{figure*}
-
-\subsection*{Empirical dataset}\label{subsec:empirical-dataset}
-
-We analyzed a mammalian dataset, with the body mass and brain mass as quantitative traits.
-The log-transformed values of body mass and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
-We removed individuals not marked as an adult, and split the data into males and females due to sexual dimorphism.
-We discarded species with only one representative.
-The mammalian nucleotide diversity is obtained from the Zoonomia project, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
-
-We also analyzed on a dataset of Primates species, with the nucleotide variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
-However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
-
-\section*{Results}\label{sec:results}
-
-To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at different scales.
-The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
-More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by nucleotide diversity of any neutral genomic region ($\pi$), denoted $\RateWhithin$, is a proxy for the mutational variance $\RateMut$.
-At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a $\Time$ generations is $\VarPhy$.
-This variance can also be normalized by the nucleotide divergence of any neutral genomic region ($d$), denoted $\RateBetween$, allowing us to obtain another proxy for $\RateMut$.
-Altogether, we defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
-Under a depleted variation between species, meaning $\NI < 1$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
-In contrast, under an inflated variation between species, meaning $\NI > 1$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.
-
-\begin{table*}[t!]
-    \centering
-    \begin{adjustbox}{width = 1.0\textwidth}
-        \begin{tabular}{||l|l||c|c||c|c||}
-            \hline
-            Assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
-            Trait encoded by few loci                        & Between species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
-            Sexual dimorphism                                & Within species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
-            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
-        \end{tabular}
-    \end{adjustbox}
-    \caption{Assumptions and their consequences on the estimation of within species variation ($\EstRateWhithin$), between species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
-    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
-    }
-    \label{table:assumptions}
-\end{table*}
-
-Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate, population size and divergence time.
-The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
-Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
-This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
-We also devised a Bayesian estimated allowing to obtain the posterior distribution of the neutrality index, and to test for evidence of diversifying selection as $\proba [\EstNI > 1]$, and evidence for stabilizing selection as $\proba [\EstNI < 1]$.
-
-\subsection*{Results against simulations}\label{subsec:results-against-simulations}
-
-The inference framework was first tested on independently simulated datasets.
-With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant mammalian empirical regime (see Materials and Methods).
-First, we assumed a constant population size ($\Ne$) and a constant mutation rate ($\MutationRate$) across the phylogeny (fig.~\ref{fig:results-simulations}A\&B)
-Simulations performed under stabilizing selection where detected as such ($\proba [\EstNI < 1] > 0.95$), no false negative was observed (blue in fig.~\ref{fig:results-simulations}A\&B).
-Simulations performed under diversifying selection were detected as such ($\proba [\EstNI > 1] > 0.975$), no false negative was observed (red in fig.~\ref{fig:results-simulations}A\&B).
-Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 77\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B), while detected as stabilizing selection in 21\% of the cases, and detected as diversifying selection in 2\% of cases.
-Second, for simulations under a fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}C\&D), simulations of diversifying selection was detected as such and stabilizing selection was detected as such in 100\% of the cases.
-For simulations under neutral evolution, 51\% of the simulations were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), 49\% of the simulations were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
-
-\begin{figure*}[!ht]
-    \centering
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.rho}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.pvalues}
-    \end{minipage}
-    \caption{
-        $1,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
-        Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-        Histogram of the ratio of between species trait variation ($\RateBetween$) over within species trait variation $\RateWhithin$, as $\NI = \RateBetween / \RateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\NI$ is greater than $1$ (right).
-        Effective population size ($\Ne$) and mutation rate ($\MutationRate$) are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
-    }
-    \label{fig:results-simulations}
-\end{figure*}
-
-\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
-For mammalian body mass, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
-The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.34$ ($[0.22--0.52]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.28$ ($[0.16--0.49]$ for $95\%$ CI).
-In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.0$ and $0.0$ respectively.
-Moreover, assuming that heritability ($\Heritability$) of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection ($ \proba [\EstNI > 1]$) becomes $0.635$ in males and $0.914$ in females.
-For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.35$ ($[0.85--2.17]$ for $95\%$ CI, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.72$ ($[1.00--2.94]$ for $95\%$ CI, $\NbrTaxa=26$).
-In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.877$ and $0.972$ respectively.
-Moreover, assuming that heritability of brain mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in both males and females.
-
-For primates body mass, we obtained $\EstNI_{\text{\Male}} = 0.56$ ($[0.40,~0.78]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.39$ ($[0.28,~0.55]$ for $95\%$ CI, $\NbrTaxa=65$).
-Assuming that heritability of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in males and $0.914$ in females.
-For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.95$ ($[1.40,~2.79]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.95$ ($[1.39,~2.79]$ for $95\%$ CI, $\NbrTaxa=65$).
-In both males and females, brain mass is under diversifying selection, with associated posterior probabilities of $1.0$.
-Importantly, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).
-
-\section*{Discussion}\label{sec:discussion}
-
-% Summary of the method and results
-In this study, we proposed a neutrality index for a quantitative trait.
-At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
-Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
-Our neutrality index for a trait, $\NI$, is the ratio of between species variation to within species variation.
-$\NI$ can be compared to 1 to seek a statistical deviation from a neutral trait.
-In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
-Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
-Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation (see section~S\ref{sec:implementation}) also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
-$ \proba [ \NI > 1 ]$).
-Importantly, our neutrality index leverages nucleotide divergence and polymorphism, as well as the heritability of traits ($\Heritability$).
-We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize the variation of trait between and within species.
-By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, while not requiring to estimate divergence time~\cite{litsios_effects_2012}.
-Instead, the normalization by nucleotide divergence and polymorphism automatically absorbs these changes~\cite{seo_estimating_2004}.
-
-% Pitfalls of the method
-We argue that the main drawback of our method is that the heritability ($\Heritability$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically harder to measure for any trait.
-Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is under-powered.
-Indeed, because $\Heritability \leq 1$ by definition, if we set $\Heritability \leq 1$, we overestimate the within species variation and thus underestimate $ \NI$ and can test for diversifying selection.
-In other words, testing for $\NI > 1$ while setting $\Heritability = 1$ means that knowing the real heritability would only increase the evidence for diversifying selection.
-Contrarily, the test for stabilizing selection is invalid if heritability is not known.
-Because our test is also based on several assumptions that might not hold on empirical data, we provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
-As an example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$.
-However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for nucleotide divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
-
-% What can and should be improved
-Our development is based on several assumptions that could be relaxed.
-First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
-Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically be relaxed by adding a time component~\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny~\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of polymorphism and divergence.
-Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within species in a unified framework.
-Also, we don't know how our test would behave in the context of gene flows and introgression, which could be investigated in future studies.
-
-% How is our test related to other tests of selection
-Our test bears analogy with other tests for neutrality developed, providing insight into its behavior.
-First, our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
-Second, because $\NI$ is compared to 1, our test bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\cite{goldman_codonbased_1994, muse_likelihood_1994}.
-As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
-Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\cite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\cite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\cite{latrille_genes_2023}.
-By explicitly modeling stabilizing selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
-
-% What can our test bring to the field
-Finally, from an empirical point of view, our test is also related to other methods that have been developed to test for selection on quantitative traits.
-At the phylogenetic scale, test for neutrality of a quantitative trait are seeking deviation from a Brownian process~\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation~\cite{rohlfs_phylogenetic_2015}.
-Ahe analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\cite{catalan_drift_2019}.
-However, the alternative model of a moving optimum was not implemented, and thus cannot be ruled out by this analysis.
-With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
-Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence ($d$) and polymorphism ($\pi$).
-Altogether, our study provides a statistical framework to test for neutrality of a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
-
-%TC:ignore
-\section*{Acknowledgements}
-\label{sec:acknowledgment}
-\textbf{Funding:}
-Université de Lausanne.
-\textbf{Author contributions:}
-Original idea: T.L.\ and N.S.;
-Model conception: T.L.\ and N.S.;
-Code: T.L.;
-Data analyses: T.L.\ and M.B.;
-Interpretation: T.L., T.G\ and N.S.;
-First draft: T.L.;
-Editing and revisions: T.L., M.B, T.G\ and N.S.;
-Project management and funding: N.S\@.
-\textbf{Competing interests:}
-The authors declare no conflicts of interest.
-\textbf{Data and materials availability:}
-Snakemake pipeline, analysis scripts and documentation are available at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}.
-
-\printbibliography
-
-\newpage
-
-\part*{Supplementary materials}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\setcounter{section}{0}
-
-\renewcommand{\baselinestretch}{1.0}\normalsize
-\tableofcontents
-\renewcommand{\baselinestretch}{1.5}\normalsize
-
-\newpage
-\section{Genetic architecture of the trait}\label{sec:simulator}
-
-\subsection{Genotype-phenotype map}\label{subsec:genotype-phenotype-map}
-
-\begin{itemize}
-    \item $\NbrLoci$ is the number of loci encoding the trait.
-    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
-    \item $\Ne$ is the effective number of individuals.
-    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
-    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the genotypic value for individual $i$.
-    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
-    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
-\end{itemize}
-
-\begin{center}
-    \captionof{figure}{Summary of trait's genetic architecture.}
-    \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
-    \label{fig:simulator-summary}
-\end{center}
-
-Within species, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
-\begin{equation}
-    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad} \VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2\label{eq:simu-genotype}
-\end{equation}
-The theoretical additive genetic variance ($\VarGenetic$) is a function of the number of loci ($\NbrLoci$) and the effect of a mutation ($a$) as:
-\begin{equation}
-    \VarGenetic = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \label{eq:simu-var-genetic}
-\end{equation}
-
-The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
-\begin{equation}
-    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2 \label{eq:simu-between}
-\end{equation}
-
-Heritability ($\Heritability$) is defined as:
-\begin{equation}
-    \Heritability = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
-\end{equation}
-Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($\Heritability$) as:
-\begin{equation}
-    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
-\end{equation}
-
-\newpage
-\section{Bayesian estimate}\label{sec:bayesian-estimate}
-
-\subsection{Multivariate Brownian process}\label{subsec:multivariate-brownian-process}
-Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
-Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
-The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-Along branch $j$ with length $d_{j}$, the Brownian process start at the ancestral node $\mathcal{A}(j)$ with value $\Brownian(\mathcal{A}(j))$, and ends at node $\mathcal{R}(j)$  with value $\Brownian(\mathcal{R}(j))$.
-The independent contrast $\contrast_{j}$ defined as change in trait along the branch normalized by $\sqrt {d_{j}}$ is a multivariate Gaussian:
-\begin{equation}
-    \label{eq:DistribBrownian}
-    \contrast_{j} = \frac{\Brownian_{\mathcal{R}(j)} - \Brownian_{\mathcal{A}(j)} }{\sqrt {d_{j}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
-\end{equation}
-The {prior} on the covariance matrix as an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
-\begin{equation}
-    \label{eq:Distribcovariance}
-    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
-\end{equation}
-
-\subsection{Sampling the covariance matrix}\label{subsec:sampling-the-covariance-matrix}
-From the independent contrast at each branch of the tree ($\contrast_{j}$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$, as:
-\begin{equation}
-    \Scattermatrix = \sum\limits_{j=1}^{\Nbranch} \contrast_{j} \MultiplyMatrix \left[\contrast_{j}\right]\tr\label{eq:bayes-scatter},
-\end{equation}
-where $\Nbranch$ is the number of branches in the tree and $\NbrTaxa$ the number of taxa.
-
-By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\Brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
-\begin{equation}
-    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)\label{eq:bayes-posterior}
-\end{equation}
-This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate_{k}$ defined by
-\begin{equation}
-    \Multivariate_{k} \sim \mathcal{N} \left( \VecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
-\end{equation}
-And from these multivariate samples, $\CovarianceMatrix$ is Gibbs sampled as:
-\begin{equation}
-    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate_{k} \MultiplyMatrix  \left[\Multivariate_{k} \right] \tr \right)\inv \label{eq:bayes-gibbs}
-\end{equation}
-
-\newpage
-\section{Bayesian and Maximum-likelihood implementation}\label{sec:implementation}
-
-Implementation is included within the \textit{BayesCode} software, available at \url{https://github.com/ThibaultLatrille/bayescode}.
-
-\subsection{Data formatting}\label{subsec:data-formatting}
-
-Running the analysis on your dataset and compute posterior probabilities requires three files:
-\begin{enumerate}
-    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers)
-    \item A file containing the mean trait values for each species.
-    \item A file containing the variation within species for each trait and the genetic variation within species (neutral markers).
-\end{enumerate}
-
-\subsubsection{Phylogenetic tree}\label{subsubsec:phylogenetic-tree}
-
-The phylogenetic tree must be in newick format, with branch lengths in number of substitutions per site (neutral markers).
-
-\subsubsection{Mean trait for each species}\label{subsubsec:mean-trait-for-each-species}
-
-The file containing mean trait values for each species must be in a tab-delimited file with the following format:
-\begin{center}
-    \begin{adjustbox}{width = 0.35\textwidth}
-        \begin{tabular}{|l|c|c|}
-            \hline
-            TaxonName            & Body\_mass & Brain\_mass \\
-            \hline
-            Panthera\_tigris     & 12.26      & 5.676       \\
-            Pithecia\_pithecia   & 7.256      & 3.436       \\
-            Colobus\_angolensis  & 9.176      & 4.284       \\
-            Saimiri\_boliviensis & 6.845      & 3.279       \\
-            $\vdots$             & $\vdots$   & $\vdots$    \\
-            \hline
-        \end{tabular}\label{tab:trait-mean}
-    \end{adjustbox}
-\end{center}
-
-The columns are:
-\begin{itemize}
-    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
-    \item As many columns as traits, without spaces or special characters in the trait.
-    \item The values can be \texttt{NaN} to indicate that the trait is not available for that taxon.
-\end{itemize}
-
-\newpage
-\subsubsection{Trait variation for each species}\label{subsubsec:trait-variation-for-each-species}
-
-The file containing trait variation for each species must be in a tab-delimited file with the following format:
-\begin{center}
-    \begin{adjustbox}{width = 1.0\textwidth}
-        \begin{tabular}{|l|c|c|c|c|c|c|}
-            \hline
-            TaxonName            & Nucleotide\_diversity & Body\_mass\_variance & Body\_mass\_heritability & Brain\_mass\_variance & Brain\_mass\_heritability \\
-            \hline
-            Pithecia\_pithecia   & 0.0016                & 0.22871              & 0.2                      & 0.00737               & 0.2                       \\
-            Colobus\_angolensis  & 0.0017                & 0.00393              & 0.2                      & 0.00416               & 0.2                       \\
-            Saimiri\_boliviensis & 0.0013                & 0.00022              & 0.2                      & 0.00045               & 0.2                       \\
-            Pygathrix\_nemaeus   & 0.0016                & 0.00347              & 0.2                      & 0.00097               & 0.2                       \\
-            $\vdots$             & $\vdots$              & $\vdots$             & $\vdots$                 & $\vdots$              & $\vdots$                  \\
-            \hline
-        \end{tabular}
-        \label{tab:trait-variance}
-    \end{adjustbox}
-\end{center}
-
-\begin{itemize}
-    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
-    \item \emph{Nucleotide\_diversity}: the nucleotide diversity within species (neutral markers), cannot be \texttt{NaN}.
-    \item As many columns as traits, without spaces or special characters in the trait.
-    \item \emph{TraitName\_variance}: the phenotypic variance of the trait within species, can be \texttt{NaN} to indicate that the trait variance is not available for that taxon.
-    \item \emph{TraitName\_heritability} (optional): the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
-    \item The columns with the suffix \texttt{\_variance} and \texttt{\_heritability} are repeated for each trait.
-    \item \emph{TraitName\_heritability\_lower} (optional): the lower bound of the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
-    \item \emph{TraitName\_heritability\_upper} (optional): the upper bound of the heritability of the trait within species, between 0 and 1, cannot be \texttt{NaN}.
-    \item If the columns with the suffix \texttt{\_heritability\_lower} and \texttt{\_heritability\_upper} are present, the heritability is randomly drawn from a uniform distribution between the lower and upper bounds.
-    \item If the columns with the suffix \texttt{\_heritability} is present, it is taken as is.
-    \item If the genetic variance (instead of phenotypic variance) is available for a trait, the heritability can be omitted and will automatically be set to 1.0.
-\end{itemize}
-
-\newpage
-\subsection{Bayesian estimation}\label{subsec:running-nodetraitsand-readnodetraits}
-
-The executable \texttt{nodetraits} from \textit{BayesCode} is used to run the Bayesian estimation of the model, and the executable \texttt{readnodetraits} is used to read the results.
-
-Assuming that the file \texttt{data/body\_size/mammals.male.tsv} contains the mean trait values for each species, the file \texttt{data/body\_size/mammals.male.var\_trait.tsv} contains the variation within species for each trait and the genetic variation within species (neutral markers), and the file \texttt{data/body\_size/mammals.male.tree} contains the phylogenetic tree, the following commands are used to run the model and read the results.
-
-\subsubsection{Running the model}\label{subsubsec:running-nodetraits}
-\texttt{nodetraits} is run with the following command:
-\begin{lstlisting}[language = sh,label={lst:nodetraits-run}]
-nodetraits  --until 2000
-            --tree data/body_size/mammals.male.tree
-            --traitsfile data/body_size/mammals.male.tsv
-            run_mammals_male
-\end{lstlisting}
-
-\subsubsection{Reading the results}\label{subsubsec:reading-the-results}
-Once the model has ran, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between species variation over within species variation with \texttt{readnodetraits}:
-\begin{lstlisting}[language = sh,label={lst:readnodetraits-rho}]
-readnodetraits --burnin 1000
-               --var_within data/body_size/mammals.male.var_trait.tsv
-               --output results_mammals_male.tsv
-               run_mammals_male
-\end{lstlisting}
-The file \texttt{data\_empirical/chain\_name.ratio.tsv} then contains the posterior mean of the ratio of between species variation over within species variation, the 95\% and 99\% credible interval, and the posterior probability that the ratio is greater than 1.
-
-\subsection{Maximum likelihood estimation}\label{subsec:maximum-likelihood-estimation}
-
-To obtain the ratio (without the posterior credible interval and probability) using maximum likelihood computation, the following python script can be used:
-\begin{lstlisting}[language = sh, label={lst:neutrality_index}]
-python3 utils/neutrality_index.py --tree data/body_size/mammals.male.tree
-                                  --traitsfile data/body_size/mammals.male.tsv
-                                  --var_within data/body_size/mammals.male.var_trait.tsv
-                                  --output results_ML_mammals_male.tsv
-\end{lstlisting}
 \end{document}
-%TC:endignore

### 2023-09-13 data_simulated/update_snake.sh

@@ -1,5 +1,5 @@
 FOLDER=$(pwd)
-for EXPERIMENT in $(pwd)/experiments/*; do
+for EXPERIMENT in $(pwd)/experiments/fluc*ing; do
   echo "Updating snakemake in: ${EXPERIMENT}"
   rm -rf "${EXPERIMENT}"/Snakefile
   ln $(pwd)/Snakefile "${EXPERIMENT}"/Snakefile

### 2023-09-13 scripts/merge_results_mammals.py

@@ -21,10 +21,25 @@ def main(tsv_ML_list: str, tsv_Bayes_list: str, output: str):
             list_df.append(df)
     df_out = pd.concat(list_df)
     # Sort by trait, dataset, sex, logT and then method
-    df_out = df_out.sort_values(by=["trait", "dataset", "sex", "logT", "h2", "method"])
+    df_out = df_out.sort_values(by=["dataset", "trait", "logT", "h2", "sex", "method"],
+                                ascending=[True, True, False, False, False, False])
     # Fill missing values with NaN
     df_out = df_out.fillna("NaN")
     df_out.to_csv(output, sep="\t", index=False, float_format="%.3f")
+
+    # Filter out the ML results
+    df_out = df_out[df_out["method"] == "Bayesian"]
+    dico_label = {"dataset": "dataset", "trait": "trait", "h2": "$h^2$", "sex": "sex",
+                  "nbr_taxa_between": "$n$", "ratio": "$\\rho$", "ratio_CI95": "$\\rho_{CI95}$",
+                  "pp_ratio_greater_1": "$\\mathbb{P}(\\rho > 1)$"}
+    columns = [i for i in dico_label if i in df_out]
+    df_out = df_out[columns]
+    df_out["dataset"] = df_out["dataset"].apply(lambda x: "Mammals" if x == "NeutralPhyloP" else x)
+    df_out["trait"] = df_out["trait"].apply(lambda x: x.replace("_g", "").replace("_", " "))
+    df_out["ratio_CI95"] = df_out["ratio_CI95"].apply(lambda x: "-".join([f"{float(i):.3f}" for i in x.split("--")]))
+    df_out = df_out.rename(columns={i: (dico_label[i] if i in dico_label else i) for i in columns})
+    df_out.to_latex(replace_last(output, ".tsv", ".tex"), index=False, float_format="%.3f",
+                    escape=False)
 
 
 if __name__ == '__main__':

### 2023-09-13 scripts/neutrality_index.py

@@ -125,17 +125,18 @@ def main(input_traits: str, input_tree: str, input_var_within: str, output_tsv: 
         if h_low in var_pop_df.columns and h_up in var_pop_df.columns:
             print(f"Found upper and lower bound for heritability of {trait}.")
             print(f"Using the mean of the confidence interval.")
-            var_pop_df[h] = var_pop_df[[h_low, h_up]].mean(axis=1)
+            trait_h = var_pop_df[[h_low, h_up]].mean(axis=1)
         elif f"{trait}_heritability" in var_pop_df.columns:
             print(f"Found heritability for {trait}.")
+            trait_h = var_pop_df[h]
         else:
             print(f"Warning: column {trait}_heritability not found in {input_var_within}.")
             print("Assuming heritability = 1.0.")
-            var_pop_df[h] = 1.0
+            trait_h = np.array([1.0] * len(var_pop_df))
         notna = (np.isfinite(var_pop_df[f"{trait}_variance"]) & (var_pop_df[f"{trait}_variance"] > 0.0))
-        notna = (notna & (np.isfinite(var_pop_df[h]) & (var_pop_df[h] >= 0.0) & (var_pop_df[h] <= 1.0)))
+        notna = (notna & (np.isfinite(trait_h) & (trait_h >= 0.0) & (trait_h <= 1.0)))
         # Computing the genetic variance (geno = h² * pheno)
-        genetic_variance_array = var_pop_df[f"{trait}_heritability"][notna] * var_pop_df[f"{trait}_variance"][notna]
+        genetic_variance_array = trait_h[notna] * var_pop_df[f"{trait}_variance"][notna]
         var_within_array = genetic_variance_array / (var_pop_df["Nucleotide_diversity"][notna])
         print(f'Found {len(genetic_variance_array)} species with a variance for {trait}.')
         var_within = np.mean(var_within_array)

### 2023-09-13 scripts/pre_processed_traits_mammals.py

@@ -112,7 +112,7 @@ def main(path_input_traits: str, path_input_nuc_div: str, path_input_dS: str, pa
             nuc_div = np.nan
         else:
             assert len(leaf_nuc_div_df) == 1
-            nuc_div = float(leaf_nuc_div_df[nuc_div_col])
+            nuc_div = float(leaf_nuc_div_df[nuc_div_col].iloc[0])
         dico_var_within["TaxonName"].append(taxa_name)
         dico_var_within[f"Nucleotide_diversity"].append(nuc_div)
         dico_traits["TaxonName"].append(taxa_name)

### 2023-09-19 manuscript/main.tex
Include N. Salamin comments

@@
 \DeclareMathOperator{\Cov}{\text{Cov}}
+\newcommand{\QstFst}{Q$_\text{ST}$--F$_\text{ST}$}
 \newcommand{\der}{\mathrm{d}}
@@
 
-[not sure of this paragraph...]
-
 Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
-Additionally, contrasting both trait and genetic differentiation across several populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
-For example, studies have shown that stabilizing selection is largely dominant in the evolution of gene expression in different populations~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
-[why only example from gene expression?]
-However, dominant selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
+Additionally, contrasting both trait and genetic differentiation across several populations, \QstFst methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
+For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006.
+However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
 
@@
 Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
-Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
+Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilising selection~\parencite{catalan_drift_2019}.
 Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
 However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
+First, a trait under stabilising selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
@@
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value~\parencite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-However, in the context of detecting selection acting on a trait, within-species variation is mostly used to describe technical errors.
-One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
+At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value to describe technical errors~\parencite{lynch_methods_1991, felsenstein_comparative_2008, hansen_interpreting_2012} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
+One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average as a signal of diversifying selection~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilising, and diversifying selection.
 We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
 Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
-For population geneticists, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
+For population geneticists, our study can be seen as the macro-evolutionary generalization of \QstFst methods to account for phylogenetic relationships between species.
 For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
@@
 
-The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ beinf a function of $\VarPhenotype$ and $\Heritability$:
+The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
 \begin{align}
@@
         Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
-        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
+        Under stabilising selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
         Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
@@
 \subsubsection*{Bayesian estimate}
-[NS: the start of this paragraph is not so clear -> to discuss]
-Even though the neutrality index can be applied to any trait, we assume here that the variation between several traits is independent.
-Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
-Here we generalize to $\Ntrait$ non-independent traits evolving along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
-Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
+
+The Bayesian framework allows to obtain the posterior distribution the neutrality index ($\EstNI$) for a given trait.
+Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
+Trait variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
+The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
 With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
@@
 From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
+The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilising selection.
+
+\subsubsection*{Applicability to empirical data}
+
+Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
+Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
+Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
+Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
+It is however possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
+Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in under-estimated $\EstNI$ since $\Heritability \leq H^2$.
+Contrarily, the test for stabilising selection is invalid if $\EstNI$ is under-estimated.
+Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
 
@@
 % nbr_generations = 13336.114128321291
-We tested the xxx [NS: explain a bit what we test here -> NI]
-We simulated phenotypic values as the sum of genotypic value and an environmental effect.
+We tested the performance of our neutrality index ($\NI$) to detect selection on a quantitative trait using simulations.
+We performed simulations under different selective regimes (neutral, stabilising, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of mutation rate (constant or fluctuating).
+Simulation are individual-based and follows a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
+Each individual phenotypic value is the sum of genotypic value and an environmental effect.
 The environmental effect was normally distributed with variance $\VarEnv$.
@@
 We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
-We followed a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
 Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
@@
 The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
-Additionally, the instant value of log-$\Ne$ was modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
-The geometric Brownian motion accounted for long-term fluctuations, while the Ornstein-Uhlenbeck introduced short-term fluctuations. [NS: not clear how you combined both here]
+Additionally, the instant value of log-$\Ne$ was modeled as a independent sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
+The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
 The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
@@
 For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
-At the species level, the additive genetic variance of the neutral trait ($\VarGenetic$) is normalized by nucleotide diversity of any neutral genomic region ($\pi$) to obtain $\RateWhithin$.
-At the scale of between-species variation, assuming a neutral trait evolving along the phylogeny, the mean trait value changes as a Brownian process.
-This variance in mean trait value can also be normalized by the nucleotide divergence of any neutral genomic region ($d$) to obtain $\RateBetween$.
-[NS: -> this is repeating the mat/met, is it really needed??]
 We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see methods), suggesting that traits for which this relationship is not verified are putatively under selection.
-Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
+Under stabilising selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
 In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
-
-[NS: this is not a result but an assumption that we have. We also already say it above. I would move it to discussion if needed]
-Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate, population size and divergence time.
-The relationship between trait variation between and within species can be used as a method to detect whether traits are evolving under a neutral model of evolution.
-Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between-species (see fig.~\ref{fig:methods}).
-This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
+Our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between-species (see fig.~\ref{fig:methods}).
+This statistic can be obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
+We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilising selection as $\proba [\EstNI < 1]$.
 
@@
     \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
-    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
+    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilising selection $\NI < 1$ are conservative or invalid due to violated assumptions.
     }
@@
 
-\subsection*{Test of selection}
-
-[NS: here again, these are not results, but a discussion]
-We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
-
-Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
-Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
-Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
-Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
-It is however possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
-Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in under-estimated $\EstNI$ since $\Heritability \leq H^2$.
-Contrarily, the test for stabilizing selection is invalid if $\EstNI$ is under-estimated.
-Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
-
 \subsection*{Results against simulations}\label{subsec:results-against-simulations}
 
-The inference framework was first tested on independently simulated datasets.
-With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant mammalian empirical regime (see Materials and Methods).
-[NS: I would remove this sentence]
-Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
-For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
-Once we introduced fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
-For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
+The inference framework was first tested on independently simulated datasets matching an empirically relevant mammalian empirical regime (see Materials and Methods).
+Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilising ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
+For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilising or diversifying selection, respectively.
+Once we introduced fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilising selection remained the same with all cases detected correctly.
+For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilising selection ($\proba [\EstNI < 1] > 0.975$).
 
@@
         $10,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
-        Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
+        Traits simulated under stabilising selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
         Histogram of estimated the ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$, as $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\EstNI$ is greater than $1$ (right).
@@
 Assuming complete heritability, brain mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
-If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
+If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\parencite{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
 Mammalian brain mass was found to be under diversifying selection with posterior probabilities of $0.877$ for males and $0.972$ for females when complete heritability was assumed.
@@
 For primate body mass, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%,
-Fixing the heritabity of body mass to $1.0$ did not change the posterior probability for males, but increased the one for female to $1.0$.
+Assuming complete heritability of body mass did not change the posterior probability for males, but increased the one for female to $1.0$.
 Evidence for diversifying selection on body mass was therefore more pronounced in Primates than in Mammals.
@@
 Estimated $\EstNI$ can be compared against the null hypothesis of neutrality by testing for deviation from the value of $1.0$.
-In other words, our statistical test seeks to find traits with extremal values in comparison to a pool of traits.
-Instead, our procedure can be applied to a single trait [NS: this is not clear as the previous sentence seems to contradict this one], estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
 Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S\ref{sec:implementation}) also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
 $ \proba [ \EstNI > 1 ]$).
-We tested our statistical procedure against simulated data and showed that for simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$) our test is able to correctly detect those regimes of selection.
-However, for simulations of a neutral trait, our test detects frequently a spurious signal of stabilizing selection ($\EstNI < 1$).
-Thus, we argue that our test should be used to detect diversifying selection, but has low accuracy to detect stabilizing selection due to many false positives.
-Importantly, our neutrality index leverages nucleotide divergence and polymorphism, as well as the heritability of traits ($\Heritability$).
+We tested our statistical procedure against simulated data and showed that for simulations under diversifying selection (test of $\EstNI > 1$) or under stabilising selection (test of $\EstNI < 1$) our test is able to correctly detect those regimes of selection.
+However, for simulations of a neutral trait, our test detects frequently a spurious signal of stabilising selection ($\EstNI < 1$).
+Thus, we argue that our test should be used to detect diversifying selection, but has low accuracy to detect stabilising selection due to many false positives.
+
+Based on these results, we argue that our method is a significant improvement over current methods to detect selection acting on a trait at the phylogenetic scale.
+Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilising selection over a Brownian process when the trait is neutral~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Thus our method does not provide improvement to detect stabilising selection at the phylogenetic scale.
+However, to detect diversifying selection, current methods modelling mean trait value are limited by the theory behind~\parencite{harmon_phylogenetic_2018}
+Namely that under diversifying selection, mean trait value will not deviate from a Brownian process, and thus can not be distinguished from neutral evolution~\parencite{hansen_translating_1996}.
+For example, test for neutrality concluded that the expression level of the majority of genes are under a neutral regime~\parencite{catalan_drift_2019}.
+Our diversity index allows to discriminate the alternative model of diversifying selection from the neutral case, inviting us to revisit these studies.
+Leveraging the ratio of between-species over within-species variations, methods have compared this ratio across a pool of traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+This strategy allows to identify outlier traits putatively under diversifying selection, but does not allow to test for selection on a single trait and requires independent confirmation~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
+Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution for a single test.
+Again, our diversity index can thus allow us to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence and polymorphism.
+
 We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize trait variation between and within species.
-By doing so, we also show using simulated data that our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, while not requiring to estimate them nor to estimate divergence time~\parencite{litsios_effects_2012}.
-Instead, the normalization by nucleotide divergence and polymorphism automatically absorbs these changes.
-[NS: we need to develop these arguments and show that the normalization works because it is similar to the dN/dS tests, etc....
-I would thus move the paragraph below on this here.
-Otherwise, this paragraph is a bit hollow]
+In the context of within species variation but across several populations, \QstFst test developed from trait and sequence allow to test for selection of a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
+Similarly, in our test, the genetic sequences from which nucleotide divergence and polymorphism are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
+Nucleotide variation allow to normalize for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRate$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Thus our test avoids to estimate them and also avoids to estimate divergence time which is otherwise necessary~\parencite{walsh_evolution_2018}.
+But importantly, by normalising with sequence variation, we also show using simulated data that our test is not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time are constant across the phylogeny, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
+Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRate$ and generation time impacts similarly trait and sequence variation.
+Thus the normalisation by nucleotide divergence and polymorphism automatically absorbs long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time which cancels out in the ratio of trait variation $\EstNI$.
+
+% How is our test related to other tests of selection
+Even though our test is developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provide insight into its behavior.
+First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations are used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
+Second, because $\NI$ is compared to 1, our test ultimately bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
+As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilising selection acting on the trait.
+Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
+With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
+First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilising selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilising selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
+By explicitly modeling stabilising selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilising selection in the context of quantitative traits to obtain a statistically more powerful test.
 
 % Pitfalls of the method
-In the context of detecting diversifying selection, we argue that the main drawback of our method is that the narrow-sense heritability ($\Heritability$) of a trait is required if one has access only to the phenotypic variance and not the additive genetic variance.
-[NS: a bit circular.
-If we have V_G, we know h^2 as V_P can be measured]
+In the context of detecting diversifying selection, we argue that the main drawback of our method is that the additive genetic variance of the trait is required instead of the phenotypic variance.
 If phenotypic variance is used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ is ultimately under-estimated.
 Similarly, using broad-sense heritability instead of narrow-sense heritability results in under-estimated $\EstNI$.
-In such context, the test of stabilizing selection ($\EstNI < 1]$) is statistically invalid.
+In such context, the test of stabilising selection ($\EstNI < 1]$) is statistically invalid.
 However, the test of diversifying selection ($\EstNI > 1$) is underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
-Because our test is also based on several assumptions that might not hold on empirical data, we also provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
-[NS: I would discuss the elements of this table and link that with some literature if possible.]
-As an example, even though we assumed complete heritability for brain mass, we can assert that brain mass is evolving under diversifying selection in mammals since $\EstNI > 1$.
-However, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass is evolving under diversifying selection, since the markers used for nucleotide divergence are not neutral which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
+As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
+However, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass is evolving under diversifying selection since the markers used for nucleotide divergence are not neutral, which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
 
@@
 These factors should be thoroughly investigated using simulations.
-Second, loci are assumed to contribute additively to the phenotype~\parencite{hill_data_2008, crow_epistasis_2010}, disregarding effects of dominance and epistasis.
-[NS: here a few words on the importance of dominance and epistasis in empirical quant genet studies would be interesting.
-And what about long-term effects at the phylogenetic level?]
-Third, the genetic architecture of the trait is assumed to be constant across the phylogeny, which might actually be variable among individuals and species.
-[NS: any literature on this (or is it the following ones)?
-I'm not aware of that right now, but we should have a look]
-Such assumption can theoretically be relaxed and changes in genetic architecture along the phylogeny could jointly be estimated ~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Second, loci are assumed to contribute additively to the phenotype.
+Although the effects of dominance and epistasis is typically weak compared the additive effects on the phenotype, their influence shall be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
+Third, the genetic architecture of the trait is assumed to be constant across the phylogeny, it might actually be variable among individuals and species~\parencite{tung_genetic_2015, huber_conservatism_2015}.
+Such assumption can theoretically be relaxed and changes in genetic architecture along the phylogeny could jointly be estimated~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
 
-% How is our test related to other tests of selection
-Even though our test is developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provide insight into its behavior.
-First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations are used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
-Second, because $\NI$ is compared to 1, our test ultimately bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
-As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
-Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
-By explicitly modeling stabilizing selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
-
 % What can our test bring to the field
-[NS: should we not start with this idea?]
-Finally, from an empirical point of view, our test is also related to other methods that have been developed to test for selection on quantitative traits.
-At the phylogenetic scale, tests for neutrality of a quantitative trait are seeking deviation from a Brownian process~\parencite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation~\parencite{rohlfs_phylogenetic_2015}.
-The analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection~\parencite{catalan_drift_2019}.
-However, the alternative model of a moving optimum was not implemented, and thus cannot be ruled out by this analysis.
-With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
-Our diversity index can thus allow us to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence ($d$) and polymorphism ($\pi$).
-Even though our method requires integrating genomic and trait variation, we also showed that our method can be applied to the empirical dataset with the illustrative example of mammals' brain and body mass.
+From an empirical point of view, even though our method requires integrating genomic and trait variation, we also showed that our method can be applied to the empirical dataset with the illustrative example of mammals' brain and body mass.
+Because our test is also based on several assumptions that might not hold on empirical data, we also provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
 Altogether, our study provides a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
@@
 \end{equation}
-The {prior} on the covariance matrix as an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
-\begin{equation}
-    \label{eq:Distribcovariance}
-    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
-\end{equation}
 
@@
 
+The {prior} on the covariance matrix is an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
+\begin{equation}
+    \label{eq:Distribcovariance}
+    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
+\end{equation}
 By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\Brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.

### 2023-09-22 manuscript/main.tex
Prep. manuscript. Find exec in various Bayescode folder

@@
 \newcommand{\Ne}{N_{\text{e}}}
-\newcommand{\dn}{d_N}
-\newcommand{\ds}{d_S}
-\newcommand{\dnds}{\dn / \ds}
+\newcommand{\dnds}{\omega}
 \newcommand{\pn}{\pi_N}
@@
 Additionally, contrasting both trait and genetic differentiation across several populations, \QstFst methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
-For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006.
+For example in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
+[why only example from gene expression?]
 However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
@@
 Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
-Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilising selection~\parencite{catalan_drift_2019}.
+Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
 Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
 However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
-First, a trait under stabilising selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
+First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
@@
 One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average as a signal of diversifying selection~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilising, and diversifying selection.
+The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
 We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
@@
 For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
+% Basically, we are testing whether the variance of means is greater than the mean of variances
+
 % Basically, we are testing whether the variance of means is greater than the mean of variances
@@
         Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
-        Under stabilising selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
+        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
         Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
@@
 From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilising selection.
+The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
 
@@
 Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in under-estimated $\EstNI$ since $\Heritability \leq H^2$.
-Contrarily, the test for stabilising selection is invalid if $\EstNI$ is under-estimated.
+Contrarily, the test for stabilizing selection is invalid if $\EstNI$ is under-estimated.
 Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
@@
 We tested the performance of our neutrality index ($\NI$) to detect selection on a quantitative trait using simulations.
-We performed simulations under different selective regimes (neutral, stabilising, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of mutation rate (constant or fluctuating).
+We performed simulations under different selective regimes (neutral, stabilizing, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of mutation rate (constant or fluctuating).
 Simulation are individual-based and follows a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
@@
 The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
-Additionally, the instant value of log-$\Ne$ was modeled as a independent sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
+Additionally to the geometric Brownian process, an Ornstein-Uhlenbeck process is overlaid to the instant value of log-$\Ne$ to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
 The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
@@
 We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see methods), suggesting that traits for which this relationship is not verified are putatively under selection.
-Under stabilising selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
+Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
 In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
@@
 This statistic can be obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
-We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilising selection as $\proba [\EstNI < 1]$.
+We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
 
@@
     \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
-    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilising selection $\NI < 1$ are conservative or invalid due to violated assumptions.
+    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
     }
@@
 The inference framework was first tested on independently simulated datasets matching an empirically relevant mammalian empirical regime (see Materials and Methods).
-Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilising ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
-For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilising or diversifying selection, respectively.
-Once we introduced fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilising selection remained the same with all cases detected correctly.
-For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilising selection ($\proba [\EstNI < 1] > 0.975$).
+Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
+For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
+Once we introduced fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
+For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
 
@@
         $10,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
-        Traits simulated under stabilising selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
+        Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
         Histogram of estimated the ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$, as $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\EstNI$ is greater than $1$ (right).
@@
 $ \proba [ \EstNI > 1 ]$).
-We tested our statistical procedure against simulated data and showed that for simulations under diversifying selection (test of $\EstNI > 1$) or under stabilising selection (test of $\EstNI < 1$) our test is able to correctly detect those regimes of selection.
-However, for simulations of a neutral trait, our test detects frequently a spurious signal of stabilising selection ($\EstNI < 1$).
-Thus, we argue that our test should be used to detect diversifying selection, but has low accuracy to detect stabilising selection due to many false positives.
+We tested our statistical procedure against simulated data and showed that for simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$) our test is able to correctly detect those regimes of selection.
+However, for simulations of a neutral trait, our test detects frequently a spurious signal of stabilizing selection ($\EstNI < 1$).
+Thus, we argue that our test should be used to detect diversifying selection, but has low accuracy to detect stabilizing selection due to many false positives.
 
+% How is our test related to other tests of selection of traits
 Based on these results, we argue that our method is a significant improvement over current methods to detect selection acting on a trait at the phylogenetic scale.
-Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilising selection over a Brownian process when the trait is neutral~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Thus our method does not provide improvement to detect stabilising selection at the phylogenetic scale.
-However, to detect diversifying selection, current methods modelling mean trait value are limited by the theory behind~\parencite{harmon_phylogenetic_2018}
-Namely that under diversifying selection, mean trait value will not deviate from a Brownian process, and thus can not be distinguished from neutral evolution~\parencite{hansen_translating_1996}.
+Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Thus our method does not provide improvement to detect stabilizing selection at the phylogenetic scale.
+However, to detect diversifying selection, current methods modelling mean trait value are limited by a theoretical challenge.
+Namely, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus can not be distinguished from neutral evolution~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
 For example, test for neutrality concluded that the expression level of the majority of genes are under a neutral regime~\parencite{catalan_drift_2019}.
-Our diversity index allows to discriminate the alternative model of diversifying selection from the neutral case, inviting us to revisit these studies.
-Leveraging the ratio of between-species over within-species variations, methods have compared this ratio across a pool of traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-This strategy allows to identify outlier traits putatively under diversifying selection, but does not allow to test for selection on a single trait and requires independent confirmation~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
+Our diversity index allows to discriminate the alternative model of diversifying selection from the neutral case, inviting us to revisit such studies.
+On top of between-species variation of the mean trait value, methods have normalized by within-species variations to detect selection, and have compared this statistic across a pool of traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+This strategy allows to identify outlier traits putatively under diversifying selection, but does not allow to test for selection on a single trait and requires an independent confirmation~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
 Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution for a single test.
@@
 
+% Why is our test different
 We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize trait variation between and within species.
@@
 Thus our test avoids to estimate them and also avoids to estimate divergence time which is otherwise necessary~\parencite{walsh_evolution_2018}.
-But importantly, by normalising with sequence variation, we also show using simulated data that our test is not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time are constant across the phylogeny, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
+But importantly, by normalizing with sequence variation, we also show using simulated data that our test is not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time are constant across the phylogeny, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
 Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRate$ and generation time impacts similarly trait and sequence variation.
-Thus the normalisation by nucleotide divergence and polymorphism automatically absorbs long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time which cancels out in the ratio of trait variation $\EstNI$.
+Thus the normalization by nucleotide divergence and polymorphism automatically absorbs long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time which cancels out in the ratio of trait variation $\EstNI$.
 
-% How is our test related to other tests of selection
+% How is our test related to other tests of selection on sequences
 Even though our test is developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provide insight into its behavior.
@@
 Second, because $\NI$ is compared to 1, our test ultimately bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
-As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilising selection acting on the trait.
+As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
 Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
 With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilising selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilising selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
-By explicitly modeling stabilising selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilising selection in the context of quantitative traits to obtain a statistically more powerful test.
+First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
+By explicitly modeling stabilizing selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
 
 % Pitfalls of the method
-In the context of detecting diversifying selection, we argue that the main drawback of our method is that the additive genetic variance of the trait is required instead of the phenotypic variance.
+In the context of detecting diversifying selection on a trait, we argue that the main drawback of our method is that the additive genetic variance of the trait is required instead of the phenotypic variance.
 If phenotypic variance is used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ is ultimately under-estimated.
 Similarly, using broad-sense heritability instead of narrow-sense heritability results in under-estimated $\EstNI$.
-In such context, the test of stabilising selection ($\EstNI < 1]$) is statistically invalid.
+In such context, the test of stabilizing selection ($\EstNI < 1]$) is statistically invalid.
 However, the test of diversifying selection ($\EstNI > 1$) is underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
 As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
-However, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass is evolving under diversifying selection since the markers used for nucleotide divergence are not neutral, which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
 
@@
 Second, loci are assumed to contribute additively to the phenotype.
-Although the effects of dominance and epistasis is typically weak compared the additive effects on the phenotype, their influence shall be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
+Although the effects of dominance and epistasis is typically weak compared the additive effects on the quantitative traits, their influence shall be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
 Third, the genetic architecture of the trait is assumed to be constant across the phylogeny, it might actually be variable among individuals and species~\parencite{tung_genetic_2015, huber_conservatism_2015}.
@@
 Because our test is also based on several assumptions that might not hold on empirical data, we also provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
+For example, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass is evolving under diversifying selection since the markers used for nucleotide divergence are not neutral, which can lead to a spurious $\EstNI > 1$.
 Altogether, our study provides a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
@@
     \label{eq:DistribBrownian}
-    \contrast_{j} = \frac{\Brownian_{\mathcal{R}(j)} - \Brownian_{\mathcal{A}(j)} }{\sqrt {d_{j}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
+    \contrast_{j} = \frac{\Brownian (\mathcal{R}(j)) - \Brownian (\mathcal{A}(j)) }{\sqrt {d_{j}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
 \end{equation}
@@
 \end{equation}
+
 By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\Brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.

### 2023-10-02 manuscript/biorxiv-main.tex
Submitted version

--- 2023-09-22 manuscript/main.tex
@@
 \DeclareMathOperator{\Cov}{\text{Cov}}
-\newcommand{\QstFst}{Q$_\text{ST}$--F$_\text{ST}$}
+\newcommand{\Qst}{Q$_\text{ST}$}
+\newcommand{\Fst}{F$_\text{ST}$}
+\newcommand{\QstFst}{\Qst--\Fst}
 \newcommand{\der}{\mathrm{d}}
@@
 
-% Including genomics, Inclusion of genomics, Linking genomics,
-\title{Leveraging genomics to detect trait selection from within and between-species variations}
-\rhead{\scshape Trait selection from within and between-species variations}
+\newsavebox{\Maketitle}
+\title{Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes}
+\rhead{\scshape Trait selection from within and between-species variation}
 
@@
 \large
-\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M. {Bastian}$^{2}$, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
+\textbf{T.
+{Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M.
+{Bastian}$^{2}$\orcidlink{0009-0003-9129-6918}, T.
+{Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N.
+{Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
 \normalsize
@@
 }
-\author{~}
+
 \begin{document}
 
+\sbox{\Maketitle}{%
+\begin{minipage}{\textwidth}
 \maketitle
+\end{minipage}}
+\noindent\usebox{\Maketitle}
 
@@
 \begin{abstract}
-    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between-species variation.
+    To quantify selection acting on a trait, methods have been developed using either within or between-species variation.
     However, methods using within-species variation do not integrate the changes at the macro-evolutionary scale.
@@
 
-Determining whether a trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
+Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
 Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
@@
 
-Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
-Additionally, contrasting both trait and genetic differentiation across several populations, \QstFst methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
-For example in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
-[why only example from gene expression?]
-However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
+Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\parencite{simons_population_2018, sella_thinking_2019}.
+Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{merila_comparison_2001, leinonen_comparative_2008}.
+A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation in different optimum trait value in the different populations~\parencite{lamy_qst_2012}.
+Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection.
+However, \QstFst\ methods have been found to require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
+Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.
+To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
+For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
+% #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
+Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\parencite{khaitovich_neutral_2004}.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
+% For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
+% However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
+% are these studies using Gst/Fst on gene expression?
+Otherwise, we should move it elsewhere or add a clear transition between the sentences
 
@@
 For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
-Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for gene with large within-species variation~\parencite{khaitovich_neutral_2004}.
+% #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
+Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for genes with large within-species variation~\parencite{khaitovich_neutral_2004}.
 Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
@@
 Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
+Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
 As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
 Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
-Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
-Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
+Deviations from the assumptions of the Brownian process are however well known.
+When trait variation is constraint because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
+Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
 However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
@@
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
-Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.
@@
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value to describe technical errors~\parencite{lynch_methods_1991, felsenstein_comparative_2008, hansen_interpreting_2012} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average as a signal of diversifying selection~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
-We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
-Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
-For population geneticists, our study can be seen as the macro-evolutionary generalization of \QstFst methods to account for phylogenetic relationships between species.
-For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
-% Basically, we are testing whether the variance of means is greater than the mean of variances
-
+At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
+Within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
+We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of molecular divergence at both scales.
+It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
+The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
+Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
+From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species.
+From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
 % Basically, we are testing whether the variance of means is greater than the mean of variances
@@
 We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
-This ratio allows to remove the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
+This ratio allows removing the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
 \begin{align}
@@
 If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
-The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given by \parencite{hansen_translating_1996}:
+The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given \parencite{hansen_translating_1996} by:
 \begin{align}
@@
 \end{align}
-That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate with which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.
+That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.
 
@@
 We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
-This ratio allows to remove the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
+This ratio allows removing the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
 \begin{align}
@@
     \caption{
-        Between species, the mean phenotypic trait value changes along the phylogeny, allowing estimating of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
-        Within-species, the genetic variance allows estimating of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
+        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
+        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
         $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
@@
 At the phylogenetic level, the variability of the mean trait value between species can be combined with the nucleotide divergence of any neutrally evolving genomic region to obtain $\RateBetween$.
-Similarly, $\RateBetween=\RateMut$ if the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogeny.
+Similarly, $\RateBetween=\RateMut$ if the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogenetic tree.
 We thus have, for a neutrally evolving trait:
@@
 
-Based on the comparative framework accounting for phylogenetic inertia~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a Maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
+Based on the comparative framework that can account for phylogenetic inertia~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
 
 \subsubsection*{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitution per site) and the topology of the phylogenetic tree.
+At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitutions per site) and the topology of the phylogenetic tree.
 The diagonal $\Distance_{\Spi,\Spi}$ represents the total distances from the root of the tree to each taxon ($\Spi$).
@@
 
-Finally, within-species variation $\EstRateBetween$ is estimated as~\parencite{omeara_testing_2006}:
+Finally, between-species variation $\EstRateBetween$ is estimated as~\parencite{omeara_testing_2006}:
 \begin{gather}
@@
 
-The Bayesian framework allows to obtain the posterior distribution the neutrality index ($\EstNI$) for a given trait.
-Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
-Trait variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
+The Bayesian framework allows obtaining the posterior distribution of the neutrality index ($\EstNI$) for a given trait.
+Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
+Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
@@
 With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
-$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is Gibbs sampled.
-For each trait and each species, prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
-Heritability and phenotypic variance for each trait is combined with nucleotide diversity to compute $\EstRateWhithin$ for each species, and is averaged out across species (as in eq.~\ref{eq:estimated-rate-pop}).
+We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
+For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
+Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\EstRateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
 From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
+The posterior distribution of $\EstNI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
 
@@
 Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
-It is however possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
-Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in under-estimated $\EstNI$ since $\Heritability \leq H^2$.
-Contrarily, the test for stabilizing selection is invalid if $\EstNI$ is under-estimated.
+It is, however, possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
+Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\EstNI$ since $\Heritability \leq H^2$.
+In contrast, the test for stabilizing selection is invalid if $\EstNI$ is underestimated.
 Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
@@
 We tested the performance of our neutrality index ($\NI$) to detect selection on a quantitative trait using simulations.
-We performed simulations under different selective regimes (neutral, stabilizing, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of mutation rate (constant or fluctuating).
-Simulation are individual-based and follows a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
-Each individual phenotypic value is the sum of genotypic value and an environmental effect.
+We performed simulations under different selective regimes (neutral, stabilizing, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of the mutation rate (constant or fluctuating).
+Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
+Each individual phenotypic value was the sum of genotypic value and an environmental effect.
 The environmental effect was normally distributed with variance $\VarEnv$.
@@
 Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
-For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
-From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
-We scaled parameters to fit plausible values for mammals.
-We used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
+For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with the mutation rate per generation $\MutationRate$.
+From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from the root to leaves of $0.18$ and average genetic diversity of $0.00276$.
+We scaled parameters in our simulations to fit plausible values for mammals.
+We thus used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
 
-The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
-Additionally to the geometric Brownian process, an Ornstein-Uhlenbeck process is overlaid to the instant value of log-$\Ne$ to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
+The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which led to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
+An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
 The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
 The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
-From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.
+From a speciation process perspective, this was equivalent to an allopatric speciation over one generation.
 
-Random genetic drift was introduced by re-sampling individuals at each generation, with each parent having a probability of being sampled that was proportional to its fitness ($W$).
-Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+A random genetic drift was introduced by resampling individuals at each generation, with each parent having a probability of being sampled that was proportional to its fitness ($W$).
+Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
 More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
@@
 
-Nucleotide diversity ($\pi$) was measured as the heterozygosity of neutral markers (not linked to the trait).
-Nucleotide divergence ($d$) was measured as the number of substitutions per site of neutral markers (not linked to the trait) along branches of the phylogenetic tree.
+Nucleotide diversity ($\pi$) was measured as the heterozygosity of neutral markers that were simulated along the phylogenetic tree but not linked to the trait simulated.
+Nucleotide divergence ($d$) was measured as the number of substitutions per site of neutral markers along the branches of the phylogenetic tree.
 The additive genetic variance was measured as phenotypic variance multiplied by heritability.
@@
         The trait's genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
-        Parents are selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
+        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
         Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRate$ to mutate.
         Drift is modeled by the resampling of parents.
-        Right panel: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
+        Right panel: examples of a trait evolving along a phylogenetic tree, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
     }
@@
 
-We analyzed a dataset of body and brain masses in mammals.
+We analyzed a dataset of body and brain masses from mammals.
 The log-transformed values of body and brain masses were taken from \textcite{tsuboi_breakdown_2018}.
@@
 For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
-We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see methods), suggesting that traits for which this relationship is not verified are putatively under selection.
+We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see Materials and Methods), suggesting that traits for which this relationship was not verified were putatively under selection.
 Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
 In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
-Our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between-species (see fig.~\ref{fig:methods}).
-This statistic can be obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
-We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
+Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
+This statistic was obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
+We also devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
 
-\begin{table*}[t!]
-    \centering
-    \begin{adjustbox}{width = 1.0\textwidth}
-        \begin{tabular}{||l|l||c|c||c|c||}
-            \hline
-            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
-            Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
-            Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
-            Markers for polymorphism under balanced selection & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
-            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
-            Multiple nucleotide substitutions at the same locus & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\\hline \hline
-        \end{tabular}
-    \end{adjustbox}
-    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
-    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
-    }
-    \label{table:assumptions}
-\end{table*}
+Our neutrality index made a series of assumptions that we described in details in the Material and Methods section.
+Table \ref{table:assumptions} summarized these assumptions and outlined possible consequences for the neutrality test that we proposed.
 
@@
 For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
-Once we introduced fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
-For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
+Once we introduced fluctuating $\Ne$ and $\MutationRate$ (Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
+For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) and none as diversifying selection.
 
@@
     \caption{
-        $10,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
+        $10,000$ simulations of trait evolution along a phylogenetic tree under different selection regimes.
         Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-        Histogram of estimated the ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$, as $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\EstNI$ is greater than $1$ (right).
-        Effective population size ($\Ne$) and mutation rate ($\MutationRate$) are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
+        Histogram of ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$ with $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and probabilities of $\EstNI$ being greater than $1$ (right).
+        Effective population size ($\Ne$) and mutation rate ($\MutationRate$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
     }
@@
     Test of diversifying selection on a mammal and a primate dataset, by splitting males (\Male) and females (\Female).
-    Traits considered are body mass or brain mass (log-transformed).
-    Heritability ($\Heritability$) is either assumed complete ($\Heritability=1.0$) or uniformly distributed between 20\% and 40\%  ($\Heritability \sim \mathcal{U}(0.2, 0.4)$).
-    $\NbrTaxa$ is the number of species in the dataset.
-    $\EstNI$ is the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\EstNI$ also computed.
-    $\proba [\EstNI > 1 ]$ is the estimated posterior probability of diversifying selection.
+    Traits considered were body mass or brain mass (log-transformed).
+    Heritability ($\Heritability$) was either assumed complete ($\Heritability=1.0$) or uniformly distributed between 20\% and 40\%  ($\Heritability \sim \mathcal{U}(0.2, 0.4)$).
+    $\NbrTaxa$ was the number of species in the dataset.
+    $\EstNI$ was the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\EstNI$ also computed.
+    $\proba [\EstNI > 1 ]$ was the estimated posterior probability of diversifying selection.
     }
@@
 
-For mammalian body and brain mass, we obtained male (\Male) and female (\Female) trait variation.
-Combined with nucleotide diversity and divergence, we estimated $\EstNI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in table~\ref{table:empirical}.
+For mammalian body and brain mass, we obtained male (\Male) and female (\Female) trait variations.
+Combined with nucleotide diversity and divergence, we estimated $\EstNI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in the Table~\ref{table:empirical}.
 Assuming complete heritability, brain mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
@@
 
-We also analyzed a similar dataset for body mass focussing this time only at Primates (see table~\ref{table:empirical}).
-For primate body mass, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%,
+We also analyzed a similar dataset for body mass focusing this time only at Primates (Table~\ref{table:empirical}).
+For primates body mass, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%.
 Assuming complete heritability of body mass did not change the posterior probability for males, but increased the one for female to $1.0$.
-Evidence for diversifying selection on body mass was therefore more pronounced in Primates than in Mammals.
-However, the genetic markers used to normalize trait variance with divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\EstNI$ (see table~\ref{table:assumptions} and methods).
+Evidence for diversifying selection on body mass was therefore more pronounced in Primates than in mammals.
+However, the genetic markers used to normalize trait variance with nucleotide divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\EstNI$ (Table~\ref{table:assumptions} and methods).
+
+\begin{table*}[t!]
+    \centering
+    \begin{adjustbox}{width = 1.0\textwidth}
+        \begin{tabular}{||l|l||c|c||c|c||}
+            \hline
+            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
+            Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
+            Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
+            Markers for polymorphism under balanced selection & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
+            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
+            Multiple nucleotide substitutions at the same locus & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\\hline \hline
+        \end{tabular}
+    \end{adjustbox}
+    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
+    The last two columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
+    }
+    \label{table:assumptions}
+\end{table*}
 
@@
 In this study, we proposed a neutrality index for a quantitative trait that can be used within a statistical framework to test for selection.
-Our neutrality index for a trait, $\NI$, is the ratio of the normalized between- to within-species variation and it allows the identification of the evolutionary regime of a quantitative trait.
-At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
-Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
-Estimated $\EstNI$ can be compared against the null hypothesis of neutrality by testing for deviation from the value of $1.0$.
-Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S\ref{sec:implementation}) also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
-$ \proba [ \EstNI > 1 ]$).
-We tested our statistical procedure against simulated data and showed that for simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$) our test is able to correctly detect those regimes of selection.
-However, for simulations of a neutral trait, our test detects frequently a spurious signal of stabilizing selection ($\EstNI < 1$).
-Thus, we argue that our test should be used to detect diversifying selection, but has low accuracy to detect stabilizing selection due to many false positives.
+Our neutrality index for a trait, $\NI$, is calculated as the ratio of the normalized within- to between-species variation and it allowed the identification of the evolutionary regime of a quantitative trait.
+At the phylogenetic scale, trait variation between species was normalized by sequence divergence obtained from a neutral set of markers.
+Similarly, trait variation within species was normalized by sequence polymorphism obtained also from a neutral set of markers.
+Our estimate of $\EstNI$ could be tested for deviation from the value of $1.0$ expected under the null hypothesis of neutrality.
+Technically, the neutrality index can be estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S\ref{sec:implementation}).
+The latter also enabled the estimation of the posterior credible interval to test for departure from a neutrally evolving trait (e.g. $ \proba [ \EstNI > 1 ]$).
+We tested our statistical procedure against simulated data and showed that our test was able to correctly detect simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$).
+However, our test detected a spurious signal of stabilizing selection ($\EstNI < 1$) when we simulated the evolution of a neutral trait.
+We thus argue that our method should be used to detect diversifying selection, but that it had low accuracy to detect stabilizing selection due to false positives.
 
 % How is our test related to other tests of selection of traits
-Based on these results, we argue that our method is a significant improvement over current methods to detect selection acting on a trait at the phylogenetic scale.
+Our results showed that our method significantly improved over currently available methods to detect selection acting on a trait at the phylogenetic scale.
 Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Thus our method does not provide improvement to detect stabilizing selection at the phylogenetic scale.
-However, to detect diversifying selection, current methods modelling mean trait value are limited by a theoretical challenge.
-Namely, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus can not be distinguished from neutral evolution~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-For example, test for neutrality concluded that the expression level of the majority of genes are under a neutral regime~\parencite{catalan_drift_2019}.
-Our diversity index allows to discriminate the alternative model of diversifying selection from the neutral case, inviting us to revisit such studies.
-On top of between-species variation of the mean trait value, methods have normalized by within-species variations to detect selection, and have compared this statistic across a pool of traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-This strategy allows to identify outlier traits putatively under diversifying selection, but does not allow to test for selection on a single trait and requires an independent confirmation~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
-Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution for a single test.
-Again, our diversity index can thus allow us to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence and polymorphism.
+Our approach could in theory be applied to detect stabilizing selection at the phylogenetic scale, but we showed that it did not have the statistical power to identify those cases.
+In contrast, we showed that our method was able to identify correctly cases of diversifying selection, which is a clear an improvement over current methods that model only mean trait value.
+Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\parencite{catalan_drift_2019}.
+Our diversity index has the advantage to discriminate the alternative model of diversifying selection from the neutral case by comparing within- and between-species variation correctly normalized to remove confounding factors.
+Our approach is not the first one to normalize between-species variation to detect selection, but this was done by using within-species variations~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} and not estimates of neutral molecular divergence as done in our study.
+These studies have further compared their statistic across a pool of traits, which allowed them to identify outlier traits putatively under diversifying selection but without testing for selection on a single trait at a time~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
+Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departures from the null model of neutral evolution for a single test.
+Our diversity index opens new avenues to revisit these studies and better test for the selective regime affecting the quantitative traits, assuming we have access to genomic datasets to estimate nucleotide divergence and polymorphism.
 
 % Why is our test different
-We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize trait variation between and within species.
-In the context of within species variation but across several populations, \QstFst test developed from trait and sequence allow to test for selection of a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
-Similarly, in our test, the genetic sequences from which nucleotide divergence and polymorphism are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
-Nucleotide variation allow to normalize for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRate$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-Thus our test avoids to estimate them and also avoids to estimate divergence time which is otherwise necessary~\parencite{walsh_evolution_2018}.
-But importantly, by normalizing with sequence variation, we also show using simulated data that our test is not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time are constant across the phylogeny, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
-Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRate$ and generation time impacts similarly trait and sequence variation.
-Thus the normalization by nucleotide divergence and polymorphism automatically absorbs long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time which cancels out in the ratio of trait variation $\EstNI$.
+The main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
+In the context of within species variation, \QstFst\ tests have been developed to compare trait and sequence across several populations to test for selection~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
+Our neutrality index also used the genetic sequences from which nucleotide divergence and polymorphism are estimated.
+Although the sequences should be neutrally evolving, they do not have to be necessarily linked to the quantitative trait under study.
+Nucleotide variation allows normalizing for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRate$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Thus our test avoids the estimation of the parameters, which are complex to correctly infer, and it also bypasses the estimation of divergence time, which was necessary in previous approaches~\parencite{walsh_evolution_2018}.
+But importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time were constant across the phylogenetic tree, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
+Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRate$ and generation time impacted similarly trait and sequence variation.
+The normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time, which canceled out in the ratio of trait variation $\EstNI$.
 
 % How is our test related to other tests of selection on sequences
-Even though our test is developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provide insight into its behavior.
-First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations are used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
-Second, because $\NI$ is compared to 1, our test ultimately bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
+Even though our test was developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provided insight into its behavior.
+First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations were used to determine the neutral expectation, and the inflation of divergence was compared to polymorphism within species.
+Second, because $\NI$ was compared to 1, our test ultimately bear analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
 As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
 Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
-By explicitly modeling stabilizing selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
+With this analogy in mind, we could leverage the vast literature discussing and interpreting the results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
+First, not rejecting the neutral null model of $\NI = 1$ did not necessarily imply that the trait was effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ did not rule out diversifying selection, but rather that this diversifying selection was not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
+By explicitly modeling stabilizing selection as a moving optimum, it would theoretically be possible to tease apart the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
 
@@
 In the context of detecting diversifying selection on a trait, we argue that the main drawback of our method is that the additive genetic variance of the trait is required instead of the phenotypic variance.
-If phenotypic variance is used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ is ultimately under-estimated.
-Similarly, using broad-sense heritability instead of narrow-sense heritability results in under-estimated $\EstNI$.
-In such context, the test of stabilizing selection ($\EstNI < 1]$) is statistically invalid.
-However, the test of diversifying selection ($\EstNI > 1$) is underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
+If phenotypic variance was used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ was ultimately underestimated.
+Similarly, using broad-sense heritability instead of narrow-sense heritability would result in underestimated $\EstNI$.
+In such context, the test of stabilizing selection ($\EstNI < 1]$) would be statistically invalid.
+However, the test of diversifying selection ($\EstNI > 1$) was underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
 As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
@@
 % What can and should be improved
-The development of our neutrality index is also based on several assumptions that could be relaxed in future studies.
-First, we cannot predict the behaviour of our test in the context of population structures, gene flow and introgression.
+The development of our neutrality index was also based on several assumptions that could be relaxed in future studies.
+First, we cannot predict the behavior of our test in the context of population structures, gene flow and introgression.
 These factors should be thoroughly investigated using simulations.
-Second, loci are assumed to contribute additively to the phenotype.
-Although the effects of dominance and epistasis is typically weak compared the additive effects on the quantitative traits, their influence shall be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
-Third, the genetic architecture of the trait is assumed to be constant across the phylogeny, it might actually be variable among individuals and species~\parencite{tung_genetic_2015, huber_conservatism_2015}.
-Such assumption can theoretically be relaxed and changes in genetic architecture along the phylogeny could jointly be estimated~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
+Second, loci were assumed to contribute additively to the phenotype.
+Although the effects of dominance and epistasis is typically weak compared to the additive effects on the quantitative traits, their influence should be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
+Third, the genetic architecture of the trait was assumed to be constant across the phylogenetic tree, whereas it might actually be variable among individuals and species~\parencite{tung_genetic_2015, huber_conservatism_2015}.
+Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Finally, our Bayesian estimation could integrate uncertainty from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
 
 % What can our test bring to the field
-From an empirical point of view, even though our method requires integrating genomic and trait variation, we also showed that our method can be applied to the empirical dataset with the illustrative example of mammals' brain and body mass.
-Because our test is also based on several assumptions that might not hold on empirical data, we also provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
-For example, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass is evolving under diversifying selection since the markers used for nucleotide divergence are not neutral, which can lead to a spurious $\EstNI > 1$.
-Altogether, our study provides a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
+From an empirical point of view, our method required integrating genomic and trait variation, which could reduce the possible datasets to be used.
+However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals brain and body mass.
+Because our test was also based on several assumptions that might not hold on empirical data, we also provided a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (Table~\ref{table:assumptions}).
+For example, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass was evolving under diversifying selection since the markers used for nucleotide divergences were not neutral, which can lead to a spurious $\EstNI > 1$.
+In conclusion, our study provided a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we believe that our new approach is a promising tool to investigate the evolution of quantitative traits.
+
+\subsection*{Acknowledgements}
+ We gratefully acknowledge the help of Nicolas Lartillot, Philippe Veber, Isabela Jeronimo do Ó, Anna Marcionetti, Julien Clavel, and Daniele Silvestro for their insightful discussions and Julien Joseph for his advice and reviews concerning this manuscript.
+
+\subsection*{Competing interests:}
+The authors declare no conflicts of interest.
+
+\subsection*{Data and materials availability:}
+The data that support the findings of this study are openly available in GitHub at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}.
+Snakemake pipeline, analysis scripts and documentation are available in the repository to replicate the study.
 
@@
 \newpage
+
+\noindent\usebox{\Maketitle}
 
@@
 Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
-Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
+Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
@@
 
-Implementation is included within the \textit{BayesCode} software, available at \censor{\url{https://github.com/ThibaultLatrille/bayescode}}.
+Implementation is included within the \textit{BayesCode} software, available at \url{https://github.com/ThibaultLatrille/bayescode}.
 
@@
 \begin{enumerate}
-    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers)
+    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers).
     \item A file containing the mean trait values for each species.
@@
 
-The phylogenetic tree must be in newick format, with branch lengths in number of substitutions per site (neutral markers).
+The phylogenetic tree must be in newick format, with branch lengths in substitutions per site (neutral markers).
 
@@
 \subsubsection{Reading the results}\label{subsubsec:reading-the-results}
-Once the model has ran, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between-species variation over within-species variation with \texttt{readnodetraits}:
+Once the model has run, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between-species variation over within-species variation with \texttt{readnodetraits}:
 \begin{lstlisting}[language = sh,label={lst:readnodetraits-rho}]

### 2023-12-06 data_simulated/RevBayesMerge.py
Compute proba OU

@@ -1,0 +1,66 @@
+#!python3
+import os
+from glob import glob
+import numpy as np
+import pandas as pd
+
+
+def minipage(size, file, name=""):
+    out = "\\begin{minipage}{" + str(size) + "\\linewidth} \n"
+    out += "\\centering " + name + "\n"
+    out += "\\includegraphics[width=\\linewidth, page=1]{" + file + "} \n"
+    out += "\\end{minipage}\n"
+    return out
+
+
+def write(exp, o):
+    o.write("\\begin{center}\n")
+    for method in ["neutral", "moving_optimum", "stabilizing"]:
+        distance_path = f"{exp}/results/inference_ML.distance.{method}.pdf"
+        assert os.path.exists(distance_path), f"Missing {distance_path}"
+        o.write(minipage(0.32, distance_path, method.replace("_", " ").title()))
+
+    for param in ["is_OU", "t_half", "sigma2"]:
+        hist_path = f"{exp}/results/inference_RevBayes.{param}.pdf"
+        assert os.path.exists(hist_path), f"Missing {hist_path}"
+        o.write(minipage(0.32, hist_path, param.replace("_", " ").title()))
+    o.write("\\end{center}\n")
+
+
+def alpha(exp):
+    tsv_path = f"{exp}/results/inference_RevBayes.tsv.gz"
+    if not os.path.exists(tsv_path):
+        print(f"Missing {tsv_path}")
+        return
+    tsv = pd.read_csv(tsv_path, sep="\t")
+    tsv["t_half"] = np.log(2) / tsv["alpha"]
+    groups = tsv.groupby("model")
+    for model, group in groups:
+        print(group["is_BM"])
+        for col in ["t_half", "is_BM"]:
+            print(f"\t{model}\t{col}\t{group[col].mean()}")
+
+
+def main():
+    tex_include = "docs/include-figures.tex"
+    tex_target = "docs/supp-mat.tex"
+    experiments = glob("experiments/*_u")
+    o = open(tex_include, 'w')
+    o.write("\\section{Experiments} \n")
+    for exp in experiments:
+        subsection = os.path.basename(exp).replace("_", " ").title()
+        o.write("\\subsection{" + subsection + "}\n")
+        write(exp, o)
+    o.close()
+    output_dir = os.path.dirname(tex_target)
[…]

### 2023-12-06 data_simulated/run.sh

@@ -1,4 +1,4 @@
 CPU=8
-for EXPERIMENT in ./*.yaml; do
+for EXPERIMENT in ./config/constant_pop_size_many_sites*.yaml; do
 python3 simulated_experiment.py -c $(basename "${EXPERIMENT}") -j ${CPU}
 done

### 2023-12-06 data_simulated/update_snake.sh

@@ -1,9 +1,11 @@
+# shellcheck disable=SC2164
+# shellcheck disable=SC2046
 FOLDER=$(pwd)
-for EXPERIMENT in $(pwd)/experiments/fluc*ing; do
+for EXPERIMENT in $(pwd)/experiments/*_u; do
   echo "Updating snakemake in: ${EXPERIMENT}"
   rm -rf "${EXPERIMENT}"/Snakefile
   ln $(pwd)/Snakefile "${EXPERIMENT}"/Snakefile
   cd "${EXPERIMENT}"
-  snakemake -j 8 --touch
+  snakemake -j 8 -k
   cd "${FOLDER}"
 done

### 2023-12-06 scripts/libraries.py

@@ -72,7 +72,10 @@ def scatter_plot(x_input, y_input, x_label, y_label, output, histy_log=False, lo
 
 def hist_plot(x_input, x_label, output, xscale="log"):
     x = {k: np.array(v) for k, v in x_input.items()}
-    xy_filtered = {k: (np.isfinite(x[k]) & (x[k] >= 0)) for k in x.keys()}
+    if xscale == "log":
+        xy_filtered = {k: (np.isfinite(x[k]) & (x[k] > 0)) for k in x.keys()}
+    else:
+        xy_filtered = {k: np.isfinite(x[k]) for k in x.keys()}
     x = {k: x[k][xy_filtered[k]] for k in x.keys()}
     output = output.replace("scatter_plot", "histogram")
     color_models = colors(x)
@@ -96,13 +99,16 @@ def hist_plot(x_input, x_label, output, xscale="log"):
                     label=f'{m.replace("_", " ").capitalize()} (mean {x_mean:.2g})')
     else:
         for id_m, m in enumerate(x):
-            ax.plot((0, 0), (0, 0), linewidth=3, color=color_models[id_m],
+            ax.plot((min_x, min_x), (0, 0), linewidth=3, color=color_models[id_m],
                     label=f'{m.replace("_", " ").capitalize()}')
 
     ax.set_xlabel(x_label, fontsize=fontsize)
     if xscale == "log":
         ax.set_xlim((0.95 * min_x, 1.05 * max_x))
         ax.set_xscale("log")
+    elif xscale == "linear":
+        ax.set_xlim((min_x, max_x))
+        ax.set_xscale("linear")
     else:
         ax.set_xlim((-0.01, 1.01))
         ax.set_yscale("log")

### 2023-12-06 scripts/plot_simulations_ML.py

@@ -9,17 +9,21 @@ from sklearn.metrics import r2_score
 
 
 def compute_distance(tree: Tree, trait: str, dico_nuc_d: dict, dico_pheno_d: dict) -> (float, float):
-    leaves = tree.get_leaves()
-    n = len(leaves)
-    x = [float(getattr(leaf, trait)) for i, leaf in enumerate(leaves)]
-
-    for i, j in itertools.product(range(n), range(n)):
-        if i >= j:
+    for node in tree.traverse("postorder"):
+        if node.is_leaf():
             continue
-        leaf_i, leaf_j = leaves[i].name, leaves[j].name
-        d = tree.get_distance(leaves[i], leaves[j]) * 4
-        dico_nuc_d[(leaf_i, leaf_j)].append(d)
-        dico_pheno_d[(leaf_i, leaf_j)].append(np.var([x[i], x[j]]))
+        leaves = node.get_leaves()
+        x = [float(getattr(leaf, trait)) for i, leaf in enumerate(leaves)]
+        n = len(leaves)
+        list_nuc, list_pheno = [], []
+        for i, j in itertools.product(range(n), range(n)):
+            if i >= j:
+                continue
+            d = tree.get_distance(leaves[i], leaves[j]) * 4
+            list_nuc.append(d)
+            list_pheno.append(np.var([x[i], x[j]]))
+        dico_nuc_d[node.name].append(np.mean(list_nuc))
+        dico_pheno_d[node.name].append(np.mean(list_pheno))
 
 
 def saturation_fit(x, a, b):
@@ -55,12 +59,18 @@ def distance_plot(x_dico, y_dico, output):
     polyfit(x_array, y_array, saturation_fit, "saturation", ax, "blue")
     polyfit(x_array, y_array, linear_fit, "linear", ax, "red")
     ax.scatter(x_array, y_array, s=12, color="black")
-    #for k in x_mean.keys():
-    #    ax.errorbar(x_mean[k], y_mean[k], color="black", alpha=0.05, linewidth=0.5,
-    #                yerr=[[y_mean[k] - y_lower[k]], [y_upper[k] - y_mean[k]]])
-    ax.set_xlabel('Nucleotide distance', weight='bold')
-    ax.set_ylabel('Phenotypic distance', weight='bold')
+    for k in x_mean.keys():
+        ax.errorbar(x_mean[k], y_mean[k], color="black", alpha=0.05, linewidth=0.5,
[…]

### 2023-12-06 scripts/plot_simulations_RevBayes.py

@@ -26,10 +26,17 @@ def main(folder, output):
             filepath = f"{folderpath}/simple_OU_RJ.log"
             df = pd.read_csv(filepath, sep="\t")
             var_dict["model"][m].append(m)
-            var_dict["is_OU"][m].append(np.mean(df["is_OU"]))
+            for col in ["alpha", "is_BM", "is_OU", "sigma2", "theta"]:
+                var_dict[col][m].append(np.mean(df[col][len(df) // 2:]))
+            var_dict["t_half"][m].append(np.log(2) / np.mean(df["alpha"][len(df) // 2:]))
 
     rename = lambda x: output.replace(".tsv.gz", x)
-    hist_plot(var_dict["is_OU"], "p[OU]", rename(f".is_OU.pdf"))
+    hist_plot(var_dict["is_OU"], "p[OU]", rename(f".is_OU.pdf"), xscale="uniform")
+    hist_plot(var_dict["is_BM"], "p[BM]", rename(f".is_BM.pdf"), xscale="uniform")
+    hist_plot(var_dict["t_half"], "t 1/2", rename(f".t_half.pdf"), xscale="log")
+    hist_plot(var_dict["alpha"], "alpha", rename(f".alpha.pdf"), xscale="log")
+    hist_plot(var_dict["sigma2"], "sigma2", rename(f".sigma2.pdf"), xscale="log")
+    hist_plot(var_dict["theta"], "theta", rename(f".theta.pdf"), xscale="linear")
 
     out_dict = defaultdict(list)
     for m in models:

### 2024-03-04 data_simulated/RevBayesMerge.py
Output alpha

@@ -1,8 +1,6 @@
 #!python3
 import os
 from glob import glob
-import numpy as np
-import pandas as pd
 
 
 def minipage(size, file, name=""):
@@ -15,30 +13,26 @@ def minipage(size, file, name=""):
 
 def write(exp, o):
     o.write("\\begin{center}\n")
+
     for method in ["neutral", "moving_optimum", "stabilizing"]:
         distance_path = f"{exp}/results/inference_ML.distance.{method}.pdf"
-        assert os.path.exists(distance_path), f"Missing {distance_path}"
+        if not os.path.exists(distance_path):
+            continue
         o.write(minipage(0.32, distance_path, method.replace("_", " ").title()))
 
     for param in ["is_OU", "t_half", "sigma2"]:
-        hist_path = f"{exp}/results/inference_RevBayes.{param}.pdf"
-        assert os.path.exists(hist_path), f"Missing {hist_path}"
+        hist_path = f"{exp}/results/inference_RevBayes.simple_OU.{param}.pdf"
+        if not os.path.exists(hist_path):
+            continue
         o.write(minipage(0.32, hist_path, param.replace("_", " ").title()))
+
+    for param in ["num_theta_changes", "t_half", "sigma2"]:
+        hist_path = f"{exp}/results/inference_RevBayes.relaxed_OU.{param}.pdf"
+        if not os.path.exists(hist_path):
+            continue
+        o.write(minipage(0.32, hist_path, param.replace("_", " ").title()))
+
     o.write("\\end{center}\n")
-
-
-def alpha(exp):
-    tsv_path = f"{exp}/results/inference_RevBayes.tsv.gz"
-    if not os.path.exists(tsv_path):
-        print(f"Missing {tsv_path}")
-        return
-    tsv = pd.read_csv(tsv_path, sep="\t")
-    tsv["t_half"] = np.log(2) / tsv["alpha"]
-    groups = tsv.groupby("model")
-    for model, group in groups:
-        print(group["is_BM"])
-        for col in ["t_half", "is_BM"]:
-            print(f"\t{model}\t{col}\t{group[col].mean()}")
 
 
 def main():
@@ -56,10 +50,6 @@ def main():
     tex_to_pdf = f"pdflatex -synctex=1 -interaction=nonstopmode -output-directory={output_dir} {tex_target}"
     os.system(tex_to_pdf)
     os.system(tex_to_pdf)
-
-    for exp in experiments:
-        print(exp)
-        alpha(exp)
 
 
 if __name__ == '__main__':

### 2024-03-04 scripts/libraries.py

@@ -76,7 +76,9 @@ def hist_plot(x_input, x_label, output, xscale="log"):
         xy_filtered = {k: (np.isfinite(x[k]) & (x[k] > 0)) for k in x.keys()}
     else:
         xy_filtered = {k: np.isfinite(x[k]) for k in x.keys()}
-    x = {k: x[k][xy_filtered[k]] for k in x.keys()}
+    x = {k: x[k][xy_filtered[k]] for k in x.keys() if len(x[k][xy_filtered[k]]) > 0}
+    if len(x) == 0:
+        return
     output = output.replace("scatter_plot", "histogram")
     color_models = colors(x)
     fig = plt.figure(figsize=(1280 / my_dpi, 640 / my_dpi), dpi=my_dpi)

### 2024-03-04 scripts/plot_simulations_RevBayes.py

@@ -10,6 +10,16 @@ from os.path import basename, isdir
 from libraries import hist_plot, scatter_plot
 
 
+def open_log(path):
+    df = pd.read_csv(path, sep="\t")
+    df["t_half"] = np.log(2) / df["alpha"]
+    return df
+
+
+def posterior(df, col="alpha", burnin=0.5):
+    return np.nanmean(df[col][int(len(df) * burnin):])
+
+
 def main(folder, output):
     os.makedirs(os.path.dirname(output), exist_ok=True)
 
@@ -20,29 +30,43 @@ def main(folder, output):
     replicates = {m: natsorted([i for i in glob(f"{p}/*") if isdir(i)]) for m, p in models_path.items()}
     assert len(set([len(g) for g in replicates.values()])) == 1
 
-    var_dict = defaultdict(lambda: defaultdict(list))
+    simple_OU_dict = defaultdict(lambda: defaultdict(list))
+    relaxed_OU_dict = defaultdict(lambda: defaultdict(list))
     for m in models:
         for f, folderpath in enumerate(replicates[m]):
-            filepath = f"{folderpath}/simple_OU_RJ.log"
-            df = pd.read_csv(filepath, sep="\t")
-            var_dict["model"][m].append(m)
-            for col in ["alpha", "is_BM", "is_OU", "sigma2", "theta"]:
-                var_dict[col][m].append(np.mean(df[col][len(df) // 2:]))
-            var_dict["t_half"][m].append(np.log(2) / np.mean(df["alpha"][len(df) // 2:]))
+            simple_df = open_log(f"{folderpath}/simple_OU_RJ.log.gz")
+            simple_OU_dict["model"][m].append(m)
+            for col in ["alpha", "is_BM", "is_OU", "sigma2", "theta", "t_half"]:
+                simple_OU_dict[col][m].append(posterior(simple_df, col=col, burnin=0.5))
+
+            relaxed_df = open_log(f"{folderpath}/relaxed_OU_RJ.log.gz")
+            relaxed_OU_dict["model"][m].append(m)
+            for col in ["alpha", "num_theta_changes", "sigma2", "theta_root", "t_half"]:
+                relaxed_OU_dict[col][m].append(posterior(relaxed_df, col=col, burnin=0.5))
 
[…]

### 2024-03-04 manuscript/main-revision.tex
Prepare for revisions

--- 2023-10-02 manuscript/biorxiv-main.tex
@@
 \usepackage{makecell}
-\usepackage{orcidlink}
 \usepackage{marvosym}
@@
 \DeclareMathOperator{\Var}{\text{Var}}
+\DeclareMathOperator{\Std}{\text{Std}}
 \DeclareMathOperator{\Cov}{\text{Cov}}
@@
 \newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
+\newcommand{\StdPhy}{\Std \left[\MeanTrait\right]}
 \newcommand{\VecZero}{\UniDimArray{0}}
@@
 \newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
-\newcommand{\MutationRate}{\mu}
+\newcommand{\MutationRatePheno}{\mu}
+\newcommand{\MutationRateNuc}{u}
 \newcommand{\SubRate}{q}
@@
 \newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
-\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
 \newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
@@
 
-\newsavebox{\Maketitle}
 \title{Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes}
@@
 
-\author{
-\large
-\textbf{T.
-{Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M.
-{Bastian}$^{2}$\orcidlink{0009-0003-9129-6918}, T.
-{Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N.
-{Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
-\normalsize
-$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
-$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
-\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
-}
-
+\author{~}
 \begin{document}
 
-\sbox{\Maketitle}{%
-\begin{minipage}{\textwidth}
 \maketitle
-\end{minipage}}
-\noindent\usebox{\Maketitle}
 
@@
 Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{merila_comparison_2001, leinonen_comparative_2008}.
-A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation in different optimum trait value in the different populations~\parencite{lamy_qst_2012}.
+A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations~\parencite{lamy_qst_2012}.
 Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection.
-However, \QstFst\ methods have been found to require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
+However, it has been found that \QstFst\ methods require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
 Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.
+
+% Pair of species
 To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
@@
 Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
-% For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
-% However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
-% are these studies using Gst/Fst on gene expression?
-Otherwise, we should move it elsewhere or add a clear transition between the sentences
-
-To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
-For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
-% #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
-Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for genes with large within-species variation~\parencite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
+Analogously, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \textcite{mcdonald_adaptative_1991} test.
+In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
 
@@
 Deviations from the assumptions of the Brownian process are however well known.
-When trait variation is constraint because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
+When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
 Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
 However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
+First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
@@
 At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-Within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
 Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
-We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of molecular divergence at both scales.
+We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
 It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
@@
 
-Prior to developing our neutrality index, we review theoretical expectations for variations of quantitative traits and genomic sequences under neutral evolution for both within- and between-species variation.
+While observing trait variations across individuals of several species, we ask if the variation within species compared to variation between species is compatible with neutral evolution or not. 
+In statistical terms, this can also be framed as: Is the variance of means equal to the mean of variances?
+The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times, and each species has a putatively different population size.
+By reviewing theoretical expectations and leveraging nucleotide sequence variations, the goal of this section is thus to obtain normalized trait variation between and within species that are equal if the trait is neutral.
+Here we denote these normalized trait variations as respectively $\RateWhithin$ for within species and as $\RateBetween$ for between species.
 
@@
 For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
-New mutations are generating trait variance and the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
 At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$:
+As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
 \begin{gather}
-    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
+    \VarMutation = 2 \MutationRatePheno \Multiply \GenArchi \label{eq:var-mutation}.
 \end{gather}
@@
     \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
-    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
+    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
 \end{align}
@@
 Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
-As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$):
+As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRateNuc$) and the effective population size ($\Ne$):
 \begin{gather}
-    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
+    \pi = 4 \Ne \Multiply \MutationRateNuc \label{eq:genetic-diversity}.
 \end{gather}
@@
 We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
-This ratio allows removing the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
+This ratio allows removing the effect of $\Ne$, giving $\RateWhithin$ as a proxy of the underlying genetic architecture:
 \begin{align}
-    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
-    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
-    & = \RateMut. \label{eq:rate-pop}
+    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi} \label{eq:def-rate-pop}, \\
+    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Ne \Multiply \MutationRateNuc} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
+    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-pop}
 \end{align}
 
-The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
+The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), as:
 \begin{align}
-    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
+    \VarGenetic & = \Heritability \Multiply \VarPhenotype.\label{eq:var-gen-pheno}
+\end{align}
+Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ as:
+\begin{align}
+    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi } \text{ from definition eq.~\ref{eq:def-rate-pop} and~\ref{eq:var-gen-pheno}}, \label{eq:def-rate-pheno-pop} \\
+    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \text{ from eq.~\ref{eq:rate-pop}} \label{eq:rate-pheno-pop}
 \end{align}
@@
     \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
-    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+    & = 4 \Time \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
 \end{align}
@@
 The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
-We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
+We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
 \begin{align}
-    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
-    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
-    & = \MutationRate. \label{eq:substitution-rate}
+    \SubRate & = 2 \Ne \Multiply \MutationRateNuc \Multiply \pfix, \\
+    & = 2 \Ne  \Multiply \MutationRateNuc  \Multiply \dfrac{1}{2\Ne}, \\
+    & = \MutationRateNuc. \label{eq:substitution-rate}
 \end{align}
@@
     d & = \Time \Multiply \SubRate \\
-    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
+    & = \Time \Multiply \MutationRateNuc \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
 \end{align}
-We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
-This ratio allows removing the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
+We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4d$).
+This ratio allows removing the effect of $\Time$, giving $\RateBetween$ as another proxy of the underlying genetic architecture:
 \begin{align}
-    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
-    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
-    & = \RateMut \label{eq:rate-phy}.
+    \RateBetween & \defEqual \frac{\VarPhy}{4 d} \label{eq:def-rate-phy},\\
+    & = \frac{4 \Time \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Time \Multiply \MutationRateNuc}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
+    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \label{eq:rate-phy}.
 \end{align}
+
+Altogether, starting from the same ancestral population, divergent lineages accumulate both phenotypic and nucleotide changes reaching fixation in the population.
+These changes ultimately result in differences across lineages.
+In equation \ref{eq:rate-phy}, we show that the variance in mean trait value between lineages ($\VarPhy$) does increase linearly with nucleotide divergence ($d$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
+In other words, the standard deviation in mean phenotypes across lineages ($\StdPhy$) becomes proportional to the square root of nucleotide divergence since the ancestral population ($\sqrt{d}$).
 
@@
 The variability between either individuals or species can be obtained for both quantitative traits and genomic sequences.
-At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$, which equals $\RateMut$ if the trait is neutrally evolving (see above).
+At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$.
 At the phylogenetic level, the variability of the mean trait value between species can be combined with the nucleotide divergence of any neutrally evolving genomic region to obtain $\RateBetween$.
-Similarly, $\RateBetween=\RateMut$ if the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogenetic tree.
-We thus have, for a neutrally evolving trait:
-\begin{gather}
-    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
-    \Rightarrow \NI \defEqual \frac{\RateBetween}{\RateWhithin} = 1.
-\end{gather}
-We define a neutrality index $\NI = \RateBetween / \RateWhithin$ that will equal $1$ for a trait evolving neutrally.
-Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
+If the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogenetic tree, we thus have:
+\begin{align}
+    \frac{\RateBetween}{\RateWhithin} &= \frac{\VarPhy}{4 d} \Multiply \frac{\pi}{\Heritability \Multiply \VarPhenotype} \text{ by definition and eq.~\ref{eq:def-rate-pheno-pop} and~\ref{eq:def-rate-phy}}\\,
+ &= \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \Multiply \frac{\MutationRateNuc}{\MutationRatePheno \Multiply \GenArchi} \text{ from eq.~\ref{eq:rate-pheno-pop} and~\ref{eq:rate-phy}}, \\
+    &= 1.
+\end{align}
+We define a neutrality index $\NI$ as:
+\begin{align}
+\NI \defEqual \frac{\RateBetween}{\RateWhithin},
+\end{align}
+ which will equal to $1$ for a trait evolving neutrally.
+Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
 Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
 
-\subsection*{Estimate}\label{subsec:estimate}
+\subsection*{Estimation}\label{subsec:estimate}
 
-Based on the comparative framework that can account for phylogenetic inertia~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
+We denote $\EstNI$, $\EstRateBetween$ and $\EstRateWhithin$ the point estimates of respectively $\NI$, $\RateBetween$ and $\RateWhithin$.
+At the phylogenetic scale, $\EstRateBetween$ (defined eq.
+\ref{eq:def-rate-phy}) can be seen as an estimate of the rate of the evolution of the quantitative trait along a phylogeny, when the tree is measured in unit of $4d$ instead of time as is usually the case in the comparative framework~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}.
+We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
 
@@
 
-\subsubsection*{Bayesian estimate}
+\subsubsection*{Multivariate Brownian process}
 
-The Bayesian framework allows obtaining the posterior distribution of the neutrality index ($\EstNI$) for a given trait.
-Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
+In the previous section, $\EstNI$ is estimated independently for each trait of interest. 
+Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree.
 Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
+Importantly, the branch length of the tree on which the Brownian process runs is measured in units of $4d$ ($d$ is the nucleotide divergence), and not time as is usually the case in the comparative method.
+The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
+
+\subsubsection*{Bayesian estimate}
+The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
+We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\parencite{latrille_inferring_2021}.
 With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
@@
 Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
-For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with the mutation rate per generation $\MutationRate$.
+For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRatePheno \Multiply \NbrLoci $, with the mutation rate per locus per generation $\MutationRatePheno$.
 From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from the root to leaves of $0.18$ and average genetic diversity of $0.00276$.
 We scaled parameters in our simulations to fit plausible values for mammals.
-We thus used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
+We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
+We set $\MutationRatePheno=\MutationRateNuc$ without loss in generality since the genetic architecture ($\NbrLoci$ and $a$) is assumed constant in the simulator.
 
-The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which led to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
+The changes in log-$\MutationRatePheno$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which led to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
 An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
@@
 
-A random genetic drift was introduced by resampling individuals at each generation, with each parent having a probability of being sampled that was proportional to its fitness ($W$).
+At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
 Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
@@
         Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
-        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRate$ to mutate.
+        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRatePheno$ to mutate.
         Drift is modeled by the resampling of parents.
@@
 We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see Materials and Methods), suggesting that traits for which this relationship was not verified were putatively under selection.
-Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
+Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained toward similar values between different species, which leads to $\NI < 1$.
 In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
@@
 The inference framework was first tested on independently simulated datasets matching an empirically relevant mammalian empirical regime (see Materials and Methods).
-Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
+Under constant population size ($\Ne$) and constant mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
 For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
-Once we introduced fluctuating $\Ne$ and $\MutationRate$ (Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
+Once we introduced fluctuating $\Ne$, $\MutationRatePheno$ and $\MutationRateNuc$(Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
 For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) and none as diversifying selection.
@@
         Histogram of ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$ with $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and probabilities of $\EstNI$ being greater than $1$ (right).
-        Effective population size ($\Ne$) and mutation rate ($\MutationRate$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
+        Effective population size ($\Ne$) and mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
     }
@@
 Our approach could in theory be applied to detect stabilizing selection at the phylogenetic scale, but we showed that it did not have the statistical power to identify those cases.
-In contrast, we showed that our method was able to identify correctly cases of diversifying selection, which is a clear an improvement over current methods that model only mean trait value.
+In contrast, we showed that our method was able to identify correctly cases of diversifying selection, which is a clear improvement over current methods that model only mean trait value.
 Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
@@
 % Why is our test different
-The main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
+As such, the main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
 In the context of within species variation, \QstFst\ tests have been developed to compare trait and sequence across several populations to test for selection~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
@@
 Although the sequences should be neutrally evolving, they do not have to be necessarily linked to the quantitative trait under study.
-Nucleotide variation allows normalizing for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRate$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Nucleotide variation allows normalizing for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
 Thus our test avoids the estimation of the parameters, which are complex to correctly infer, and it also bypasses the estimation of divergence time, which was necessary in previous approaches~\parencite{walsh_evolution_2018}.
-But importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time were constant across the phylogenetic tree, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
-Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRate$ and generation time impacted similarly trait and sequence variation.
-The normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time, which canceled out in the ratio of trait variation $\EstNI$.
+But importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$, $\MutationRateNuc$ and generation time were constant across the phylogenetic tree, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
+In fact, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, $\MutationRatePheno$, $\MutationRateNuc$ and generation time, which canceled out in the ratio of trait variation $\EstNI$.
 
@@
 Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Finally, our Bayesian estimation could integrate uncertainty from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
+Finally, from a statistical perspective, our Bayesian estimation could integrate uncertainty from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
 
@@
 
-\subsection*{Acknowledgements}
- We gratefully acknowledge the help of Nicolas Lartillot, Philippe Veber, Isabela Jeronimo do Ó, Anna Marcionetti, Julien Clavel, and Daniele Silvestro for their insightful discussions and Julien Joseph for his advice and reviews concerning this manuscript.
-
-\subsection*{Competing interests:}
-The authors declare no conflicts of interest.
-
-\subsection*{Data and materials availability:}
-The data that support the findings of this study are openly available in GitHub at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}.
-Snakemake pipeline, analysis scripts and documentation are available in the repository to replicate the study.
-
 %TC:ignore
@@
 \newpage
-
-\noindent\usebox{\Maketitle}
 
@@
 \begin{equation}
-    \VarGenetic = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \label{eq:simu-var-genetic}
+    \VarGenetic = 4 \Ne \Multiply \MutationRatePheno \Multiply \NbrLoci \Multiply a^2 \label{eq:simu-var-genetic}
 \end{equation}
@@
 \begin{equation}
-    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
+    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRatePheno \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
 \end{equation}

### 2024-03-05 manuscript/figureS2.tex
Added figures for revision

\documentclass[10pt]{beamer}
\geometry{margin=0pt, nohead, nofoot, paperwidth=1080pt,paperheight=375pt}
\usepackage{pdfpages}
\usepackage{graphicx}
\pdfinclusioncopyfonts=1
\usepackage{pgfplots,pgf}
\usepgfplotslibrary{fillbetween}
\usebackgroundtemplate{\tikz\node[opacity=0]{};}
\setbeamertemplate{footline}[frame number]{}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}{}
\usefonttheme{serif}

\begin{document}

    \centering
    \begin{minipage}{0.32\textwidth}
        \begin{center}
            \LARGE $\mu =\mu_{\text{ref}} / 2 $
        \end{center}
        \includegraphics[width=\textwidth, page=1] {artworks/distance.neutral.low_u.pdf}
    \end{minipage}
    \begin{minipage}{0.32\textwidth}
        \begin{center}
            \LARGE $\mu =\mu_{\text{ref}} \cdot 4 $
        \end{center}
        \includegraphics[width=\textwidth, page=1] {artworks/distance.neutral.high_u.pdf}
    \end{minipage}
    \begin{minipage}{0.32\textwidth}
        \begin{center}
            \LARGE $\mu =\mu_{\text{ref}} \cdot 20 $
        \end{center}
        \includegraphics[width=\textwidth, page=1] {artworks/distance.neutral.very_high_u.pdf}
    \end{minipage}
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/pvalues.low_u.pdf}
    \end{minipage}
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/pvalues.high_u.pdf}
    \end{minipage}
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/pvalues.very_high_u.pdf}
    \end{minipage}

\end{document}

### 2024-03-05 manuscript/main-revision.tex
Added figures for revision

@@
 % Research Article (7500 words)
-\section*{Introduction}\label{sec:introduction}
+\section{Introduction}\label{sec:introduction}
 
@@
 
-\section*{Materials and Methods}\label{sec:materials-and-methods}
-\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
+\section{Materials and Methods}\label{sec:materials-and-methods}
+\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
 
@@
 
-\subsubsection*{Within-species trait variations}
+\subsubsection{Within-species trait variations}
 
@@
 
-\subsubsection*{Between-species trait variations}
+\subsubsection{Between-species trait variations}
 
@@
     \centering
-    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
+    \includegraphics[width=\textwidth, page=1] {figure1}
     \caption{
@@
 
-\subsubsection*{Neutrality index}
+\subsubsection{Neutrality index}
 
@@
 
-\subsection*{Estimation}\label{subsec:estimate}
+\subsection{Estimation}\label{subsec:estimate}
 
@@
 
-\subsubsection*{Maximum likelihood estimate}
+\subsubsection{Maximum likelihood estimate}
 At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitutions per site) and the topology of the phylogenetic tree.
@@
 
-\subsubsection*{Multivariate Brownian process}
+\subsubsection{Multivariate Brownian process}
 
@@
 
-\subsubsection*{Bayesian estimate}
+\subsubsection{Bayesian estimate}
 The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
@@
 
-\subsubsection*{Applicability to empirical data}
+\subsubsection{Applicability to empirical data}
 
@@
 
-\subsection*{Simulation}\label{subsec:simulations}
+\subsection{Simulation}\label{subsec:simulations}
 
@@
     \centering
-    \includegraphics[width=\textwidth, page=1] {artworks/fig-simulator}
+    \includegraphics[width=\textwidth, page=1] {figure2}
     \caption{
@@
 
-\subsection*{Empirical dataset}\label{subsec:empirical-dataset}
+\subsection{Empirical dataset}\label{subsec:empirical-dataset}
 
@@
 
-\section*{Results}\label{sec:results}
+\section{Results}\label{sec:results}
 
-\subsection*{Neutrality index}
+\subsection{Neutrality index}
 
@@
 
-\subsection*{Results against simulations}\label{subsec:results-against-simulations}
+\subsection{Results against simulations}\label{subsec:results-against-simulations}
 
@@
     \centering
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.rho}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.pvalues}
-    \end{minipage}
+    \includegraphics[width=\textwidth, page=1] {figure3}
     \caption{
@@
 
-\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
+\subsection{Results on empirical data}\label{subsec:results-on-empirical-data}
 
@@
 
-\section*{Discussion}\label{sec:discussion}
+\section{Discussion}\label{sec:discussion}
 
@@
     \captionof{figure}{summary of trait's genetic architecture.}
-    \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
+    \includegraphics[width=0.6\textwidth, page=1] {figureS1}
     \label{fig:simulator-summary}

### 2024-03-05 copy_artworks.sh

@@ -3,3 +3,9 @@
 \cp -f ./data_simulated/experiments/constant_pop_size/results/inference_bayesian.Phenotype.pvalues.pdf ./manuscript/artworks/cst_L5000.pvalues.pdf
 \cp -f ./data_simulated/experiments/fluctuating/results/inference_bayesian.Phenotype.pdf ./manuscript/artworks/fluNe_L5000.rho.pdf
 \cp -f ./data_simulated/experiments/fluctuating/results/inference_bayesian.Phenotype.pvalues.pdf ./manuscript/artworks/fluNe_L5000.pvalues.pdf
+\cp -f ./data_simulated/experiments/constant_pop_size_low_u/results/inference_ML.distance.neutral.pdf ./manuscript/artworks/distance.neutral.low_u.pdf
+\cp -f ./data_simulated/experiments/constant_pop_size_high_u/results/inference_ML.distance.neutral.pdf ./manuscript/artworks/distance.neutral.high_u.pdf
+\cp -f ./data_simulated/experiments/constant_pop_size_very_high_u/results/inference_ML.distance.neutral.pdf ./manuscript/artworks/distance.neutral.very_high_u.pdf
+\cp -f ./data_simulated/experiments/constant_pop_size_low_u/results/inference_bayesian.Phenotype.pvalues.pdf ./manuscript/artworks/pvalues.low_u.pdf
+\cp -f ./data_simulated/experiments/constant_pop_size_high_u/results/inference_bayesian.Phenotype.pvalues.pdf ./manuscript/artworks/pvalues.high_u.pdf
+\cp -f ./data_simulated/experiments/constant_pop_size_very_high_u/results/inference_bayesian.Phenotype.pvalues.pdf ./manuscript/artworks/pvalues.very_high_u.pdf

### 2024-03-05 manuscript/compile.sh

@@ -1,0 +1,1 @@
+latexdiff main.tex main-revision.tex > diff.tex

### 2024-03-06 manuscript/main-revision.tex
Implemented revisions

@@
 \DeclareMathOperator{\E}{\mathbb{E}}
-\DeclareMathOperator{\Var}{\text{Var}}
-\DeclareMathOperator{\Std}{\text{Std}}
-\DeclareMathOperator{\Cov}{\text{Cov}}
+\DeclareMathOperator{\Var}{\text{var}}
+\DeclareMathOperator{\Cov}{\text{cov}}
 \newcommand{\Qst}{Q$_\text{ST}$}
@@
 \newcommand{\NbrTaxa}{n}
-\newcommand{\Time}{t}
+\newcommand{\Time}{T}
+\newcommand{\GenTime}{\tau}
+\newcommand{\NbrGen}{t_{\Spi, \Spj}}
+\newcommand{\NucDiv}{d_{\Spi, \Spj}}
 \newcommand{\Trait}{P}
@@
 \newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
-\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
+\newcommand{\MeanTrait}{\bar{\Trait}}
 \newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
-\newcommand{\Root}{0}
-\newcommand{\RootTrait}{\Trait_{\Root}}
-\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
-\newcommand{\StdPhy}{\Std \left[\MeanTrait\right]}
+\newcommand{\RootTrait}{\theta}
+\newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
 \newcommand{\VecZero}{\UniDimArray{0}}
@@
 Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{merila_comparison_2001, leinonen_comparative_2008}.
-A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations~\parencite{lamy_qst_2012}.
-Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection.
+A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
+Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\parencite{lamy_qst_2012}.
 However, it has been found that \QstFst\ methods require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
+Alternatively, methods derived the statistical distribution of population means generated by random genetic drift and used the probability density of this distribution to test whether the observed pattern could be generated by drift alone~\parencite{ovaskainen_new_2011}.
 Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.
@@
 To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
-For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
-% #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
+For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes reaching fixation in the population.
+These changes ultimately result in different mean trait values across lineages.
+Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\parencite{lande_genetic_1980, turelli_heritable_1984}.
 Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\parencite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
 Analogously, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \textcite{mcdonald_adaptative_1991} test.
 In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
 
@@
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
+Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\parencite{grabowski_cautionary_2023}.
 Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
@@
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
+At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
 Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
@@
 Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
-From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species.
+From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species~\parencite{ovaskainen_new_2011}.
 From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
@@
 For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
-At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
+For a diploid individual, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
 As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
@@
 
-For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating in the population divided by the length of the region.
+For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all loci~\parencite{tajima_statistical_1989}.
 Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
@@
 
-\subsubsection{Between-species trait variations}
-
-For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
-If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
-The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given \parencite{hansen_translating_1996} by:
+\subsubsection{Between-species trait variations}\label{subsec:between-species-var}
+For a given species $\Spi$, we denote by $\MeanTrait_{\Spi}$ the mean value of the trait across the individuals.
+If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
+Given a phylogenetic tree, for a pair of species $\Spi$ and $\Spj$ from this tree, we denote as $\NbrGen$ the number of generations between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \textcite{hansen_translating_1996} by:
 \begin{align}
-    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
-    & = 4 \Time \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \NbrGen \\
+    & = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
 \end{align}
-
+where $\NbrGen$ is the number of generation between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
 Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
 The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
-We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
+We can derive the substitution rate per generation per loci ($\SubRate$) as the number of mutations per generation per loci ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{kimura_evolutionary_1968}, giving:
 \begin{align}
@@
 \end{align}
-That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.
+That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\parencite[review]{mccandlish_modeling_2014}.
 
-After $\Time$ generations and assuming that no multiple substitutions occurred at the same site, the nucleotide divergence $d$, which is the fraction of the genomic region that generated a substitution, will be $\Time$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
+Next, we denote $\NucDiv$ as the nucleotide divergence between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+In other words, $\NucDiv$ is the number of observed substitutions per loci during the $\NbrGen$ generations.
+Assuming that no multiple substitutions occurred at the same site, $\NucDiv$ is $\NbrGen$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
 \begin{align}
-    d & = \Time \Multiply \SubRate \\
-    & = \Time \Multiply \MutationRateNuc \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
+    \NucDiv & = \NbrGen \Multiply \SubRate \\
+    & = \NbrGen \Multiply \MutationRateNuc \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
 \end{align}
-We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4d$).
-This ratio allows removing the effect of $\Time$, giving $\RateBetween$ as another proxy of the underlying genetic architecture:
+We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4\NucDiv$).
+This ratio allows removing the effect of the number of generations ($\NbrGen$), giving $\RateBetween$ as another proxy of the underlying genetic architecture:
 \begin{align}
-    \RateBetween & \defEqual \frac{\VarPhy}{4 d} \label{eq:def-rate-phy},\\
-    & = \frac{4 \Time \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Time \Multiply \MutationRateNuc}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
+    \RateBetween & \defEqual \frac{\VarPhy}{4 \NucDiv} \label{eq:def-rate-phy},\\
+    & = \frac{4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \NbrGen \Multiply \MutationRateNuc}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
     & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \label{eq:rate-phy}.
@@
 
-Altogether, starting from the same ancestral population, divergent lineages accumulate both phenotypic and nucleotide changes reaching fixation in the population.
-These changes ultimately result in differences across lineages.
-In equation \ref{eq:rate-phy}, we show that the variance in mean trait value between lineages ($\VarPhy$) does increase linearly with nucleotide divergence ($d$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
-In other words, the standard deviation in mean phenotypes across lineages ($\StdPhy$) becomes proportional to the square root of nucleotide divergence since the ancestral population ($\sqrt{d}$).
+In equation \ref{eq:rate-phy}, we show that the covariance in mean trait value between a pair of species ($\VarPhy$) does increase linearly with shared nucleotide divergence ($\NucDiv$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
+Importantly, since the number of generations is the ratio of time and generation time (average time between two consecutive generations), removing the effect of the number of generations in equation \ref{eq:rate-phy} also removes the effect of both time and generation time.
 
@@
 \begin{align}
-    \frac{\RateBetween}{\RateWhithin} &= \frac{\VarPhy}{4 d} \Multiply \frac{\pi}{\Heritability \Multiply \VarPhenotype} \text{ by definition and eq.~\ref{eq:def-rate-pheno-pop} and~\ref{eq:def-rate-phy}}\\,
+    \frac{\RateBetween}{\RateWhithin} &= \frac{\VarPhy}{4 \NucDiv} \Multiply \frac{\pi}{\Heritability \Multiply \VarPhenotype} \text{ by definition and eq.~\ref{eq:def-rate-pheno-pop} and~\ref{eq:def-rate-phy}},\\
     &= \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \Multiply \frac{\MutationRateNuc}{\MutationRatePheno \Multiply \GenArchi} \text{ from eq.~\ref{eq:rate-pheno-pop} and~\ref{eq:rate-phy}}, \\
@@
  which will equal to $1$ for a trait evolving neutrally.
-Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
-Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
+Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$), generation time ($\GenTime$) or time of divergence ($\NbrGen$) need to be estimated.
+Moreover, the nucleotide sequence from which $\pi$ and $\NucDiv$ are obtained should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
 
@@
 We denote $\EstNI$, $\EstRateBetween$ and $\EstRateWhithin$ the point estimates of respectively $\NI$, $\RateBetween$ and $\RateWhithin$.
-At the phylogenetic scale, $\EstRateBetween$ (defined eq.
-\ref{eq:def-rate-phy}) can be seen as an estimate of the rate of the evolution of the quantitative trait along a phylogeny, when the tree is measured in unit of $4d$ instead of time as is usually the case in the comparative framework~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}.
+For each species with data available, $\RateWhithin$ as defined in eq.~\ref{eq:def-rate-pheno-pop} can be seen as a replicate sample.
+Thus, $\EstRateWhithin$ can be obtained by averaging out across all the sampled species.
+On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}.
+Generally, $\EstRateWhithin$ can thus be seen as an estimate of the rate of the evolution of the quantitative trait along a phylogeny, when the tree is measured in units of $4d$ ($d$ is the nucleotide divergence).
+As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\EstRateWhithin$.
 We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
@@
 \subsubsection{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitutions per site) and the topology of the phylogenetic tree.
-The diagonal $\Distance_{\Spi,\Spi}$ represents the total distances from the root of the tree to each taxon ($\Spi$).
-The off-diagonal elements ($\Distance_{\Spi,\Spj} = \Distance_{\Spj,\Spi}$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the symmetric distance matrix computed from the branch lengths and the topology of the phylogenetic tree.
+The diagonal $\Distance_{\Spi,\Spi}$ represents the total nucleotide divergence from the root of the tree to each taxon ($\Spi$).
+The off-diagonal elements ($\Distance_{\Spi,\Spj} = \NucDiv$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$, as in section \ref{subsec:between-species-var}.
 The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
@@
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-Importantly, the branch length of the tree on which the Brownian process runs is measured in units of $4d$ ($d$ is the nucleotide divergence), and not time as is usually the case in the comparative method.
+The branch length of the tree on which the Brownian process runs is measured in units of $4d$ ($d$ is the nucleotide divergence).
 The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
@@
 It is, however, possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
-Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\EstNI$ since $\Heritability \leq H^2$.
-In contrast, the test for stabilizing selection is invalid if $\EstNI$ is underestimated.
+Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and falling within the range of 0.2-0.5 in a vast majority of phenotypes tested~\parencite{hansen_heritability_2011, hansen_evolvability_2021}.
+Alternatively, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\EstNI$ since $\Heritability \leq H^2$.
+If available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.
+
+In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\EstNI$ is underestimated.
 Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
@@
 The environmental effect was normally distributed with variance $\VarEnv$.
-We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
+We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and for the theoretical formulation see section~S\ref{subsec:genotype-phenotype-map} and fig.~\ref{fig:simulator-summary}).
 We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
@@
 We scaled parameters in our simulations to fit plausible values for mammals.
-We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
+We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
 We set $\MutationRatePheno=\MutationRateNuc$ without loss in generality since the genetic architecture ($\NbrLoci$ and $a$) is assumed constant in the simulator.
 
-The changes in log-$\MutationRatePheno$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which led to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
+The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modeled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
+These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which if counted across $13,500$ generations leads to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$.
+In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $1.0$.
 An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
@@
         Left panel: For a given individual, the trait phenotypic value is the sum of genotypic value and a environmental effect (standard deviation $\VarEnv$).
-        The trait's genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
+        The trait's genotypic value is encoded by $\NbrLoci$ independent loci (meaning no linkage), with each locus contributing additively to the genotypic value.
         Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
@@
 We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
-We discarded species with only one representative sample.
-The mammalian nucleotide diversity was obtained from the Zoonomia project~\parencite{genereux_comparative_2020}, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
+We discarded species with only one representative since phenotypic variance cannot be estimated.
+The mammalian genomic data are gathered from the Zoonomia project~\parencite{genereux_comparative_2020}.
+More specifically, nucleotide divergence is estimated on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
 
@@
 However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
+As such, the evidence for $\EstNI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers for divergence can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
 
@@
 Our neutrality index made a series of assumptions that we described in details in the Material and Methods section.
-Table \ref{table:assumptions} summarized these assumptions and outlined possible consequences for the neutrality test that we proposed.
+Table~\ref{table:assumptions} summarized these assumptions and outlined possible consequences for the neutrality test that we proposed.
 
@@
             \hline
-            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
+            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\EstNI > 1$ & Test $\EstNI < 1$ \\ \hline \hline
             Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
             Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Phenotypic plasticity & Trait responding to individual environments  & Overestimated & -- & Conservative & Invalid  \\ \hline
             Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
@@
 However, our test detected a spurious signal of stabilizing selection ($\EstNI < 1$) when we simulated the evolution of a neutral trait.
+An assumption in our test is that the neutral phenotypic trait is evolving as a Brownian process and is unbounded.
+However, the phenotype is encoded by a genetic architecture and is thus ultimately bounded, even more so if the trait is encoded by a few loci (see table~\ref{table:assumptions}).
+At the macro-evolutionary scale, phenotypic divergence should plateau at some point, resulting in a reduced between-species trait variation.
+We argue that this effect can result in a spurious signal of stabilizing selection ($\EstNI < 1$), especially for deeper phylogeny (figure~\ref{fig:supp-distance} and section~S\ref{sec:supp-distance})
 We thus argue that our method should be used to detect diversifying selection, but that it had low accuracy to detect stabilizing selection due to false positives.
@@
 For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\parencite{catalan_drift_2019}.
-Our diversity index has the advantage to discriminate the alternative model of diversifying selection from the neutral case by comparing within- and between-species variation correctly normalized to remove confounding factors.
-Our approach is not the first one to normalize between-species variation to detect selection, but this was done by using within-species variations~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} and not estimates of neutral molecular divergence as done in our study.
-These studies have further compared their statistic across a pool of traits, which allowed them to identify outlier traits putatively under diversifying selection but without testing for selection on a single trait at a time~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
-Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departures from the null model of neutral evolution for a single test.
-Our diversity index opens new avenues to revisit these studies and better test for the selective regime affecting the quantitative traits, assuming we have access to genomic datasets to estimate nucleotide divergence and polymorphism.
+Instead, our diversity index has the advantage to discriminate the alternative model of diversifying selection from the neutral case by comparing within- and between-species variation while correctly normalizing them using nucleotide markers.
+Our approach is not the first one coupling between-species and within-species variations, and those approaches employ different strategies to detect selection.
+First, one empirical strategy is to compare the ratio of between to within variation across a pool of traits, which allow to identify outlier traits putatively under diversifying selection~\parencite{rohlfs_modeling_2014}.
+However, this method does not allow formally testing for diversifying selection, and requires many traits such as expression level data to seek outliers genes~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
+Second, other methods leverage Lande’s generalized genetic distance (LGGD), which relate the ratio of between to within variations to population-genetic parameters~\parencite{lynch_analysis_1990, lemos_evolutionary_2001, lemos_rates_2005, weaver_were_2007, porto_rate_2015}.
+Specifically, by either assuming constancy or by leveraging estimates of effective population size ($\Ne$) and number of generations between species, these methods can test for departures from the null model of neutral evolution for a single trait.
+Such methods have been successful in identifying specific instances of diversifying selection\parencite{schroeder_evolution_2017, machado_preeminent_2022} and near-drift~\parencite{machado_using_2023}.
+However, $\Ne$ and the number of generations are complex parameters to correctly infer, and is usually done for a pair of species or few species, and ultimately requires large genomic datasets and heavy statistical methods~\parencite{wilder_contribution_2023}.
+Instead, our diversity index opens new avenues to revisit these studies testing for the selective regime affecting the quantitative traits, by formally incorporating nucleotide divergence and polymorphism, bypassing estimation of $\Ne$, generation time and calibration of ancestral node ages~\parencite{machado_using_2023}.
 
@@
 As such, the main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
-In the context of within species variation, \QstFst\ tests have been developed to compare trait and sequence across several populations to test for selection~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
-Our neutrality index also used the genetic sequences from which nucleotide divergence and polymorphism are estimated.
-Although the sequences should be neutrally evolving, they do not have to be necessarily linked to the quantitative trait under study.
-Nucleotide variation allows normalizing for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-Thus our test avoids the estimation of the parameters, which are complex to correctly infer, and it also bypasses the estimation of divergence time, which was necessary in previous approaches~\parencite{walsh_evolution_2018}.
-But importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$, $\MutationRateNuc$ and generation time were constant across the phylogenetic tree, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
-In fact, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, $\MutationRatePheno$, $\MutationRateNuc$ and generation time, which canceled out in the ratio of trait variation $\EstNI$.
+In this context, our test bears many similarities to \QstFst\ tests that have been developed to test for selection of a trait across several populations while also leveraging sequence variation~\parencite{martin_multivariate_2008, ovaskainen_new_2011, leinonen_qst_2013}.
+Our method can be seen as an extension at the phylogenetic scale, where although the sequences used should be neutrally evolving, they can be obtained from different sampled individuals than for the trait.
+Importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$ and mutation rates were constant across the phylogenetic tree, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
+Indeed, under the neutral case of evolution, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, generation time and mutation rates, which canceled out in the neutrality index $\EstNI$.
+
+In the context of phylogenetic comparative methods, modeling mean trait evolution as a function of nucleotide divergence ($d$) instead of time has more general consequences.
+As an example, for a neutrally evolving trait, since generation time will change along the phylogeny, we argue that using time-calibrated trees can produce biases, such that $d$-scaled trees should be used instead~\parencite{litsios_effects_2012}.
+Using nucleotide divergence would also remove the potential effect of molecular clocks and calibration assumptions required to estimate ancestral node ages.
+We argue, that the soundness of studying trait evolution on $d$-scaled trees can be evaluated by absolute fit of a model to the data~\parencite{pennell_model_2015}.
+More generally, genomic information could potentially be seen as a way to disentangle congruence models~\parencite{louca_extant_2020}, or as prior for methods that detect shifts in adaptive regimes~\parencite{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020},.
 
@@
 Even though our test was developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provided insight into its behavior.
-First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations were used to determine the neutral expectation, and the inflation of divergence was compared to polymorphism within species.
-Second, because $\NI$ was compared to 1, our test ultimately bear analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
+First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences in a pair of species, except that the non-synonymous versus synonymous distinction is replaced by one between quantitative trait versus neutral genomic sequence.
+Second, at the phylogenetic scale across several species, our test also bears analogy to codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
 As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
@@
 As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
+If available, any prior knowledge on heritability can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection~\parencite{hansen_heritability_2011, hansen_evolvability_2021}.
 
@@
 From an empirical point of view, our method required integrating genomic and trait variation, which could reduce the possible datasets to be used.
-However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals brain and body mass.
+However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals' brain and body mass, showing signals of diversifying selection.
+The consensus on macro-evolutionary studies, assuming constancy of $\Ne$, generation time and mutation rates, is that empirical rates of evolution calculated on phylogenetic trees and the fossil record are far inferior to the expected under drift~\parencite{lynch_analysis_1990, uyeda_millionyear_2011}.
+Our finding of diversifying selection on body and brain mass could be seen as an argument against that interpretation.
+In fact, rates of nucleotide evolution also show a tendency for slowing down on a longer timescale~\parencite{rolland_conceptual_2023}.
+One possible interpretation is that normalization by nucleotide divergence could absorb this observed slowing rate of evolution.
+Altogether, further empirical and theoretical studies are required to disentangle this discrepancy between these different interpretations.
 Because our test was also based on several assumptions that might not hold on empirical data, we also provided a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (Table~\ref{table:assumptions}).
@@
     \captionof{figure}{summary of trait's genetic architecture.}
-    \includegraphics[width=0.6\textwidth, page=1] {figureS1}
+    \includegraphics[width=\textwidth, page=1] {figureS1}
     \label{fig:simulator-summary}
@@
 \begin{enumerate}
-    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers).
+    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers), from which the values of nucleotide divergence ($d$) used in denominator of eq.~\ref{eq:estimated-rate-phy} are used.
     \item A file containing the mean trait values for each species.
@@
 \end{lstlisting}
+
+\section{Saturation of phenotypic divergence}\label{sec:supp-distance}
+
+The phenotype is encoded by a genetic architecture and is thus ultimately bounded.
+At the macro-evolutionary scale, phenotypic divergence should plateau at some point, ultimately resulting in a $\RateBetween$.
+
+\begin{center}
+    \captionof{figure}{Saturation of phenotypic divergence.}
+    \includegraphics[width=1.0\textwidth, page=1] {figureS2}
+    \label{fig:supp-distance}
+\end{center}
+\begin{itemize}
+    \item Left column: $1,000$ simulations with low divergence between species (half that of mammals).
+    \item Middle column: $1,000$ simulations with high divergence between species (4 times that of mammals).
+    \item Right column: $200$ simulations with very high divergence between species (20 times that of mammals).
+    \item Top row: Simulation of neutral trait, phenotypic divergence between species as a function of the nucleotide divergence.
+Phenotypic divergence is computed between two species as the covariance between the trait ($\VarPhy$), and the nucleotide divergence is computed as the number of substitutions per site shared between the two species ($d$).
+Each point is a pair of species, and the bounds of the intervals in grey are the 2.5\% and 97.5\% quantiles across the replicates.
+The blue line is the linear regression and the red line is the saturation model ($y = \alpha \Multiply x / (\beta + x)$).
+    \item Bottom row: Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
+Histogram of probabilities of $\EstNI$ being greater than $1$.
+\end{itemize}
+Under a model of neutral trait evolution, when mutation rate increases, or equivalently the divergence between species increases, the phenotypic divergence between species saturates faster than the nucleotide divergence.
+This saturation effect can result in a spurious signal of stabilizing selection ($\EstNI < 1$) for deeper phylogeny when the trait is evolving neutrally.
 \end{document}

### 2024-03-13 manuscript/main-revision.tex
Revision round 1

@@
-%! BibTeX Compiler = biber
+%! BibTeX Compiler = bibtex
 %TC:ignore
@@
 \usepackage{biorxiv}
-\usepackage[backend=biber,eprint=false,isbn=false,url=false,giveninits=true, style=apa,date=year]{biblatex}
-\DeclareDelimFormat{nameyeardelim}{\addcomma\space}
-\addbibresource{references.bib}
 
@@
 \newcommand{\pfix}{\proba_{\text{fix}}}
+\newcommand{\Indiv}{k}
+\newcommand{\Branch}{b}
+\newcommand{\WishartIDD}{m}
 \newcommand{\Spi}{i}
@@
 \newcommand{\NbrTaxa}{n}
-\newcommand{\Time}{T}
-\newcommand{\GenTime}{\tau}
+\newcommand{\Time}{t}
 \newcommand{\NbrGen}{t_{\Spi, \Spj}}
@@
 \newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
-\newcommand{\RootTrait}{\theta}
+\newcommand{\RootTrait}{\phi}
 \newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
@@
 \newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
+\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
 \newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
@@
 \newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
-\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
-\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
+\newcommand{\EstRateBetween}{\widehat{\sigma}^2_{\mathrm{B}}}
+\newcommand{\EstRateWhithin}{\widehat{\sigma}^2_{\mathrm{W}}}
 \newcommand{\NI}{\rho}
@@
 The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.
+All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\citep{lynch_genetics_1998, walsh_evolution_2018}.
 
-Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\parencite{simons_population_2018, sella_thinking_2019}.
-Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{merila_comparison_2001, leinonen_comparative_2008}.
-A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
-Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\parencite{lamy_qst_2012}.
-However, it has been found that \QstFst\ methods require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
-Alternatively, methods derived the statistical distribution of population means generated by random genetic drift and used the probability density of this distribution to test whether the observed pattern could be generated by drift alone~\parencite{ovaskainen_new_2011}.
-Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.
+Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
+Given this genetic architecture of traits, from two diverging populations, it is possible to distinguish which traits have evolved under natural selection in controlled experimental settings, by performing genetic cross between individuals~\citep{fraser_detecting_2020}.
+Across several populations, by contrasting both trait differentiation (\Qst) and genetic differentiation (\Fst), so-called \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
+\Qst\ higher than \Fst\ is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
+Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\citep{lamy_qst_2012}.
+Other frameworks explicitly model genetic drift as a random process generating both trait and genetic differences between individuals and populations.
+This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
+However, regardless of the strengths and weaknesses of each method\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.
 
@@
 To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
-For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes reaching fixation in the population.
+For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes that will reach fixation in the population.
 These changes ultimately result in different mean trait values across lineages.
-Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\parencite{lande_genetic_1980, turelli_heritable_1984}.
-Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\parencite{khaitovich_neutral_2004}.
-Analogously, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \textcite{mcdonald_adaptative_1991} test.
+Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984}.
+Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
+As an analogy, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \citet{mcdonald_adaptative_1991} test.
 In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.
 
 % Phylogenetics (mean trait evolution)
-Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
-As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
+Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{hansen_translating_1996}.
+As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
 Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
 Deviations from the assumptions of the Brownian process are however well known.
-When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
-Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
-However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
+When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{catalan_drift_2019}.
+Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
+However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
+First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
-Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\parencite{grabowski_cautionary_2023}.
-Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
+Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.
@@
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
+Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
 Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
@@
 Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
-From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species~\parencite{ovaskainen_new_2011}.
-From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
-% Basically, we are testing whether the variance of means is greater than the mean of variances
+From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary generalization to account for phylogenetic relationships between species.
+From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
 
-\section{Materials and Methods}\label{sec:materials-and-methods}
-\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
+\section*{Materials and Methods}\label{sec:materials-and-methods}
+\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
 
@@
 In statistical terms, this can also be framed as: Is the variance of means equal to the mean of variances?
-The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times, and each species has a putatively different population size.
+The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times.
 By reviewing theoretical expectations and leveraging nucleotide sequence variations, the goal of this section is thus to obtain normalized trait variation between and within species that are equal if the trait is neutral.
@@
 
-\subsubsection{Within-species trait variations}
+\subsubsection*{Within-species trait variations}
 
@@
 For a diploid individual, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
+As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per locus per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
 \begin{gather}
@@
 
-While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
-For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
+While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\citep{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
+For a neutral trait at equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
 \begin{align}
@@
 
-For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all loci~\parencite{tajima_statistical_1989}.
+For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all nucleotide sites~\citep{tajima_statistical_1989}.
 Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
-As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRateNuc$) and the effective population size ($\Ne$):
+As shown in \citet{tajima_statistical_1989}, $\pi$ is a function of the mutation rate ($\MutationRateNuc$, per nucleotide site per generation) and the effective population size ($\Ne$):
 \begin{gather}
@@
 
-We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
-This ratio allows removing the effect of $\Ne$, giving $\RateWhithin$ as a proxy of the underlying genetic architecture:
+To remove the effect of $\Ne$, we define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
+After simplification, $\RateWhithin$ is then solely a function of the underlying genetic architecture as:
 \begin{align}
@@
 
-The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), as:
+If $\VarGenetic$ is not empirically accessible, it can be related to the observed phenotypic variance ($\VarPhenotype$), multiplied by narrow-sense heritability of the trait ($\Heritability$), as~\citep{hill_data_2008}:
 \begin{align}
@@
 \end{align}
-Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ as:
+Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ instead of $\VarGenetic$ as:
 \begin{align}
@@
 
-\subsubsection{Between-species trait variations}\label{subsec:between-species-var}
+\subsubsection*{Between-species trait variations}\label{subsec:between-species-var}
 For a given species $\Spi$, we denote by $\MeanTrait_{\Spi}$ the mean value of the trait across the individuals.
-If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
+If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
 Given a phylogenetic tree, for a pair of species $\Spi$ and $\Spj$ from this tree, we denote as $\NbrGen$ the number of generations between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
-Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \textcite{hansen_translating_1996} by:
+Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \citet{hansen_translating_1996} by:
 \begin{align}
     \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \NbrGen \\
-    & = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+    & = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}}.
 \end{align}
-where $\NbrGen$ is the number of generation between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
 Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
-The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
-We can derive the substitution rate per generation per loci ($\SubRate$) as the number of mutations per generation per loci ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{kimura_evolutionary_1968}, giving:
+The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$~\citep{kimura_probability_1962}.
+We can derive the substitution rate ($\SubRate$, per nucleotide site per generation) as the number of newly arisen mutations ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{kimura_evolutionary_1968}, giving:
 \begin{align}
@@
 \end{align}
-That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\parencite[review]{mccandlish_modeling_2014}.
+That is, if mutations are neutral, the rate of substitution per generation within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\citep[review]{mccandlish_modeling_2014}.
 
 Next, we denote $\NucDiv$ as the nucleotide divergence between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
-In other words, $\NucDiv$ is the number of observed substitutions per loci during the $\NbrGen$ generations.
-Assuming that no multiple substitutions occurred at the same site, $\NucDiv$ is $\NbrGen$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
+In other words, $\NucDiv$ is the expected number of substitutions per nucleotide site during the $\NbrGen$ generations.
+Assuming that no multiple substitutions occurred at the same site, $\NucDiv$ is the number of generations ($\NbrGen$) multiplied by the nucleotide substitution rate per generation ($\SubRate$):
 \begin{align}
@@
 \end{align}
-We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4\NucDiv$).
-This ratio allows removing the effect of the number of generations ($\NbrGen$), giving $\RateBetween$ as another proxy of the underlying genetic architecture:
+To remove the effect of the number of generations ($\NbrGen$) first, and to also equate to $\RateWhithin$ (eq.~\ref{eq:rate-pop}), we define $\RateBetween$ as the covariance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4\NucDiv$).
+After simplification, $\RateBetween$ is also solely a function of the underlying genetic architecture as:
 \begin{align}
@@
     & = \frac{4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \NbrGen \Multiply \MutationRateNuc}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
-    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \label{eq:rate-phy}.
+    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-phy}
 \end{align}
 
-In equation \ref{eq:rate-phy}, we show that the covariance in mean trait value between a pair of species ($\VarPhy$) does increase linearly with shared nucleotide divergence ($\NucDiv$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
-Importantly, since the number of generations is the ratio of time and generation time (average time between two consecutive generations), removing the effect of the number of generations in equation \ref{eq:rate-phy} also removes the effect of both time and generation time.
+In equation~\ref{eq:rate-phy}, we show that the covariance in mean trait value between a pair of species ($\VarPhy$) does increase linearly with shared nucleotide divergence ($\NucDiv$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
+Importantly, since the number of generations is the ratio of time divided by generation time (average time between two consecutive generations), removing the effect of the number of generations in equation~\ref{eq:rate-phy} also removes the effect of both time and generation time.
 
@@
  which will equal to $1$ for a trait evolving neutrally.
-Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$), generation time ($\GenTime$) or time of divergence ($\NbrGen$) need to be estimated.
+Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$), generation time or time of divergence ($\NbrGen$) need to be estimated.
 Moreover, the nucleotide sequence from which $\pi$ and $\NucDiv$ are obtained should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
 
-\subsection{Estimation}\label{subsec:estimate}
+\subsection*{Estimation}\label{subsec:estimate}
 
@@
 Thus, $\EstRateWhithin$ can be obtained by averaging out across all the sampled species.
-On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}.
-Generally, $\EstRateWhithin$ can thus be seen as an estimate of the rate of the evolution of the quantitative trait along a phylogeny, when the tree is measured in units of $4d$ ($d$ is the nucleotide divergence).
+On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}.
+Generally, $\EstRateWhithin$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
 As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\EstRateWhithin$.
@@
 
-\subsubsection{Maximum likelihood estimate}
+\subsubsection*{Maximum likelihood estimate}
 At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the symmetric distance matrix computed from the branch lengths and the topology of the phylogenetic tree.
 The diagonal $\Distance_{\Spi,\Spi}$ represents the total nucleotide divergence from the root of the tree to each taxon ($\Spi$).
-The off-diagonal elements ($\Distance_{\Spi,\Spj} = \NucDiv$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$, as in section \ref{subsec:between-species-var}.
-The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
+The off-diagonal elements ($\Distance_{\Spi,\Spj} = \NucDiv$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$, as in equation~\ref{eq:distance}.
+The mean trait value at the root of the tree ($\RootTrait$) can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\citep{omeara_testing_2006}:
 \begin{gather}
@@
 
-Finally, between-species variation $\EstRateBetween$ is estimated as~\parencite{omeara_testing_2006}:
+Finally, between-species variation $\EstRateBetween$ is estimated as~\citep{omeara_testing_2006}:
 \begin{gather}
@@
 
-\subsubsection{Multivariate Brownian process}
+\subsubsection*{Multivariate Brownian process}
 
@@
 Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree.
-Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
+Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-The branch length of the tree on which the Brownian process runs is measured in units of $4d$ ($d$ is the nucleotide divergence).
+The branch lengths of the tree used to model the Brownian process runs is measured in units of $4d$ ($d$ being the nucleotide divergence).
 The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
 
-\subsubsection{Bayesian estimate}
+\subsubsection*{Bayesian estimate}
 The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
-We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\parencite{latrille_inferring_2021}.
+We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\citep{latrille_inferring_2021}.
 With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
@@
 
-\subsubsection{Applicability to empirical data}
+\subsubsection*{Applicability to empirical data}
 
@@
 It is, however, possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
-Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and falling within the range of 0.2-0.5 in a vast majority of phenotypes tested~\parencite{hansen_heritability_2011, hansen_evolvability_2021}.
+Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and fall within the range of 0.2-0.5 in a vast majority of phenotypic traits tested~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
 Alternatively, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\EstNI$ since $\Heritability \leq H^2$.
@@
 
-\subsection{Simulation}\label{subsec:simulations}
+\subsection*{Simulation}\label{subsec:simulations}
 
@@
 We scaled parameters in our simulations to fit plausible values for mammals.
-We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
+We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per site per generation  and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
 We set $\MutationRatePheno=\MutationRateNuc$ without loss in generality since the genetic architecture ($\NbrLoci$ and $a$) is assumed constant in the simulator.
@@
 The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modeled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
-These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which if counted across $13,500$ generations leads to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$.
+These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$.
 In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $1.0$.
@@
 At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
-Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
 More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
@@
 Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
-Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\parencite{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
+Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
 Neutral evolution was implemented by fixing the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.
@@
 The additive genetic variance was measured as phenotypic variance multiplied by heritability.
-Heritability was estimated from the slopes of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\parencite{lynch_genetics_1998} averaged over the last 10 simulated generations.
+Heritability was estimated from the slopes of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\citep{lynch_genetics_1998} averaged over the last 10 simulated generations.
 Heritability was thus not a given parameter of the simulations, but rather measured as it would be in empirical data.
@@
 
-\subsection{Empirical dataset}\label{subsec:empirical-dataset}
+\subsection*{Empirical dataset}\label{subsec:empirical-dataset}
 
 We analyzed a dataset of body and brain masses from mammals.
-The log-transformed values of body and brain masses were taken from \textcite{tsuboi_breakdown_2018}.
+The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
 We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
 We discarded species with only one representative since phenotypic variance cannot be estimated.
-The mammalian genomic data are gathered from the Zoonomia project~\parencite{genereux_comparative_2020}.
-More specifically, nucleotide divergence is estimated on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
+The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
+More specifically, nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \citet{wilder_contribution_2023}.
 
-We also analyzed a dataset of primate species, with the nucleotide variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
+We also analyzed a dataset of primate species, with the nucleotide variation obtained from \citet{kuderna_global_2023} and the quantitative trait variation also from \citet{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
 However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
-As such, the evidence for $\EstNI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers for divergence can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
+As such, the evidence for $\EstNI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers included in the estimate of divergence can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
 
-\section{Results}\label{sec:results}
+\section*{Results}\label{sec:results}
 
-\subsection{Neutrality index}
+\subsection*{Neutrality index}
 
@@
 We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see Materials and Methods), suggesting that traits for which this relationship was not verified were putatively under selection.
-Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained toward similar values between different species, which leads to $\NI < 1$.
-In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
+Under stabilizing selection, the variation between species is depleted because the mean trait is maintained toward similar values between different species, which leads to $\NI < 1$.
+In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\citep{hansen_stabilizing_1997}, which leads to $\NI > 1$.
 Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
@@
 
-\subsection{Results against simulations}\label{subsec:results-against-simulations}
+\subsection*{Results against simulations}\label{subsec:results-against-simulations}
 
@@
 
-\subsection{Results on empirical data}\label{subsec:results-on-empirical-data}
+\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
 
@@
 Assuming complete heritability, brain mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
-If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\parencite{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
+If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\citep{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
 Mammalian brain mass was found to be under diversifying selection with posterior probabilities of $0.877$ for males and $0.972$ for females when complete heritability was assumed.
@@
 However, our test detected a spurious signal of stabilizing selection ($\EstNI < 1$) when we simulated the evolution of a neutral trait.
-An assumption in our test is that the neutral phenotypic trait is evolving as a Brownian process and is unbounded.
-However, the phenotype is encoded by a genetic architecture and is thus ultimately bounded, even more so if the trait is encoded by a few loci (see table~\ref{table:assumptions}).
+An assumption of our test is that the neutral phenotypic trait is evolving as a Brownian process and is, therefore, unbounded.
+However, the phenotype is encoded by a genetic architecture, and is thus ultimately bounded, even more so if the trait is encoded by a few loci (see table~\ref{table:assumptions}).
 At the macro-evolutionary scale, phenotypic divergence should plateau at some point, resulting in a reduced between-species trait variation.
@@
 Our results showed that our method significantly improved over currently available methods to detect selection acting on a trait at the phylogenetic scale.
-Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 Our approach could in theory be applied to detect stabilizing selection at the phylogenetic scale, but we showed that it did not have the statistical power to identify those cases.
 In contrast, we showed that our method was able to identify correctly cases of diversifying selection, which is a clear improvement over current methods that model only mean trait value.
-Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\parencite{catalan_drift_2019}.
+Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
+For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\citep{catalan_drift_2019}.
 Instead, our diversity index has the advantage to discriminate the alternative model of diversifying selection from the neutral case by comparing within- and between-species variation while correctly normalizing them using nucleotide markers.
 Our approach is not the first one coupling between-species and within-species variations, and those approaches employ different strategies to detect selection.
-First, one empirical strategy is to compare the ratio of between to within variation across a pool of traits, which allow to identify outlier traits putatively under diversifying selection~\parencite{rohlfs_modeling_2014}.
-However, this method does not allow formally testing for diversifying selection, and requires many traits such as expression level data to seek outliers genes~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
-Second, other methods leverage Lande’s generalized genetic distance (LGGD), which relate the ratio of between to within variations to population-genetic parameters~\parencite{lynch_analysis_1990, lemos_evolutionary_2001, lemos_rates_2005, weaver_were_2007, porto_rate_2015}.
-Specifically, by either assuming constancy or by leveraging estimates of effective population size ($\Ne$) and number of generations between species, these methods can test for departures from the null model of neutral evolution for a single trait.
-Such methods have been successful in identifying specific instances of diversifying selection\parencite{schroeder_evolution_2017, machado_preeminent_2022} and near-drift~\parencite{machado_using_2023}.
-However, $\Ne$ and the number of generations are complex parameters to correctly infer, and is usually done for a pair of species or few species, and ultimately requires large genomic datasets and heavy statistical methods~\parencite{wilder_contribution_2023}.
-Instead, our diversity index opens new avenues to revisit these studies testing for the selective regime affecting the quantitative traits, by formally incorporating nucleotide divergence and polymorphism, bypassing estimation of $\Ne$, generation time and calibration of ancestral node ages~\parencite{machado_using_2023}.
+First, one empirical strategy is to compare the ratio of between to within variation across a pool of traits, which allow to identify outlier traits putatively under diversifying selection~\citep{rohlfs_modeling_2014}.
+However, this method does not formally allow testing for diversifying selection, and requires many traits such as expression level data to seek outliers genes~\citep{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
+Second, other methods leverage Lande’s generalized genetic distance (LGGD), which relate the ratio of between to within variations to population-genetic parameters~\citep{lynch_analysis_1990, lemos_evolutionary_2001, lemos_rates_2005, weaver_were_2007, porto_rate_2015}.
+Specifically, by leveraging estimates of effective population size ($\Ne$) and number of generations between species, or alternatively by assuming their constancy, these methods can test for departures from the null model of neutral evolution for a single trait.
+Such methods have been successful in identifying specific instances of diversifying selection\citep{schroeder_evolution_2017, machado_preeminent_2022} and near-drift~\citep{machado_using_2023}.
+However, $\Ne$ and the number of generations are complex parameters to correctly infer, and is usually done for a pair or only a few species, and ultimately requires large genomic datasets and heavy statistical methods~\citep{wilder_contribution_2023}.
+Instead, our diversity index opens new avenues to revisit these studies testing for the selective regime affecting the quantitative traits, by formally incorporating nucleotide divergence and polymorphism, bypassing estimation of $\Ne$, generation time and calibration of ancestral node ages~\citep{machado_using_2023}.
 
@@
 As such, the main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
-In this context, our test bears many similarities to \QstFst\ tests that have been developed to test for selection of a trait across several populations while also leveraging sequence variation~\parencite{martin_multivariate_2008, ovaskainen_new_2011, leinonen_qst_2013}.
+In this context, our test bears many similarities to \QstFst\ tests (and their derivatives) that have been developed to test for selection of a trait across several populations while also leveraging sequence variation~\citep{martin_multivariate_2008, leinonen_qst_2013} or co-ancestry between individuals~\citep{ovaskainen_new_2011}.
 Our method can be seen as an extension at the phylogenetic scale, where although the sequences used should be neutrally evolving, they can be obtained from different sampled individuals than for the trait.
-Importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$ and mutation rates were constant across the phylogenetic tree, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
+Importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$ and mutation rates were constant across the phylogenetic tree, an unmet assumption empirically~\citep{bergeron_evolution_2023, wilder_contribution_2023}.
 Indeed, under the neutral case of evolution, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, generation time and mutation rates, which canceled out in the neutrality index $\EstNI$.
@@
 In the context of phylogenetic comparative methods, modeling mean trait evolution as a function of nucleotide divergence ($d$) instead of time has more general consequences.
-As an example, for a neutrally evolving trait, since generation time will change along the phylogeny, we argue that using time-calibrated trees can produce biases, such that $d$-scaled trees should be used instead~\parencite{litsios_effects_2012}.
-Using nucleotide divergence would also remove the potential effect of molecular clocks and calibration assumptions required to estimate ancestral node ages.
-We argue, that the soundness of studying trait evolution on $d$-scaled trees can be evaluated by absolute fit of a model to the data~\parencite{pennell_model_2015}.
-More generally, genomic information could potentially be seen as a way to disentangle congruence models~\parencite{louca_extant_2020}, or as prior for methods that detect shifts in adaptive regimes~\parencite{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020},.
+As an example, trait variation is often modeled as a Brownian process running on a time-calibrated tree, which can produce biases~\citep{litsios_effects_2012}.
+Indeed, for a neutrally evolving trait, trait variation depends directly on the number of generations, which in turn correlates with time.
+But, since species generation time might vary along the phylogenetic tree, $d$-scaled trees absorbing changes in generation time should be used instead of time-scaled trees.
+Using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks).
+We argue, that the soundness of studying trait evolution on $d$-scaled trees can be evaluated by the absolute fit of a model to the data~\citep{pennell_model_2015}.
+More generally, genomic information could potentially be seen as a way to disentangle congruence models~\citep{louca_extant_2020}, or as prior for methods that detect shifts in adaptive regimes~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020},.
 
@@
 Even though our test was developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provided insight into its behavior.
-First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences in a pair of species, except that the non-synonymous versus synonymous distinction is replaced by one between quantitative trait versus neutral genomic sequence.
-Second, at the phylogenetic scale across several species, our test also bears analogy to codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
+First, we acknowledge that our test took inspiration from the \citet{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences in a pair of species, except that the non-synonymous versus synonymous distinction is replaced by the comparison between quantitative trait and neutral genomic sequence.
+Second, at the phylogenetic scale, when comparison is done across several species, our test also bears analogy to codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\citep{goldman_codonbased_1994, muse_likelihood_1994}.
 As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
 Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we could leverage the vast literature discussing and interpreting the results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ did not necessarily imply that the trait was effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ did not rule out diversifying selection, but rather that this diversifying selection was not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
+With this analogy in mind, we could leverage the vast literature discussing and interpreting the results of these tests and their pitfalls~\citep{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
+First, not rejecting the neutral null model of $\NI = 1$ did not necessarily imply that the trait was effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\citep{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ did not rule out diversifying selection, but rather that this diversifying selection was not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\citep{latrille_genes_2023}.
 By explicitly modeling stabilizing selection as a moving optimum, it would theoretically be possible to tease apart the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
@@
 As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
-If available, any prior knowledge on heritability can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection~\parencite{hansen_heritability_2011, hansen_evolvability_2021}.
+If available, any prior knowledge on heritability can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
 
@@
 Second, loci were assumed to contribute additively to the phenotype.
-Although the effects of dominance and epistasis is typically weak compared to the additive effects on the quantitative traits, their influence should be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
-Third, the genetic architecture of the trait was assumed to be constant across the phylogenetic tree, whereas it might actually be variable among individuals and species~\parencite{tung_genetic_2015, huber_conservatism_2015}.
-Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Although the effects of dominance and epistasis is typically weak compared to the additive effects on the quantitative traits, their influence should be assessed~\citep{hill_data_2008, crow_epistasis_2010}.
+Third, the genetic architecture of the trait was assumed to be constant across the phylogenetic tree, whereas it might actually be variable among individuals and species~\citep{tung_genetic_2015, huber_conservatism_2015}.
+Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\citep{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 Finally, from a statistical perspective, our Bayesian estimation could integrate uncertainty from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
@@
 However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals' brain and body mass, showing signals of diversifying selection.
-The consensus on macro-evolutionary studies, assuming constancy of $\Ne$, generation time and mutation rates, is that empirical rates of evolution calculated on phylogenetic trees and the fossil record are far inferior to the expected under drift~\parencite{lynch_analysis_1990, uyeda_millionyear_2011}.
+The consensus on macro-evolutionary studies, assuming constancy of $\Ne$, generation time and mutation rates, is that empirical rates of evolution calculated on phylogenetic trees and the fossil record are far inferior to the expected under drift~\citep{lynch_analysis_1990, uyeda_millionyear_2011}.
 Our finding of diversifying selection on body and brain mass could be seen as an argument against that interpretation.
-In fact, rates of nucleotide evolution also show a tendency for slowing down on a longer timescale~\parencite{rolland_conceptual_2023}.
+In fact, rates of nucleotide evolution also show a tendency for slowing down on a longer timescale~\citep{rolland_conceptual_2023}.
 One possible interpretation is that normalization by nucleotide divergence could absorb this observed slowing rate of evolution.
@@
 %TC:ignore
-\printbibliography
+\bibliographystyle{natbib}
+\bibliography{references_bibtex}
 
@@
     \item $\Ne$ is the effective number of individuals.
-    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
-    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the genotypic value for individual $i$.
-    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
-    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
+    \item $g_{\Indiv,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $\Indiv \in \{1, \hdots, \Ne\}$.
+    \item $G_{\Indiv} = \sum_{l=1}^{\NbrLoci} a_l \times g_{\Indiv,l}$ is the genotypic value for individual $\Indiv$.
+    \item $\xi_{\Indiv} \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual $\Indiv$.
+    \item $\Trait_{\Indiv} = G_{\Indiv}+ \xi_i$ is the phenotype for individual $\Indiv$.
 \end{itemize}
@@
     \captionof{figure}{summary of trait's genetic architecture.}
-    \includegraphics[width=\textwidth, page=1] {figureS1}
+    \includegraphics[width=0.6\textwidth, page=1] {figureS1}
     \label{fig:simulator-summary}
@@
 \begin{equation}
-    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad} \VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2\label{eq:simu-genotype}
+    \bar{G} = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne} G_{\Indiv} \text{\quad and \quad} \VarGenetic = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne}\left(G_{\Indiv}- \bar{G} \right)^2\label{eq:simu-genotype}
 \end{equation}
@@
 \begin{equation}
-    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2 \label{eq:simu-between}
+    \bar{\Trait} = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne} \Trait_{\Indiv}\text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne}\left(\Trait_{\Indiv} - \bar{\Trait} \right)^2 \label{eq:simu-between}
 \end{equation}
@@
 \subsection{Multivariate Brownian process}\label{subsec:multivariate-brownian-process}
-Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
+Here we model $\Ntrait$ traits evolving along the phylogenetic tree that are correlated between them.
 Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-Along branch $j$ with length $d_{j}$, the Brownian process start at the ancestral node $\mathcal{A}(j)$ with value $\Brownian(\mathcal{A}(j))$, and ends at node $\mathcal{R}(j)$  with value $\Brownian(\mathcal{R}(j))$.
-The independent contrast $\contrast_{j}$ defined as change in trait along the branch normalized by $\sqrt {d_{j}}$ is a multivariate Gaussian:
+Along branch $\Branch$ with length $d_{\Branch}$, the Brownian process start at the ancestral node $\mathcal{A}(\Branch)$ with value $\Brownian(\mathcal{A}(\Branch))$, and ends at node $\mathcal{R}(\Branch)$  with value $\Brownian(\mathcal{R}(\Branch))$.
+The independent contrast $\contrast_{\Branch}$ defined as change in trait along the branch normalized by $\sqrt {d_{\Branch}}$ is a multivariate Gaussian:
 \begin{equation}
     \label{eq:DistribBrownian}
-    \contrast_{j} = \frac{\Brownian (\mathcal{R}(j)) - \Brownian (\mathcal{A}(j)) }{\sqrt {d_{j}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
+    \contrast_{\Branch} = \frac{\Brownian (\mathcal{R}(\Branch)) - \Brownian (\mathcal{A}(\Branch)) }{\sqrt {d_{\Branch}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
 \end{equation}
@@
 \subsection{Sampling the covariance matrix}\label{subsec:sampling-the-covariance-matrix}
-From the independent contrast at each branch of the tree ($\contrast_{j}$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$, as:
+From the independent contrast at each branch of the tree ($\contrast_{\Branch}$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$, as:
 \begin{equation}
-    \Scattermatrix = \sum\limits_{j=1}^{\Nbranch} \contrast_{j} \MultiplyMatrix \left[\contrast_{j}\right]\tr\label{eq:bayes-scatter},
+    \Scattermatrix = \sum\limits_{\Branch=1}^{\Nbranch} \contrast_{\Branch} \MultiplyMatrix \contrast_{\Branch}\tr\label{eq:bayes-scatter},
 \end{equation}
@@
 \end{equation}
-This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate_{k}$ defined by
+This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate_{\WishartIDD}$ defined by
 \begin{equation}
-    \Multivariate_{k} \sim \mathcal{N} \left( \VecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
+    \Multivariate_{\WishartIDD} \sim \mathcal{N} \left( \VecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
 \end{equation}
@@
 \begin{equation}
-    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate_{k} \MultiplyMatrix  \left[\Multivariate_{k} \right] \tr \right)\inv \label{eq:bayes-gibbs}
+    \CovarianceMatrix = \left( \sum\limits_{\WishartIDD=1}^{\WishartPostDf} \Multivariate_{\WishartIDD} \MultiplyMatrix  \left[\Multivariate_{\WishartIDD} \right] \tr \right)\inv \label{eq:bayes-gibbs}
 \end{equation}
@@
 
-\subsubsection{Phylogenetic tree}\label{subsubsec:phylogenetic-tree}
+\subsubsection{Phylogenetic tree}
 
@@
 
-\subsubsection{Mean trait for each species}\label{subsubsec:mean-trait-for-each-species}
+\subsubsection{Mean trait for each species}
 
@@
 \newpage
-\subsubsection{Trait variation for each species}\label{subsubsec:trait-variation-for-each-species}
+\subsubsection{Trait variation for each species}
 
@@
 
-\subsubsection{Running the model}\label{subsubsec:running-nodetraits}
+\subsubsection{Running the model}
 \texttt{nodetraits} is run with the following command:
@@
 
-\subsubsection{Reading the results}\label{subsubsec:reading-the-results}
+\subsubsection{Reading the results}
 Once the model has run, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between-species variation over within-species variation with \texttt{readnodetraits}:
@@
 
+\newpage
 \section{Saturation of phenotypic divergence}\label{sec:supp-distance}

### 2024-07-03 manuscript/main-first-submission.tex
Revision round 2 manuscript files.

--- 2023-10-02 manuscript/biorxiv-main.tex
@@
 \newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
-\newcommand{\MutationRate}{\mu}
+\newcommand{\MutationRatePheno}{\mu}
 \newcommand{\SubRate}{q}
@@
 The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.
+All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\citep{lynch_genetics_1998, walsh_evolution_2018}.
 
-Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\parencite{simons_population_2018, sella_thinking_2019}.
-Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{merila_comparison_2001, leinonen_comparative_2008}.
-A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation in different optimum trait value in the different populations~\parencite{lamy_qst_2012}.
+Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
+Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
+A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation in different optimum trait value in the different populations~\citep{lamy_qst_2012}.
 Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection.
-However, \QstFst\ methods have been found to require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
+However, \QstFst\ methods have been found to require many populations~\citep{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\citep{pujol_are_2008, edelaar_comparisons_2011}.
 Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.
 To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
-For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
+For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\citep{lande_genetic_1980, turelli_heritable_1984}.
 % #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
-Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\parencite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
-% For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
-% However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
+Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.
+% For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\citep{whitehead_neutral_2006, gilad_natural_2006}.
+% However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\citep{signor_evolution_2018, price_detecting_2022}.
 % are these studies using Gst/Fst on gene expression?
@@
 To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
-For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
+For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\citep{lande_genetic_1980, turelli_heritable_1984}.
 % #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
-Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for genes with large within-species variation~\parencite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
+Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for genes with large within-species variation~\citep{khaitovich_neutral_2004}.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.
 
 % Phylogenetics (mean trait evolution)
-Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
-As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
+Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{hansen_translating_1996}.
+As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
 Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
 Deviations from the assumptions of the Brownian process are however well known.
-When trait variation is constraint because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
-Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
-However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
+When trait variation is constraint because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{catalan_drift_2019}.
+Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
+However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
+First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
-Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.
@@
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-Within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
+Within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
 Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
@@
 From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species.
-From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
+From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
 % Basically, we are testing whether the variance of means is greater than the mean of variances
@@
 At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$:
+As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and $\RateMut$:
 \begin{gather}
-    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
+    \VarMutation = 2 \MutationRatePheno \Multiply \RateMut \label{eq:var-mutation}.
 \end{gather}
 
-While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
-For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
+While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\citep{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
+For a neutral trait at equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
 \begin{align}
     \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
-    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
+    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
 \end{align}
@@
 Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
-As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$):
+As shown in \citet{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and the effective population size ($\Ne$):
 \begin{gather}
-    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
+    \pi = 4 \Ne \Multiply \MutationRatePheno \label{eq:genetic-diversity}.
 \end{gather}
@@
 We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
-This ratio allows removing the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
+This ratio allows removing the effect of $\Ne$ and $\MutationRatePheno$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
 \begin{align}
     \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
-    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
+    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \RateMut}{4 \Ne \Multiply \MutationRatePheno} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
     & = \RateMut. \label{eq:rate-pop}
@@
 
-The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
+The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \citep{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
 \begin{align}
@@
 For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
-If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
-The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given \parencite{hansen_translating_1996} by:
+If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
+The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given \citep{hansen_translating_1996} by:
 \begin{align}
     \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
-    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+    & = 4 \Time \Multiply \MutationRatePheno \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
 \end{align}
@@
 Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
-The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
-We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
+The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \citep{kimura_probability_1962}.
+We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRatePheno$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{mccandlish_modeling_2014}, giving:
 \begin{align}
-    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
-    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
-    & = \MutationRate. \label{eq:substitution-rate}
+    \SubRate & = 2 \Ne \Multiply \MutationRatePheno \Multiply \pfix, \\
+    & = 2 \Ne  \Multiply \MutationRatePheno  \Multiply \dfrac{1}{2\Ne}, \\
+    & = \MutationRatePheno. \label{eq:substitution-rate}
 \end{align}
-That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.
+That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\citep{kimura_evolutionary_1968}.
 
@@
     d & = \Time \Multiply \SubRate \\
-    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
+    & = \Time \Multiply \MutationRatePheno \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
 \end{align}
 We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
-This ratio allows removing the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
+This ratio allows removing the effect of $\Time$ and $\MutationRatePheno$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
 \begin{align}
     \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
-    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
+    & = \frac{4 \Time \Multiply \MutationRatePheno \Multiply \RateMut}{4 \Time \Multiply \MutationRatePheno}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
     & = \RateMut \label{eq:rate-phy}.
@@
 We define a neutrality index $\NI = \RateBetween / \RateWhithin$ that will equal $1$ for a trait evolving neutrally.
-Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
+Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRatePheno$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
 Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
@@
 
-Based on the comparative framework that can account for phylogenetic inertia~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
+Based on the comparative framework that can account for phylogenetic inertia~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
 
@@
 The off-diagonal elements ($\Distance_{\Spi,\Spj} = \Distance_{\Spj,\Spi}$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
-The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
+The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\citep{omeara_testing_2006}:
 \begin{gather}
@@
 The Bayesian framework allows obtaining the posterior distribution of the neutrality index ($\EstNI$) for a given trait.
-Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
-Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
+Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree using the \textit{BayesCode} software~\citep{latrille_inferring_2021}.
+Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
@@
 Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
-For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with the mutation rate per generation $\MutationRate$.
+For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRatePheno \Multiply \NbrLoci $, with the mutation rate per generation $\MutationRatePheno$.
 From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from the root to leaves of $0.18$ and average genetic diversity of $0.00276$.
 We scaled parameters in our simulations to fit plausible values for mammals.
-We thus used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
+We thus used a mutation rate of $\MutationRatePheno=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
 
-The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which led to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
+The changes in log-$\MutationRatePheno$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which led to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
 An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
@@
 A random genetic drift was introduced by resampling individuals at each generation, with each parent having a probability of being sampled that was proportional to its fitness ($W$).
-Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
 More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
@@
 We analyzed a dataset of body and brain masses from mammals.
-The log-transformed values of body and brain masses were taken from \textcite{tsuboi_breakdown_2018}.
+The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
 We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
 We discarded species with only one representative sample.
-The mammalian nucleotide diversity was obtained from the Zoonomia project~\parencite{genereux_comparative_2020}, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
+The mammalian nucleotide diversity was obtained from the Zoonomia project~\citep{genereux_comparative_2020}, with nucleotide divergence obtained on a set of neutral markers in \citet{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \citet{wilder_contribution_2023}.
 
-We also analyzed a dataset of primate species, with the nucleotide variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
+We also analyzed a dataset of primate species, with the nucleotide variation obtained from \citet{kuderna_global_2023} and the quantitative trait variation also from \citet{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
 However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
@@
 Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
-In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
+In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\citep{hansen_stabilizing_1997}, which leads to $\NI > 1$.
 Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
@@
 The inference framework was first tested on independently simulated datasets matching an empirically relevant mammalian empirical regime (see Materials and Methods).
-Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
+Under constant population size ($\Ne$) and constant mutation rate ($\MutationRatePheno$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
 For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
-Once we introduced fluctuating $\Ne$ and $\MutationRate$ (Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
+Once we introduced fluctuating $\Ne$ and $\MutationRatePheno$ (Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
 For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) and none as diversifying selection.
@@
         Histogram of ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$ with $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and probabilities of $\EstNI$ being greater than $1$ (right).
-        Effective population size ($\Ne$) and mutation rate ($\MutationRate$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
+        Effective population size ($\Ne$) and mutation rate ($\MutationRatePheno$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
     }
@@
 Our results showed that our method significantly improved over currently available methods to detect selection acting on a trait at the phylogenetic scale.
-Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 Our approach could in theory be applied to detect stabilizing selection at the phylogenetic scale, but we showed that it did not have the statistical power to identify those cases.
 In contrast, we showed that our method was able to identify correctly cases of diversifying selection, which is a clear an improvement over current methods that model only mean trait value.
-Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\parencite{catalan_drift_2019}.
+Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
+For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\citep{catalan_drift_2019}.
 Our diversity index has the advantage to discriminate the alternative model of diversifying selection from the neutral case by comparing within- and between-species variation correctly normalized to remove confounding factors.
-Our approach is not the first one to normalize between-species variation to detect selection, but this was done by using within-species variations~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} and not estimates of neutral molecular divergence as done in our study.
-These studies have further compared their statistic across a pool of traits, which allowed them to identify outlier traits putatively under diversifying selection but without testing for selection on a single trait at a time~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
+Our approach is not the first one to normalize between-species variation to detect selection, but this was done by using within-species variations~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} and not estimates of neutral molecular divergence as done in our study.
+These studies have further compared their statistic across a pool of traits, which allowed them to identify outlier traits putatively under diversifying selection but without testing for selection on a single trait at a time~\citep{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
 Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departures from the null model of neutral evolution for a single test.
@@
 The main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
-In the context of within species variation, \QstFst\ tests have been developed to compare trait and sequence across several populations to test for selection~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
+In the context of within species variation, \QstFst\ tests have been developed to compare trait and sequence across several populations to test for selection~\citep{martin_multivariate_2008, leinonen_qst_2013}.
 Our neutrality index also used the genetic sequences from which nucleotide divergence and polymorphism are estimated.
 Although the sequences should be neutrally evolving, they do not have to be necessarily linked to the quantitative trait under study.
-Nucleotide variation allows normalizing for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRate$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-Thus our test avoids the estimation of the parameters, which are complex to correctly infer, and it also bypasses the estimation of divergence time, which was necessary in previous approaches~\parencite{walsh_evolution_2018}.
-But importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time were constant across the phylogenetic tree, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
-Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRate$ and generation time impacted similarly trait and sequence variation.
-The normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time, which canceled out in the ratio of trait variation $\EstNI$.
+Nucleotide variation allows normalizing for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRatePheno$) and generation time~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
+Thus our test avoids the estimation of the parameters, which are complex to correctly infer, and it also bypasses the estimation of divergence time, which was necessary in previous approaches~\citep{walsh_evolution_2018}.
+But importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$, $\MutationRatePheno$ and generation time were constant across the phylogenetic tree, an unmet assumption empirically~\citep{bergeron_evolution_2023, wilder_contribution_2023}.
+Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRatePheno$ and generation time impacted similarly trait and sequence variation.
+The normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, $\MutationRatePheno$ and generation time, which canceled out in the ratio of trait variation $\EstNI$.
 
@@
 Even though our test was developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provided insight into its behavior.
-First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations were used to determine the neutral expectation, and the inflation of divergence was compared to polymorphism within species.
-Second, because $\NI$ was compared to 1, our test ultimately bear analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
+First, we acknowledge that our test took inspiration from the \citet{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations were used to determine the neutral expectation, and the inflation of divergence was compared to polymorphism within species.
+Second, because $\NI$ was compared to 1, our test ultimately bear analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\citep{goldman_codonbased_1994, muse_likelihood_1994}.
 As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
 Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we could leverage the vast literature discussing and interpreting the results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ did not necessarily imply that the trait was effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ did not rule out diversifying selection, but rather that this diversifying selection was not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
+With this analogy in mind, we could leverage the vast literature discussing and interpreting the results of these tests and their pitfalls~\citep{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
+First, not rejecting the neutral null model of $\NI = 1$ did not necessarily imply that the trait was effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\citep{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ did not rule out diversifying selection, but rather that this diversifying selection was not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\citep{latrille_genes_2023}.
 By explicitly modeling stabilizing selection as a moving optimum, it would theoretically be possible to tease apart the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
@@
 Second, loci were assumed to contribute additively to the phenotype.
-Although the effects of dominance and epistasis is typically weak compared to the additive effects on the quantitative traits, their influence should be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
-Third, the genetic architecture of the trait was assumed to be constant across the phylogenetic tree, whereas it might actually be variable among individuals and species~\parencite{tung_genetic_2015, huber_conservatism_2015}.
-Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Although the effects of dominance and epistasis is typically weak compared to the additive effects on the quantitative traits, their influence should be assessed~\citep{hill_data_2008, crow_epistasis_2010}.
+Third, the genetic architecture of the trait was assumed to be constant across the phylogenetic tree, whereas it might actually be variable among individuals and species~\citep{tung_genetic_2015, huber_conservatism_2015}.
+Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\citep{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
 Finally, our Bayesian estimation could integrate uncertainty from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
@@
 
-\subsection*{Acknowledgements}
- We gratefully acknowledge the help of Nicolas Lartillot, Philippe Veber, Isabela Jeronimo do Ó, Anna Marcionetti, Julien Clavel, and Daniele Silvestro for their insightful discussions and Julien Joseph for his advice and reviews concerning this manuscript.
-
-\subsection*{Competing interests:}
-The authors declare no conflicts of interest.
-
-\subsection*{Data and materials availability:}
-The data that support the findings of this study are openly available in GitHub at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}.
-Snakemake pipeline, analysis scripts and documentation are available in the repository to replicate the study.
-
 %TC:ignore
-\printbibliography
+\bibliographystyle{natbib}
+\bibliography{references_bibtex}
 
 \newpage
-
-\noindent\usebox{\Maketitle}
 

### 2024-07-03 manuscript/main-revision-2.tex
Revision round 2 manuscript files.

--- 2024-03-13 manuscript/main-revision.tex
@@
 This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
-However, regardless of the strengths and weaknesses of each method\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.
+However, regardless of the strengths and weaknesses of each method~\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.
 
@@
 These changes ultimately result in different mean trait values across lineages.
-Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984}.
+Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984, felsenstein_phylogenies_1988a}.
 Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
@@
 Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{hansen_translating_1996}.
+Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, hansen_translating_1996}.
 As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
@@
 Deviations from the assumptions of the Brownian process are however well known.
-When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{catalan_drift_2019}.
+When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, catalan_drift_2019}.
 Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
@@
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
-Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
-Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Second, a better fit of a Brownian could be due to a trait evolving with a rate too low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
+And third, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.
@@
 At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+Across many species, within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within species variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
 Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
@@
 It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
-The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
+The ratio that we propose can be considered as a neutrality index for any quantitative trait ~\citep{lynch_rate_1990}, while articulating trait and nucleotide variation within and between species.
 Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
-From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary generalization to account for phylogenetic relationships between species.
+From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary analog to account for phylogenetic relationships between species.
 From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
@@
     \caption{
-        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
-        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
-        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
-        Under neutral evolution, $\EstNI$ is expected to be equal to one.
-        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
-        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
+        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\RateBetween$, which is normalized by nucleotide divergence.
+        Within species, the genetic variance allows the estimation of within-species trait variation, $\RateWhithin$, which is  normalized by nucleotide diversity.
+        $\NI$ is the ratio of $\RateBetween$ over $\RateWhithin$.
+        Under neutral evolution, $\NI$ is expected to be equal to one.
+        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\NI$ greater than one.
+        Under stabilizing selection, the trait is homogeneous between species, leading to $\NI$ smaller than one.
         Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
@@
 
-We denote $\EstNI$, $\EstRateBetween$ and $\EstRateWhithin$ the point estimates of respectively $\NI$, $\RateBetween$ and $\RateWhithin$.
+We hereby seek to obtain point estimates of $\RateBetween$, $\RateWhithin$ and ultimately $\NI$.
 For each species with data available, $\RateWhithin$ as defined in eq.~\ref{eq:def-rate-pheno-pop} can be seen as a replicate sample.
-Thus, $\EstRateWhithin$ can be obtained by averaging out across all the sampled species.
+Thus, $\RateWhithin$ can be obtained by averaging out across all the sampled species.
 On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}.
-Generally, $\EstRateWhithin$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
-As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\EstRateWhithin$.
+Generally, $\RateBetween$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
+As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\RateWhithin$.
 We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
@@
 
-Finally, between-species variation $\EstRateBetween$ is estimated as~\citep{omeara_testing_2006}:
+Finally, between-species variation $\RateBetween$ is estimated as~\citep{omeara_testing_2006}:
 \begin{gather}
-    \EstRateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
+    \RateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
 \label{eq:estimated-rate-phy}
@@
 The ratio of $\VarGeneticSpi$ over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
-Averaged across all species, we obtain the estimate $\EstRateWhithin$ as:
+Averaged across all species, we obtain the estimate $\RateWhithin$ as:
 \begin{gather}
-    \EstRateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}.
+    \RateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}.
 \label{eq:estimated-rate-pop}
@@
 \begin{gather}
-    \EstNI = \frac{\EstRateBetween}{\EstRateWhithin}. \label{eq:estimated-NI}
+    \NI = \frac{\RateBetween}{\RateWhithin}. \label{eq:estimated-NI}
 \end{gather}
@@
 
-In the previous section, $\EstNI$ is estimated independently for each trait of interest.
-Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree.
-Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
+In the previous section, $\NI$ is estimated independently for each trait of interest.
+Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree, since simultaneously estimating all $\RateBetween$ allows improving their estimation~\citep{adams_multivariate_2018}.
+More specifically, trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
 The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
 The branch lengths of the tree used to model the Brownian process runs is measured in units of $4d$ ($d$ being the nucleotide divergence).
-The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
+The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\RateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
+Of note, modeling trait evolution as a multi-dimensional process is reliable only if $\Ntrait \ll \NbrTaxa$, meaning that the number of species is largely superior to the number of traits~\citep{adams_multivariate_2018}.
+Thus, relying on a $\Ntrait$-dimensional process should be reserved for a handful of allometric traits (e.g. brain mass and body mass).
+If $\Ntrait$ is large, the traits are better tested independently each with a $1$-dimensional Brownian process, which is a specific case of the multi-dimensional process.
 
 \subsubsection*{Bayesian estimate}
-The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
+The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\NI$) for traits of interest.
 We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\citep{latrille_inferring_2021}.
@@
 For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
-Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\EstRateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
-From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
-The posterior distribution of $\EstNI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
+Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\RateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
+From $\RateWhithin$ estimated independently for each trait and the diagonal elements of $\CovarianceMatrix$ (i.e.
+the $\RateBetween$ for each trait), the posterior distribution of $\NI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
+The posterior distribution of $\NI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\NI > 1 ]$ to test for evidence of diversifying selection and $\proba [\NI < 1 ]$ to test for evidence of stabilizing selection.
 
@@
 Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
-Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
-It is, however, possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
+Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\NI$.
+It is, however, possible to test for diversifying selection because testing for $\NI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
+Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\NI$ since $\Heritability \leq H^2$ and thus can be used to detect diversifying selection if $\Heritability$ is not available.
 Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and fall within the range of 0.2-0.5 in a vast majority of phenotypic traits tested~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
-Alternatively, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\EstNI$ since $\Heritability \leq H^2$.
-If available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.
-
-In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\EstNI$ is underestimated.
+Thus, if available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.
+
+In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\NI$ is underestimated.
 Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
@@
 Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
-Neutral evolution was implemented by fixing the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.
+Neutral evolution was implemented by flattening the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.
 
@@
 However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
-As such, the evidence for $\EstNI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers included in the estimate of divergence can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
+As such, the evidence for $\NI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers included in the estimate of divergence can lead to a spurious $\NI > 1$ (see table~\ref{table:assumptions}).
 
@@
 Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
-This statistic was obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
-We also devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
+This statistic was obtained as a maximum likelihood estimate from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
+We also devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\NI > 1]$, and stabilizing selection as $\proba [\NI < 1]$.
 
@@
 The inference framework was first tested on independently simulated datasets matching an empirically relevant mammalian empirical regime (see Materials and Methods).
-Under constant population size ($\Ne$) and constant mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
-For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
+Under constant population size ($\Ne$) and constant mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\NI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\NI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
+For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\NI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
 Once we introduced fluctuating $\Ne$, $\MutationRatePheno$ and $\MutationRateNuc$(Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
-For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) and none as diversifying selection.
+For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\NI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\NI < 1] > 0.975$) and none as diversifying selection.
 
@@
         Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-        Histogram of ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$ with $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and probabilities of $\EstNI$ being greater than $1$ (right).
+        Histogram of ratio of between-species trait variation ($\RateBetween$) over within-species trait variation $\RateWhithin$ with $\NI = \RateBetween / \RateWhithin$ estimated from each simulated data (left) and probabilities of $\NI$ being greater than $1$ (right).
         Effective population size ($\Ne$) and mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
@@
         \toprule
-        Dataset & Trait & $\Heritability$ & Sex & $\NbrTaxa$ & $\EstNI$ & $95\%$ CI for $\EstNI$ & $\proba [\EstNI > 1 ]$ \\ \hline
+        Dataset & Trait & $\Heritability$ & Sex & $\NbrTaxa$ & $\NI$ & $95\%$ CI for $\NI$ & $\proba [\NI > 1 ]$ \\ \hline
         \midrule
@@
     $\NbrTaxa$ was the number of species in the dataset.
-    $\EstNI$ was the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\EstNI$ also computed.
-    $\proba [\EstNI > 1 ]$ was the estimated posterior probability of diversifying selection.
+    $\NI$ was the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\NI$ also computed.
+    $\proba [\NI > 1 ]$ was the estimated posterior probability of diversifying selection.
     }
@@
 For mammalian body and brain mass, we obtained male (\Male) and female (\Female) trait variations.
-Combined with nucleotide diversity and divergence, we estimated $\EstNI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in the Table~\ref{table:empirical}.
-Assuming complete heritability, brain mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
+Combined with nucleotide diversity and divergence, we estimated $\NI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in the Table~\ref{table:empirical}.
+Assuming complete heritability, body mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
 If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\citep{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
@@
 We also analyzed a similar dataset for body mass focusing this time only at Primates (Table~\ref{table:empirical}).
-For primates body mass, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%.
-Assuming complete heritability of body mass did not change the posterior probability for males, but increased the one for female to $1.0$.
-Evidence for diversifying selection on body mass was therefore more pronounced in Primates than in mammals.
-However, the genetic markers used to normalize trait variance with nucleotide divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\EstNI$ (Table~\ref{table:assumptions} and methods).
+For primates body mass, assuming complete heritability, body mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females, exactly as in the mammal dataset.
+However, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%.
+For brain mass, assuming complete heritability or not (between 20\% and 40\%) did not change the posterior probability of diversifying selection, which was $1.0$.
+Evidence for diversifying selection on both brain and body mass was therefore more pronounced in Primates than in mammals.
+However, the genetic markers used to normalize trait variance with nucleotide divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\NI$ (Table~\ref{table:assumptions} and methods).
 
@@
             \hline
-            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\EstNI > 1$ & Test $\EstNI < 1$ \\ \hline \hline
+            Broken assumption                                       & Consequences                                       & $\RateWhithin$   & $\RateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
             Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
@@
     \end{adjustbox}
-    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
+    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\RateWhithin$), between-species variation ($\RateBetween$), and on the neutrality index $\NI = \RateBetween/\RateWhithin$.
     The last two columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
@@
 Similarly, trait variation within species was normalized by sequence polymorphism obtained also from a neutral set of markers.
-Our estimate of $\EstNI$ could be tested for deviation from the value of $1.0$ expected under the null hypothesis of neutrality.
+Our estimate of $\NI$ could be tested for deviation from the value of $1.0$ expected under the null hypothesis of neutrality.
 Technically, the neutrality index can be estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S\ref{sec:implementation}).
-The latter also enabled the estimation of the posterior credible interval to test for departure from a neutrally evolving trait (e.g. $ \proba [ \EstNI > 1 ]$).
-We tested our statistical procedure against simulated data and showed that our test was able to correctly detect simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$).
-However, our test detected a spurious signal of stabilizing selection ($\EstNI < 1$) when we simulated the evolution of a neutral trait.
+The latter also enabled the estimation of the posterior credible interval to test for departure from a neutrally evolving trait (e.g. $ \proba [ \NI > 1 ]$).
+We tested our statistical procedure against simulated data and showed that our test was able to correctly detect simulations under diversifying selection (test of $\NI > 1$) or under stabilizing selection (test of $\NI < 1$).
+However, our test detected a spurious signal of stabilizing selection ($\NI < 1$) when we simulated the evolution of a neutral trait.
 An assumption of our test is that the neutral phenotypic trait is evolving as a Brownian process and is, therefore, unbounded.
-However, the phenotype is encoded by a genetic architecture, and is thus ultimately bounded, even more so if the trait is encoded by a few loci (see table~\ref{table:assumptions}).
-At the macro-evolutionary scale, phenotypic divergence should plateau at some point, resulting in a reduced between-species trait variation.
-We argue that this effect can result in a spurious signal of stabilizing selection ($\EstNI < 1$), especially for deeper phylogeny (figure~\ref{fig:supp-distance} and section~S\ref{sec:supp-distance})
+However, the phenotype may be bounded by what the genetic architecture can produce, and this could cause a slowdown of phenotypic divergence over time due to the erosion of possible phenotypic changes at the underlying loci.
+Typically, such an effect depends on the number of alleles per locus, whether new mutations are generating new alleles or instead reverting to previous alleles.
+Altogether, in our simulation setting under a constant genetic architecture with a fixed number of loci, such a slowdown of phenotypic divergence can result in a spurious signal of stabilizing selection ($\NI < 1$), especially for deeper phylogeny (figure~\ref{fig:supp-distance} and section~S\ref{sec:supp-distance})
 We thus argue that our method should be used to detect diversifying selection, but that it had low accuracy to detect stabilizing selection due to false positives.
@@
 However, this method does not formally allow testing for diversifying selection, and requires many traits such as expression level data to seek outliers genes~\citep{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
-Second, other methods leverage Lande’s generalized genetic distance (LGGD), which relate the ratio of between to within variations to population-genetic parameters~\citep{lynch_analysis_1990, lemos_evolutionary_2001, lemos_rates_2005, weaver_were_2007, porto_rate_2015}.
+Second, other methods leverage Lande’s generalized genetic distance (LGGD), which relate the ratio of between to within variations to population-genetic parameters~\citep{lande_quantitative_1979, lynch_analysis_1990, lemos_evolutionary_2001, lemos_rates_2005, weaver_were_2007, porto_rate_2015}.
 Specifically, by leveraging estimates of effective population size ($\Ne$) and number of generations between species, or alternatively by assuming their constancy, these methods can test for departures from the null model of neutral evolution for a single trait.
-Such methods have been successful in identifying specific instances of diversifying selection\citep{schroeder_evolution_2017, machado_preeminent_2022} and near-drift~\citep{machado_using_2023}.
+Such methods have been successful in identifying specific instances of diversifying selection~\citep{schroeder_evolution_2017, machado_preeminent_2022} and near-drift~\citep{machado_using_2023}.
 However, $\Ne$ and the number of generations are complex parameters to correctly infer, and is usually done for a pair or only a few species, and ultimately requires large genomic datasets and heavy statistical methods~\citep{wilder_contribution_2023}.
@@
 In this context, our test bears many similarities to \QstFst\ tests (and their derivatives) that have been developed to test for selection of a trait across several populations while also leveraging sequence variation~\citep{martin_multivariate_2008, leinonen_qst_2013} or co-ancestry between individuals~\citep{ovaskainen_new_2011}.
-Our method can be seen as an extension at the phylogenetic scale, where although the sequences used should be neutrally evolving, they can be obtained from different sampled individuals than for the trait.
+Our method can be seen as an analog at the phylogenetic scale, where although the sequences used should be neutrally evolving, they can be obtained from different sampled individuals than for the trait.
 Importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$ and mutation rates were constant across the phylogenetic tree, an unmet assumption empirically~\citep{bergeron_evolution_2023, wilder_contribution_2023}.
-Indeed, under the neutral case of evolution, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, generation time and mutation rates, which canceled out in the neutrality index $\EstNI$.
+Indeed, under the neutral case of evolution, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, generation time and mutation rates, which canceled out in the neutrality index $\NI$.
 
@@
 In the context of detecting diversifying selection on a trait, we argue that the main drawback of our method is that the additive genetic variance of the trait is required instead of the phenotypic variance.
-If phenotypic variance was used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ was ultimately underestimated.
-Similarly, using broad-sense heritability instead of narrow-sense heritability would result in underestimated $\EstNI$.
-In such context, the test of stabilizing selection ($\EstNI < 1]$) would be statistically invalid.
-However, the test of diversifying selection ($\EstNI > 1$) was underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
-As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
+If phenotypic variance was used instead of additive genetic variance to estimate $\NI$, meaning that we assumed complete heritability, the neutrality index $\NI$ was ultimately underestimated.
+Similarly, using broad-sense heritability instead of narrow-sense heritability would result in underestimated $\NI$.
+In such context, the test of stabilizing selection ($\NI < 1]$) would be statistically invalid.
+However, the test of diversifying selection ($\NI > 1$) was underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
+As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\NI > 1$.
 If available, any prior knowledge on heritability can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
+Additionally, phenotypic plasticity also affects the genotype-phenotype relationship with intricate consequences for our test of selection.
+First, at the level of within species variation, individuals might occupy different patches with different environments.
+Responding to these individual environmental conditions, phenotypic plasticity would then result in increased trait variation within species.
+In this scenario, as hypothesized in \citet{rohlfs_phylogenetic_2015}, phenotypic plasticity then leads to a reduced ratio of between to within species variations, thus ultimately leading to our tests of diversifying selection being underpowered although not invalid.
+Alternatively, it is also possible that different species are experiencing different macro-environments, for example with species spread along a latitudinal or elevation gradient, with different temperatures or precipitation.
+These species could thus have different mean phenotypes solely because of phenotypic plasticity, while such changes are not encoded in their genome~\citep{stamp_relative_2020, schraiber_heritability_2023}.
+Such an effect can lead to $\NI > 1$ erroneously interpreting diversifying selection.
+The test of $\NI > 1$ would however be correct that the changes in mean phenotypes across species is due to change in environment, albeit in such a case not encoded by the genotype of individuals but due to phenotypic plasticity.
 
@@
 From an empirical point of view, our method required integrating genomic and trait variation, which could reduce the possible datasets to be used.
-However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals' brain and body mass, showing signals of diversifying selection.
-The consensus on macro-evolutionary studies, assuming constancy of $\Ne$, generation time and mutation rates, is that empirical rates of evolution calculated on phylogenetic trees and the fossil record are far inferior to the expected under drift~\citep{lynch_analysis_1990, uyeda_millionyear_2011}.
+However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals’ brain and body mass, both showing signals of diversifying selection.
+As such, this result corroborates studies relying solely on changes in mean trait values across mammals, showing strong statistical support for several distinct evolutionary regimes for body- and brain mass~\citep{mitov_automatic_2019}.
+Interestingly, our strongest signal is for brain mass, corroborating studies in hominids where skull size (related to brain mass) is the only trait that exceeded the expected rate of phenotypic evolution under a neutral model~\citep{lynch_rate_1990}.
+Hence, one first interpretation here is that brain mass might be an exceptional case among many phenotypic traits (e.g. dental and skeletal measures).
+Second, from a macro-evolutionary perspective, the consensus is that empirical rates of evolution calculated on phylogenetic trees and the fossil record are far inferior to the expected under drift~\citep{lynch_analysis_1990, uyeda_millionyear_2011}, where such methods assume constancy of $\Ne$, generation time and mutation rates.
 Our finding of diversifying selection on body and brain mass could be seen as an argument against that interpretation.
@@
 One possible interpretation is that normalization by nucleotide divergence could absorb this observed slowing rate of evolution.
-Altogether, further empirical and theoretical studies are required to disentangle this discrepancy between these different interpretations.
+Altogether, further empirical and theoretical studies are required to disentangle this discrepancy between these different results and interpretations
 Because our test was also based on several assumptions that might not hold on empirical data, we also provided a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (Table~\ref{table:assumptions}).
-For example, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass was evolving under diversifying selection since the markers used for nucleotide divergences were not neutral, which can lead to a spurious $\EstNI > 1$.
+For example, at the primate scale, the evidence for $\NI > 1$ does not necessarily imply that the brain mass was evolving under diversifying selection since the markers used for nucleotide divergences were not neutral, which can lead to a spurious $\NI > 1$.
 In conclusion, our study provided a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we believe that our new approach is a promising tool to investigate the evolution of quantitative traits.
@@
 \bibliographystyle{natbib}
-\bibliography{references_bibtex}
+\begin{thebibliography}{}
+
+    \bibitem[Adams and Collyer(2018)Adams and Collyer]{adams_multivariate_2018}
+    Adams, D.~C. and Collyer, M.~L. 2018.
+    \newblock Multivariate {{Phylogenetic Comparative Methods}}: {{Evaluations}},
+        {{Comparisons}}, and {{Recommendations}}.
+    \newblock {\em Systematic Biology\/}, {67}(1): 14--31.
+
+    \bibitem[Anisimova and Kosiol(2009)Anisimova and
+    Kosiol]{anisimova_investigating_2009}
+    Anisimova, M. and Kosiol, C. 2009.
+    \newblock Investigating protein-coding sequence evolution with probabilistic
+    codon substitution models.
+    \newblock {\em Molecular Biology and Evolution\/}, {26}(2): 255--271.
+
+    \bibitem[Arnold {\em et~al.}(2008)Arnold, B{\"u}rger, Hohenlohe, Ajie, and
+    Jones]{arnold_understanding_2008}
+    Arnold, S.~J., B{\"u}rger, R., Hohenlohe, P.~A., Ajie, B.~C., and Jones, A.~G.
+    2008.
+    \newblock Understanding the evolution and stability of the {{G-Matrix}}.
+    \newblock {\em Evolution\/}, {62}(10): 2451--2461.
+
+    \bibitem[Barton {\em et~al.}(2017)Barton, Etheridge, and
+    V{\'e}ber]{barton_infinitesimal_2017}
+    Barton, N.~H., Etheridge, A.~M., and V{\'e}ber, A. 2017.
+    \newblock The infinitesimal model: {{Definition}}, derivation, and
+    implications.
+    \newblock {\em Theoretical Population Biology\/}, {118}: 50--73.
+
+    \bibitem[Bergeron {\em et~al.}(2023)Bergeron, Besenbacher, Zheng, Li,
+        Bertelsen, Quintard, Hoffman, Li, St.~Leger, Shao, Stiller, Gilbert,
+        Schierup, and Zhang]{bergeron_evolution_2023}
+    Bergeron, L.~A., Besenbacher, S., Zheng, J., Li, P., Bertelsen, M.~F.,
+    Quintard, B., Hoffman, J.~I., Li, Z., St.~Leger, J., Shao, C., Stiller, J.,
+    Gilbert, M. T.~P., Schierup, M.~H., and Zhang, G. 2023.
+    \newblock Evolution of the germline mutation rate across vertebrates.
+    \newblock {\em Nature\/}, pages 1--7.
+
+    \bibitem[Blanquart and Bataillon(2016)Blanquart and
+    Bataillon]{blanquart_epistasis_2016}
+    Blanquart, F. and Bataillon, T. 2016.
+    \newblock Epistasis and the structure of fitness landscapes: {{Are}}
+    experimental fitness landscapes compatible with fisher's geometric model?
+    \newblock {\em Genetics\/}, {203}(2): 847--862.
+
+    \bibitem[Catal{\'a}n {\em et~al.}(2019)Catal{\'a}n, Briscoe, and
+    H{\"o}hna]{catalan_drift_2019}
+    Catal{\'a}n, A., Briscoe, A.~D., and H{\"o}hna, S. 2019.
+    \newblock Drift and directional selection are the evolutionary forces driving
+    gene expression divergence in eye and brain tissue of {{Heliconius}}
+    butterflies.
+    \newblock {\em Genetics\/}, {213}(2): 581--594.
+
+    \bibitem[Cooper {\em et~al.}(2016)Cooper, Thomas, Venditti, Meade, and
+    Freckleton]{cooper_cautionary_2016}
+    Cooper, N., Thomas, G.~H., Venditti, C., Meade, A., and Freckleton, R.~P. 2016.
+    \newblock A cautionary note on the use of {{Ornstein Uhlenbeck}} models in
+    macroevolutionary studies.
+    \newblock {\em Biological Journal of the Linnean Society\/}, {118}(1): 64--77.
+
+    \bibitem[Crow(2010)Crow]{crow_epistasis_2010}
+    Crow, J.~F. 2010.
+    \newblock On epistasis: Why it is unimportant in polygenic directional
+    selection.
+    \newblock {\em Philosophical Transactions of the Royal Society B: Biological
+    Sciences\/}, {365}(1544): 1241--1244.
+
+    \bibitem[Edelaar {\em et~al.}(2011)Edelaar, Burraco, and
+        {Gomez-Mestre}]{edelaar_comparisons_2011}
+    Edelaar, P., Burraco, P., and {Gomez-Mestre}, I. 2011.
+    \newblock Comparisons between {{QST}} and {{FST}}---how wrong have we been?
+    \newblock {\em Molecular Ecology\/}, {20}(23): 4830--4839.
+
+    \bibitem[Felsenstein(1985)Felsenstein]{felsenstein_phylogenies_1985}
+    Felsenstein, J. 1985.
+    \newblock Phylogenies and the {{Comparative Method}}.
+    \newblock {\em The American Naturalist\/}, {125}(1): 1--15.
+
+    \bibitem[Felsenstein(1988)Felsenstein]{felsenstein_phylogenies_1988a}
+    Felsenstein, J. 1988.
+    \newblock Phylogenies and quantitative characters.
+    \newblock {\em Annual Review of Ecology and Systematics\/}, {19}(1): 445--471.
+
+    \bibitem[Felsenstein(2008)Felsenstein]{felsenstein_comparative_2008}
+    Felsenstein, J. 2008.
+    \newblock Comparative methods with sampling error and within-species variation:
+    Contrasts revisited and revised.
+    \newblock {\em The American Naturalist\/}, {171}(6): 713--725.
+
+    \bibitem[Foley {\em et~al.}(2023)Foley, Mason, Harris, Bredemeyer, Damas,
+        Lewin, Eizirik, Gatesy, Karlsson, {Lindblad-Toh}, {Zoonomia Consortium},
+        Springer, and Murphy]{foley_genomic_2023}
+    Foley, N.~M., Mason, V.~C., Harris, A.~J., Bredemeyer, K.~R., Damas, J., Lewin,
+    H.~A., Eizirik, E., Gatesy, J., Karlsson, E.~K., {Lindblad-Toh}, K.,
+        {Zoonomia Consortium}, Springer, M.~S., and Murphy, W.~J. 2023.
+    \newblock A genomic timescale for placental mammal evolution.
+    \newblock {\em Science\/}, {380}(6643): eabl8189.
+
+    \bibitem[Fraser(2020)Fraser]{fraser_detecting_2020}
+    Fraser, H.~B. 2020.
+    \newblock Detecting selection with a genetic cross.
+    \newblock {\em Proceedings of the National Academy of Sciences\/}, {117}(36):
+    22323--22330.
+
+    \bibitem[Gaboriau {\em et~al.}(2020)Gaboriau, Mendes, Joly, Silvestro, and
+    Salamin]{gaboriau_multiplatform_2020}
+    Gaboriau, T., Mendes, F.~K., Joly, S., Silvestro, D., and Salamin, N. 2020.
+    \newblock A multi-platform package for the analysis of intra- and interspecific
+    trait evolution.
+    \newblock {\em Methods in Ecology and Evolution\/}, {11}(11): 1439--1447.
+
+    \bibitem[Gaboriau {\em et~al.}(2023)Gaboriau, Tobias, Silvestro, and
+    Salamin]{gaboriau_exploring_2023}
+    Gaboriau, T., Tobias, J.~A., Silvestro, D., and Salamin, N. 2023.
+    \newblock Exploring the {{Macroevolutionary Signature}} of {{Asymmetric
+    Inheritance}} at {{Speciation}}.
+
+    \bibitem[Genereux {\em et~al.}(2020)Genereux, Serres, Armstrong, Johnson,
+        Marinescu, Mur{\'e}n, Juan, Bejerano, Casewell, Chemnick, Damas, Di~Palma,
+        Diekhans, Fiddes, Garber, Gladyshev, Goodman, Haerty, Houck, Hubley, Kivioja,
+        Koepfli, Kuderna, Lander, Meadows, Murphy, Nash, Noh, Nweeia, Pfenning,
+        Pollard, Ray, Shapiro, Smit, Springer, Steiner, Swofford, Taipale, Teeling,
+            {Turner-Maier}, Alfoldi, Birren, Ryder, Lewin, Paten, {Marques-Bonet},
+            {Lindblad-Toh}, Karlsson, and {Zoonomia
+        Consortium}]{genereux_comparative_2020}
+    Genereux, D.~P., Serres, A., Armstrong, J., Johnson, J., Marinescu, V.~D.,
+    Mur{\'e}n, E., Juan, D., Bejerano, G., Casewell, N.~R., Chemnick, L.~G.,
+    Damas, J., Di~Palma, F., Diekhans, M., Fiddes, I.~T., Garber, M., Gladyshev,
+    V.~N., Goodman, L., Haerty, W., Houck, M.~L., Hubley, R., Kivioja, T.,
+    Koepfli, K.-P., Kuderna, L. F.~K., Lander, E.~S., Meadows, J. R.~S., Murphy,
+    W.~J., Nash, W., Noh, H.~J., Nweeia, M., Pfenning, A.~R., Pollard, K.~S.,
+    Ray, D.~A., Shapiro, B., Smit, A. F.~A., Springer, M.~S., Steiner, C.~C.,
+    Swofford, R., Taipale, J., Teeling, E.~C., {Turner-Maier}, J., Alfoldi, J.,
+    Birren, B., Ryder, O.~A., Lewin, H.~A., Paten, B., {Marques-Bonet}, T.,
+        {Lindblad-Toh}, K., Karlsson, E.~K., and {Zoonomia Consortium} 2020.
+    \newblock A comparative genomics multitool for scientific discovery and
+    conservation.
+    \newblock {\em Nature\/}, {587}(7833): 240--245.
+
+    \bibitem[Gillard {\em et~al.}(2021)Gillard, Gr{\o}nvold, R{\o}s{\ae}g, Holen,
+        Monsen, Koop, Rondeau, Gundappa, Mendoza, Macqueen, Rohlfs, Sandve, and
+        Hvidsten]{gillard_comparative_2021}
+    Gillard, G.~B., Gr{\o}nvold, L., R{\o}s{\ae}g, L.~L., Holen, M.~M., Monsen,
+        {\O}., Koop, B.~F., Rondeau, E.~B., Gundappa, M.~K., Mendoza, J., Macqueen,
+    D.~J., Rohlfs, R.~V., Sandve, S.~R., and Hvidsten, T.~R. 2021.
+    \newblock Comparative regulomics supports pervasive selection on gene dosage
+    following whole genome duplication.
+    \newblock {\em Genome Biology\/}, {22}(1): 103.
+
+    \bibitem[Goldman and Yang(1994)Goldman and Yang]{goldman_codonbased_1994}
+    Goldman, N. and Yang, Z. 1994.
+    \newblock A codon-based model of nucleotide substitution for protein-coding
+        {{DNA}} sequences.
+    \newblock {\em Molecular Biology and Evolution\/}, {11}(5): 725--736.
+
+    \bibitem[Grabowski {\em et~al.}(2023)Grabowski, Pienaar, Voje, Andersson,
+            {Fuentes-Gonz{\'a}lez}, Kopperud, Moen, Tsuboi, Uyeda, and
+        Hansen]{grabowski_cautionary_2023}
+    Grabowski, M., Pienaar, J., Voje, K.~L., Andersson, S., {Fuentes-Gonz{\'a}lez},
+    J., Kopperud, B.~T., Moen, D.~S., Tsuboi, M., Uyeda, J., and Hansen, T.~F.
+    2023.
+    \newblock A {{Cautionary Note}} on ``{{A Cautionary Note}} on the {{Use}} of
+        {{Ornstein Uhlenbeck Models}} in {{Macroevolutionary Studies}}''.
+    \newblock {\em Systematic Biology\/}, {72}(4): 955--963.
+
+    \bibitem[Hansen(1997)Hansen]{hansen_stabilizing_1997}
+    Hansen, T.~F. 1997.
+    \newblock Stabilizing selection and the comparative analysis of adaptation.
+    \newblock {\em Evolution\/}, {51}(5): 1341--1351.
+
+    \bibitem[Hansen and Bartoszek(2012)Hansen and
+    Bartoszek]{hansen_interpreting_2012}
+    Hansen, T.~F. and Bartoszek, K. 2012.
+    \newblock Interpreting the evolutionary regression: The interplay between
+    observational and biological errors in phylogenetic comparative studies.
+    \newblock {\em Systematic Biology\/}, {61}(3): 413--425.
+
+    \bibitem[Hansen and Martins(1996)Hansen and Martins]{hansen_translating_1996}
+    Hansen, T.~F. and Martins, E.~P. 1996.
+    \newblock Translating between microevolutionary process and macroevolutionary
+    patterns: The correlation structure of interspecific data.
+    \newblock {\em Evolution\/}, {50}(4): 1404--1417.
+
+    \bibitem[Hansen and P{\'e}labon(2021)Hansen and
+    P{\'e}labon]{hansen_evolvability_2021}
+    Hansen, T.~F. and P{\'e}labon, C. 2021.
+    \newblock Evolvability: {{A Quantitative-Genetics Perspective}}.
+    \newblock {\em Annual Review of Ecology, Evolution, and Systematics\/},
+        {52}(1): 153--175.
+
+    \bibitem[Hansen {\em et~al.}(2011)Hansen, P{\'e}labon, and
+    Houle]{hansen_heritability_2011}
+    Hansen, T.~F., P{\'e}labon, C., and Houle, D. 2011.
+    \newblock Heritability is not {{Evolvability}}.
+    \newblock {\em Evolutionary Biology\/}, {38}(3): 258--277.
+
+    \bibitem[Harmon(2018)Harmon]{harmon_phylogenetic_2018}
+    Harmon, L. 2018.
+    \newblock Phylogenetic comparative methods: Learning from trees.
+
+    \bibitem[Hill {\em et~al.}(2008)Hill, Goddard, and Visscher]{hill_data_2008}
+    Hill, W.~G., Goddard, M.~E., and Visscher, P.~M. 2008.
+    \newblock Data and {{Theory Point}} to {{Mainly Additive Genetic Variance}} for
+        {{Complex Traits}}.
+    \newblock {\em PLOS Genetics\/}, {4}(2): e1000008.
+
+    \bibitem[Hohenlohe and Arnold(2008)Hohenlohe and Arnold]{hohenlohe_mipod_2008}
+    Hohenlohe, P.~A. and Arnold, S.~J. 2008.
+    \newblock {{MIPoD}}: {{A}} hypothesis-testing framework for microevolutionary
+    inference from patterns of divergence.
+    \newblock {\em The American Naturalist\/}, {171}(3): 366--385.
+
+    \bibitem[Hu {\em et~al.}(2022)Hu, Park, and Reecy]{hu_bringing_2022}
+    Hu, Z.-L., Park, C.~A., and Reecy, J.~M. 2022.
+    \newblock Bringing the {{Animal QTLdb}} and {{CorrDB}} into the future: Meeting
+    new challenges and providing updated services.
+    \newblock {\em Nucleic Acids Research\/}, {50}(D1): D956--D961.
+
+    \bibitem[Huber {\em et~al.}(2015)Huber, Whibley, Poul, Navarro, Martin, Baxter,
+        Shah, Gilles, Wirth, McMillan, and Joron]{huber_conservatism_2015}
+    Huber, B., Whibley, A., Poul, Y.~L., Navarro, N., Martin, A., Baxter, S., Shah,
+    A., Gilles, B., Wirth, T., McMillan, W.~O., and Joron, M. 2015.
+    \newblock Conservatism and novelty in the genetic architecture of adaptation in
+        {{Heliconius}} butterflies.
+    \newblock {\em Heredity\/}, {114}(5): 515--524.
+
+    \bibitem[Huelsenbeck and Rannala(2003)Huelsenbeck and
+    Rannala]{huelsenbeck_detecting_2003}
+    Huelsenbeck, J.~P. and Rannala, B. 2003.
+    \newblock Detecting correlation between characters in a comparative analysis
+    with uncertain phylogeny.
+    \newblock {\em Evolution\/}, {57}(6): 1237--1247.
+
+    \bibitem[Ingram and Mahler(2013)Ingram and Mahler]{ingram_surface_2013}
+    Ingram, T. and Mahler, D. 2013.
+    \newblock {{SURFACE}}: Detecting convergent evolution from comparative data by
+    fitting {{Ornstein-Uhlenbeck}} models with stepwise {{Akaike Information
+    Criterion}}.
+    \newblock {\em Methods in Ecology and Evolution\/}, {4}(5): 416--425.
+
+    \bibitem[Jensen {\em et~al.}(2019)Jensen, Payseur, Stephan, Aquadro, Lynch,
+        Charlesworth, and Charlesworth]{jensen_importance_2019}
+    Jensen, J.~D., Payseur, B.~A., Stephan, W., Aquadro, C.~F., Lynch, M.,
+    Charlesworth, D., and Charlesworth, B. 2019.
+    \newblock The importance of the {{Neutral Theory}} in 1968 and 50 years on:
+        {{A}} response to {{Kern}} and {{Hahn}} 2018.
+    \newblock {\em Evolution\/}, {73}(1): 111--114.
+
+    \bibitem[Khabbazian {\em et~al.}(2016)Khabbazian, Kriebel, Rohe, and
+    An{\'e}]{khabbazian_fast_2016}
+    Khabbazian, M., Kriebel, R., Rohe, K., and An{\'e}, C. 2016.
+    \newblock Fast and accurate detection of evolutionary shifts in
+        {{Ornstein}}--{{Uhlenbeck}} models.
+    \newblock {\em Methods in Ecology and Evolution\/}, {7}(7): 811--824.
+
+    \bibitem[Khaitovich {\em et~al.}(2004)Khaitovich, Weiss, Lachmann, Hellmann,
+        Enard, Muetzel, Wirkner, Ansorge, and P{\"a}{\"a}bo]{khaitovich_neutral_2004}
+    Khaitovich, P., Weiss, G., Lachmann, M., Hellmann, I., Enard, W., Muetzel, B.,
+    Wirkner, U., Ansorge, W., and P{\"a}{\"a}bo, S. 2004.
+    \newblock A neutral model of transcriptome evolution.
+    \newblock {\em PLOS Biology\/}, {2}(5): e132.
+
+    \bibitem[Kimura(1962)Kimura]{kimura_probability_1962}
+    Kimura, M. 1962.
+    \newblock On the probability of fixation of mutant genes in a population.
+    \newblock {\em Genetics\/}, {47}(6): 713--719.
+
+    \bibitem[Kimura(1968)Kimura]{kimura_evolutionary_1968}
+    Kimura, M. 1968.
+    \newblock Evolutionary rate at the molecular level.
+    \newblock {\em Nature\/}, {217}(5129): 624--626.
+
+    \bibitem[Kostikova {\em et~al.}(2016)Kostikova, Silvestro, Pearman, and
+    Salamin]{kostikova_bridging_2016}
+    Kostikova, A., Silvestro, D., Pearman, P.~B., and Salamin, N. 2016.
+    \newblock Bridging inter- and intraspecific trait evolution with a hierarchical
+    bayesian approach.
+    \newblock {\em Systematic Biology\/}, {65}(3): 417--431.
+
+    \bibitem[Kuderna {\em et~al.}(2023)Kuderna, Gao, Janiak, Kuhlwilm, Orkin,
+        Bataillon, Manu, Valenzuela, Bergman, Rousselle, Silva, Agueda, Blanc, Gut,
+            {de Vries}, Goodhead, Harris, Raveendran, Jensen, Chuma, Horvath, Hvilsom,
+        Juan, Frandsen, Schraiber, {de Melo}, Bertuol, Byrne, Sampaio, Farias,
+        Valsecchi, Messias, {da Silva}, Trivedi, Rossi, Hrbek, Andriaholinirina,
+        Rabarivola, Zaramody, Jolly, {Phillips-Conroy}, Wilkerson, Abee, Simmons,
+            {Fernandez-Duque}, Kanthaswamy, Shiferaw, Wu, Zhou, Shao, Zhang, Keyyu,
+        Knauf, Le, Lizano, Merker, Navarro, Nadler, Khor, Lee, Tan, Lim, Kitchener,
+        Zinner, Gut, Melin, Guschanski, Schierup, Beck, Umapathy, Roos, Boubli,
+        Rogers, Farh, and Marques~Bonet]{kuderna_global_2023}
+    Kuderna, L. F.~K., Gao, H., Janiak, M.~C., Kuhlwilm, M., Orkin, J.~D.,
+    Bataillon, T., Manu, S., Valenzuela, A., Bergman, J., Rousselle, M., Silva,
+    F.~E., Agueda, L., Blanc, J., Gut, M., {de Vries}, D., Goodhead, I., Harris,
+    R.~A., Raveendran, M., Jensen, A., Chuma, I.~S., Horvath, J.~E., Hvilsom, C.,
+    Juan, D., Frandsen, P., Schraiber, J.~G., {de Melo}, F.~R., Bertuol, F.,
+    Byrne, H., Sampaio, I., Farias, I., Valsecchi, J., Messias, M., {da Silva},
+    M. N.~F., Trivedi, M., Rossi, R., Hrbek, T., Andriaholinirina, N.,
+    Rabarivola, C.~J., Zaramody, A., Jolly, C.~J., {Phillips-Conroy}, J.,
+    Wilkerson, G., Abee, C., Simmons, J.~H., {Fernandez-Duque}, E., Kanthaswamy,
+    S., Shiferaw, F., Wu, D., Zhou, L., Shao, Y., Zhang, G., Keyyu, J.~D., Knauf,
+    S., Le, M.~D., Lizano, E., Merker, S., Navarro, A., Nadler, T., Khor, C.~C.,
+    Lee, J., Tan, P., Lim, W.~K., Kitchener, A.~C., Zinner, D., Gut, I., Melin,
+    A.~D., Guschanski, K., Schierup, M.~H., Beck, R. M.~D., Umapathy, G., Roos,
+    C., Boubli, J.~P., Rogers, J., Farh, K. K.-H., and Marques~Bonet, T. 2023.
+    \newblock A global catalog of whole-genome diversity from 233 primate species.
+    \newblock {\em Science\/}, {380}(6648): 906--913.
+
+    \bibitem[Lamy {\em et~al.}(2012)Lamy, Plomion, Kremer, and
+    Delzon]{lamy_qst_2012}
+    Lamy, J.-B., Plomion, C., Kremer, A., and Delzon, S. 2012.
+    \newblock {{QST}} {$<$} {{FST As}} a signature of canalization.
+    \newblock {\em Molecular Ecology\/}, {21}(23): 5646--5655.
+
+    \bibitem[Lande(1979)Lande]{lande_quantitative_1979}
+    Lande, R. 1979.
+    \newblock Quantitative genetic analysis of multivariate evolution, applied to
+    brain: Body size allometry.
+    \newblock {\em Evolution\/}, {33}(1): 402--416.
+
+    \bibitem[Lande(1980a)Lande]{lande_genetic_1980}
+    Lande, R. 1980a.
+    \newblock Genetic variation and phenotypic evolution during allopatric
+    speciation.
+    \newblock {\em The American Naturalist\/}, {116}(4): 463--479.
+
+    \bibitem[Lande(1980b)Lande]{lande_sexual_1980}
+    Lande, R. 1980b.
+    \newblock Sexual dimorphism, sexual selection, and adaptation in polygenic
+    characters.
+    \newblock {\em Evolution\/}, {34}(2): 292--305.
+
+    \bibitem[Lartillot and Delsuc(2012)Lartillot and Delsuc]{lartillot_joint_2012}
+    Lartillot, N. and Delsuc, F. 2012.
+    \newblock Joint reconstruction of divergence times and life-history evolution
+    in placental mammals using a phylogenetic covariance model.
+    \newblock {\em Evolution\/}, {66}(6): 1773--1787.
+
+    \bibitem[Lartillot and Poujol(2011)Lartillot and
+    Poujol]{lartillot_phylogenetic_2011}
+    Lartillot, N. and Poujol, R. 2011.
+    \newblock A phylogenetic model for investigating correlated evolution of
+    substitution rates and continuous phenotypic characters.
+    \newblock {\em Molecular Biology and Evolution\/}, {28}(1): 729--744.
+
+    \bibitem[Latrille {\em et~al.}(2021)Latrille, Lanore, and
+    Lartillot]{latrille_inferring_2021}
+    Latrille, T., Lanore, V., and Lartillot, N. 2021.
+    \newblock Inferring long-term effective population size with
+    mutation--selection models.
+    \newblock {\em Molecular Biology and Evolution\/}, {38}(10): 4573--4587.
+
+    \bibitem[Latrille {\em et~al.}(2023)Latrille, Rodrigue, and
+    Lartillot]{latrille_genes_2023}
+    Latrille, T., Rodrigue, N., and Lartillot, N. 2023.
+    \newblock Genes and sites under adaptation at the phylogenetic scale also
+    exhibit adaptation at the population-genetic scale.
+    \newblock {\em Proceedings of the National Academy of Sciences of the United
+    States of America\/}, {120}(11): e2214977120.
+
+    \bibitem[Leinonen {\em et~al.}(2008)Leinonen, O'hara, Cano, and
+    Meril{\"a}]{leinonen_comparative_2008}
+    Leinonen, T., O'hara, R.~B., Cano, J.~M., and Meril{\"a}, J. 2008.
+    \newblock Comparative studies of quantitative trait and neutral marker
+    divergence: A meta-analysis.
+    \newblock {\em Journal of Evolutionary Biology\/}, {21}(1): 1--17.
+
+    \bibitem[Leinonen {\em et~al.}(2013)Leinonen, McCairns, O'Hara, and
+    Meril{\"a}]{leinonen_qst_2013}
+    Leinonen, T., McCairns, R. J.~S., O'Hara, R.~B., and Meril{\"a}, J. 2013.
+    \newblock {{QST}}--{{FST}} comparisons: Evolutionary and ecological insights
+    from genomic heterogeneity.
+    \newblock {\em Nature Reviews Genetics\/}, {14}(3): 179--190.
+
+    \bibitem[Lemos {\em et~al.}(2001)Lemos, Marroig, and
+    Cerqueira]{lemos_evolutionary_2001}
+    Lemos, B., Marroig, G., and Cerqueira, R. 2001.
+    \newblock Evolutionary rates and stabilizing selection in large-bodied opossum
+    skulls ({{Didelphimorphia}}: {{Didelphidae}}).
+    \newblock {\em Journal of Zoology\/}, {255}(2): 181--189.
+
+    \bibitem[Lemos {\em et~al.}(2005)Lemos, Meiklejohn, C{\'a}ceres, and
+    Hartl]{lemos_rates_2005}
+    Lemos, B., Meiklejohn, C.~D., C{\'a}ceres, M., and Hartl, D.~L. 2005.
+    \newblock Rates of {{Divergence}} in {{Gene Expression Profiles}} of
+        {{Primates}}, {{Mice}}, and {{Flies}}: {{Stabilizing Selection}} and
+        {{Variability Among Functional Categories}}.
+    \newblock {\em Evolution\/}, {59}(1): 126--137.
+
+    \bibitem[Litsios and Salamin(2012)Litsios and Salamin]{litsios_effects_2012}
+    Litsios, G. and Salamin, N. 2012.
+    \newblock Effects of {{Phylogenetic Signal}} on {{Ancestral State
+    Reconstruction}}.
+    \newblock {\em Systematic Biology\/}, {61}(3): 533--538.
+
+    \bibitem[Louca and Pennell(2020)Louca and Pennell]{louca_extant_2020}
+    Louca, S. and Pennell, M.~W. 2020.
+    \newblock Extant timetrees are consistent with a myriad of diversification
+    histories.
+    \newblock {\em Nature\/}, {580}(7804): 502--505.
+
+    \bibitem[Lynch(1990)Lynch]{lynch_rate_1990}
+    Lynch, M. 1990.
+    \newblock The {{Rate}} of {{Morphological Evolution}} in {{Mammals}} from the
+        {{Standpoint}} of the {{Neutral Expectation}}.
+    \newblock {\em The American Naturalist\/}, {136}(6): 727--741.
+
+    \bibitem[Lynch(1991)Lynch]{lynch_methods_1991}
+    Lynch, M. 1991.
+    \newblock Methods for the analysis of comparative data in evolutionary biology.
+    \newblock {\em Evolution\/}, {45}(5): 1065--1080.
+
+    \bibitem[Lynch and Crease(1990)Lynch and Crease]{lynch_analysis_1990}
+    Lynch, M. and Crease, T.~J. 1990.
+    \newblock The analysis of population survey data on {{DNA}} sequence variation.
+    \newblock {\em Molecular Biology and Evolution\/}, {7}(4): 377--394.
+
+    \bibitem[Lynch and Walsh(1998)Lynch and Walsh]{lynch_genetics_1998}
+    Lynch, M. and Walsh, B. 1998.
+    \newblock {\em Genetics and Analysis of Quantitative Traits\/}, volume~1.
+    \newblock Sinauer Sunderland, MA.
+
+    \bibitem[Lynch {\em et~al.}(1998)Lynch, Latta, Hicks, and
+    Giorgianni]{lynch_mutation_1998}
+    Lynch, M., Latta, L., Hicks, J., and Giorgianni, M. 1998.
+    \newblock Mutation, selection, and the maintenance of life-history variation in
+    a natural population.
+    \newblock {\em Evolution\/}, {52}(3): 727--733.
+
+    \bibitem[Machado {\em et~al.}(2022)Machado, Marroig, and
+    Hubbe]{machado_preeminent_2022}
+    Machado, F.~A., Marroig, G., and Hubbe, A. 2022.
+    \newblock The pre-eminent role of directional selection in generating extreme
+    morphological change in glyptodonts ({{Cingulata}}; {{Xenarthra}}).
+    \newblock {\em Proceedings of the Royal Society B: Biological Sciences\/},
+        {289}(1967): 20212521.
+
+    \bibitem[Machado {\em et~al.}(2023)Machado, Mongle, Slater, Penna, Wisniewski,
+        Soffin, Dutra, and Uyeda]{machado_using_2023}
+    Machado, F.~A., Mongle, C.~S., Slater, G., Penna, A., Wisniewski, A., Soffin,
+    A., Dutra, V., and Uyeda, J.~C. 2023.
+    \newblock Using developmental rules to align microevolution with
+    macroevolution.
+
+    \bibitem[Martin {\em et~al.}(2008)Martin, Chapuis, and
+    Goudet]{martin_multivariate_2008}
+    Martin, G., Chapuis, E., and Goudet, J. 2008.
+    \newblock Multivariate {{Qst}}--{{Fst Comparisons}}: {{A Neutrality Test}} for
+    the {{Evolution}} of the {{G Matrix}} in {{Structured Populations}}.
+    \newblock {\em Genetics\/}, {180}(4): 2135--2149.
+
+    \bibitem[McCandlish and Stoltzfus(2014)McCandlish and
+    Stoltzfus]{mccandlish_modeling_2014}
+    McCandlish, D.~M. and Stoltzfus, A. 2014.
+    \newblock Modeling evolution using the probability of fixation: {{History}} and
+    implications.
+    \newblock {\em Quarterly Review of Biology\/}, {89}(3): 225--252.
+
+    \bibitem[McDonald and Kreitman(1991)McDonald and
+    Kreitman]{mcdonald_adaptative_1991}
+    McDonald, J.~H. and Kreitman, M. 1991.
+    \newblock Adaptative protein evolution at {{Adh}} locus in {{Drosophila}}.
+    \newblock {\em Nature\/}, {351}(6328): 652--654.
+
+    \bibitem[Meril{\"a} and Crnokrak(2001)Meril{\"a} and
+    Crnokrak]{merila_comparison_2001}
+    Meril{\"a}, J. and Crnokrak, P. 2001.
+    \newblock Comparison of genetic differentiation at marker loci and quantitative
+    traits.
+    \newblock {\em Journal of Evolutionary Biology\/}, {14}(6): 892--903.
+
+    \bibitem[Mitov {\em et~al.}(2019)Mitov, Bartoszek, and
+    Stadler]{mitov_automatic_2019}
+    Mitov, V., Bartoszek, K., and Stadler, T. 2019.
+    \newblock Automatic generation of evolutionary hypotheses using mixed
+        {{Gaussian}} phylogenetic models.
+    \newblock {\em Proceedings of the National Academy of Sciences\/}, {116}(34):
+    16921--16926.
+
+    \bibitem[Mitov {\em et~al.}(2020)Mitov, Bartoszek, Asimomitis, and
+    Stadler]{mitov_fast_2020}
+    Mitov, V., Bartoszek, K., Asimomitis, G., and Stadler, T. 2020.
+    \newblock Fast likelihood calculation for multivariate {{Gaussian}}
+    phylogenetic models with shifts.
+    \newblock {\em Theoretical Population Biology\/}, {131}: 66--78.
+
+    \bibitem[Muse and Gaut(1994)Muse and Gaut]{muse_likelihood_1994}
+    Muse, S.~V. and Gaut, B.~S. 1994.
+    \newblock A likelihood approach for comparing synonymous and nonsynonymous
+    nucleotide substitution rates, with application to the chloroplast genome.
+    \newblock {\em Molecular Biology and Evolution\/}, {1}(5): 715--724.
+
+    \bibitem[Nielsen(2005)Nielsen]{nielsen_molecular_2005}
+    Nielsen, R. 2005.
+    \newblock Molecular signatures of natural selection.
+    \newblock {\em Annual Review of Genetics\/}, {39}(1): 197--218.
+
+    \bibitem[O'Meara {\em et~al.}(2006)O'Meara, An{\'e}, Sanderson, and
+    Wainwright]{omeara_testing_2006}
+    O'Meara, B.~C., An{\'e}, C., Sanderson, M.~J., and Wainwright, P.~C. 2006.
+    \newblock Testing for {{Different Rates}} of {{Continuous Trait Evolution Using
+    Likelihood}}.
+    \newblock {\em Evolution\/}, {60}(5): 922--933.
+
+    \bibitem[Ovaskainen {\em et~al.}(2011)Ovaskainen, Karhunen, Zheng, Arias, and
+    Meril{\"a}]{ovaskainen_new_2011}
+    Ovaskainen, O., Karhunen, M., Zheng, C., Arias, J. M.~C., and Meril{\"a}, J.
+    2011.
+    \newblock A {{New Method}} to {{Uncover Signatures}} of {{Divergent}} and
+        {{Stabilizing Selection}} in {{Quantitative Traits}}.
+    \newblock {\em Genetics\/}, {189}(2): 621--632.
+
+    \bibitem[Pennell {\em et~al.}(2015)Pennell, FitzJohn, Cornwell, and
+    Harmon]{pennell_model_2015}
+    Pennell, M.~W., FitzJohn, R.~G., Cornwell, W.~K., and Harmon, L.~J. 2015.
+    \newblock Model {{Adequacy}} and the {{Macroevolution}} of {{Angiosperm
+    Functional Traits}}.
+    \newblock {\em The American Naturalist\/}, {186}(2): E33--E50.
+
+    \bibitem[Porto {\em et~al.}(2015)Porto, Sebasti{\~a}o, Pavan, VandeBerg,
+        Marroig, and Cheverud]{porto_rate_2015}
+    Porto, A., Sebasti{\~a}o, H., Pavan, S.~E., VandeBerg, J.~L., Marroig, G., and
+    Cheverud, J.~M. 2015.
+    \newblock Rate of evolutionary change in cranial morphology of the marsupial
+    genus {{Monodelphis}} is constrained by the availability of additive genetic
+    variation.
+    \newblock {\em Journal of Evolutionary Biology\/}, {28}(4): 973--985.
+
+    \bibitem[Price {\em et~al.}(2022)Price, Palmer~Droguett, Taylor, Kim, Place,
+        Rogers, Mank, Cooney, and Wright]{price_detecting_2022}
+    Price, P.~D., Palmer~Droguett, D.~H., Taylor, J.~A., Kim, D.~W., Place, E.~S.,
+    Rogers, T.~F., Mank, J.~E., Cooney, C.~R., and Wright, A.~E. 2022.
+    \newblock Detecting signatures of selection on gene expression.
+    \newblock {\em Nature Ecology \& Evolution\/}, pages 1--11.
+
+    \bibitem[Pujol {\em et~al.}(2008)Pujol, Wilson, Ross, and
+    Pannell]{pujol_are_2008}
+    Pujol, B., Wilson, A.~J., Ross, R. I.~C., and Pannell, J.~R. 2008.
+    \newblock Are {{QST}}--{{FST}} comparisons for natural populations meaningful?
+    \newblock {\em Molecular Ecology\/}, {17}(22): 4782--4785.
+
+    \bibitem[Rohlfs and Nielsen(2015)Rohlfs and Nielsen]{rohlfs_phylogenetic_2015}
+    Rohlfs, R.~V. and Nielsen, R. 2015.
+    \newblock Phylogenetic {{ANOVA}}: {{The}} expression variance and evolution
+    model for quantitative trait evolution.
+    \newblock {\em Systematic Biology\/}, {64}(5): 695--708.
+
+    \bibitem[Rohlfs {\em et~al.}(2014)Rohlfs, Harrigan, and
+    Nielsen]{rohlfs_modeling_2014}
+    Rohlfs, R.~V., Harrigan, P., and Nielsen, R. 2014.
+    \newblock Modeling gene expression evolution with an extended
+        {{Ornstein}}--{{Uhlenbeck}} process accounting for within-species variation.
+    \newblock {\em Molecular Biology and Evolution\/}, {31}(1): 201--211.
+
+    \bibitem[Rolland {\em et~al.}(2023)Rolland, {Henao-Diaz}, Doebeli, Germain,
+        Harmon, Knowles, Liow, Mank, Machac, Otto, Pennell, Salamin, Silvestro,
+        Sugawara, Uyeda, Wagner, and Schluter]{rolland_conceptual_2023}
+    Rolland, J., {Henao-Diaz}, L.~F., Doebeli, M., Germain, R., Harmon, L.~J.,
+    Knowles, L.~L., Liow, L.~H., Mank, J.~E., Machac, A., Otto, S.~P., Pennell,
+    M., Salamin, N., Silvestro, D., Sugawara, M., Uyeda, J., Wagner, C.~E., and
+    Schluter, D. 2023.
+    \newblock Conceptual and empirical bridges between micro- and macroevolution.
+    \newblock {\em Nature Ecology \& Evolution\/}, pages 1--13.
+
+    \bibitem[Schraiber and Edge(2023)Schraiber and
+    Edge]{schraiber_heritability_2023}
+    Schraiber, J.~G. and Edge, M.~D. 2023.
+    \newblock Heritability within groups is uninformative about differences among
+    groups: Cases from behavioral, evolutionary, and statistical genetics.
+
+    \bibitem[Schroeder and {von Cramon-Taubadel}(2017)Schroeder and {von
+    Cramon-Taubadel}]{schroeder_evolution_2017}
+    Schroeder, L. and {von Cramon-Taubadel}, N. 2017.
+    \newblock The evolution of hominoid cranial diversity: {{A}} quantitative
+    genetic approach.
+    \newblock {\em Evolution\/}, {71}(11): 2634--2649.
+
+    \bibitem[Sella and Barton(2019)Sella and Barton]{sella_thinking_2019}
+    Sella, G. and Barton, N.~H. 2019.
+    \newblock Thinking about the evolution of complex traits in the era of
+    genome-wide association studies.
+    \newblock {\em Annual Review of Genomics and Human Genetics\/}, {20}(1):
+    461--493.
+
+    \bibitem[Silvestro {\em et~al.}(2015)Silvestro, Kostikova, Litsios, Pearman,
+        and Salamin]{silvestro_measurement_2015}
+    Silvestro, D., Kostikova, A., Litsios, G., Pearman, P.~B., and Salamin, N.
+    2015.
+    \newblock Measurement errors should always be incorporated in phylogenetic
+    comparative analysis.
+    \newblock {\em Methods in Ecology and Evolution\/}, {6}(3): 340--346.
+
+    \bibitem[Silvestro {\em et~al.}(2019)Silvestro, Tejedor, {Serrano-Serrano},
+        Loiseau, Rossier, Rolland, Zizka, H{\"o}hna, Antonelli, and
+        Salamin]{silvestro_early_2019}
+    Silvestro, D., Tejedor, M.~F., {Serrano-Serrano}, M.~L., Loiseau, O., Rossier,
+    V., Rolland, J., Zizka, A., H{\"o}hna, S., Antonelli, A., and Salamin, N.
+    2019.
+    \newblock Early {{Arrival}} and {{Climatically-Linked Geographic Expansion}} of
+        {{New World Monkeys}} from {{Tiny African Ancestors}}.
+    \newblock {\em Systematic Biology\/}, {68}(1): 78--92.
+
+    \bibitem[Simons {\em et~al.}(2018)Simons, Bullaughey, Hudson, and
+    Sella]{simons_population_2018}
+    Simons, Y.~B., Bullaughey, K., Hudson, R.~R., and Sella, G. 2018.
+    \newblock A population genetic interpretation of {{GWAS}} findings for human
+    quantitative traits.
+    \newblock {\em PLOS Biology\/}, {16}(3): e2002985.
+
+    \bibitem[Stamp and Hadfield(2020)Stamp and Hadfield]{stamp_relative_2020}
+    Stamp, M.~A. and Hadfield, J.~D. 2020.
+    \newblock The relative importance of plasticity versus genetic differentiation
+    in explaining between population differences; a meta-analysis.
+    \newblock {\em Ecology Letters\/}, {23}(10): 1432--1441.
+
+    \bibitem[Tajima(1989)Tajima]{tajima_statistical_1989}
+    Tajima, F. 1989.
+    \newblock Statistical method for testing the neutral mutation hypothesis by
+        {{DNA}} polymorphism.
+    \newblock {\em Genetics\/}, {123}(3): 585--595.
+
+    \bibitem[Tenaillon(2014)Tenaillon]{tenaillon_utility_2014}
+    Tenaillon, O. 2014.
+    \newblock The utility of fisher's geometric model in evolutionary genetics.
+    \newblock {\em Annual Review of Ecology, Evolution, and Systematics\/},
+        {45}(1): 179--201.
+
+    \bibitem[Tsuboi {\em et~al.}(2018)Tsuboi, {van der Bijl}, Kopperud,
+        Erritz{\o}e, Voje, Kotrschal, Yopak, Collin, Iwaniuk, and
+        Kolm]{tsuboi_breakdown_2018}
+    Tsuboi, M., {van der Bijl}, W., Kopperud, B.~T., Erritz{\o}e, J., Voje, K.~L.,
+    Kotrschal, A., Yopak, K.~E., Collin, S.~P., Iwaniuk, A.~N., and Kolm, N.
+    2018.
+    \newblock Breakdown of brain--body allometry and the encephalization of birds
+    and mammals.
+    \newblock {\em Nature Ecology \& Evolution\/}, {2}(9): 1492--1500.
+
+    \bibitem[Tung {\em et~al.}(2015)Tung, Zhou, Alberts, Stephens, and
+    Gilad]{tung_genetic_2015}
+    Tung, J., Zhou, X., Alberts, S.~C., Stephens, M., and Gilad, Y. 2015.
+    \newblock The genetic architecture of gene expression levels in wild baboons.
+    \newblock {\em eLife\/}, {4}: e04729.
+
+    \bibitem[Turelli(1984)Turelli]{turelli_heritable_1984}
+    Turelli, M. 1984.
+    \newblock Heritable genetic variation via mutation-selection balance:
+        {{Lerch}}'s zeta meets the abdominal bristle.
+    \newblock {\em Theoretical Population Biology\/}, {25}(2): 138--193.
+
+    \bibitem[Turelli(2017)Turelli]{turelli_commentary_2017}
+    Turelli, M. 2017.
+    \newblock Commentary: {{Fisher}}'s infinitesimal model: {{A}} story for the
+    ages.
+    \newblock {\em Theoretical Population Biology\/}, {118}: 46--49.
+
+    \bibitem[Uyeda and Harmon(2014)Uyeda and Harmon]{uyeda_novel_2014}
+    Uyeda, J.~C. and Harmon, L.~J. 2014.
+    \newblock A {{Novel Bayesian Method}} for {{Inferring}} and {{Interpreting}}
+    the {{Dynamics}} of {{Adaptive Landscapes}} from {{Phylogenetic Comparative
+    Data}}.
+    \newblock {\em Systematic Biology\/}, {63}(6): 902--918.
+
+    \bibitem[Uyeda {\em et~al.}(2011)Uyeda, Hansen, Arnold, and
+    Pienaar]{uyeda_millionyear_2011}
+    Uyeda, J.~C., Hansen, T.~F., Arnold, S.~J., and Pienaar, J. 2011.
+    \newblock The million-year wait for macroevolutionary bursts.
+    \newblock {\em Proceedings of the National Academy of Sciences\/}, {108}(38):
+    15908--15913.
+
+    \bibitem[Walsh and Lynch(2018)Walsh and Lynch]{walsh_evolution_2018}
+    Walsh, B. and Lynch, M. 2018.
+    \newblock {\em Evolution and {{Selection}} of {{Quantitative Traits}}\/}.
+    \newblock Oxford University Press.
+
+    \bibitem[Weaver {\em et~al.}(2007)Weaver, Roseman, and
+    Stringer]{weaver_were_2007}
+    Weaver, T.~D., Roseman, C.~C., and Stringer, C.~B. 2007.
+    \newblock Were neandertal and modern human cranial differences produced by
+    natural selection or genetic drift?
+    \newblock {\em Journal of Human Evolution\/}, {53}(2): 135--145.
+
+    \bibitem[Wilder {\em et~al.}(2023)Wilder, Supple, Subramanian, Mudide,
+        Swofford, {Serres-Armero}, Steiner, Koepfli, Genereux, Karlsson,
+            {Lindblad-Toh}, {Marques-Bonet}, Munoz~Fuentes, Foley, Meyer, {Zoonomia
+        Consortium}, Ryder, and Shapiro]{wilder_contribution_2023}
+    Wilder, A.~P., Supple, M.~A., Subramanian, A., Mudide, A., Swofford, R.,
+        {Serres-Armero}, A., Steiner, C., Koepfli, K.-P., Genereux, D.~P., Karlsson,
+    E.~K., {Lindblad-Toh}, K., {Marques-Bonet}, T., Munoz~Fuentes, V., Foley, K.,
+    Meyer, W.~K., {Zoonomia Consortium}, Ryder, O.~A., and Shapiro, B. 2023.
+    \newblock The contribution of historical processes to contemporary extinction
+    risk in placental mammals.
+    \newblock {\em Science\/}, {380}(6643): eabn5856.
+
+\end{thebibliography}
 
@@
 
-The phenotype is encoded by a genetic architecture and is thus ultimately bounded.
-At the macro-evolutionary scale, phenotypic divergence should plateau at some point, ultimately resulting in a $\RateBetween$.
+In our simulation setting, the genetic architecture underlying the phenotype is not changing along the phylogeny.
+As a result, the phenotype may be bounded by what such genetic architecture can produce, and this could cause a slowdown of phenotypic divergence over time, ultimately resulting in a decreased $\RateBetween$.
 
@@
     \item Bottom row: Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-Histogram of probabilities of $\EstNI$ being greater than $1$.
+Histogram of probabilities of $\NI$ being greater than $1$.
 \end{itemize}
 Under a model of neutral trait evolution, when mutation rate increases, or equivalently the divergence between species increases, the phenotypic divergence between species saturates faster than the nucleotide divergence.
-This saturation effect can result in a spurious signal of stabilizing selection ($\EstNI < 1$) for deeper phylogeny when the trait is evolving neutrally.
+This saturation effect can result in a spurious signal of stabilizing selection ($\NI < 1$) for deeper phylogeny when the trait is evolving neutrally.
 \end{document}

### 2024-07-03 manuscript/compile.sh

@@ -1,1 +1,2 @@
-latexdiff main.tex main-revision.tex > diff.tex
+latexdiff main-first-submission.tex main-revision.tex > diff.tex
+latexdiff main-revision.tex main.tex > diff-revision.tex

### 2024-07-05 manuscript/bayescode_nodetraits.tex
Cleaned unecessary files. Updated README.

--- 2024-03-13 manuscript/main-revision.tex
@@
 \renewcommand{\arraystretch}{1.2}
-\linenumbers
 \frenchspacing
@@
 
-\maketitle
+\section{Bayesian implementation}\label{sec:implementation}
 
-% Abstract (≤ 250 words)
-%TC:endignore
-\begin{abstract}
-    To quantify selection acting on a trait, methods have been developed using either within or between-species variation.
-    However, methods using within-species variation do not integrate the changes at the macro-evolutionary scale.
-    Conversely, current methods using between-species variation usually discard within-species variation, thus not accounting for processes at the micro-evolutionary scale.
-    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
-    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
-    As such, it does not require estimation of population size nor of time of speciation for normalization.
-    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
-    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
-    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
-\end{abstract}
-
-\keywords{Quantitative genetics \and Trait evolution \and Selection \and Phylogenetics \and Population genetics }
-
-% Research Article (7500 words)
-\section*{Introduction}\label{sec:introduction}
-
-Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
-Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
-The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\citep{lynch_genetics_1998, walsh_evolution_2018}.
-
-Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
-Given this genetic architecture of traits, from two diverging populations, it is possible to distinguish which traits have evolved under natural selection in controlled experimental settings, by performing genetic cross between individuals~\citep{fraser_detecting_2020}.
-Across several populations, by contrasting both trait differentiation (\Qst) and genetic differentiation (\Fst), so-called \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
-\Qst\ higher than \Fst\ is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
-Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\citep{lamy_qst_2012}.
-Other frameworks explicitly model genetic drift as a random process generating both trait and genetic differences between individuals and populations.
-This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
-However, regardless of the strengths and weaknesses of each method\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.
-
-% Pair of species
-To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
-For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes that will reach fixation in the population.
-These changes ultimately result in different mean trait values across lineages.
-Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984}.
-Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
-As an analogy, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \citet{mcdonald_adaptative_1991} test.
-In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.
-
-% Phylogenetics (mean trait evolution)
-Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{hansen_translating_1996}.
-As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
-Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
-Deviations from the assumptions of the Brownian process are however well known.
-When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{catalan_drift_2019}.
-Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
-However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
-In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
-Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
-Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.
-
-% At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
-We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
-It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
-The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
-Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
-From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary generalization to account for phylogenetic relationships between species.
-From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
-
-\section*{Materials and Methods}\label{sec:materials-and-methods}
-\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
-
-While observing trait variations across individuals of several species, we ask if the variation within species compared to variation between species is compatible with neutral evolution or not.
-In statistical terms, this can also be framed as: Is the variance of means equal to the mean of variances?
-The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times.
-By reviewing theoretical expectations and leveraging nucleotide sequence variations, the goal of this section is thus to obtain normalized trait variation between and within species that are equal if the trait is neutral.
-Here we denote these normalized trait variations as respectively $\RateWhithin$ for within species and as $\RateBetween$ for between species.
-
-\subsubsection*{Within-species trait variations}
-
-For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
-For a diploid individual, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per locus per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
-\begin{gather}
-    \VarMutation = 2 \MutationRatePheno \Multiply \GenArchi \label{eq:var-mutation}.
-\end{gather}
-
-While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\citep{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
-For a neutral trait at equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
-\begin{align}
-    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
-    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
-\end{align}
-
-For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all nucleotide sites~\citep{tajima_statistical_1989}.
-Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
-As shown in \citet{tajima_statistical_1989}, $\pi$ is a function of the mutation rate ($\MutationRateNuc$, per nucleotide site per generation) and the effective population size ($\Ne$):
-\begin{gather}
-    \pi = 4 \Ne \Multiply \MutationRateNuc \label{eq:genetic-diversity}.
-\end{gather}
-
-To remove the effect of $\Ne$, we define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
-After simplification, $\RateWhithin$ is then solely a function of the underlying genetic architecture as:
-\begin{align}
-    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi} \label{eq:def-rate-pop}, \\
-    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Ne \Multiply \MutationRateNuc} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
-    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-pop}
-\end{align}
-
-If $\VarGenetic$ is not empirically accessible, it can be related to the observed phenotypic variance ($\VarPhenotype$), multiplied by narrow-sense heritability of the trait ($\Heritability$), as~\citep{hill_data_2008}:
-\begin{align}
-    \VarGenetic & = \Heritability \Multiply \VarPhenotype.\label{eq:var-gen-pheno}
-\end{align}
-Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ instead of $\VarGenetic$ as:
-\begin{align}
-    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi } \text{ from definition eq.~\ref{eq:def-rate-pop} and~\ref{eq:var-gen-pheno}}, \label{eq:def-rate-pheno-pop} \\
-    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \text{ from eq.~\ref{eq:rate-pop}} \label{eq:rate-pheno-pop}
-\end{align}
-
-\subsubsection*{Between-species trait variations}\label{subsec:between-species-var}
-For a given species $\Spi$, we denote by $\MeanTrait_{\Spi}$ the mean value of the trait across the individuals.
-If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
-Given a phylogenetic tree, for a pair of species $\Spi$ and $\Spj$ from this tree, we denote as $\NbrGen$ the number of generations between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
-Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \citet{hansen_translating_1996} by:
-\begin{align}
-    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \NbrGen \\
-    & = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}}.
-\end{align}
-Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
-The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$~\citep{kimura_probability_1962}.
-We can derive the substitution rate ($\SubRate$, per nucleotide site per generation) as the number of newly arisen mutations ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{kimura_evolutionary_1968}, giving:
-\begin{align}
-    \SubRate & = 2 \Ne \Multiply \MutationRateNuc \Multiply \pfix, \\
-    & = 2 \Ne  \Multiply \MutationRateNuc  \Multiply \dfrac{1}{2\Ne}, \\
-    & = \MutationRateNuc. \label{eq:substitution-rate}
-\end{align}
-That is, if mutations are neutral, the rate of substitution per generation within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\citep[review]{mccandlish_modeling_2014}.
-
-Next, we denote $\NucDiv$ as the nucleotide divergence between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
-In other words, $\NucDiv$ is the expected number of substitutions per nucleotide site during the $\NbrGen$ generations.
-Assuming that no multiple substitutions occurred at the same site, $\NucDiv$ is the number of generations ($\NbrGen$) multiplied by the nucleotide substitution rate per generation ($\SubRate$):
-\begin{align}
-    \NucDiv & = \NbrGen \Multiply \SubRate \\
-    & = \NbrGen \Multiply \MutationRateNuc \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
-\end{align}
-To remove the effect of the number of generations ($\NbrGen$) first, and to also equate to $\RateWhithin$ (eq.~\ref{eq:rate-pop}), we define $\RateBetween$ as the covariance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4\NucDiv$).
-After simplification, $\RateBetween$ is also solely a function of the underlying genetic architecture as:
-\begin{align}
-    \RateBetween & \defEqual \frac{\VarPhy}{4 \NucDiv} \label{eq:def-rate-phy},\\
-    & = \frac{4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \NbrGen \Multiply \MutationRateNuc}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
-    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-phy}
-\end{align}
-
-In equation~\ref{eq:rate-phy}, we show that the covariance in mean trait value between a pair of species ($\VarPhy$) does increase linearly with shared nucleotide divergence ($\NucDiv$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
-Importantly, since the number of generations is the ratio of time divided by generation time (average time between two consecutive generations), removing the effect of the number of generations in equation~\ref{eq:rate-phy} also removes the effect of both time and generation time.
-
-\begin{figure*}[!ht]
-    \centering
-    \includegraphics[width=\textwidth, page=1] {figure1}
-    \caption{
-        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
-        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
-        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
-        Under neutral evolution, $\EstNI$ is expected to be equal to one.
-        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
-        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
-        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
-    }
-    \label{fig:methods}
-\end{figure*}
-
-\subsubsection*{Neutrality index}
-
-The variability between either individuals or species can be obtained for both quantitative traits and genomic sequences.
-At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$.
-At the phylogenetic level, the variability of the mean trait value between species can be combined with the nucleotide divergence of any neutrally evolving genomic region to obtain $\RateBetween$.
-If the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogenetic tree, we thus have:
-\begin{align}
-    \frac{\RateBetween}{\RateWhithin} &= \frac{\VarPhy}{4 \NucDiv} \Multiply \frac{\pi}{\Heritability \Multiply \VarPhenotype} \text{ by definition and eq.~\ref{eq:def-rate-pheno-pop} and~\ref{eq:def-rate-phy}},\\
-    &= \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \Multiply \frac{\MutationRateNuc}{\MutationRatePheno \Multiply \GenArchi} \text{ from eq.~\ref{eq:rate-pheno-pop} and~\ref{eq:rate-phy}}, \\
-    &= 1.
-\end{align}
-We define a neutrality index $\NI$ as:
-\begin{align}
-\NI \defEqual \frac{\RateBetween}{\RateWhithin},
-\end{align}
- which will equal to $1$ for a trait evolving neutrally.
-Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$), generation time or time of divergence ($\NbrGen$) need to be estimated.
-Moreover, the nucleotide sequence from which $\pi$ and $\NucDiv$ are obtained should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
-
-\subsection*{Estimation}\label{subsec:estimate}
-
-We denote $\EstNI$, $\EstRateBetween$ and $\EstRateWhithin$ the point estimates of respectively $\NI$, $\RateBetween$ and $\RateWhithin$.
-For each species with data available, $\RateWhithin$ as defined in eq.~\ref{eq:def-rate-pheno-pop} can be seen as a replicate sample.
-Thus, $\EstRateWhithin$ can be obtained by averaging out across all the sampled species.
-On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}.
-Generally, $\EstRateWhithin$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
-As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\EstRateWhithin$.
-We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
-
-\subsubsection*{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the symmetric distance matrix computed from the branch lengths and the topology of the phylogenetic tree.
-The diagonal $\Distance_{\Spi,\Spi}$ represents the total nucleotide divergence from the root of the tree to each taxon ($\Spi$).
-The off-diagonal elements ($\Distance_{\Spi,\Spj} = \NucDiv$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$, as in equation~\ref{eq:distance}.
-The mean trait value at the root of the tree ($\RootTrait$) can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\citep{omeara_testing_2006}:
-\begin{gather}
-    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
-\end{gather}
-where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.
-
-Finally, between-species variation $\EstRateBetween$ is estimated as~\citep{omeara_testing_2006}:
-\begin{gather}
-    \EstRateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
-\label{eq:estimated-rate-phy}
-\end{gather}
-
-For a given species $\Spi$ with inter-individual data available, additive genetic variance of a trait ($\VarGeneticSpi$) is the product of heritability ($\Heritability_{i}$) and phenotypic variance ($V_{\Trait, i}$).
-The ratio of $\VarGeneticSpi$ over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
-Averaged across all species, we obtain the estimate $\EstRateWhithin$ as:
-\begin{gather}
-    \EstRateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}.
-\label{eq:estimated-rate-pop}
-\end{gather}
-
-As depicted in fig.~\ref{fig:methods}, the neutrality index is estimated as:
-\begin{gather}
-    \EstNI = \frac{\EstRateBetween}{\EstRateWhithin}. \label{eq:estimated-NI}
-\end{gather}
-
-\subsubsection*{Multivariate Brownian process}
-
-In the previous section, $\EstNI$ is estimated independently for each trait of interest.
-Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree.
-Trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
-The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-The branch lengths of the tree used to model the Brownian process runs is measured in units of $4d$ ($d$ being the nucleotide divergence).
-The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
-
-\subsubsection*{Bayesian estimate}
-The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
-We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\citep{latrille_inferring_2021}.
-With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
-We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
-For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
-Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\EstRateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
-From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
-The posterior distribution of $\EstNI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
-
-\subsubsection*{Applicability to empirical data}
-
-Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
-Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
-Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
-Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
-It is, however, possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
-Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and fall within the range of 0.2-0.5 in a vast majority of phenotypic traits tested~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
-Alternatively, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\EstNI$ since $\Heritability \leq H^2$.
-If available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.
-
-In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\EstNI$ is underestimated.
-Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
-
-\subsection*{Simulation}\label{subsec:simulations}
-
-% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
-% pS = 0.0027577478571428568
-% Tree length (dS) = 2.340579
-% Tree length (My) = 1259.4490200000002
-% Root age (My) = 98.9489413888889
-% Root age (dS) = 0.18388820079995308
-% nbr_sites_var = 27.577478571428568
-% u = 1.3788739285714283e-05
-% nbr_generations = 13336.114128321291
-We tested the performance of our neutrality index ($\NI$) to detect selection on a quantitative trait using simulations.
-We performed simulations under different selective regimes (neutral, stabilizing, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of the mutation rate (constant or fluctuating).
-Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
-Each individual phenotypic value was the sum of genotypic value and an environmental effect.
-The environmental effect was normally distributed with variance $\VarEnv$.
-We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and for the theoretical formulation see section~S\ref{subsec:genotype-phenotype-map} and fig.~\ref{fig:simulator-summary}).
-We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
-Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
-For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRatePheno \Multiply \NbrLoci $, with the mutation rate per locus per generation $\MutationRatePheno$.
-From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from the root to leaves of $0.18$ and average genetic diversity of $0.00276$.
-We scaled parameters in our simulations to fit plausible values for mammals.
-We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per site per generation  and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
-We set $\MutationRatePheno=\MutationRateNuc$ without loss in generality since the genetic architecture ($\NbrLoci$ and $a$) is assumed constant in the simulator.
-
-The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modeled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
-These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$.
-In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $1.0$.
-An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
-The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
-The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
-From a speciation process perspective, this was equivalent to an allopatric speciation over one generation.
-
-At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
-Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
-More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
-Mutations were considered as a displacement of the phenotype in the multidimensional space.
-Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
-Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
-Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
-Neutral evolution was implemented by fixing the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.
-
-Nucleotide diversity ($\pi$) was measured as the heterozygosity of neutral markers that were simulated along the phylogenetic tree but not linked to the trait simulated.
-Nucleotide divergence ($d$) was measured as the number of substitutions per site of neutral markers along the branches of the phylogenetic tree.
-The additive genetic variance was measured as phenotypic variance multiplied by heritability.
-Heritability was estimated from the slopes of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\citep{lynch_genetics_1998} averaged over the last 10 simulated generations.
-Heritability was thus not a given parameter of the simulations, but rather measured as it would be in empirical data.
-
-\begin{figure*}[!ht]
-    \centering
-    \includegraphics[width=\textwidth, page=1] {figure2}
-    \caption{
-        Wright-Fisher simulations with mutation, selection and drift.
-        Left panel: For a given individual, the trait phenotypic value is the sum of genotypic value and a environmental effect (standard deviation $\VarEnv$).
-        The trait's genotypic value is encoded by $\NbrLoci$ independent loci (meaning no linkage), with each locus contributing additively to the genotypic value.
-        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
-        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRatePheno$ to mutate.
-        Drift is modeled by the resampling of parents.
-        Right panel: examples of a trait evolving along a phylogenetic tree, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
-    }
-    \label{fig:simulator}
-\end{figure*}
-
-\subsection*{Empirical dataset}\label{subsec:empirical-dataset}
-
-We analyzed a dataset of body and brain masses from mammals.
-The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
-We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
-We discarded species with only one representative since phenotypic variance cannot be estimated.
-The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
-More specifically, nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \citet{wilder_contribution_2023}.
-
-We also analyzed a dataset of primate species, with the nucleotide variation obtained from \citet{kuderna_global_2023} and the quantitative trait variation also from \citet{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
-However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
-As such, the evidence for $\EstNI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers included in the estimate of divergence can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).
-
-\section*{Results}\label{sec:results}
-
-\subsection*{Neutrality index}
-
-For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
-We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see Materials and Methods), suggesting that traits for which this relationship was not verified were putatively under selection.
-Under stabilizing selection, the variation between species is depleted because the mean trait is maintained toward similar values between different species, which leads to $\NI < 1$.
-In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\citep{hansen_stabilizing_1997}, which leads to $\NI > 1$.
-Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
-This statistic was obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
-We also devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
-
-Our neutrality index made a series of assumptions that we described in details in the Material and Methods section.
-Table~\ref{table:assumptions} summarized these assumptions and outlined possible consequences for the neutrality test that we proposed.
-
-\subsection*{Results against simulations}\label{subsec:results-against-simulations}
-
-The inference framework was first tested on independently simulated datasets matching an empirically relevant mammalian empirical regime (see Materials and Methods).
-Under constant population size ($\Ne$) and constant mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
-For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
-Once we introduced fluctuating $\Ne$, $\MutationRatePheno$ and $\MutationRateNuc$(Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
-For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) and none as diversifying selection.
-
-\begin{figure*}[!ht]
-    \centering
-    \includegraphics[width=\textwidth, page=1] {figure3}
-    \caption{
-        $10,000$ simulations of trait evolution along a phylogenetic tree under different selection regimes.
-        Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-        Histogram of ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$ with $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and probabilities of $\EstNI$ being greater than $1$ (right).
-        Effective population size ($\Ne$) and mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
-    }
-    \label{fig:results-simulations}
-\end{figure*}
-
-\begin{table*}[t!]
-    \centering
-    \begin{adjustbox}{width = 0.75\textwidth}
-        \begin{tabular}{||l|l|l|c|c|c|c|c||}
-        \toprule
-        Dataset & Trait & $\Heritability$ & Sex & $\NbrTaxa$ & $\EstNI$ & $95\%$ CI for $\EstNI$ & $\proba [\EstNI > 1 ]$ \\ \hline
-        \midrule
-        Mammals & Body mass & 1.0 & \Male & 36 & 0.340 & 0.217-0.523 & 0.000 \\ \hline
-        Mammals & Body mass & 1.0 & \Female & 26 & 0.277 & 0.160-0.490 & 0.000 \\ \hline
-        Mammals & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 36 & 1.124 & 0.721-1.754 & 0.635 \\ \hline
-        Mammals & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 26 & 0.936 & 0.523-1.715 & 0.324 \\ \hline \hline
-        Mammals & Brain mass & 1.0 & \Male & 36 & 1.351 & 0.851-2.173 & 0.877 \\ \hline
-        Mammals & Brain mass & 1.0 & \Female & 26 & 1.727 & 0.991-2.938 & 0.972 \\ \hline
-        Mammals & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 36 & 4.527 & 2.831-7.091 & 1.000 \\ \hline
-        Mammals & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 26 & 6.001 & 3.288-10.941 & 1.000 \\ \hline \hline
-        Primates & Body mass & 1.0 & \Male & 71 & 0.558 & 0.401-0.784 & 0.000 \\ \hline
-        Primates & Body mass & 1.0 & \Female & 65 & 0.389 & 0.278-0.547 & 0.000 \\ \hline
-        Primates & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 71 & 1.875 & 1.288-2.695 & 1.000 \\ \hline
-        Primates & Body mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 65 & 1.296 & 0.899-1.821 & 0.914 \\ \hline \hline
-        Primates & Brain mass & 1.0 & \Male & 71 & 1.929 & 1.395-2.616 & 1.000 \\ \hline
-        Primates & Brain mass & 1.0 & \Female & 65 & 1.950 & 1.399-2.790 & 1.000 \\ \hline
-        Primates & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Male & 71 & 6.479 & 4.658-8.944 & 1.000 \\ \hline
-        Primates & Brain mass & $\mathcal{U}(0.2, 0.4)$ & \Female & 65 & 6.522 & 4.664-9.294 & 1.000 \\
-        \bottomrule
-        \end{tabular}
-    \end{adjustbox}
-    \caption{
-    Test of diversifying selection on a mammal and a primate dataset, by splitting males (\Male) and females (\Female).
-    Traits considered were body mass or brain mass (log-transformed).
-    Heritability ($\Heritability$) was either assumed complete ($\Heritability=1.0$) or uniformly distributed between 20\% and 40\%  ($\Heritability \sim \mathcal{U}(0.2, 0.4)$).
-    $\NbrTaxa$ was the number of species in the dataset.
-    $\EstNI$ was the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\EstNI$ also computed.
-    $\proba [\EstNI > 1 ]$ was the estimated posterior probability of diversifying selection.
-    }
-    \label{table:empirical}
-\end{table*}
-
-\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
-
-For mammalian body and brain mass, we obtained male (\Male) and female (\Female) trait variations.
-Combined with nucleotide diversity and divergence, we estimated $\EstNI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in the Table~\ref{table:empirical}.
-Assuming complete heritability, brain mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
-If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\citep{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
-Mammalian brain mass was found to be under diversifying selection with posterior probabilities of $0.877$ for males and $0.972$ for females when complete heritability was assumed.
-Assuming a uniform distribution between 20\% and 40\% for heritability led to posterior probabilities of diversifying selection of $1.0$ for both males and females.
-
-We also analyzed a similar dataset for body mass focusing this time only at Primates (Table~\ref{table:empirical}).
-For primates body mass, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%.
-Assuming complete heritability of body mass did not change the posterior probability for males, but increased the one for female to $1.0$.
-Evidence for diversifying selection on body mass was therefore more pronounced in Primates than in mammals.
-However, the genetic markers used to normalize trait variance with nucleotide divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\EstNI$ (Table~\ref{table:assumptions} and methods).
-
-\begin{table*}[t!]
-    \centering
-    \begin{adjustbox}{width = 1.0\textwidth}
-        \begin{tabular}{||l|l||c|c||c|c||}
-            \hline
-            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\EstNI > 1$ & Test $\EstNI < 1$ \\ \hline \hline
-            Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
-            Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Phenotypic plasticity & Trait responding to individual environments  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
-            Markers for polymorphism under balanced selection & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
-            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
-            Multiple nucleotide substitutions at the same locus & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\\hline \hline
-        \end{tabular}
-    \end{adjustbox}
-    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
-    The last two columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
-    }
-    \label{table:assumptions}
-\end{table*}
-
-\section*{Discussion}\label{sec:discussion}
-
-% Summary of the method and results
-In this study, we proposed a neutrality index for a quantitative trait that can be used within a statistical framework to test for selection.
-Our neutrality index for a trait, $\NI$, is calculated as the ratio of the normalized within- to between-species variation and it allowed the identification of the evolutionary regime of a quantitative trait.
-At the phylogenetic scale, trait variation between species was normalized by sequence divergence obtained from a neutral set of markers.
-Similarly, trait variation within species was normalized by sequence polymorphism obtained also from a neutral set of markers.
-Our estimate of $\EstNI$ could be tested for deviation from the value of $1.0$ expected under the null hypothesis of neutrality.
-Technically, the neutrality index can be estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S\ref{sec:implementation}).
-The latter also enabled the estimation of the posterior credible interval to test for departure from a neutrally evolving trait (e.g. $ \proba [ \EstNI > 1 ]$).
-We tested our statistical procedure against simulated data and showed that our test was able to correctly detect simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$).
-However, our test detected a spurious signal of stabilizing selection ($\EstNI < 1$) when we simulated the evolution of a neutral trait.
-An assumption of our test is that the neutral phenotypic trait is evolving as a Brownian process and is, therefore, unbounded.
-However, the phenotype is encoded by a genetic architecture, and is thus ultimately bounded, even more so if the trait is encoded by a few loci (see table~\ref{table:assumptions}).
-At the macro-evolutionary scale, phenotypic divergence should plateau at some point, resulting in a reduced between-species trait variation.
-We argue that this effect can result in a spurious signal of stabilizing selection ($\EstNI < 1$), especially for deeper phylogeny (figure~\ref{fig:supp-distance} and section~S\ref{sec:supp-distance})
-We thus argue that our method should be used to detect diversifying selection, but that it had low accuracy to detect stabilizing selection due to false positives.
-
-% How is our test related to other tests of selection of traits
-Our results showed that our method significantly improved over currently available methods to detect selection acting on a trait at the phylogenetic scale.
-Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Our approach could in theory be applied to detect stabilizing selection at the phylogenetic scale, but we showed that it did not have the statistical power to identify those cases.
-In contrast, we showed that our method was able to identify correctly cases of diversifying selection, which is a clear improvement over current methods that model only mean trait value.
-Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
-For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\citep{catalan_drift_2019}.
-Instead, our diversity index has the advantage to discriminate the alternative model of diversifying selection from the neutral case by comparing within- and between-species variation while correctly normalizing them using nucleotide markers.
-Our approach is not the first one coupling between-species and within-species variations, and those approaches employ different strategies to detect selection.
-First, one empirical strategy is to compare the ratio of between to within variation across a pool of traits, which allow to identify outlier traits putatively under diversifying selection~\citep{rohlfs_modeling_2014}.
-However, this method does not formally allow testing for diversifying selection, and requires many traits such as expression level data to seek outliers genes~\citep{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
-Second, other methods leverage Lande’s generalized genetic distance (LGGD), which relate the ratio of between to within variations to population-genetic parameters~\citep{lynch_analysis_1990, lemos_evolutionary_2001, lemos_rates_2005, weaver_were_2007, porto_rate_2015}.
-Specifically, by leveraging estimates of effective population size ($\Ne$) and number of generations between species, or alternatively by assuming their constancy, these methods can test for departures from the null model of neutral evolution for a single trait.
-Such methods have been successful in identifying specific instances of diversifying selection\citep{schroeder_evolution_2017, machado_preeminent_2022} and near-drift~\citep{machado_using_2023}.
-However, $\Ne$ and the number of generations are complex parameters to correctly infer, and is usually done for a pair or only a few species, and ultimately requires large genomic datasets and heavy statistical methods~\citep{wilder_contribution_2023}.
-Instead, our diversity index opens new avenues to revisit these studies testing for the selective regime affecting the quantitative traits, by formally incorporating nucleotide divergence and polymorphism, bypassing estimation of $\Ne$, generation time and calibration of ancestral node ages~\citep{machado_using_2023}.
-
-% Why is our test different
-As such, the main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
-In this context, our test bears many similarities to \QstFst\ tests (and their derivatives) that have been developed to test for selection of a trait across several populations while also leveraging sequence variation~\citep{martin_multivariate_2008, leinonen_qst_2013} or co-ancestry between individuals~\citep{ovaskainen_new_2011}.
-Our method can be seen as an extension at the phylogenetic scale, where although the sequences used should be neutrally evolving, they can be obtained from different sampled individuals than for the trait.
-Importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$ and mutation rates were constant across the phylogenetic tree, an unmet assumption empirically~\citep{bergeron_evolution_2023, wilder_contribution_2023}.
-Indeed, under the neutral case of evolution, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, generation time and mutation rates, which canceled out in the neutrality index $\EstNI$.
-
-In the context of phylogenetic comparative methods, modeling mean trait evolution as a function of nucleotide divergence ($d$) instead of time has more general consequences.
-As an example, trait variation is often modeled as a Brownian process running on a time-calibrated tree, which can produce biases~\citep{litsios_effects_2012}.
-Indeed, for a neutrally evolving trait, trait variation depends directly on the number of generations, which in turn correlates with time.
-But, since species generation time might vary along the phylogenetic tree, $d$-scaled trees absorbing changes in generation time should be used instead of time-scaled trees.
-Using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks).
-We argue, that the soundness of studying trait evolution on $d$-scaled trees can be evaluated by the absolute fit of a model to the data~\citep{pennell_model_2015}.
-More generally, genomic information could potentially be seen as a way to disentangle congruence models~\citep{louca_extant_2020}, or as prior for methods that detect shifts in adaptive regimes~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020},.
-
-% How is our test related to other tests of selection on sequences
-Even though our test was developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provided insight into its behavior.
-First, we acknowledge that our test took inspiration from the \citet{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences in a pair of species, except that the non-synonymous versus synonymous distinction is replaced by the comparison between quantitative trait and neutral genomic sequence.
-Second, at the phylogenetic scale, when comparison is done across several species, our test also bears analogy to codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\citep{goldman_codonbased_1994, muse_likelihood_1994}.
-As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
-Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we could leverage the vast literature discussing and interpreting the results of these tests and their pitfalls~\citep{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ did not necessarily imply that the trait was effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\citep{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ did not rule out diversifying selection, but rather that this diversifying selection was not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\citep{latrille_genes_2023}.
-By explicitly modeling stabilizing selection as a moving optimum, it would theoretically be possible to tease apart the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
-
-% Pitfalls of the method
-In the context of detecting diversifying selection on a trait, we argue that the main drawback of our method is that the additive genetic variance of the trait is required instead of the phenotypic variance.
-If phenotypic variance was used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ was ultimately underestimated.
-Similarly, using broad-sense heritability instead of narrow-sense heritability would result in underestimated $\EstNI$.
-In such context, the test of stabilizing selection ($\EstNI < 1]$) would be statistically invalid.
-However, the test of diversifying selection ($\EstNI > 1$) was underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
-As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
-If available, any prior knowledge on heritability can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
-
-% What can and should be improved
-The development of our neutrality index was also based on several assumptions that could be relaxed in future studies.
-First, we cannot predict the behavior of our test in the context of population structures, gene flow and introgression.
-These factors should be thoroughly investigated using simulations.
-Second, loci were assumed to contribute additively to the phenotype.
-Although the effects of dominance and epistasis is typically weak compared to the additive effects on the quantitative traits, their influence should be assessed~\citep{hill_data_2008, crow_epistasis_2010}.
-Third, the genetic architecture of the trait was assumed to be constant across the phylogenetic tree, whereas it might actually be variable among individuals and species~\citep{tung_genetic_2015, huber_conservatism_2015}.
-Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\citep{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Finally, from a statistical perspective, our Bayesian estimation could integrate uncertainty from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
-
-% What can our test bring to the field
-From an empirical point of view, our method required integrating genomic and trait variation, which could reduce the possible datasets to be used.
-However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals' brain and body mass, showing signals of diversifying selection.
-The consensus on macro-evolutionary studies, assuming constancy of $\Ne$, generation time and mutation rates, is that empirical rates of evolution calculated on phylogenetic trees and the fossil record are far inferior to the expected under drift~\citep{lynch_analysis_1990, uyeda_millionyear_2011}.
-Our finding of diversifying selection on body and brain mass could be seen as an argument against that interpretation.
-In fact, rates of nucleotide evolution also show a tendency for slowing down on a longer timescale~\citep{rolland_conceptual_2023}.
-One possible interpretation is that normalization by nucleotide divergence could absorb this observed slowing rate of evolution.
-Altogether, further empirical and theoretical studies are required to disentangle this discrepancy between these different interpretations.
-Because our test was also based on several assumptions that might not hold on empirical data, we also provided a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (Table~\ref{table:assumptions}).
-For example, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass was evolving under diversifying selection since the markers used for nucleotide divergences were not neutral, which can lead to a spurious $\EstNI > 1$.
-In conclusion, our study provided a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we believe that our new approach is a promising tool to investigate the evolution of quantitative traits.
-
-%TC:ignore
-\bibliographystyle{natbib}
-\bibliography{references_bibtex}
-
-\newpage
-
-\part*{Supplementary materials}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\setcounter{section}{0}
-
-\renewcommand{\baselinestretch}{1.0}\normalsize
-\tableofcontents
-\renewcommand{\baselinestretch}{1.5}\normalsize
-
-\newpage
-\section{Genetic architecture of the trait}\label{sec:simulator}
-
-\subsection{Genotype-phenotype map}\label{subsec:genotype-phenotype-map}
-
-\begin{itemize}
-    \item $\NbrLoci$ is the number of loci encoding the trait.
-    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
-    \item $\Ne$ is the effective number of individuals.
-    \item $g_{\Indiv,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $\Indiv \in \{1, \hdots, \Ne\}$.
-    \item $G_{\Indiv} = \sum_{l=1}^{\NbrLoci} a_l \times g_{\Indiv,l}$ is the genotypic value for individual $\Indiv$.
-    \item $\xi_{\Indiv} \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual $\Indiv$.
-    \item $\Trait_{\Indiv} = G_{\Indiv}+ \xi_i$ is the phenotype for individual $\Indiv$.
-\end{itemize}
-
-\begin{center}
-    \captionof{figure}{summary of trait's genetic architecture.}
-    \includegraphics[width=0.6\textwidth, page=1] {figureS1}
-    \label{fig:simulator-summary}
-\end{center}
-
-within-species, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
-\begin{equation}
-    \bar{G} = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne} G_{\Indiv} \text{\quad and \quad} \VarGenetic = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne}\left(G_{\Indiv}- \bar{G} \right)^2\label{eq:simu-genotype}
-\end{equation}
-The theoretical additive genetic variance ($\VarGenetic$) is a function of the number of loci ($\NbrLoci$) and the effect of a mutation ($a$) as:
-\begin{equation}
-    \VarGenetic = 4 \Ne \Multiply \MutationRatePheno \Multiply \NbrLoci \Multiply a^2 \label{eq:simu-var-genetic}
-\end{equation}
-
-The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
-\begin{equation}
-    \bar{\Trait} = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne} \Trait_{\Indiv}\text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne}\left(\Trait_{\Indiv} - \bar{\Trait} \right)^2 \label{eq:simu-between}
-\end{equation}
-
-Heritability ($\Heritability$) is defined as:
-\begin{equation}
-    \Heritability = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
-\end{equation}
-Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($\Heritability$) as:
-\begin{equation}
-    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRatePheno \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
-\end{equation}
-
-\newpage
-\section{Bayesian estimate}\label{sec:bayesian-estimate}
-
-\subsection{Multivariate Brownian process}\label{subsec:multivariate-brownian-process}
-Here we model $\Ntrait$ traits evolving along the phylogenetic tree that are correlated between them.
-Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology.
-The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-Along branch $\Branch$ with length $d_{\Branch}$, the Brownian process start at the ancestral node $\mathcal{A}(\Branch)$ with value $\Brownian(\mathcal{A}(\Branch))$, and ends at node $\mathcal{R}(\Branch)$  with value $\Brownian(\mathcal{R}(\Branch))$.
-The independent contrast $\contrast_{\Branch}$ defined as change in trait along the branch normalized by $\sqrt {d_{\Branch}}$ is a multivariate Gaussian:
-\begin{equation}
-    \label{eq:DistribBrownian}
-    \contrast_{\Branch} = \frac{\Brownian (\mathcal{R}(\Branch)) - \Brownian (\mathcal{A}(\Branch)) }{\sqrt {d_{\Branch}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
-\end{equation}
-
-\subsection{Sampling the covariance matrix}\label{subsec:sampling-the-covariance-matrix}
-From the independent contrast at each branch of the tree ($\contrast_{\Branch}$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$, as:
-\begin{equation}
-    \Scattermatrix = \sum\limits_{\Branch=1}^{\Nbranch} \contrast_{\Branch} \MultiplyMatrix \contrast_{\Branch}\tr\label{eq:bayes-scatter},
-\end{equation}
-where $\Nbranch$ is the number of branches in the tree and $\NbrTaxa$ the number of taxa.
-
-The {prior} on the covariance matrix is an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
-\begin{equation}
-    \label{eq:Distribcovariance}
-    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
-\end{equation}
-
-By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\Brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
-\begin{equation}
-    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)\label{eq:bayes-posterior}
-\end{equation}
-This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate_{\WishartIDD}$ defined by
-\begin{equation}
-    \Multivariate_{\WishartIDD} \sim \mathcal{N} \left( \VecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
-\end{equation}
-And from these multivariate samples, $\CovarianceMatrix$ is Gibbs sampled as:
-\begin{equation}
-    \CovarianceMatrix = \left( \sum\limits_{\WishartIDD=1}^{\WishartPostDf} \Multivariate_{\WishartIDD} \MultiplyMatrix  \left[\Multivariate_{\WishartIDD} \right] \tr \right)\inv \label{eq:bayes-gibbs}
-\end{equation}
-
-\newpage
-\section{Bayesian and Maximum-likelihood implementation}\label{sec:implementation}
-
-Implementation is included within the \textit{BayesCode} software, available at \censor{\url{https://github.com/ThibaultLatrille/bayescode}}.
+Bayesian implementation is included within the \textit{BayesCode} software, available at \url{https://github.com/ThibaultLatrille/bayescode}.
 
@@
 \begin{enumerate}
-    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers), from which the values of nucleotide divergence ($d$) used in denominator of eq.~\ref{eq:estimated-rate-phy} are used.
+    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers), from which the values of nucleotide divergence ($d$) are used.
     \item A file containing the mean trait values for each species.
@@
 
-\subsection{Maximum likelihood estimation}\label{subsec:maximum-likelihood-estimation}
-
-To obtain the ratio (without the posterior credible interval and probability) using maximum likelihood computation, the following python script can be used:
-\begin{lstlisting}[language = sh, label={lst:neutrality_index}]
-python3 utils/neutrality_index.py --tree data/body_size/mammals.male.tree
-                                  --traitsfile data/body_size/mammals.male.tsv
-                                  --var_within data/body_size/mammals.male.var_trait.tsv
-                                  --output results_ML_mammals_male.tsv
-\end{lstlisting}
-
-\newpage
-\section{Saturation of phenotypic divergence}\label{sec:supp-distance}
-
-The phenotype is encoded by a genetic architecture and is thus ultimately bounded.
-At the macro-evolutionary scale, phenotypic divergence should plateau at some point, ultimately resulting in a $\RateBetween$.
-
-\begin{center}
-    \captionof{figure}{Saturation of phenotypic divergence.}
-    \includegraphics[width=1.0\textwidth, page=1] {figureS2}
-    \label{fig:supp-distance}
-\end{center}
-\begin{itemize}
-    \item Left column: $1,000$ simulations with low divergence between species (half that of mammals).
-    \item Middle column: $1,000$ simulations with high divergence between species (4 times that of mammals).
-    \item Right column: $200$ simulations with very high divergence between species (20 times that of mammals).
-    \item Top row: Simulation of neutral trait, phenotypic divergence between species as a function of the nucleotide divergence.
-Phenotypic divergence is computed between two species as the covariance between the trait ($\VarPhy$), and the nucleotide divergence is computed as the number of substitutions per site shared between the two species ($d$).
-Each point is a pair of species, and the bounds of the intervals in grey are the 2.5\% and 97.5\% quantiles across the replicates.
-The blue line is the linear regression and the red line is the saturation model ($y = \alpha \Multiply x / (\beta + x)$).
-    \item Bottom row: Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-Histogram of probabilities of $\EstNI$ being greater than $1$.
-\end{itemize}
-Under a model of neutral trait evolution, when mutation rate increases, or equivalently the divergence between species increases, the phenotypic divergence between species saturates faster than the nucleotide divergence.
-This saturation effect can result in a spurious signal of stabilizing selection ($\EstNI < 1$) for deeper phylogeny when the trait is evolving neutrally.
 \end{document}

### 2024-07-05 manuscript/main.tex
Cleaned unecessary files. Updated README.

@@
 \usepackage{caption}
-\usepackage{censor}
 \usepackage{xcolor, colortbl}
@@
 \DeclareMathOperator{\E}{\mathbb{E}}
-\DeclareMathOperator{\Var}{\text{Var}}
-\DeclareMathOperator{\Cov}{\text{Cov}}
-\newcommand{\QstFst}{Q$_\text{ST}$--F$_\text{ST}$}
+\DeclareMathOperator{\Var}{\text{var}}
+\DeclareMathOperator{\Cov}{\text{cov}}
+\newcommand{\Qst}{Q$_\text{ST}$}
+\newcommand{\Fst}{F$_\text{ST}$}
+\newcommand{\QstFst}{\Qst--\Fst}
 \newcommand{\der}{\mathrm{d}}
@@
 \newcommand{\Time}{t}
+\newcommand{\NbrGen}{t_{\Spi, \Spj}}
+\newcommand{\NucDiv}{d_{\Spi, \Spj}}
 \newcommand{\Trait}{P}
@@
 \newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
-\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
+\newcommand{\MeanTrait}{\bar{\Trait}}
 \newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
-\newcommand{\Root}{0}
-\newcommand{\RootTrait}{\Trait_{\Root}}
-\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
+\newcommand{\RootTrait}{\phi}
+\newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
 \newcommand{\VecZero}{\UniDimArray{0}}
@@
 
-% Including genomics, Inclusion of genomics, Linking genomics,
-\title{Leveraging genomics to detect trait selection from within and between-species variations}
-\rhead{\scshape Trait selection from within and between-species variations}
-
-\author{
-\large
-\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M. {Bastian}$^{2}$, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
-\normalsize
-$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
-$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
-\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
-}
+\title{Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes}
+\rhead{\scshape Trait selection from within and between-species variation}
+
 \author{~}
@@
 
-Determining whether a trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
+Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
 Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
 The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
-All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.
-
-Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
-Additionally, contrasting both trait and genetic differentiation across several populations, \QstFst methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
-For example in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
-[why only example from gene expression?]
-However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.
-
-To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
-For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
-Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for gene with large within-species variation~\parencite{khaitovich_neutral_2004}.
-Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
+All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\citep{lynch_genetics_1998, walsh_evolution_2018}.
+
+Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
+Given this genetic architecture of traits, from two diverging populations, it is possible to distinguish which traits have evolved under natural selection in controlled experimental settings, by performing genetic cross between individuals~\citep{fraser_detecting_2020}.
+Across several populations, by contrasting both trait differentiation (\Qst) and genetic differentiation (\Fst), so-called \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
+\Qst\ higher than \Fst\ is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
+Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\citep{lamy_qst_2012}.
+Other frameworks explicitly model genetic drift as a random process generating both trait and genetic differences between individuals and populations.
+This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
+However, regardless of the strengths and weaknesses of each method~\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.
+
+% Pair of species
+To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
+For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes that will reach fixation in the population.
+These changes ultimately result in different mean trait values across lineages.
+Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984, felsenstein_phylogenies_1988a}.
+Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
+As an analogy, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \citet{mcdonald_adaptative_1991} test.
+In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
+Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.
 
 % Phylogenetics (mean trait evolution)
-Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
-Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
-As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
+Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
+Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, hansen_translating_1996}.
+As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
 Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
-Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
-Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
-However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
-First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
+Deviations from the assumptions of the Brownian process are however well known.
+When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, catalan_drift_2019}.
+Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
+However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
+First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
 In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
-Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Second, a better fit of a Brownian could be due to a trait evolving with a rate too low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
+And third, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
 Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.
@@
 % At the frontier Phylogenetics/Population-genetics
-At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value to describe technical errors~\parencite{lynch_methods_1991, felsenstein_comparative_2008, hansen_interpreting_2012} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
-One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average as a signal of diversifying selection~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
-We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
-Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
-For population geneticists, our study can be seen as the macro-evolutionary generalization of \QstFst methods to account for phylogenetic relationships between species.
-For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
-% Basically, we are testing whether the variance of means is greater than the mean of variances
-
-% Basically, we are testing whether the variance of means is greater than the mean of variances
+At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
+Across many species, within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within species variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
+Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
+We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
+It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
+The ratio that we propose can be considered as a neutrality index for any quantitative trait ~\citep{lynch_rate_1990}, while articulating trait and nucleotide variation within and between species.
+Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
+From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary analog to account for phylogenetic relationships between species.
+From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
 
@@
 
-Prior to developing our neutrality index, we review theoretical expectations for variations of quantitative traits and genomic sequences under neutral evolution for both within- and between-species variation.
+While observing trait variations across individuals of several species, we ask if the variation within species compared to variation between species is compatible with neutral evolution or not.
+In statistical terms, this can also be framed as: Is the variance of means equal to the mean of variances?
+The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times.
+By reviewing theoretical expectations and leveraging nucleotide sequence variations, the goal of this section is thus to obtain normalized trait variation between and within species that are equal if the trait is neutral.
+Here we denote these normalized trait variations as respectively $\RateWhithin$ for within species and as $\RateBetween$ for between species.
 
@@
 For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
-New mutations are generating trait variance and the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
-At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
-As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$:
+For a diploid individual, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
+As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per locus per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
 \begin{gather}
-    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
+    \VarMutation = 2 \MutationRatePheno \Multiply \GenArchi \label{eq:var-mutation}.
 \end{gather}
 
-While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
-For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
+While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\citep{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
+For a neutral trait at equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
 \begin{align}
     \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
-    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
+    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
 \end{align}
 
-For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating in the population divided by the length of the region.
+For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all nucleotide sites~\citep{tajima_statistical_1989}.
 Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
-As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$):
+As shown in \citet{tajima_statistical_1989}, $\pi$ is a function of the mutation rate ($\MutationRateNuc$, per nucleotide site per generation) and the effective population size ($\Ne$):
 \begin{gather}
-    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
+    \pi = 4 \Ne \Multiply \MutationRateNuc \label{eq:genetic-diversity}.
 \end{gather}
 
-We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
-This ratio allows to remove the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
+To remove the effect of $\Ne$, we define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
+After simplification, $\RateWhithin$ is then solely a function of the underlying genetic architecture as:
 \begin{align}
-    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
-    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
-    & = \RateMut. \label{eq:rate-pop}
+    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi} \label{eq:def-rate-pop}, \\
+    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Ne \Multiply \MutationRateNuc} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
+    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-pop}
 \end{align}
 
-The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
+If $\VarGenetic$ is not empirically accessible, it can be related to the observed phenotypic variance ($\VarPhenotype$), multiplied by narrow-sense heritability of the trait ($\Heritability$), as~\citep{hill_data_2008}:
 \begin{align}
-    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
+    \VarGenetic & = \Heritability \Multiply \VarPhenotype.\label{eq:var-gen-pheno}
 \end{align}
-
-\subsubsection*{Between-species trait variations}
-
-For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
-If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
-The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given by \parencite{hansen_translating_1996}:
+Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ instead of $\VarGenetic$ as:
 \begin{align}
-    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
-    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
+    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi } \text{ from definition eq.~\ref{eq:def-rate-pop} and~\ref{eq:var-gen-pheno}}, \label{eq:def-rate-pheno-pop} \\
+    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \text{ from eq.~\ref{eq:rate-pop}} \label{eq:rate-pheno-pop}
 \end{align}
 
+\subsubsection*{Between-species trait variations}\label{subsec:between-species-var}
+For a given species $\Spi$, we denote by $\MeanTrait_{\Spi}$ the mean value of the trait across the individuals.
+If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
+Given a phylogenetic tree, for a pair of species $\Spi$ and $\Spj$ from this tree, we denote as $\NbrGen$ the number of generations between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \citet{hansen_translating_1996} by:
+\begin{align}
+    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \NbrGen \\
+    & = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}}.
+\end{align}
 Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
-The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
-We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
+The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$~\citep{kimura_probability_1962}.
+We can derive the substitution rate ($\SubRate$, per nucleotide site per generation) as the number of newly arisen mutations ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{kimura_evolutionary_1968}, giving:
 \begin{align}
-    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
-    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
-    & = \MutationRate. \label{eq:substitution-rate}
+    \SubRate & = 2 \Ne \Multiply \MutationRateNuc \Multiply \pfix, \\
+    & = 2 \Ne  \Multiply \MutationRateNuc  \Multiply \dfrac{1}{2\Ne}, \\
+    & = \MutationRateNuc. \label{eq:substitution-rate}
 \end{align}
-That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate with which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.
-
-After $\Time$ generations and assuming that no multiple substitutions occurred at the same site, the nucleotide divergence $d$, which is the fraction of the genomic region that generated a substitution, will be $\Time$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
+That is, if mutations are neutral, the rate of substitution per generation within a genomic region equals the rate at which new mutations arise per generation for the same genomic region, reviewed in \citet{mccandlish_modeling_2014}.
+
+Next, we denote $\NucDiv$ as the nucleotide divergence between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+In other words, $\NucDiv$ is the expected number of substitutions per nucleotide site during the $\NbrGen$ generations.
+Assuming that no multiple substitutions occurred at the same site, $\NucDiv$ is the number of generations ($\NbrGen$) multiplied by the nucleotide substitution rate per generation ($\SubRate$):
 \begin{align}
-    d & = \Time \Multiply \SubRate \\
-    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
+    \NucDiv & = \NbrGen \Multiply \SubRate \\
+    & = \NbrGen \Multiply \MutationRateNuc \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
 \end{align}
-We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
-This ratio allows to remove the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
+To remove the effect of the number of generations ($\NbrGen$) first, and to also equate to $\RateWhithin$ (eq.~\ref{eq:rate-pop}), we define $\RateBetween$ as the covariance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4\NucDiv$).
+After simplification, $\RateBetween$ is also solely a function of the underlying genetic architecture as:
 \begin{align}
-    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
-    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
-    & = \RateMut \label{eq:rate-phy}.
+    \RateBetween & \defEqual \frac{\VarPhy}{4 \NucDiv} \label{eq:def-rate-phy},\\
+    & = \frac{4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \NbrGen \Multiply \MutationRateNuc}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
+    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-phy}
 \end{align}
+
+In equation~\ref{eq:rate-phy}, we show that the covariance in mean trait value between a pair of species ($\VarPhy$) does increase linearly with shared nucleotide divergence ($\NucDiv$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
+Importantly, since the number of generations is the ratio of time divided by generation time (average time between two consecutive generations), removing the effect of the number of generations in equation~\ref{eq:rate-phy} also removes the effect of both time and generation time.
 
@@
     \centering
-    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
+    \includegraphics[width=\textwidth, page=1] {figure1}
     \caption{
-        Between species, the mean phenotypic trait value changes along the phylogeny, allowing estimating of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
-        Within-species, the genetic variance allows estimating of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
-        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
-        Under neutral evolution, $\EstNI$ is expected to be equal to one.
-        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
-        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
+        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\RateBetween$, which is normalized by nucleotide divergence.
+        Within species, the genetic variance allows the estimation of within-species trait variation, $\RateWhithin$, which is  normalized by nucleotide diversity.
+        $\NI$ is the ratio of $\RateBetween$ over $\RateWhithin$.
+        Under neutral evolution, $\NI$ is expected to be equal to one.
+        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\NI$ greater than one.
+        Under stabilizing selection, the trait is homogeneous between species, leading to $\NI$ smaller than one.
         Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
@@
 The variability between either individuals or species can be obtained for both quantitative traits and genomic sequences.
-At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$, which equals $\RateMut$ if the trait is neutrally evolving (see above).
+At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$.
 At the phylogenetic level, the variability of the mean trait value between species can be combined with the nucleotide divergence of any neutrally evolving genomic region to obtain $\RateBetween$.
-Similarly, $\RateBetween=\RateMut$ if the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogeny.
-We thus have, for a neutrally evolving trait:
-\begin{gather}
-    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
-    \Rightarrow \NI \defEqual \frac{\RateBetween}{\RateWhithin} = 1.
-\end{gather}
-We define a neutrality index $\NI = \RateBetween / \RateWhithin$ that will equal $1$ for a trait evolving neutrally.
-Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
-Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
-
-\subsection*{Estimate}\label{subsec:estimate}
-
-Based on the comparative framework accounting for phylogenetic inertia~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a Maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
+If the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogenetic tree, we thus have:
+\begin{align}
+    \frac{\RateBetween}{\RateWhithin} &= \frac{\VarPhy}{4 \NucDiv} \Multiply \frac{\pi}{\Heritability \Multiply \VarPhenotype} \text{ by definition and eq.~\ref{eq:def-rate-pheno-pop} and~\ref{eq:def-rate-phy}},\\
+    &= \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \Multiply \frac{\MutationRateNuc}{\MutationRatePheno \Multiply \GenArchi} \text{ from eq.~\ref{eq:rate-pheno-pop} and~\ref{eq:rate-phy}}, \\
+    &= 1.
+\end{align}
+We define a neutrality index $\NI$ as:
+\begin{align}
+\NI \defEqual \frac{\RateBetween}{\RateWhithin},
+\end{align}
+ which will equal to $1$ for a trait evolving neutrally.
+Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$), generation time or time of divergence ($\NbrGen$) need to be estimated.
+Moreover, the nucleotide sequence from which $\pi$ and $\NucDiv$ are obtained should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
+
+\subsection*{Estimation}\label{subsec:estimate}
+
+We hereby seek to obtain point estimates of $\RateBetween$, $\RateWhithin$ and ultimately $\NI$.
+For each species with data available, $\RateWhithin$ as defined in eq.~\ref{eq:def-rate-pheno-pop} can be seen as a replicate sample.
+Thus, $\RateWhithin$ can be obtained by averaging out across all the sampled species.
+On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}.
+Generally, $\RateBetween$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
+As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\RateWhithin$.
+We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.
 
 \subsubsection*{Maximum likelihood estimate}
-At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitution per site) and the topology of the phylogenetic tree.
-The diagonal $\Distance_{\Spi,\Spi}$ represents the total distances from the root of the tree to each taxon ($\Spi$).
-The off-diagonal elements ($\Distance_{\Spi,\Spj} = \Distance_{\Spj,\Spi}$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
-The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
+At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the symmetric distance matrix computed from the branch lengths and the topology of the phylogenetic tree.
+The diagonal $\Distance_{\Spi,\Spi}$ represents the total nucleotide divergence from the root of the tree to each taxon ($\Spi$).
+The off-diagonal elements ($\Distance_{\Spi,\Spj} = \NucDiv$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$, as in equation~\ref{eq:distance}.
+The mean trait value at the root of the tree ($\RootTrait$) can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\citep{omeara_testing_2006}:
 \begin{gather}
@@
 
-Finally, within-species variation $\EstRateBetween$ is estimated as~\parencite{omeara_testing_2006}:
+Finally, between-species variation $\RateBetween$ is estimated as~\citep{omeara_testing_2006}:
 \begin{gather}
-    \EstRateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
+    \RateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}.
 \label{eq:estimated-rate-phy}
@@
 \begin{gather}
-    \EstNI = \frac{\EstRateBetween}{\EstRateWhithin}. \label{eq:estimated-NI}
+    \NI = \frac{\RateBetween}{\RateWhithin}. \label{eq:estimated-NI}
 \end{gather}
 
+\subsubsection*{Multivariate Brownian process}
+
+In the previous section, $\NI$ is estimated independently for each trait of interest.
+Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree, since simultaneously estimating all $\RateBetween$ allows improving their estimation~\citep{adams_multivariate_2018}.
+More specifically, trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
+The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
+The branch lengths of the tree used to model the Brownian process runs is measured in units of $4d$ ($d$ being the nucleotide divergence).
+The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\RateBetween$ (see section~S2.1).
+Of note, modeling trait evolution as a multi-dimensional process is reliable only if $\Ntrait \ll \NbrTaxa$, meaning that the number of species is largely superior to the number of traits~\citep{adams_multivariate_2018}.
+Thus, relying on a $\Ntrait$-dimensional process should be reserved for a handful of allometric traits (e.g. brain mass and body mass).
+If $\Ntrait$ is large, the traits are better tested independently each with a $1$-dimensional Brownian process, which is a specific case of the multi-dimensional process.
+
 \subsubsection*{Bayesian estimate}
-
-The Bayesian framework allows to obtain the posterior distribution the neutrality index ($\EstNI$) for a given trait.
-Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
-Trait variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
-The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
-With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
-$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is Gibbs sampled.
-For each trait and each species, prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
-Heritability and phenotypic variance for each trait is combined with nucleotide diversity to compute $\EstRateWhithin$ for each species, and is averaged out across species (as in eq.~\ref{eq:estimated-rate-pop}).
-From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
-The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.
+The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\NI$) for traits of interest.
+We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\citep{latrille_inferring_2021}.
+With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S2.2).
+We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
+For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
+Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\RateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
+From $\RateWhithin$ estimated independently for each trait and the diagonal elements of $\CovarianceMatrix$ (i.e.
+the $\RateBetween$ for each trait), the posterior distribution of $\NI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
+The posterior distribution of $\NI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\NI > 1 ]$ to test for evidence of diversifying selection and $\proba [\NI < 1 ]$ to test for evidence of stabilizing selection.
 
@@
 Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
-Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
-It is however possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
-Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in under-estimated $\EstNI$ since $\Heritability \leq H^2$.
-Contrarily, the test for stabilizing selection is invalid if $\EstNI$ is under-estimated.
+Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\NI$.
+It is, however, possible to test for diversifying selection because testing for $\NI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
+Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\NI$ since $\Heritability \leq H^2$ and thus can be used to detect diversifying selection if $\Heritability$ is not available.
+Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and fall within the range of 0.2-0.5 in a vast majority of phenotypic traits tested~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
+Thus, if available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.
+
+In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\NI$ is underestimated.
 Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.
@@
 We tested the performance of our neutrality index ($\NI$) to detect selection on a quantitative trait using simulations.
-We performed simulations under different selective regimes (neutral, stabilizing, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of mutation rate (constant or fluctuating).
-Simulation are individual-based and follows a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
-Each individual phenotypic value is the sum of genotypic value and an environmental effect.
+We performed simulations under different selective regimes (neutral, stabilizing, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of the mutation rate (constant or fluctuating).
+Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
+Each individual phenotypic value was the sum of genotypic value and an environmental effect.
 The environmental effect was normally distributed with variance $\VarEnv$.
-We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
-We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
+We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and for the theoretical formulation see section~S1.1 and fig.~S1).
+We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S1.1).
 Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
-For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
-From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
-We scaled parameters to fit plausible values for mammals.
-We used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
-
-The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
-Additionally to the geometric Brownian process, an Ornstein-Uhlenbeck process is overlaid to the instant value of log-$\Ne$ to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
+For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRatePheno \Multiply \NbrLoci $, with the mutation rate per locus per generation $\MutationRatePheno$.
+From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from the root to leaves of $0.18$ and average genetic diversity of $0.00276$.
+We scaled parameters in our simulations to fit plausible values for mammals.
+We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per site per generation  and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
+We set $\MutationRatePheno=\MutationRateNuc$ without loss in generality since the genetic architecture ($\NbrLoci$ and $a$) is assumed constant in the simulator.
+
+The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modeled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
+These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$.
+In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $1.0$.
+An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
 The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
 The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
-From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.
-
-Random genetic drift was introduced by re-sampling individuals at each generation, with each parent having a probability of being sampled that was proportional to its fitness ($W$).
-Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
+From a speciation process perspective, this was equivalent to an allopatric speciation over one generation.
+
+At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
+Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
 More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
@@
 Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
-Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\parencite{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
-Neutral evolution was implemented by fixing the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.
-
-Nucleotide diversity ($\pi$) was measured as the heterozygosity of neutral markers (not linked to the trait).
-Nucleotide divergence ($d$) was measured as the number of substitutions per site of neutral markers (not linked to the trait) along branches of the phylogenetic tree.
+Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
+Neutral evolution was implemented by flattening the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.
+
+Nucleotide diversity ($\pi$) was measured as the heterozygosity of neutral markers that were simulated along the phylogenetic tree but not linked to the trait simulated.
+Nucleotide divergence ($d$) was measured as the number of substitutions per site of neutral markers along the branches of the phylogenetic tree.
 The additive genetic variance was measured as phenotypic variance multiplied by heritability.
-Heritability was estimated from the slopes of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\parencite{lynch_genetics_1998} averaged over the last 10 simulated generations.
+Heritability was estimated from the slopes of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\citep{lynch_genetics_1998} averaged over the last 10 simulated generations.
 Heritability was thus not a given parameter of the simulations, but rather measured as it would be in empirical data.
@@
         Left panel: For a given individual, the trait phenotypic value is the sum of genotypic value and a environmental effect (standard deviation $\VarEnv$).
-        The trait's genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
-        Parents are selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
-        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRate$ to mutate.
+        The trait's genotypic value is encoded by $\NbrLoci$ independent loci (meaning no linkage), with each locus contributing additively to the genotypic value.
+        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
+        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRatePheno$ to mutate.
         Drift is modeled by the resampling of parents.
-        Right panel: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
+        Right panel: examples of a trait evolving along a phylogenetic tree, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
     }
@@
 
-We analyzed a dataset of body and brain masses in mammals.
-The log-transformed values of body and brain masses were taken from \textcite{tsuboi_breakdown_2018}.
+We analyzed a dataset of body and brain masses from mammals.
+The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
 We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
-We discarded species with only one representative sample.
-The mammalian nucleotide diversity was obtained from the Zoonomia project~\parencite{genereux_comparative_2020}, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.
-
-We also analyzed a dataset of primate species, with the nucleotide variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
+We discarded species with only one representative since phenotypic variance cannot be estimated.
+The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
+More specifically, nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \citet{wilder_contribution_2023}.
+
+We also analyzed a dataset of primate species, with the nucleotide variation obtained from \citet{kuderna_global_2023} and the quantitative trait variation also from \citet{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
 However, the primate nucleotide divergence was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.
+As such, the evidence for $\NI > 1$ does not necessarily imply that the trait is evolving under diversifying selection since non-neutral markers included in the estimate of divergence can lead to a spurious $\NI > 1$ (see table~\ref{table:assumptions}).
 
@@
 For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
-We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see methods), suggesting that traits for which this relationship is not verified are putatively under selection.
-Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
-In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
-Our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between-species (see fig.~\ref{fig:methods}).
-This statistic can be obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
-We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilizing selection as $\proba [\EstNI < 1]$.
-
-\begin{table*}[t!]
-    \centering
-    \begin{adjustbox}{width = 1.0\textwidth}
-        \begin{tabular}{||l|l||c|c||c|c||}
-            \hline
-            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
-            Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
-            Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
-            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
-            Markers for polymorphism under balanced selection & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
-            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
-            Multiple nucleotide substitutions at the same locus & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\\hline \hline
-        \end{tabular}
-    \end{adjustbox}
-    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
-    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
-    }
-    \label{table:assumptions}
-\end{table*}
+We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see Materials and Methods), suggesting that traits for which this relationship was not verified were putatively under selection.
+Under stabilizing selection, the variation between species is depleted because the mean trait is maintained toward similar values between different species, which leads to $\NI < 1$.
+In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\citep{hansen_stabilizing_1997}, which leads to $\NI > 1$.
+Our neutrality index for a quantitative trait leveraged the data for any number of species, and took advantage of the signal over the whole phylogenetic tree, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (Fig.~\ref{fig:methods}).
+This statistic was obtained as a maximum likelihood estimate from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
+We also devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\NI > 1]$, and stabilizing selection as $\proba [\NI < 1]$.
+
+Our neutrality index made a series of assumptions that we described in details in the Material and Methods section.
+Table~\ref{table:assumptions} summarized these assumptions and outlined possible consequences for the neutrality test that we proposed.
 
@@
 The inference framework was first tested on independently simulated datasets matching an empirically relevant mammalian empirical regime (see Materials and Methods).
-Under constant population size ($\Ne$) and constant mutation rate ($\MutationRate$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\EstNI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\EstNI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
-For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
-Once we introduced fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
-For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).
+Under constant population size ($\Ne$) and constant mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) across the phylogenetic tree (fig.~\ref{fig:results-simulations}, top row), we found no false negative for simulations of stabilizing ($\proba [\NI < 1] > 0.975$; blue in fig.~\ref{fig:results-simulations}) or diversifying ($\proba [\NI > 1] > 0.975$; red in fig.~\ref{fig:results-simulations}) selection.
+For simulations under neutral evolution, 77\% of those were correctly identified ($0.025 \leq \proba [\NI > 1] \leq 0.975$; yellow in fig.~\ref{fig:results-simulations}), while 21\% and 2\% were wrongly detected as stabilizing or diversifying selection, respectively.
+Once we introduced fluctuating $\Ne$, $\MutationRatePheno$ and $\MutationRateNuc$(Fig.~\ref{fig:results-simulations}, bottom row), our ability to identify simulations under either diversifying or stabilizing selection remained the same with all cases detected correctly.
+For simulations under neutral evolution, 51\% of the simulations were correctly detected ($0.025 \leq \proba [\NI > 1] \leq 0.975$), while 49\% were detected as stabilizing selection ($\proba [\NI < 1] > 0.975$) and none as diversifying selection.
 
@@
     \centering
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.rho}
-    \end{minipage}
-    \begin{minipage}{0.49\textwidth}
-        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.pvalues}
-    \end{minipage}
+    \includegraphics[width=\textwidth, page=1] {figure3}
     \caption{
-        $10,000$ replicate simulations of trait evolution along a phylogeny under different selection regimes.
+        $10,000$ simulations of trait evolution along a phylogenetic tree under different selection regimes.
         Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
-        Histogram of estimated the ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$, as $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and the histogram of probabilities that the estimated $\EstNI$ is greater than $1$ (right).
-        Effective population size ($\Ne$) and mutation rate ($\MutationRate$) are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
+        Histogram of ratio of between-species trait variation ($\RateBetween$) over within-species trait variation $\RateWhithin$ with $\NI = \RateBetween / \RateWhithin$ estimated from each simulated data (left) and probabilities of $\NI$ being greater than $1$ (right).
+        Effective population size ($\Ne$) and mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
     }
@@
     Test of diversifying selection on a mammal and a primate dataset, by splitting males (\Male) and females (\Female).
-    Traits considered are body mass or brain mass (log-transformed).
-    Heritability ($\Heritability$) is either assumed complete ($\Heritability=1.0$) or uniformly distributed between 20\% and 40\%  ($\Heritability \sim \mathcal{U}(0.2, 0.4)$).
-    $\NbrTaxa$ is the number of species in the dataset.
-    $\EstNI$ is the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\EstNI$ also computed.
-    $\proba [\EstNI > 1 ]$ is the estimated posterior probability of diversifying selection.
+    Traits considered were body mass or brain mass (log-transformed).
+    Heritability ($\Heritability$) was either assumed complete ($\Heritability=1.0$) or uniformly distributed between 20\% and 40\%  ($\Heritability \sim \mathcal{U}(0.2, 0.4)$).
+    $\NbrTaxa$ was the number of species in the dataset.
+    $\NI$ was the posterior estimate of our neutrality index, with the $95\%$ credible interval (CI) for $\NI$ also computed.
+    $\proba [\NI > 1 ]$ was the estimated posterior probability of diversifying selection.
     }
@@
 
-For mammalian body and brain mass, we obtained male (\Male) and female (\Female) trait variation.
-Combined with nucleotide diversity and divergence, we estimated $\EstNI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in table~\ref{table:empirical}.
-Assuming complete heritability, brain mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
-If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\parencite{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
+For mammalian body and brain mass, we obtained male (\Male) and female (\Female) trait variations.
+Combined with nucleotide diversity and divergence, we estimated $\NI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in the Table~\ref{table:empirical}.
+Assuming complete heritability, body mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
+If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\citep{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
 Mammalian brain mass was found to be under diversifying selection with posterior probabilities of $0.877$ for males and $0.972$ for females when complete heritability was assumed.
@@
 
-We also analyzed a similar dataset for body mass focussing this time only at Primates (see table~\ref{table:empirical}).
-For primate body mass, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%,
-Assuming complete heritability of body mass did not change the posterior probability for males, but increased the one for female to $1.0$.
-Evidence for diversifying selection on body mass was therefore more pronounced in Primates than in Mammals.
-However, the genetic markers used to normalize trait variance with divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\EstNI$ (see table~\ref{table:assumptions} and methods).
+We also analyzed a similar dataset for body mass focusing this time only at Primates (Table~\ref{table:empirical}).
+For primates body mass, assuming complete heritability, body mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females, exactly as in the mammal dataset.
+However, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%.
+For brain mass, assuming complete heritability or not (between 20\% and 40\%) did not change the posterior probability of diversifying selection, which was $1.0$.
+Evidence for diversifying selection on both brain and body mass was therefore more pronounced in Primates than in mammals.
+However, the genetic markers used to normalize trait variance with nucleotide divergence were not necessarily neutral, which could create spurious false positives by artificially inflating $\NI$ (Table~\ref{table:assumptions} and methods).
+
+\begin{table*}[t!]
+    \centering
+    \begin{adjustbox}{width = 1.0\textwidth}
+        \begin{tabular}{||l|l||c|c||c|c||}
+            \hline
+            Broken assumption                                       & Consequences                                       & $\RateWhithin$   & $\RateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
+            Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
+            Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Phenotypic plasticity & Trait responding to individual environments  & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
+            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
+            Markers for polymorphism under balanced selection & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
+            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
+            Multiple nucleotide substitutions at the same locus & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\\hline \hline
+        \end{tabular}
+    \end{adjustbox}
+    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\RateWhithin$), between-species variation ($\RateBetween$), and on the neutrality index $\NI = \RateBetween/\RateWhithin$.
+    The last two columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
+    }
+    \label{table:assumptions}
+\end{table*}
 
@@
 In this study, we proposed a neutrality index for a quantitative trait that can be used within a statistical framework to test for selection.
-Our neutrality index for a trait, $\NI$, is the ratio of the normalized between- to within-species variation and it allows the identification of the evolutionary regime of a quantitative trait.
-At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
-Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
-Estimated $\EstNI$ can be compared against the null hypothesis of neutrality by testing for deviation from the value of $1.0$.
-Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S\ref{sec:implementation}) also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g.
-$ \proba [ \EstNI > 1 ]$).
-We tested our statistical procedure against simulated data and showed that for simulations under diversifying selection (test of $\EstNI > 1$) or under stabilizing selection (test of $\EstNI < 1$) our test is able to correctly detect those regimes of selection.
-However, for simulations of a neutral trait, our test detects frequently a spurious signal of stabilizing selection ($\EstNI < 1$).
-Thus, we argue that our test should be used to detect diversifying selection, but has low accuracy to detect stabilizing selection due to many false positives.
+Our neutrality index for a trait, $\NI$, is calculated as the ratio of the normalized within- to between-species variation and it allowed the identification of the evolutionary regime of a quantitative trait.
+At the phylogenetic scale, trait variation between species was normalized by sequence divergence obtained from a neutral set of markers.
+Similarly, trait variation within species was normalized by sequence polymorphism obtained also from a neutral set of markers.
+Our estimate of $\NI$ could be tested for deviation from the value of $1.0$ expected under the null hypothesis of neutrality.
+Technically, the neutrality index can be estimated either as a maximum likelihood point estimate, or as a mean posterior estimate from a Bayesian implementation (see section~S3).
+The latter also enabled the estimation of the posterior credible interval to test for departure from a neutrally evolving trait (e.g. $ \proba [ \NI > 1 ]$).
+We tested our statistical procedure against simulated data and showed that our test was able to correctly detect simulations under diversifying selection (test of $\NI > 1$) or under stabilizing selection (test of $\NI < 1$).
+However, our test detected a spurious signal of stabilizing selection ($\NI < 1$) when we simulated the evolution of a neutral trait.
+An assumption of our test is that the neutral phenotypic trait is evolving as a Brownian process and is, therefore, unbounded.
+However, the phenotype may be bounded by what the genetic architecture can produce, and this could cause a slowdown of phenotypic divergence over time due to the erosion of possible phenotypic changes at the underlying loci.
+Typically, such an effect depends on the number of alleles per locus, whether new mutations are generating new alleles or instead reverting to previous alleles.
+Altogether, in our simulation setting under a constant genetic architecture with a fixed number of loci, such a slowdown of phenotypic divergence can result in a spurious signal of stabilizing selection ($\NI < 1$), especially for deeper phylogeny (figure~S2 and section~S4)
+We thus argue that our method should be used to detect diversifying selection, but that it had low accuracy to detect stabilizing selection due to false positives.
 
 % How is our test related to other tests of selection of traits
-Based on these results, we argue that our method is a significant improvement over current methods to detect selection acting on a trait at the phylogenetic scale.
-Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Thus our method does not provide improvement to detect stabilizing selection at the phylogenetic scale.
-However, to detect diversifying selection, current methods modelling mean trait value are limited by a theoretical challenge.
-Namely, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus can not be distinguished from neutral evolution~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-For example, test for neutrality concluded that the expression level of the majority of genes are under a neutral regime~\parencite{catalan_drift_2019}.
-Our diversity index allows to discriminate the alternative model of diversifying selection from the neutral case, inviting us to revisit such studies.
-On top of between-species variation of the mean trait value, methods have normalized by within-species variations to detect selection, and have compared this statistic across a pool of traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
-This strategy allows to identify outlier traits putatively under diversifying selection, but does not allow to test for selection on a single trait and requires an independent confirmation~\parencite{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
-Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution for a single test.
-Again, our diversity index can thus allow us to revisit these studies, assuming we have access to genomic datasets to estimate nucleotide divergence and polymorphism.
+Our results showed that our method significantly improved over currently available methods to detect selection acting on a trait at the phylogenetic scale.
+Current methods relying on evolution of the mean trait value between species also tend to statistically prefer a model of stabilizing selection over a Brownian process when the trait is neutral~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Our approach could in theory be applied to detect stabilizing selection at the phylogenetic scale, but we showed that it did not have the statistical power to identify those cases.
+In contrast, we showed that our method was able to identify correctly cases of diversifying selection, which is a clear improvement over current methods that model only mean trait value.
+Indeed, under diversifying selection, mean trait value will not deviate from a Brownian process, and thus cannot be distinguished from neutral evolution~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
+For example, testing the selective regime in the expression level of the majority of genes led to the selection of a Brownian process as the prefered model and the interpretation that the expression was evolving neutrally~\citep{catalan_drift_2019}.
+Instead, our diversity index has the advantage to discriminate the alternative model of diversifying selection from the neutral case by comparing within- and between-species variation while correctly normalizing them using nucleotide markers.
+Our approach is not the first one coupling between-species and within-species variations, and those approaches employ different strategies to detect selection.
+First, one empirical strategy is to compare the ratio of between to within variation across a pool of traits, which allow to identify outlier traits putatively under diversifying selection~\citep{rohlfs_modeling_2014}.
+However, this method does not formally allow testing for diversifying selection, and requires many traits such as expression level data to seek outliers genes~\citep{rohlfs_phylogenetic_2015, gillard_comparative_2021}.
+Second, other methods leverage Lande’s generalized genetic distance (LGGD), which relate the ratio of between to within variations to population-genetic parameters~\citep{lande_quantitative_1979, lynch_analysis_1990, lemos_evolutionary_2001, lemos_rates_2005, weaver_were_2007, porto_rate_2015}.
+Specifically, by leveraging estimates of effective population size ($\Ne$) and number of generations between species, or alternatively by assuming their constancy, these methods can test for departures from the null model of neutral evolution for a single trait.
+Such methods have been successful in identifying specific instances of diversifying selection~\citep{schroeder_evolution_2017, machado_preeminent_2022} and near-drift~\citep{machado_using_2023}.
+However, $\Ne$ and the number of generations are complex parameters to correctly infer, and is usually done for a pair or only a few species, and ultimately requires large genomic datasets and heavy statistical methods~\citep{wilder_contribution_2023}.
+Instead, our diversity index opens new avenues to revisit these studies testing for the selective regime affecting the quantitative traits, by formally incorporating nucleotide divergence and polymorphism, bypassing estimation of $\Ne$, generation time and calibration of ancestral node ages~\citep{machado_using_2023}.
 
 % Why is our test different
-We argue that the main novelty of this study is being able to use nucleotide divergence and polymorphism to normalize trait variation between and within species.
-In the context of within species variation but across several populations, \QstFst test developed from trait and sequence allow to test for selection of a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
-Similarly, in our test, the genetic sequences from which nucleotide divergence and polymorphism are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.
-Nucleotide variation allow to normalize for diversity driven by confounding factors such as population sizes ($\Ne$), mutation rates ($\MutationRate$) and generation time~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
-Thus our test avoids to estimate them and also avoids to estimate divergence time which is otherwise necessary~\parencite{walsh_evolution_2018}.
-But importantly, by normalizing with sequence variation, we also show using simulated data that our test is not sensitive to the assumption that $\Ne$, $\MutationRate$ and generation time are constant across the phylogeny, an unmet assumption empirically~\parencite{bergeron_evolution_2023, wilder_contribution_2023}.
-Indeed, under the neutral case of evolution, changes in $\Ne$, $\MutationRate$ and generation time impacts similarly trait and sequence variation.
-Thus the normalization by nucleotide divergence and polymorphism automatically absorbs long-term and short-term changes in $\Ne$, $\MutationRate$ and generation time which cancels out in the ratio of trait variation $\EstNI$.
+As such, the main novelty of our study was to use the nucleotide divergence and polymorphism to normalize trait variation between and within species.
+In this context, our test bears many similarities to \QstFst\ tests (and their derivatives) that have been developed to test for selection of a trait across several populations while also leveraging sequence variation~\citep{martin_multivariate_2008, leinonen_qst_2013} or co-ancestry between individuals~\citep{ovaskainen_new_2011}.
+Our method can be seen as an analog at the phylogenetic scale, where although the sequences used should be neutrally evolving, they can be obtained from different sampled individuals than for the trait.
+Importantly, by normalizing with sequence variation, we also showed using simulated data that our test was not sensitive to the assumption that $\Ne$ and mutation rates were constant across the phylogenetic tree, an unmet assumption empirically~\citep{bergeron_evolution_2023, wilder_contribution_2023}.
+Indeed, under the neutral case of evolution, the normalization by nucleotide divergence and polymorphism automatically absorbed long-term and short-term changes in $\Ne$, generation time and mutation rates, which canceled out in the neutrality index $\NI$.
+
+In the context of phylogenetic comparative methods, modeling mean trait evolution as a function of nucleotide divergence ($d$) instead of time has more general consequences.
+As an example, trait variation is often modeled as a Brownian process running on a time-calibrated tree, which can produce biases~\citep{litsios_effects_2012}.
+Indeed, for a neutrally evolving trait, trait variation depends directly on the number of generations, which in turn correlates with time.
+But, since species generation time might vary along the phylogenetic tree, $d$-scaled trees absorbing changes in generation time should be used instead of time-scaled trees.
+Using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks).
+We argue, that the soundness of studying trait evolution on $d$-scaled trees can be evaluated by the absolute fit of a model to the data~\citep{pennell_model_2015}.
+More generally, genomic information could potentially be seen as a way to disentangle congruence models~\citep{louca_extant_2020}, or as prior for methods that detect shifts in adaptive regimes~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020},.
 
 % How is our test related to other tests of selection on sequences
-Even though our test is developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provide insight into its behavior.
-First, we acknowledge that our test took inspiration from the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, where synonymous mutations are used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
-Second, because $\NI$ is compared to 1, our test ultimately bears analogy to the codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\parencite{goldman_codonbased_1994, muse_likelihood_1994}.
+Even though our test was developed for a quantitative trait, analogies with other tests of selection developed for molecular sequences also provided insight into its behavior.
+First, we acknowledge that our test took inspiration from the \citet{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences in a pair of species, except that the non-synonymous versus synonymous distinction is replaced by the comparison between quantitative trait and neutral genomic sequence.
+Second, at the phylogenetic scale, when comparison is done across several species, our test also bears analogy to codon-based test of selection, where the ratio of non-synonymous to synonymous substitutions ($\dnds$) is compared to 1~\citep{goldman_codonbased_1994, muse_likelihood_1994}.
 As $\dnds < 1$ is interpreted as purifying selection acting on the protein, $\NI < 1$ is interpreted as stabilizing selection acting on the trait.
 Similarly, the interpretation of adaptation for $\dnds > 1$ is analogous to diversifying selection for $\NI > 1$.
-With this analogy in mind, we can leverage the vast bibliography discussing and interpreting results of these tests and their pitfalls~\parencite{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
-First, not rejecting the neutral null model of $\NI = 1$ does not necessarily imply that the trait is effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\parencite{nielsen_molecular_2005}.
-Second, empirical evidence for $\NI < 1$ does not rule out diversifying selection, but rather that this diversifying selection is not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\parencite{latrille_genes_2023}.
-By explicitly modeling stabilizing selection as a moving optimum, it would theoretically allow to tease out the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
+With this analogy in mind, we could leverage the vast literature discussing and interpreting the results of these tests and their pitfalls~\citep{nielsen_molecular_2005, anisimova_investigating_2009, jensen_importance_2019}.
+First, not rejecting the neutral null model of $\NI = 1$ did not necessarily imply that the trait was effectively neutral, since diversifying and stabilizing selection could compensate each other resulting in $\NI = 1$, analogously to $\dnds=1$ under a mix of adaptation and purifying selection~\citep{nielsen_molecular_2005}.
+Second, empirical evidence for $\NI < 1$ did not rule out diversifying selection, but rather that this diversifying selection was not strong enough to overcome the stabilizing selection, similarly to strong purifying selection resulting $\dnds < 1$ even though those genes and sites are under adaptation~\citep{latrille_genes_2023}.
+By explicitly modeling stabilizing selection as a moving optimum, it would theoretically be possible to tease apart the effect of diversifying and stabilizing selection in the context of quantitative traits to obtain a statistically more powerful test.
 
@@
 In the context of detecting diversifying selection on a trait, we argue that the main drawback of our method is that the additive genetic variance of the trait is required instead of the phenotypic variance.
-If phenotypic variance is used instead of additive genetic variance to estimate $\EstNI$, meaning that we assumed complete heritability, the neutrality index $\EstNI$ is ultimately under-estimated.
-Similarly, using broad-sense heritability instead of narrow-sense heritability results in under-estimated $\EstNI$.
-In such context, the test of stabilizing selection ($\EstNI < 1]$) is statistically invalid.
-However, the test of diversifying selection ($\EstNI > 1$) is underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
-As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\EstNI > 1$.
+If phenotypic variance was used instead of additive genetic variance to estimate $\NI$, meaning that we assumed complete heritability, the neutrality index $\NI$ was ultimately underestimated.
+Similarly, using broad-sense heritability instead of narrow-sense heritability would result in underestimated $\NI$.
+In such context, the test of stabilizing selection ($\NI < 1]$) would be statistically invalid.
+However, the test of diversifying selection ($\NI > 1$) was underpowered although not invalided, meaning that absence of evidence would not be evidence of absence.
+As an example, even though we assumed complete heritability for brain mass, we uncovered diversifying selection in mammals since $\NI > 1$.
+If available, any prior knowledge on heritability can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
+Additionally, phenotypic plasticity also affects the genotype-phenotype relationship with intricate consequences for our test of selection.
+First, at the level of within species variation, individuals might occupy different patches with different environments.
+Responding to these individual environmental conditions, phenotypic plasticity would then result in increased trait variation within species.
+In this scenario, as hypothesized in \citet{rohlfs_phylogenetic_2015}, phenotypic plasticity then leads to a reduced ratio of between to within species variations, thus ultimately leading to our tests of diversifying selection being underpowered although not invalid.
+Alternatively, it is also possible that different species are experiencing different macro-environments, for example with species spread along a latitudinal or elevation gradient, with different temperatures or precipitation.
+These species could thus have different mean phenotypes solely because of phenotypic plasticity, while such changes are not encoded in their genome~\citep{stamp_relative_2020, schraiber_heritability_2023}.
+Such an effect can lead to $\NI > 1$ erroneously interpreting diversifying selection.
+The test of $\NI > 1$ would however be correct that the changes in mean phenotypes across species is due to change in environment, albeit in such a case not encoded by the genotype of individuals but due to phenotypic plasticity.
 
 % What can and should be improved
-The development of our neutrality index is also based on several assumptions that could be relaxed in future studies.
-First, we cannot predict the behaviour of our test in the context of population structures, gene flow and introgression.
+The development of our neutrality index was also based on several assumptions that could be relaxed in future studies.
+First, we cannot predict the behavior of our test in the context of population structures, gene flow and introgression.
 These factors should be thoroughly investigated using simulations.
-Second, loci are assumed to contribute additively to the phenotype.
-Although the effects of dominance and epistasis is typically weak compared the additive effects on the quantitative traits, their influence shall be assessed~\parencite{hill_data_2008, crow_epistasis_2010}.
-Third, the genetic architecture of the trait is assumed to be constant across the phylogeny, it might actually be variable among individuals and species~\parencite{tung_genetic_2015, huber_conservatism_2015}.
-Such assumption can theoretically be relaxed and changes in genetic architecture along the phylogeny could jointly be estimated~\parencite{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
-Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
+Second, loci were assumed to contribute additively to the phenotype.
+Although the effects of dominance and epistasis is typically weak compared to the additive effects on the quantitative traits, their influence should be assessed~\citep{hill_data_2008, crow_epistasis_2010}.
+Third, the genetic architecture of the trait was assumed to be constant across the phylogenetic tree, whereas it might actually be variable among individuals and species~\citep{tung_genetic_2015, huber_conservatism_2015}.
+Such an assumption can theoretically be relaxed and changes in genetic architecture along the phylogenetic tree could jointly be estimated~\citep{arnold_understanding_2008, hohenlohe_mipod_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
+Finally, from a statistical perspective, our Bayesian estimation could integrate uncertainty from the estimation of genetic variation, using sequences as input instead of estimated values of nucleotide diversity and divergence.
 
 % What can our test bring to the field
-From an empirical point of view, even though our method requires integrating genomic and trait variation, we also showed that our method can be applied to the empirical dataset with the illustrative example of mammals' brain and body mass.
-Because our test is also based on several assumptions that might not hold on empirical data, we also provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
-For example, at the primate scale, the evidence for $\EstNI > 1$ does not necessarily imply that the brain mass is evolving under diversifying selection since the markers used for nucleotide divergence are not neutral, which can lead to a spurious $\EstNI > 1$.
-Altogether, our study provides a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we argue that it is a promising tool to investigate the evolution of quantitative traits.
+From an empirical point of view, our method required integrating genomic and trait variation, which could reduce the possible datasets to be used.
+However, such datasets will become more and more accessible and we showed the applicability of our method by applying it to the illustrative example of mammals’ brain and body mass, both showing signals of diversifying selection.
+As such, this result corroborates studies relying solely on changes in mean trait values across mammals, showing strong statistical support for several distinct evolutionary regimes for body- and brain mass~\citep{mitov_automatic_2019}.
+Interestingly, our strongest signal is for brain mass, corroborating studies in hominids where skull size (related to brain mass) is the only trait that exceeded the expected rate of phenotypic evolution under a neutral model~\citep{lynch_rate_1990}.
+Hence, one first interpretation here is that brain mass might be an exceptional case among many phenotypic traits (e.g. dental and skeletal measures).
+Second, from a macro-evolutionary perspective, the consensus is that empirical rates of evolution calculated on phylogenetic trees and the fossil record are far inferior to the expected under drift~\citep{lynch_analysis_1990, uyeda_millionyear_2011}, where such methods assume constancy of $\Ne$, generation time and mutation rates.
+Our finding of diversifying selection on body and brain mass could be seen as an argument against that interpretation.
+In fact, rates of nucleotide evolution also show a tendency for slowing down on a longer timescale~\citep{rolland_conceptual_2023}.
+One possible interpretation is that normalization by nucleotide divergence could absorb this observed slowing rate of evolution.
+Altogether, further empirical and theoretical studies are required to disentangle this discrepancy between these different results and interpretations
+Because our test was also based on several assumptions that might not hold on empirical data, we also provided a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (Table~\ref{table:assumptions}).
+For example, at the primate scale, the evidence for $\NI > 1$ does not necessarily imply that the brain mass was evolving under diversifying selection since the markers used for nucleotide divergences were not neutral, which can lead to a spurious $\NI > 1$.
+In conclusion, our study provided a statistical framework to test for diversifying selection acting on a quantitative trait while integrating the trove of genomic data available both within and between species, and we believe that our new approach is a promising tool to investigate the evolution of quantitative traits.
 
 %TC:ignore
-\printbibliography
-
-\newpage
-
-\part*{Supplementary materials}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\setcounter{section}{0}
-
-\renewcommand{\baselinestretch}{1.0}\normalsize
-\tableofcontents
-\renewcommand{\baselinestretch}{1.5}\normalsize
-
-\newpage
-\section{Genetic architecture of the trait}\label{sec:simulator}
-
-\subsection{Genotype-phenotype map}\label{subsec:genotype-phenotype-map}
-
-\begin{itemize}
-    \item $\NbrLoci$ is the number of loci encoding the trait.
-    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
-    \item $\Ne$ is the effective number of individuals.
-    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
-    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the genotypic value for individual $i$.
-    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
-    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
-\end{itemize}
-
-\begin{center}
-    \captionof{figure}{summary of trait's genetic architecture.}
-    \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
-    \label{fig:simulator-summary}
-\end{center}
-
-within-species, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
-\begin{equation}
-    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad} \VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2\label{eq:simu-genotype}
-\end{equation}
-The theoretical additive genetic variance ($\VarGenetic$) is a function of the number of loci ($\NbrLoci$) and the effect of a mutation ($a$) as:
-\begin{equation}
-    \VarGenetic = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \label{eq:simu-var-genetic}
-\end{equation}
-
-The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
-\begin{equation}
-    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2 \label{eq:simu-between}
-\end{equation}
-
-Heritability ($\Heritability$) is defined as:
-\begin{equation}
-    \Heritability = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
-\end{equation}
-Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($\Heritability$) as:
-\begin{equation}
-    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
-\end{equation}
-
-\newpage
-\section{Bayesian estimate}\label{sec:bayesian-estimate}
-
-\subsection{Multivariate Brownian process}\label{subsec:multivariate-brownian-process}
-Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
-Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
-The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
-Along branch $j$ with length $d_{j}$, the Brownian process start at the ancestral node $\mathcal{A}(j)$ with value $\Brownian(\mathcal{A}(j))$, and ends at node $\mathcal{R}(j)$  with value $\Brownian(\mathcal{R}(j))$.
-The independent contrast $\contrast_{j}$ defined as change in trait along the branch normalized by $\sqrt {d_{j}}$ is a multivariate Gaussian:
-\begin{equation}
-    \label{eq:DistribBrownian}
-    \contrast_{j} = \frac{\Brownian (\mathcal{R}(j)) - \Brownian (\mathcal{A}(j)) }{\sqrt {d_{j}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
-\end{equation}
-
-\subsection{Sampling the covariance matrix}\label{subsec:sampling-the-covariance-matrix}
-From the independent contrast at each branch of the tree ($\contrast_{j}$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$, as:
-\begin{equation}
-    \Scattermatrix = \sum\limits_{j=1}^{\Nbranch} \contrast_{j} \MultiplyMatrix \left[\contrast_{j}\right]\tr\label{eq:bayes-scatter},
-\end{equation}
-where $\Nbranch$ is the number of branches in the tree and $\NbrTaxa$ the number of taxa.
-
-The {prior} on the covariance matrix is an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
-\begin{equation}
-    \label{eq:Distribcovariance}
-    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
-\end{equation}
-
-By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\Brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
-\begin{equation}
-    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)\label{eq:bayes-posterior}
-\end{equation}
-This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate_{k}$ defined by
-\begin{equation}
-    \Multivariate_{k} \sim \mathcal{N} \left( \VecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
-\end{equation}
-And from these multivariate samples, $\CovarianceMatrix$ is Gibbs sampled as:
-\begin{equation}
-    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate_{k} \MultiplyMatrix  \left[\Multivariate_{k} \right] \tr \right)\inv \label{eq:bayes-gibbs}
-\end{equation}
-
-\newpage
-\section{Bayesian and Maximum-likelihood implementation}\label{sec:implementation}
-
-Implementation is included within the \textit{BayesCode} software, available at \censor{\url{https://github.com/ThibaultLatrille/bayescode}}.
-
-\subsection{Data formatting}\label{subsec:data-formatting}
-
-Running the analysis on your dataset and compute posterior probabilities requires three files:
-\begin{enumerate}
-    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers)
-    \item A file containing the mean trait values for each species.
-    \item A file containing the variation within-species for each trait and the genetic variation within-species (neutral markers).
-\end{enumerate}
-
-\subsubsection{Phylogenetic tree}\label{subsubsec:phylogenetic-tree}
-
-The phylogenetic tree must be in newick format, with branch lengths in number of substitutions per site (neutral markers).
-
-\subsubsection{Mean trait for each species}\label{subsubsec:mean-trait-for-each-species}
-
-The file containing mean trait values for each species must be in a tab-delimited file with the following format:
-\begin{center}
-    \begin{adjustbox}{width = 0.35\textwidth}
-        \begin{tabular}{|l|c|c|}
-            \hline
-            TaxonName            & Body\_mass & Brain\_mass \\
-            \hline
-            Panthera\_tigris     & 12.26      & 5.676       \\
-            Pithecia\_pithecia   & 7.256      & 3.436       \\
-            Colobus\_angolensis  & 9.176      & 4.284       \\
-            Saimiri\_boliviensis & 6.845      & 3.279       \\
-            $\vdots$             & $\vdots$   & $\vdots$    \\
-            \hline
-        \end{tabular}\label{tab:trait-mean}
-    \end{adjustbox}
-\end{center}
-
-The columns are:
-\begin{itemize}
-    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
-    \item As many columns as traits, without spaces or special characters in the trait.
-    \item The values can be \texttt{NaN} to indicate that the trait is not available for that taxon.
-\end{itemize}
-
-\newpage
-\subsubsection{Trait variation for each species}\label{subsubsec:trait-variation-for-each-species}
-
-The file containing trait variation for each species must be in a tab-delimited file with the following format:
-\begin{center}
-    \begin{adjustbox}{width = 1.0\textwidth}
-        \begin{tabular}{|l|c|c|c|c|c|c|}
-            \hline
-            TaxonName            & Nucleotide\_diversity & Body\_mass\_variance & Body\_mass\_heritability & Brain\_mass\_variance & Brain\_mass\_heritability \\
-            \hline
-            Pithecia\_pithecia   & 0.0016                & 0.22871              & 0.2                      & 0.00737               & 0.2                       \\
-            Colobus\_angolensis  & 0.0017                & 0.00393              & 0.2                      & 0.00416               & 0.2                       \\
-            Saimiri\_boliviensis & 0.0013                & 0.00022              & 0.2                      & 0.00045               & 0.2                       \\
-            Pygathrix\_nemaeus   & 0.0016                & 0.00347              & 0.2                      & 0.00097               & 0.2                       \\
-            $\vdots$             & $\vdots$              & $\vdots$             & $\vdots$                 & $\vdots$              & $\vdots$                  \\
-            \hline
-        \end{tabular}
-        \label{tab:trait-variance}
-    \end{adjustbox}
-\end{center}
-
-\begin{itemize}
-    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
-    \item \emph{Nucleotide\_diversity}: the nucleotide diversity within-species (neutral markers), cannot be \texttt{NaN}.
-    \item As many columns as traits, without spaces or special characters in the trait.
-    \item \emph{TraitName\_variance}: the phenotypic variance of the trait within-species, can be \texttt{NaN} to indicate that the trait variance is not available for that taxon.
-    \item \emph{TraitName\_heritability} (optional): the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
-    \item The columns with the suffix \texttt{\_variance} and \texttt{\_heritability} are repeated for each trait.
-    \item \emph{TraitName\_heritability\_lower} (optional): the lower bound of the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
-    \item \emph{TraitName\_heritability\_upper} (optional): the upper bound of the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
-    \item If the columns with the suffix \texttt{\_heritability\_lower} and \texttt{\_heritability\_upper} are present, the heritability is randomly drawn from a uniform distribution between the lower and upper bounds.
-    \item If the columns with the suffix \texttt{\_heritability} is present, it is taken as is.
-    \item If the additive genetic variance (instead of phenotypic variance) is available for a trait, the heritability can be omitted and will automatically be set to 1.0.
-\end{itemize}
-
-\newpage
-\subsection{Bayesian estimation}\label{subsec:running-nodetraitsand-readnodetraits}
-
-The executable \texttt{nodetraits} from \textit{BayesCode} is used to run the Bayesian estimation of the model, and the executable \texttt{readnodetraits} is used to read the results.
-
-Assuming that the file \texttt{data/body\_size/mammals.male.tsv} contains the mean trait values for each species, the file \texttt{data/body\_size/mammals.male.var\_trait.tsv} contains the variation within-species for each trait and the genetic variation within-species (neutral markers), and the file \texttt{data/body\_size/mammals.male.tree} contains the phylogenetic tree, the following commands are used to run the model and read the results.
-
-\subsubsection{Running the model}\label{subsubsec:running-nodetraits}
-\texttt{nodetraits} is run with the following command:
-\begin{lstlisting}[language = sh,label={lst:nodetraits-run}]
-nodetraits  --until 2000
-            --tree data/body_size/mammals.male.tree
-            --traitsfile data/body_size/mammals.male.tsv
-            run_mammals_male
-\end{lstlisting}
-
-\subsubsection{Reading the results}\label{subsubsec:reading-the-results}
-Once the model has ran, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between-species variation over within-species variation with \texttt{readnodetraits}:
-\begin{lstlisting}[language = sh,label={lst:readnodetraits-rho}]
-readnodetraits --burnin 1000
-               --var_within data/body_size/mammals.male.var_trait.tsv
-               --output results_mammals_male.tsv
-               run_mammals_male
-\end{lstlisting}
-The file \texttt{data\_empirical/chain\_name.ratio.tsv} then contains the posterior mean of the ratio of between-species variation over within-species variation, the 95\% and 99\% credible interval, and the posterior probability that the ratio is greater than 1.
-
-\subsection{Maximum likelihood estimation}\label{subsec:maximum-likelihood-estimation}
-
-To obtain the ratio (without the posterior credible interval and probability) using maximum likelihood computation, the following python script can be used:
-\begin{lstlisting}[language = sh, label={lst:neutrality_index}]
-python3 utils/neutrality_index.py --tree data/body_size/mammals.male.tree
-                                  --traitsfile data/body_size/mammals.male.tsv
-                                  --var_within data/body_size/mammals.male.var_trait.tsv
-                                  --output results_ML_mammals_male.tsv
-\end{lstlisting}
+\bibliographystyle{natbib}
+\begin{thebibliography}{}
+
+    \bibitem[Adams and Collyer(2018)Adams and Collyer]{adams_multivariate_2018}
+    Adams, D.~C. and Collyer, M.~L. 2018.
+    \newblock Multivariate {{Phylogenetic Comparative Methods}}: {{Evaluations}},
+        {{Comparisons}}, and {{Recommendations}}.
+    \newblock {\em Systematic Biology\/}, {67}(1): 14--31.
+
+    \bibitem[Anisimova and Kosiol(2009)Anisimova and
+    Kosiol]{anisimova_investigating_2009}
+    Anisimova, M. and Kosiol, C. 2009.
+    \newblock Investigating protein-coding sequence evolution with probabilistic
+    codon substitution models.
+    \newblock {\em Molecular Biology and Evolution\/}, {26}(2): 255--271.
+
+    \bibitem[Arnold {\em et~al.}(2008)Arnold, B{\"u}rger, Hohenlohe, Ajie, and
+    Jones]{arnold_understanding_2008}
+    Arnold, S.~J., B{\"u}rger, R., Hohenlohe, P.~A., Ajie, B.~C., and Jones, A.~G.
+    2008.
+    \newblock Understanding the evolution and stability of the {{G-Matrix}}.
+    \newblock {\em Evolution\/}, {62}(10): 2451--2461.
+
+    \bibitem[Barton {\em et~al.}(2017)Barton, Etheridge, and
+    V{\'e}ber]{barton_infinitesimal_2017}
+    Barton, N.~H., Etheridge, A.~M., and V{\'e}ber, A. 2017.
+    \newblock The infinitesimal model: {{Definition}}, derivation, and
+    implications.
+    \newblock {\em Theoretical Population Biology\/}, {118}: 50--73.
+
+    \bibitem[Bergeron {\em et~al.}(2023)Bergeron, Besenbacher, Zheng, Li,
+        Bertelsen, Quintard, Hoffman, Li, St.~Leger, Shao, Stiller, Gilbert,
+        Schierup, and Zhang]{bergeron_evolution_2023}
+    Bergeron, L.~A., Besenbacher, S., Zheng, J., Li, P., Bertelsen, M.~F.,
+    Quintard, B., Hoffman, J.~I., Li, Z., St.~Leger, J., Shao, C., Stiller, J.,
+    Gilbert, M. T.~P., Schierup, M.~H., and Zhang, G. 2023.
+    \newblock Evolution of the germline mutation rate across vertebrates.
+    \newblock {\em Nature\/}, pages 1--7.
+
+    \bibitem[Blanquart and Bataillon(2016)Blanquart and
+    Bataillon]{blanquart_epistasis_2016}
+    Blanquart, F. and Bataillon, T. 2016.
+    \newblock Epistasis and the structure of fitness landscapes: {{Are}}
+    experimental fitness landscapes compatible with fisher's geometric model?
+    \newblock {\em Genetics\/}, {203}(2): 847--862.
+
+    \bibitem[Catal{\'a}n {\em et~al.}(2019)Catal{\'a}n, Briscoe, and
+    H{\"o}hna]{catalan_drift_2019}
+    Catal{\'a}n, A., Briscoe, A.~D., and H{\"o}hna, S. 2019.
+    \newblock Drift and directional selection are the evolutionary forces driving
+    gene expression divergence in eye and brain tissue of {{Heliconius}}
+    butterflies.
+    \newblock {\em Genetics\/}, {213}(2): 581--594.
+
+    \bibitem[Cooper {\em et~al.}(2016)Cooper, Thomas, Venditti, Meade, and
+    Freckleton]{cooper_cautionary_2016}
+    Cooper, N., Thomas, G.~H., Venditti, C., Meade, A., and Freckleton, R.~P. 2016.
+    \newblock A cautionary note on the use of {{Ornstein Uhlenbeck}} models in
+    macroevolutionary studies.
+    \newblock {\em Biological Journal of the Linnean Society\/}, {118}(1): 64--77.
+
+    \bibitem[Crow(2010)Crow]{crow_epistasis_2010}
+    Crow, J.~F. 2010.
+    \newblock On epistasis: Why it is unimportant in polygenic directional
+    selection.
+    \newblock {\em Philosophical Transactions of the Royal Society B: Biological
+    Sciences\/}, {365}(1544): 1241--1244.
+
+    \bibitem[Edelaar {\em et~al.}(2011)Edelaar, Burraco, and
+        {Gomez-Mestre}]{edelaar_comparisons_2011}
+    Edelaar, P., Burraco, P., and {Gomez-Mestre}, I. 2011.
+    \newblock Comparisons between {{QST}} and {{FST}}---how wrong have we been?
+    \newblock {\em Molecular Ecology\/}, {20}(23): 4830--4839.
+
+    \bibitem[Felsenstein(1985)Felsenstein]{felsenstein_phylogenies_1985}
+    Felsenstein, J. 1985.
+    \newblock Phylogenies and the {{Comparative Method}}.
+    \newblock {\em The American Naturalist\/}, {125}(1): 1--15.
+
+    \bibitem[Felsenstein(1988)Felsenstein]{felsenstein_phylogenies_1988a}
+    Felsenstein, J. 1988.
+    \newblock Phylogenies and quantitative characters.
+    \newblock {\em Annual Review of Ecology and Systematics\/}, {19}(1): 445--471.
+
+    \bibitem[Felsenstein(2008)Felsenstein]{felsenstein_comparative_2008}
+    Felsenstein, J. 2008.
+    \newblock Comparative methods with sampling error and within-species variation:
+    Contrasts revisited and revised.
+    \newblock {\em The American Naturalist\/}, {171}(6): 713--725.
+
+    \bibitem[Foley {\em et~al.}(2023)Foley, Mason, Harris, Bredemeyer, Damas,
+        Lewin, Eizirik, Gatesy, Karlsson, {Lindblad-Toh}, {Zoonomia Consortium},
+        Springer, and Murphy]{foley_genomic_2023}
+    Foley, N.~M., Mason, V.~C., Harris, A.~J., Bredemeyer, K.~R., Damas, J., Lewin,
+    H.~A., Eizirik, E., Gatesy, J., Karlsson, E.~K., {Lindblad-Toh}, K.,
+        {Zoonomia Consortium}, Springer, M.~S., and Murphy, W.~J. 2023.
+    \newblock A genomic timescale for placental mammal evolution.
+    \newblock {\em Science\/}, {380}(6643): eabl8189.
+
+    \bibitem[Fraser(2020)Fraser]{fraser_detecting_2020}
+    Fraser, H.~B. 2020.
+    \newblock Detecting selection with a genetic cross.
+    \newblock {\em Proceedings of the National Academy of Sciences\/}, {117}(36):
+    22323--22330.
+
+    \bibitem[Gaboriau {\em et~al.}(2020)Gaboriau, Mendes, Joly, Silvestro, and
+    Salamin]{gaboriau_multiplatform_2020}
+    Gaboriau, T., Mendes, F.~K., Joly, S., Silvestro, D., and Salamin, N. 2020.
+    \newblock A multi-platform package for the analysis of intra- and interspecific
+    trait evolution.
+    \newblock {\em Methods in Ecology and Evolution\/}, {11}(11): 1439--1447.
+
+    \bibitem[Gaboriau {\em et~al.}(2023)Gaboriau, Tobias, Silvestro, and
+    Salamin]{gaboriau_exploring_2023}
+    Gaboriau, T., Tobias, J.~A., Silvestro, D., and Salamin, N. 2023.
+    \newblock Exploring the {{Macroevolutionary Signature}} of {{Asymmetric
+    Inheritance}} at {{Speciation}}.
+
+    \bibitem[Genereux {\em et~al.}(2020)Genereux, Serres, Armstrong, Johnson,
+        Marinescu, Mur{\'e}n, Juan, Bejerano, Casewell, Chemnick, Damas, Di~Palma,
+        Diekhans, Fiddes, Garber, Gladyshev, Goodman, Haerty, Houck, Hubley, Kivioja,
+        Koepfli, Kuderna, Lander, Meadows, Murphy, Nash, Noh, Nweeia, Pfenning,
+        Pollard, Ray, Shapiro, Smit, Springer, Steiner, Swofford, Taipale, Teeling,
+            {Turner-Maier}, Alfoldi, Birren, Ryder, Lewin, Paten, {Marques-Bonet},
+            {Lindblad-Toh}, Karlsson, and {Zoonomia
+        Consortium}]{genereux_comparative_2020}
+    Genereux, D.~P., Serres, A., Armstrong, J., Johnson, J., Marinescu, V.~D.,
+    Mur{\'e}n, E., Juan, D., Bejerano, G., Casewell, N.~R., Chemnick, L.~G.,
+    Damas, J., Di~Palma, F., Diekhans, M., Fiddes, I.~T., Garber, M., Gladyshev,
+    V.~N., Goodman, L., Haerty, W., Houck, M.~L., Hubley, R., Kivioja, T.,
+    Koepfli, K.-P., Kuderna, L. F.~K., Lander, E.~S., Meadows, J. R.~S., Murphy,
+    W.~J., Nash, W., Noh, H.~J., Nweeia, M., Pfenning, A.~R., Pollard, K.~S.,
+    Ray, D.~A., Shapiro, B., Smit, A. F.~A., Springer, M.~S., Steiner, C.~C.,
+    Swofford, R., Taipale, J., Teeling, E.~C., {Turner-Maier}, J., Alfoldi, J.,
+    Birren, B., Ryder, O.~A., Lewin, H.~A., Paten, B., {Marques-Bonet}, T.,
+        {Lindblad-Toh}, K., Karlsson, E.~K., and {Zoonomia Consortium} 2020.
+    \newblock A comparative genomics multitool for scientific discovery and
+    conservation.
+    \newblock {\em Nature\/}, {587}(7833): 240--245.
+
+    \bibitem[Gillard {\em et~al.}(2021)Gillard, Gr{\o}nvold, R{\o}s{\ae}g, Holen,
+        Monsen, Koop, Rondeau, Gundappa, Mendoza, Macqueen, Rohlfs, Sandve, and
+        Hvidsten]{gillard_comparative_2021}
+    Gillard, G.~B., Gr{\o}nvold, L., R{\o}s{\ae}g, L.~L., Holen, M.~M., Monsen,
+        {\O}., Koop, B.~F., Rondeau, E.~B., Gundappa, M.~K., Mendoza, J., Macqueen,
+    D.~J., Rohlfs, R.~V., Sandve, S.~R., and Hvidsten, T.~R. 2021.
+    \newblock Comparative regulomics supports pervasive selection on gene dosage
+    following whole genome duplication.
+    \newblock {\em Genome Biology\/}, {22}(1): 103.
+
+    \bibitem[Goldman and Yang(1994)Goldman and Yang]{goldman_codonbased_1994}
+    Goldman, N. and Yang, Z. 1994.
+    \newblock A codon-based model of nucleotide substitution for protein-coding
+        {{DNA}} sequences.
+    \newblock {\em Molecular Biology and Evolution\/}, {11}(5): 725--736.
+
+    \bibitem[Grabowski {\em et~al.}(2023)Grabowski, Pienaar, Voje, Andersson,
+            {Fuentes-Gonz{\'a}lez}, Kopperud, Moen, Tsuboi, Uyeda, and
+        Hansen]{grabowski_cautionary_2023}
+    Grabowski, M., Pienaar, J., Voje, K.~L., Andersson, S., {Fuentes-Gonz{\'a}lez},
+    J., Kopperud, B.~T., Moen, D.~S., Tsuboi, M., Uyeda, J., and Hansen, T.~F.
+    2023.
+    \newblock A {{Cautionary Note}} on ``{{A Cautionary Note}} on the {{Use}} of
+        {{Ornstein Uhlenbeck Models}} in {{Macroevolutionary Studies}}''.
+    \newblock {\em Systematic Biology\/}, {72}(4): 955--963.
+
+    \bibitem[Hansen(1997)Hansen]{hansen_stabilizing_1997}
+    Hansen, T.~F. 1997.
+    \newblock Stabilizing selection and the comparative analysis of adaptation.
+    \newblock {\em Evolution\/}, {51}(5): 1341--1351.
+
+    \bibitem[Hansen and Bartoszek(2012)Hansen and
+    Bartoszek]{hansen_interpreting_2012}
+    Hansen, T.~F. and Bartoszek, K. 2012.
+    \newblock Interpreting the evolutionary regression: The interplay between
+    observational and biological errors in phylogenetic comparative studies.
+    \newblock {\em Systematic Biology\/}, {61}(3): 413--425.
+
+    \bibitem[Hansen and Martins(1996)Hansen and Martins]{hansen_translating_1996}
+    Hansen, T.~F. and Martins, E.~P. 1996.
+    \newblock Translating between microevolutionary process and macroevolutionary
+    patterns: The correlation structure of interspecific data.
+    \newblock {\em Evolution\/}, {50}(4): 1404--1417.
+
+    \bibitem[Hansen and P{\'e}labon(2021)Hansen and
+    P{\'e}labon]{hansen_evolvability_2021}
+    Hansen, T.~F. and P{\'e}labon, C. 2021.
+    \newblock Evolvability: {{A Quantitative-Genetics Perspective}}.
+    \newblock {\em Annual Review of Ecology, Evolution, and Systematics\/},
+        {52}(1): 153--175.
+
+    \bibitem[Hansen {\em et~al.}(2011)Hansen, P{\'e}labon, and
+    Houle]{hansen_heritability_2011}
+    Hansen, T.~F., P{\'e}labon, C., and Houle, D. 2011.
+    \newblock Heritability is not {{Evolvability}}.
+    \newblock {\em Evolutionary Biology\/}, {38}(3): 258--277.
+
+    \bibitem[Harmon(2018)Harmon]{harmon_phylogenetic_2018}
+    Harmon, L. 2018.
+    \newblock Phylogenetic comparative methods: Learning from trees.
+
+    \bibitem[Hill {\em et~al.}(2008)Hill, Goddard, and Visscher]{hill_data_2008}
+    Hill, W.~G., Goddard, M.~E., and Visscher, P.~M. 2008.
+    \newblock Data and {{Theory Point}} to {{Mainly Additive Genetic Variance}} for
+        {{Complex Traits}}.
+    \newblock {\em PLOS Genetics\/}, {4}(2): e1000008.
+
+    \bibitem[Hohenlohe and Arnold(2008)Hohenlohe and Arnold]{hohenlohe_mipod_2008}
+    Hohenlohe, P.~A. and Arnold, S.~J. 2008.
+    \newblock {{MIPoD}}: {{A}} hypothesis-testing framework for microevolutionary
+    inference from patterns of divergence.
+    \newblock {\em The American Naturalist\/}, {171}(3): 366--385.
+
+    \bibitem[Hu {\em et~al.}(2022)Hu, Park, and Reecy]{hu_bringing_2022}
+    Hu, Z.-L., Park, C.~A., and Reecy, J.~M. 2022.
+    \newblock Bringing the {{Animal QTLdb}} and {{CorrDB}} into the future: Meeting
+    new challenges and providing updated services.
+    \newblock {\em Nucleic Acids Research\/}, {50}(D1): D956--D961.
+
+    \bibitem[Huber {\em et~al.}(2015)Huber, Whibley, Poul, Navarro, Martin, Baxter,
+        Shah, Gilles, Wirth, McMillan, and Joron]{huber_conservatism_2015}
+    Huber, B., Whibley, A., Poul, Y.~L., Navarro, N., Martin, A., Baxter, S., Shah,
+    A., Gilles, B., Wirth, T., McMillan, W.~O., and Joron, M. 2015.
+    \newblock Conservatism and novelty in the genetic architecture of adaptation in
+        {{Heliconius}} butterflies.
+    \newblock {\em Heredity\/}, {114}(5): 515--524.
+
+    \bibitem[Huelsenbeck and Rannala(2003)Huelsenbeck and
+    Rannala]{huelsenbeck_detecting_2003}
+    Huelsenbeck, J.~P. and Rannala, B. 2003.
+    \newblock Detecting correlation between characters in a comparative analysis
+    with uncertain phylogeny.
+    \newblock {\em Evolution\/}, {57}(6): 1237--1247.
+
+    \bibitem[Ingram and Mahler(2013)Ingram and Mahler]{ingram_surface_2013}
+    Ingram, T. and Mahler, D. 2013.
+    \newblock {{SURFACE}}: Detecting convergent evolution from comparative data by
+    fitting {{Ornstein-Uhlenbeck}} models with stepwise {{Akaike Information
+    Criterion}}.
+    \newblock {\em Methods in Ecology and Evolution\/}, {4}(5): 416--425.
+
+    \bibitem[Jensen {\em et~al.}(2019)Jensen, Payseur, Stephan, Aquadro, Lynch,
+        Charlesworth, and Charlesworth]{jensen_importance_2019}
+    Jensen, J.~D., Payseur, B.~A., Stephan, W., Aquadro, C.~F., Lynch, M.,
+    Charlesworth, D., and Charlesworth, B. 2019.
+    \newblock The importance of the {{Neutral Theory}} in 1968 and 50 years on:
+        {{A}} response to {{Kern}} and {{Hahn}} 2018.
+    \newblock {\em Evolution\/}, {73}(1): 111--114.
+
+    \bibitem[Khabbazian {\em et~al.}(2016)Khabbazian, Kriebel, Rohe, and
+    An{\'e}]{khabbazian_fast_2016}
+    Khabbazian, M., Kriebel, R., Rohe, K., and An{\'e}, C. 2016.
+    \newblock Fast and accurate detection of evolutionary shifts in
+        {{Ornstein}}--{{Uhlenbeck}} models.
+    \newblock {\em Methods in Ecology and Evolution\/}, {7}(7): 811--824.
+
+    \bibitem[Khaitovich {\em et~al.}(2004)Khaitovich, Weiss, Lachmann, Hellmann,
+        Enard, Muetzel, Wirkner, Ansorge, and P{\"a}{\"a}bo]{khaitovich_neutral_2004}
+    Khaitovich, P., Weiss, G., Lachmann, M., Hellmann, I., Enard, W., Muetzel, B.,
+    Wirkner, U., Ansorge, W., and P{\"a}{\"a}bo, S. 2004.
+    \newblock A neutral model of transcriptome evolution.
+    \newblock {\em PLOS Biology\/}, {2}(5): e132.
+
+    \bibitem[Kimura(1962)Kimura]{kimura_probability_1962}
+    Kimura, M. 1962.
+    \newblock On the probability of fixation of mutant genes in a population.
+    \newblock {\em Genetics\/}, {47}(6): 713--719.
+
+    \bibitem[Kimura(1968)Kimura]{kimura_evolutionary_1968}
+    Kimura, M. 1968.
+    \newblock Evolutionary rate at the molecular level.
+    \newblock {\em Nature\/}, {217}(5129): 624--626.
+
+    \bibitem[Kostikova {\em et~al.}(2016)Kostikova, Silvestro, Pearman, and
+    Salamin]{kostikova_bridging_2016}
+    Kostikova, A., Silvestro, D., Pearman, P.~B., and Salamin, N. 2016.
+    \newblock Bridging inter- and intraspecific trait evolution with a hierarchical
+    bayesian approach.
+    \newblock {\em Systematic Biology\/}, {65}(3): 417--431.
+
+    \bibitem[Kuderna {\em et~al.}(2023)Kuderna, Gao, Janiak, Kuhlwilm, Orkin,
+        Bataillon, Manu, Valenzuela, Bergman, Rousselle, Silva, Agueda, Blanc, Gut,
+            {de Vries}, Goodhead, Harris, Raveendran, Jensen, Chuma, Horvath, Hvilsom,
+        Juan, Frandsen, Schraiber, {de Melo}, Bertuol, Byrne, Sampaio, Farias,
+        Valsecchi, Messias, {da Silva}, Trivedi, Rossi, Hrbek, Andriaholinirina,
+        Rabarivola, Zaramody, Jolly, {Phillips-Conroy}, Wilkerson, Abee, Simmons,
+            {Fernandez-Duque}, Kanthaswamy, Shiferaw, Wu, Zhou, Shao, Zhang, Keyyu,
+        Knauf, Le, Lizano, Merker, Navarro, Nadler, Khor, Lee, Tan, Lim, Kitchener,
+        Zinner, Gut, Melin, Guschanski, Schierup, Beck, Umapathy, Roos, Boubli,
+        Rogers, Farh, and Marques~Bonet]{kuderna_global_2023}
+    Kuderna, L. F.~K., Gao, H., Janiak, M.~C., Kuhlwilm, M., Orkin, J.~D.,
+    Bataillon, T., Manu, S., Valenzuela, A., Bergman, J., Rousselle, M., Silva,
+    F.~E., Agueda, L., Blanc, J., Gut, M., {de Vries}, D., Goodhead, I., Harris,
+    R.~A., Raveendran, M., Jensen, A., Chuma, I.~S., Horvath, J.~E., Hvilsom, C.,
+    Juan, D., Frandsen, P., Schraiber, J.~G., {de Melo}, F.~R., Bertuol, F.,
+    Byrne, H., Sampaio, I., Farias, I., Valsecchi, J., Messias, M., {da Silva},
+    M. N.~F., Trivedi, M., Rossi, R., Hrbek, T., Andriaholinirina, N.,
+    Rabarivola, C.~J., Zaramody, A., Jolly, C.~J., {Phillips-Conroy}, J.,
+    Wilkerson, G., Abee, C., Simmons, J.~H., {Fernandez-Duque}, E., Kanthaswamy,
+    S., Shiferaw, F., Wu, D., Zhou, L., Shao, Y., Zhang, G., Keyyu, J.~D., Knauf,
+    S., Le, M.~D., Lizano, E., Merker, S., Navarro, A., Nadler, T., Khor, C.~C.,
+    Lee, J., Tan, P., Lim, W.~K., Kitchener, A.~C., Zinner, D., Gut, I., Melin,
+    A.~D., Guschanski, K., Schierup, M.~H., Beck, R. M.~D., Umapathy, G., Roos,
+    C., Boubli, J.~P., Rogers, J., Farh, K. K.-H., and Marques~Bonet, T. 2023.
+    \newblock A global catalog of whole-genome diversity from 233 primate species.
+    \newblock {\em Science\/}, {380}(6648): 906--913.
+
+    \bibitem[Lamy {\em et~al.}(2012)Lamy, Plomion, Kremer, and
+    Delzon]{lamy_qst_2012}
+    Lamy, J.-B., Plomion, C., Kremer, A., and Delzon, S. 2012.
+    \newblock {{QST}} {$<$} {{FST As}} a signature of canalization.
+    \newblock {\em Molecular Ecology\/}, {21}(23): 5646--5655.
+
+    \bibitem[Lande(1979)Lande]{lande_quantitative_1979}
+    Lande, R. 1979.
+    \newblock Quantitative genetic analysis of multivariate evolution, applied to
+    brain: Body size allometry.
+    \newblock {\em Evolution\/}, {33}(1): 402--416.
+
+    \bibitem[Lande(1980a)Lande]{lande_genetic_1980}
+    Lande, R. 1980a.
+    \newblock Genetic variation and phenotypic evolution during allopatric
+    speciation.
+    \newblock {\em The American Naturalist\/}, {116}(4): 463--479.
+
+    \bibitem[Lande(1980b)Lande]{lande_sexual_1980}
+    Lande, R. 1980b.
+    \newblock Sexual dimorphism, sexual selection, and adaptation in polygenic
+    characters.
+    \newblock {\em Evolution\/}, {34}(2): 292--305.
+
+    \bibitem[Lartillot and Delsuc(2012)Lartillot and Delsuc]{lartillot_joint_2012}
+    Lartillot, N. and Delsuc, F. 2012.
+    \newblock Joint reconstruction of divergence times and life-history evolution
+    in placental mammals using a phylogenetic covariance model.
+    \newblock {\em Evolution\/}, {66}(6): 1773--1787.
+
+    \bibitem[Lartillot and Poujol(2011)Lartillot and
+    Poujol]{lartillot_phylogenetic_2011}
+    Lartillot, N. and Poujol, R. 2011.
+    \newblock A phylogenetic model for investigating correlated evolution of
+    substitution rates and continuous phenotypic characters.
+    \newblock {\em Molecular Biology and Evolution\/}, {28}(1): 729--744.
+
+    \bibitem[Latrille {\em et~al.}(2021)Latrille, Lanore, and
+    Lartillot]{latrille_inferring_2021}
+    Latrille, T., Lanore, V., and Lartillot, N. 2021.
+    \newblock Inferring long-term effective population size with
+    mutation--selection models.
+    \newblock {\em Molecular Biology and Evolution\/}, {38}(10): 4573--4587.
+
+    \bibitem[Latrille {\em et~al.}(2023)Latrille, Rodrigue, and
+    Lartillot]{latrille_genes_2023}
+    Latrille, T., Rodrigue, N., and Lartillot, N. 2023.
+    \newblock Genes and sites under adaptation at the phylogenetic scale also
+    exhibit adaptation at the population-genetic scale.
+    \newblock {\em Proceedings of the National Academy of Sciences of the United
+    States of America\/}, {120}(11): e2214977120.
+
+    \bibitem[Leinonen {\em et~al.}(2008)Leinonen, O'hara, Cano, and
+    Meril{\"a}]{leinonen_comparative_2008}
+    Leinonen, T., O'hara, R.~B., Cano, J.~M., and Meril{\"a}, J. 2008.
+    \newblock Comparative studies of quantitative trait and neutral marker
+    divergence: A meta-analysis.
+    \newblock {\em Journal of Evolutionary Biology\/}, {21}(1): 1--17.
+
+    \bibitem[Leinonen {\em et~al.}(2013)Leinonen, McCairns, O'Hara, and
+    Meril{\"a}]{leinonen_qst_2013}
+    Leinonen, T., McCairns, R. J.~S., O'Hara, R.~B., and Meril{\"a}, J. 2013.
+    \newblock {{QST}}--{{FST}} comparisons: Evolutionary and ecological insights
+    from genomic heterogeneity.
+    \newblock {\em Nature Reviews Genetics\/}, {14}(3): 179--190.
+
+    \bibitem[Lemos {\em et~al.}(2001)Lemos, Marroig, and
+    Cerqueira]{lemos_evolutionary_2001}
+    Lemos, B., Marroig, G., and Cerqueira, R. 2001.
+    \newblock Evolutionary rates and stabilizing selection in large-bodied opossum
+    skulls ({{Didelphimorphia}}: {{Didelphidae}}).
+    \newblock {\em Journal of Zoology\/}, {255}(2): 181--189.
+
+    \bibitem[Lemos {\em et~al.}(2005)Lemos, Meiklejohn, C{\'a}ceres, and
+    Hartl]{lemos_rates_2005}
+    Lemos, B., Meiklejohn, C.~D., C{\'a}ceres, M., and Hartl, D.~L. 2005.
+    \newblock Rates of {{Divergence}} in {{Gene Expression Profiles}} of
+        {{Primates}}, {{Mice}}, and {{Flies}}: {{Stabilizing Selection}} and
+        {{Variability Among Functional Categories}}.
+    \newblock {\em Evolution\/}, {59}(1): 126--137.
+
+    \bibitem[Litsios and Salamin(2012)Litsios and Salamin]{litsios_effects_2012}
+    Litsios, G. and Salamin, N. 2012.
+    \newblock Effects of {{Phylogenetic Signal}} on {{Ancestral State
+    Reconstruction}}.
+    \newblock {\em Systematic Biology\/}, {61}(3): 533--538.
+
+    \bibitem[Louca and Pennell(2020)Louca and Pennell]{louca_extant_2020}
+    Louca, S. and Pennell, M.~W. 2020.
+    \newblock Extant timetrees are consistent with a myriad of diversification
+    histories.
+    \newblock {\em Nature\/}, {580}(7804): 502--505.
+
+    \bibitem[Lynch(1990)Lynch]{lynch_rate_1990}
+    Lynch, M. 1990.
+    \newblock The {{Rate}} of {{Morphological Evolution}} in {{Mammals}} from the
+        {{Standpoint}} of the {{Neutral Expectation}}.
+    \newblock {\em The American Naturalist\/}, {136}(6): 727--741.
+
+    \bibitem[Lynch(1991)Lynch]{lynch_methods_1991}
+    Lynch, M. 1991.
+    \newblock Methods for the analysis of comparative data in evolutionary biology.
+    \newblock {\em Evolution\/}, {45}(5): 1065--1080.
+
+    \bibitem[Lynch and Crease(1990)Lynch and Crease]{lynch_analysis_1990}
+    Lynch, M. and Crease, T.~J. 1990.
+    \newblock The analysis of population survey data on {{DNA}} sequence variation.
+    \newblock {\em Molecular Biology and Evolution\/}, {7}(4): 377--394.
+
+    \bibitem[Lynch and Walsh(1998)Lynch and Walsh]{lynch_genetics_1998}
+    Lynch, M. and Walsh, B. 1998.
+    \newblock {\em Genetics and Analysis of Quantitative Traits\/}, volume~1.
+    \newblock Sinauer Sunderland, MA.
+
+    \bibitem[Lynch {\em et~al.}(1998)Lynch, Latta, Hicks, and
+    Giorgianni]{lynch_mutation_1998}
+    Lynch, M., Latta, L., Hicks, J., and Giorgianni, M. 1998.
+    \newblock Mutation, selection, and the maintenance of life-history variation in
+    a natural population.
+    \newblock {\em Evolution\/}, {52}(3): 727--733.
+
+    \bibitem[Machado {\em et~al.}(2022)Machado, Marroig, and
+    Hubbe]{machado_preeminent_2022}
+    Machado, F.~A., Marroig, G., and Hubbe, A. 2022.
+    \newblock The pre-eminent role of directional selection in generating extreme
+    morphological change in glyptodonts ({{Cingulata}}; {{Xenarthra}}).
+    \newblock {\em Proceedings of the Royal Society B: Biological Sciences\/},
+        {289}(1967): 20212521.
+
+    \bibitem[Machado {\em et~al.}(2023)Machado, Mongle, Slater, Penna, Wisniewski,
+        Soffin, Dutra, and Uyeda]{machado_using_2023}
+    Machado, F.~A., Mongle, C.~S., Slater, G., Penna, A., Wisniewski, A., Soffin,
+    A., Dutra, V., and Uyeda, J.~C. 2023.
+    \newblock Using developmental rules to align microevolution with
+    macroevolution.
+
+    \bibitem[Martin {\em et~al.}(2008)Martin, Chapuis, and
+    Goudet]{martin_multivariate_2008}
+    Martin, G., Chapuis, E., and Goudet, J. 2008.
+    \newblock Multivariate {{Qst}}--{{Fst Comparisons}}: {{A Neutrality Test}} for
+    the {{Evolution}} of the {{G Matrix}} in {{Structured Populations}}.
+    \newblock {\em Genetics\/}, {180}(4): 2135--2149.
+
+    \bibitem[McCandlish and Stoltzfus(2014)McCandlish and
+    Stoltzfus]{mccandlish_modeling_2014}
+    McCandlish, D.~M. and Stoltzfus, A. 2014.
+    \newblock Modeling evolution using the probability of fixation: {{History}} and
+    implications.
+    \newblock {\em Quarterly Review of Biology\/}, {89}(3): 225--252.
+
+    \bibitem[McDonald and Kreitman(1991)McDonald and
+    Kreitman]{mcdonald_adaptative_1991}
+    McDonald, J.~H. and Kreitman, M. 1991.
+    \newblock Adaptative protein evolution at {{Adh}} locus in {{Drosophila}}.
+    \newblock {\em Nature\/}, {351}(6328): 652--654.
+
+    \bibitem[Meril{\"a} and Crnokrak(2001)Meril{\"a} and
+    Crnokrak]{merila_comparison_2001}
+    Meril{\"a}, J. and Crnokrak, P. 2001.
+    \newblock Comparison of genetic differentiation at marker loci and quantitative
+    traits.
+    \newblock {\em Journal of Evolutionary Biology\/}, {14}(6): 892--903.
+
+    \bibitem[Mitov {\em et~al.}(2019)Mitov, Bartoszek, and
+    Stadler]{mitov_automatic_2019}
+    Mitov, V., Bartoszek, K., and Stadler, T. 2019.
+    \newblock Automatic generation of evolutionary hypotheses using mixed
+        {{Gaussian}} phylogenetic models.
+    \newblock {\em Proceedings of the National Academy of Sciences\/}, {116}(34):
+    16921--16926.
+
+    \bibitem[Mitov {\em et~al.}(2020)Mitov, Bartoszek, Asimomitis, and
+    Stadler]{mitov_fast_2020}
+    Mitov, V., Bartoszek, K., Asimomitis, G., and Stadler, T. 2020.
+    \newblock Fast likelihood calculation for multivariate {{Gaussian}}
+    phylogenetic models with shifts.
+    \newblock {\em Theoretical Population Biology\/}, {131}: 66--78.
+
+    \bibitem[Muse and Gaut(1994)Muse and Gaut]{muse_likelihood_1994}
+    Muse, S.~V. and Gaut, B.~S. 1994.
+    \newblock A likelihood approach for comparing synonymous and nonsynonymous
+    nucleotide substitution rates, with application to the chloroplast genome.
+    \newblock {\em Molecular Biology and Evolution\/}, {1}(5): 715--724.
+
+    \bibitem[Nielsen(2005)Nielsen]{nielsen_molecular_2005}
+    Nielsen, R. 2005.
+    \newblock Molecular signatures of natural selection.
+    \newblock {\em Annual Review of Genetics\/}, {39}(1): 197--218.
+
+    \bibitem[O'Meara {\em et~al.}(2006)O'Meara, An{\'e}, Sanderson, and
+    Wainwright]{omeara_testing_2006}
+    O'Meara, B.~C., An{\'e}, C., Sanderson, M.~J., and Wainwright, P.~C. 2006.
+    \newblock Testing for {{Different Rates}} of {{Continuous Trait Evolution Using
+    Likelihood}}.
+    \newblock {\em Evolution\/}, {60}(5): 922--933.
+
+    \bibitem[Ovaskainen {\em et~al.}(2011)Ovaskainen, Karhunen, Zheng, Arias, and
+    Meril{\"a}]{ovaskainen_new_2011}
+    Ovaskainen, O., Karhunen, M., Zheng, C., Arias, J. M.~C., and Meril{\"a}, J.
+    2011.
+    \newblock A {{New Method}} to {{Uncover Signatures}} of {{Divergent}} and
+        {{Stabilizing Selection}} in {{Quantitative Traits}}.
+    \newblock {\em Genetics\/}, {189}(2): 621--632.
+
+    \bibitem[Pennell {\em et~al.}(2015)Pennell, FitzJohn, Cornwell, and
+    Harmon]{pennell_model_2015}
+    Pennell, M.~W., FitzJohn, R.~G., Cornwell, W.~K., and Harmon, L.~J. 2015.
+    \newblock Model {{Adequacy}} and the {{Macroevolution}} of {{Angiosperm
+    Functional Traits}}.
+    \newblock {\em The American Naturalist\/}, {186}(2): E33--E50.
+
+    \bibitem[Porto {\em et~al.}(2015)Porto, Sebasti{\~a}o, Pavan, VandeBerg,
+        Marroig, and Cheverud]{porto_rate_2015}
+    Porto, A., Sebasti{\~a}o, H., Pavan, S.~E., VandeBerg, J.~L., Marroig, G., and
+    Cheverud, J.~M. 2015.
+    \newblock Rate of evolutionary change in cranial morphology of the marsupial
+    genus {{Monodelphis}} is constrained by the availability of additive genetic
+    variation.
+    \newblock {\em Journal of Evolutionary Biology\/}, {28}(4): 973--985.
+
+    \bibitem[Price {\em et~al.}(2022)Price, Palmer~Droguett, Taylor, Kim, Place,
+        Rogers, Mank, Cooney, and Wright]{price_detecting_2022}
+    Price, P.~D., Palmer~Droguett, D.~H., Taylor, J.~A., Kim, D.~W., Place, E.~S.,
+    Rogers, T.~F., Mank, J.~E., Cooney, C.~R., and Wright, A.~E. 2022.
+    \newblock Detecting signatures of selection on gene expression.
+    \newblock {\em Nature Ecology \& Evolution\/}, pages 1--11.
+
+    \bibitem[Pujol {\em et~al.}(2008)Pujol, Wilson, Ross, and
+    Pannell]{pujol_are_2008}
+    Pujol, B., Wilson, A.~J., Ross, R. I.~C., and Pannell, J.~R. 2008.
+    \newblock Are {{QST}}--{{FST}} comparisons for natural populations meaningful?
+    \newblock {\em Molecular Ecology\/}, {17}(22): 4782--4785.
+
+    \bibitem[Rohlfs and Nielsen(2015)Rohlfs and Nielsen]{rohlfs_phylogenetic_2015}
+    Rohlfs, R.~V. and Nielsen, R. 2015.
+    \newblock Phylogenetic {{ANOVA}}: {{The}} expression variance and evolution
+    model for quantitative trait evolution.
+    \newblock {\em Systematic Biology\/}, {64}(5): 695--708.
+
+    \bibitem[Rohlfs {\em et~al.}(2014)Rohlfs, Harrigan, and
+    Nielsen]{rohlfs_modeling_2014}
+    Rohlfs, R.~V., Harrigan, P., and Nielsen, R. 2014.
+    \newblock Modeling gene expression evolution with an extended
+        {{Ornstein}}--{{Uhlenbeck}} process accounting for within-species variation.
+    \newblock {\em Molecular Biology and Evolution\/}, {31}(1): 201--211.
+
+    \bibitem[Rolland {\em et~al.}(2023)Rolland, {Henao-Diaz}, Doebeli, Germain,
+        Harmon, Knowles, Liow, Mank, Machac, Otto, Pennell, Salamin, Silvestro,
+        Sugawara, Uyeda, Wagner, and Schluter]{rolland_conceptual_2023}
+    Rolland, J., {Henao-Diaz}, L.~F., Doebeli, M., Germain, R., Harmon, L.~J.,
+    Knowles, L.~L., Liow, L.~H., Mank, J.~E., Machac, A., Otto, S.~P., Pennell,
+    M., Salamin, N., Silvestro, D., Sugawara, M., Uyeda, J., Wagner, C.~E., and
+    Schluter, D. 2023.
+    \newblock Conceptual and empirical bridges between micro- and macroevolution.
+    \newblock {\em Nature Ecology \& Evolution\/}, pages 1--13.
+
+    \bibitem[Schraiber and Edge(2023)Schraiber and
+    Edge]{schraiber_heritability_2023}
+    Schraiber, J.~G. and Edge, M.~D. 2023.
+    \newblock Heritability within groups is uninformative about differences among
+    groups: Cases from behavioral, evolutionary, and statistical genetics.
+
+    \bibitem[Schroeder and {von Cramon-Taubadel}(2017)Schroeder and {von
+    Cramon-Taubadel}]{schroeder_evolution_2017}
+    Schroeder, L. and {von Cramon-Taubadel}, N. 2017.
+    \newblock The evolution of hominoid cranial diversity: {{A}} quantitative
+    genetic approach.
+    \newblock {\em Evolution\/}, {71}(11): 2634--2649.
+
+    \bibitem[Sella and Barton(2019)Sella and Barton]{sella_thinking_2019}
+    Sella, G. and Barton, N.~H. 2019.
+    \newblock Thinking about the evolution of complex traits in the era of
+    genome-wide association studies.
+    \newblock {\em Annual Review of Genomics and Human Genetics\/}, {20}(1):
+    461--493.
+
+    \bibitem[Silvestro {\em et~al.}(2015)Silvestro, Kostikova, Litsios, Pearman,
+        and Salamin]{silvestro_measurement_2015}
+    Silvestro, D., Kostikova, A., Litsios, G., Pearman, P.~B., and Salamin, N.
+    2015.
+    \newblock Measurement errors should always be incorporated in phylogenetic
+    comparative analysis.
+    \newblock {\em Methods in Ecology and Evolution\/}, {6}(3): 340--346.
+
+    \bibitem[Silvestro {\em et~al.}(2019)Silvestro, Tejedor, {Serrano-Serrano},
+        Loiseau, Rossier, Rolland, Zizka, H{\"o}hna, Antonelli, and
+        Salamin]{silvestro_early_2019}
+    Silvestro, D., Tejedor, M.~F., {Serrano-Serrano}, M.~L., Loiseau, O., Rossier,
+    V., Rolland, J., Zizka, A., H{\"o}hna, S., Antonelli, A., and Salamin, N.
+    2019.
+    \newblock Early {{Arrival}} and {{Climatically-Linked Geographic Expansion}} of
+        {{New World Monkeys}} from {{Tiny African Ancestors}}.
+    \newblock {\em Systematic Biology\/}, {68}(1): 78--92.
+
+    \bibitem[Simons {\em et~al.}(2018)Simons, Bullaughey, Hudson, and
+    Sella]{simons_population_2018}
+    Simons, Y.~B., Bullaughey, K., Hudson, R.~R., and Sella, G. 2018.
+    \newblock A population genetic interpretation of {{GWAS}} findings for human
+    quantitative traits.
+    \newblock {\em PLOS Biology\/}, {16}(3): e2002985.
+
+    \bibitem[Stamp and Hadfield(2020)Stamp and Hadfield]{stamp_relative_2020}
+    Stamp, M.~A. and Hadfield, J.~D. 2020.
+    \newblock The relative importance of plasticity versus genetic differentiation
+    in explaining between population differences; a meta-analysis.
+    \newblock {\em Ecology Letters\/}, {23}(10): 1432--1441.
+
+    \bibitem[Tajima(1989)Tajima]{tajima_statistical_1989}
+    Tajima, F. 1989.
+    \newblock Statistical method for testing the neutral mutation hypothesis by
+        {{DNA}} polymorphism.
+    \newblock {\em Genetics\/}, {123}(3): 585--595.
+
+    \bibitem[Tenaillon(2014)Tenaillon]{tenaillon_utility_2014}
+    Tenaillon, O. 2014.
+    \newblock The utility of fisher's geometric model in evolutionary genetics.
+    \newblock {\em Annual Review of Ecology, Evolution, and Systematics\/},
+        {45}(1): 179--201.
+
+    \bibitem[Tsuboi {\em et~al.}(2018)Tsuboi, {van der Bijl}, Kopperud,
+        Erritz{\o}e, Voje, Kotrschal, Yopak, Collin, Iwaniuk, and
+        Kolm]{tsuboi_breakdown_2018}
+    Tsuboi, M., {van der Bijl}, W., Kopperud, B.~T., Erritz{\o}e, J., Voje, K.~L.,
+    Kotrschal, A., Yopak, K.~E., Collin, S.~P., Iwaniuk, A.~N., and Kolm, N.
+    2018.
+    \newblock Breakdown of brain--body allometry and the encephalization of birds
+    and mammals.
+    \newblock {\em Nature Ecology \& Evolution\/}, {2}(9): 1492--1500.
+
+    \bibitem[Tung {\em et~al.}(2015)Tung, Zhou, Alberts, Stephens, and
+    Gilad]{tung_genetic_2015}
+    Tung, J., Zhou, X., Alberts, S.~C., Stephens, M., and Gilad, Y. 2015.
+    \newblock The genetic architecture of gene expression levels in wild baboons.
+    \newblock {\em eLife\/}, {4}: e04729.
+
+    \bibitem[Turelli(1984)Turelli]{turelli_heritable_1984}
+    Turelli, M. 1984.
+    \newblock Heritable genetic variation via mutation-selection balance:
+        {{Lerch}}'s zeta meets the abdominal bristle.
+    \newblock {\em Theoretical Population Biology\/}, {25}(2): 138--193.
+
+    \bibitem[Turelli(2017)Turelli]{turelli_commentary_2017}
+    Turelli, M. 2017.
+    \newblock Commentary: {{Fisher}}'s infinitesimal model: {{A}} story for the
+    ages.
+    \newblock {\em Theoretical Population Biology\/}, {118}: 46--49.
+
+    \bibitem[Uyeda and Harmon(2014)Uyeda and Harmon]{uyeda_novel_2014}
+    Uyeda, J.~C. and Harmon, L.~J. 2014.
+    \newblock A {{Novel Bayesian Method}} for {{Inferring}} and {{Interpreting}}
+    the {{Dynamics}} of {{Adaptive Landscapes}} from {{Phylogenetic Comparative
+    Data}}.
+    \newblock {\em Systematic Biology\/}, {63}(6): 902--918.
+
+    \bibitem[Uyeda {\em et~al.}(2011)Uyeda, Hansen, Arnold, and
+    Pienaar]{uyeda_millionyear_2011}
+    Uyeda, J.~C., Hansen, T.~F., Arnold, S.~J., and Pienaar, J. 2011.
+    \newblock The million-year wait for macroevolutionary bursts.
+    \newblock {\em Proceedings of the National Academy of Sciences\/}, {108}(38):
+    15908--15913.
+
+    \bibitem[Walsh and Lynch(2018)Walsh and Lynch]{walsh_evolution_2018}
+    Walsh, B. and Lynch, M. 2018.
+    \newblock {\em Evolution and {{Selection}} of {{Quantitative Traits}}\/}.
+    \newblock Oxford University Press.
+
+    \bibitem[Weaver {\em et~al.}(2007)Weaver, Roseman, and
+    Stringer]{weaver_were_2007}
+    Weaver, T.~D., Roseman, C.~C., and Stringer, C.~B. 2007.
+    \newblock Were neandertal and modern human cranial differences produced by
+    natural selection or genetic drift?
+    \newblock {\em Journal of Human Evolution\/}, {53}(2): 135--145.
+
+    \bibitem[Wilder {\em et~al.}(2023)Wilder, Supple, Subramanian, Mudide,
+        Swofford, {Serres-Armero}, Steiner, Koepfli, Genereux, Karlsson,
+            {Lindblad-Toh}, {Marques-Bonet}, Munoz~Fuentes, Foley, Meyer, {Zoonomia
+        Consortium}, Ryder, and Shapiro]{wilder_contribution_2023}
+    Wilder, A.~P., Supple, M.~A., Subramanian, A., Mudide, A., Swofford, R.,
+        {Serres-Armero}, A., Steiner, C., Koepfli, K.-P., Genereux, D.~P., Karlsson,
+    E.~K., {Lindblad-Toh}, K., {Marques-Bonet}, T., Munoz~Fuentes, V., Foley, K.,
+    Meyer, W.~K., {Zoonomia Consortium}, Ryder, O.~A., and Shapiro, B. 2023.
+    \newblock The contribution of historical processes to contemporary extinction
+    risk in placental mammals.
+    \newblock {\em Science\/}, {380}(6643): eabn5856.
+
+\end{thebibliography}
 \end{document}

### 2024-07-05 manuscript/supp-mat.tex
Updated manuscript
Cleaned unecessary files. Updated README.

@@
+%! BibTeX Compiler = bibtex
 %TC:ignore
@@
 \definecolor{DARKBLUELINK}{HTML}{0B0080}
+\definecolor{LIGHTGREY}{gray}{0.9}
 \PassOptionsToPackage{hyphens}{url}
 \usepackage[colorlinks=false]{hyperref}
+% for linking between references, figures, TOC, etc in the pdf document
 \hypersetup{colorlinks,
@@
     urlcolor=BLUELINK
-}
+} % Color citation links in purple
 \PassOptionsToPackage{unicode}{hyperref}
@@
 
-\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
-\addbibresource{references.bib}
+\usepackage{biorxiv}
 
-\usepackage[margin=50pt]{geometry}
+\usepackage{url}
 \usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
 \usepackage{lmodern}
-\usepackage{bm,bbold}
-\usepackage{verbatim}
-\usepackage{float}
+\usepackage{xfrac, nicefrac}
+\usepackage{bm}
 \usepackage{listings, enumerate, enumitem}
 \usepackage[export]{adjustbox}
+\usepackage{graphicx}
+\usepackage{bbold}
+\usepackage{pdfpages}
+\pdfinclusioncopyfonts=1
+\usepackage{lineno}
 \usepackage{tabu}
-\usepackage{longtable}
-\tabulinesep=0.6mm
-\newcommand\cellwidth{\TX@col@width}
 \usepackage{hhline}
-\setlength{\arrayrulewidth}{1.2pt}
 \usepackage{multicol,multirow,array}
 \usepackage{etoolbox}
-\AtBeginEnvironment{tabu}{\footnotesize}
 \usepackage{booktabs}
-\usepackage{graphicx}
-\pdfinclusioncopyfonts=1
+\usepackage{makecell}
+\usepackage{marvosym}
+\usepackage{orcidlink}
 
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
+% -- Defining colors:
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
+\lstdefinestyle{mystyle}{
+    backgroundcolor=\color{backcolour},
+    basicstyle=\ttfamily\scriptsize\bfseries,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=t,
+    keepspaces=true,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2
+}% -- Setting up the custom style:
+\lstset{style=mystyle}
+\captionsetup[table]{hypcap=false}
+\captionsetup[figure]{hypcap=false}
 
-\input{notations.tex}
+\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
+\newcommand{\Multiply}{\cdot}
+\newcommand{\MultiplyMatrix}{\times}
+\newcommand{\UniDimArray}[1]{\bm{#1}}
+\newcommand{\BiDimArray}[1]{\bm{#1}}
+\newcommand{\tr}{^{\intercal}}
+\newcommand{\inv}{^{-1}}
+\DeclareMathOperator{\E}{\mathbb{E}}
+\DeclareMathOperator{\Var}{\text{var}}
+\DeclareMathOperator{\Cov}{\text{cov}}
+\newcommand{\Qst}{Q$_\text{ST}$}
+\newcommand{\Fst}{F$_\text{ST}$}
+\newcommand{\QstFst}{\Qst--\Fst}
+\newcommand{\der}{\mathrm{d}}
+\newcommand{\e}{\text{e}}
+\newcommand{\Ne}{N_{\text{e}}}
+\newcommand{\dnds}{\omega}
+\newcommand{\pn}{\pi_N}
+\newcommand{\ps}{\pi_S}
+\newcommand{\pnps}{\pn / \ps}
+\newcommand{\proba}{\mathbb{P}}
+\newcommand{\pfix}{\proba_{\text{fix}}}
+\newcommand{\Indiv}{k}
+\newcommand{\Branch}{b}
+\newcommand{\WishartIDD}{m}
+\newcommand{\Spi}{i}
+\newcommand{\Spj}{j}
+\newcommand{\NbrTaxa}{n}
+\newcommand{\Time}{t}
+\newcommand{\NbrGen}{t_{\Spi, \Spj}}
+\newcommand{\NucDiv}{d_{\Spi, \Spj}}
+\newcommand{\Trait}{P}
+\newcommand{\Heritability}{h^2}
+\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
+\newcommand{\MeanTrait}{\bar{\Trait}}
+\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
+\newcommand{\RootTrait}{\phi}
+\newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
+\newcommand{\VecZero}{\UniDimArray{0}}
+\newcommand{\VecOne}{\UniDimArray{1}}
+\newcommand{\Distance}{\BiDimArray{D}}
+\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
+\newcommand{\MutationRatePheno}{\mu}
+\newcommand{\MutationRateNuc}{u}
+\newcommand{\SubRate}{q}
+\newcommand{\NbrLoci}{L}
+\newcommand{\VarPhenotype}{V_{\Trait}}
+\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
+\newcommand{\VarGenetic}{V_{\mathrm{A}}}
+\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
+\newcommand{\VarEnv}{V_{\mathrm{E}}}
+\newcommand{\VarMutation}{V_{\mathrm{M}}}
+\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
+\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
+\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
+\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
+\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
+\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
+\newcommand{\EstRateBetween}{\widehat{\sigma}^2_{\mathrm{B}}}
+\newcommand{\EstRateWhithin}{\widehat{\sigma}^2_{\mathrm{W}}}
+\newcommand{\NI}{\rho}
+\newcommand{\EstNI}{\widehat{\rho}}
+\newcommand{\StdSelection}{\sigma}
+\newcommand{\VarSelection}{\StdSelection^2}
 
-\title{Detection of selection for a quantitative traits through the scaled mutational variance at the phylogenetic and population-genetic level}
+% Tree
+\newcommand{\Nbranch}{2 \NbrTaxa - 2}
+\newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
+\newcommand{\Ntrait}{K}
+\newcommand{\contrast}{\UniDimArray{C}}
+\newcommand{\Covariancematrix}{\Sigma}
+\newcommand{\CovarianceMatrix}{\BiDimArray{\Covariancematrix}}
+\newcommand{\Precisionmatrix}{\Omega}
+\newcommand{\PrecisionMatrix}{\BiDimArray{\Precisionmatrix}}
+\newcommand{\Identitymatrix}{\BiDimArray{I}}
+\newcommand{\brownian}{\mathcal{B}}
+\newcommand{\Brownian}{\UniDimArray{\brownian}}
+\newcommand{\Scattermatrix}{\BiDimArray{A}}
+\newcommand{\Multivariate}{\UniDimArray{Z}}
+
+\renewcommand{\baselinestretch}{1.5}
+\renewcommand{\arraystretch}{1.2}
+\linenumbers
+\frenchspacing
+
+\title{Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes}
+\rhead{\scshape Trait selection from within and between-species variation}
 
@@
     \large
-    \textbf{T. {Latrille}$^{1}$, T. {Gaboriau}$^{1}$, N. {Salamin}$^{1}$}\\
+    \textbf{T.
+{Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M.
+{Bastian}$^{2}$\orcidlink{0009-0003-9129-6918}, T.
+{Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N.
+{Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
     \normalsize
-    $^{1}$Université de Lausanne, Lausanne, Switzerland\\
+    $^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
+    $^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
     \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
@@
 
-\date{}
+\begin{document}
 
-\begin{document}
 \maketitle
+
 \part*{Supplementary materials}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\setcounter{section}{0}
+
+\renewcommand{\baselinestretch}{1.0}\normalsize
 \tableofcontents
-    \clearpage
+\renewcommand{\baselinestretch}{1.5}\normalsize
 
-    \section{Theoretical derivation}
-    \subsection{Definitions}
+\newpage
+\section{Genetic architecture of the trait}\label{sec:simulator}
 
-    We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $\species$.
+\subsection{Genotype-phenotype map}\label{subsec:genotype-phenotype-map}
 
-    \begin{align}
-        \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
-    \end{align}
+\begin{itemize}
+    \item $\NbrLoci$ is the number of loci encoding the trait.
+    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
+    \item $\Ne$ is the effective number of individuals.
+    \item $g_{\Indiv,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $\Indiv \in \{1, \hdots, \Ne\}$.
+    \item $G_{\Indiv} = \sum_{l=1}^{\NbrLoci} a_l \times g_{\Indiv,l}$ is the genotypic value for individual $\Indiv$.
+    \item $\xi_{\Indiv} \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual $\Indiv$.
+    \item $\Trait_{\Indiv} = G_{\Indiv}+ \xi_i$ is the phenotype for individual $\Indiv$.
+\end{itemize}
 
-    \subsection{Mutation-selection-drift equilibrium}
+\begin{center}
+    \captionof{figure}{summary of trait's genetic architecture.}
+    \includegraphics[width=0.6\textwidth, page=1] {figureS1}
+    \label{fig:simulator-summary}
+\end{center}
 
-    Selection, whether positive or negative, depletes standing variation in the population and reduces to genetic variance.
-    Hence, at equilibrium between mutation, selection and drift, the observed genetic variance is lower than the theoretical neutral variance:
-    \begin{align}
-        \VarGenetic & < 2 \Ne \VarMutation \\
-        \iff \VarGenetic & < 4 \Ne \NbrLoci \MutationRate \E \left[ a^2 \right]
-    \end{align}
+within-species, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
+\begin{equation}
+    \bar{G} = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne} G_{\Indiv} \text{\quad and \quad} \VarGenetic = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne}\left(G_{\Indiv}- \bar{G} \right)^2\label{eq:simu-genotype}
+\end{equation}
+The theoretical additive genetic variance ($\VarGenetic$) is a function of the number of loci ($\NbrLoci$) and the effect of a mutation ($a$) as:
+\begin{equation}
+    \VarGenetic = 4 \Ne \Multiply \MutationRatePheno \Multiply \NbrLoci \Multiply a^2 \label{eq:simu-var-genetic}
+\end{equation}
 
-    Under the House of cards model of stabilizing selection\cite{kingman_simple_1978}, the trait's fitness is normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$.
-    Moreover, the trait is encoded by a few loci with large effects and the genetic variance at equilibrium between mutation, selection and drift is:
-    \begin{gather}
-        \VarGenetic = 2 \NbrLoci \MutationRate \VarSelection_S \label{eq-mutsel-HC}.
-    \end{gather}
+The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
+\begin{equation}
+    \bar{\Trait} = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne} \Trait_{\Indiv}\text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{\Indiv=1}^{\Ne}\left(\Trait_{\Indiv} - \bar{\Trait} \right)^2 \label{eq:simu-between}
+\end{equation}
 
-    Alternatively, under the Gaussian model of stabilizing selection\cite{lande_natural_1976}, the trait's fitness is also normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$.
-    However, the trait is encoded by a large number of loci traits with small effect and the genetic variance at equilibrium is:
-    \begin{gather}
-        \VarGenetic = \sqrt{2 \NbrLoci \VarMutation \VarSelection_S} \label{eq-mutsel-G}.
-    \end{gather}
+Heritability ($\Heritability$) is defined as:
+\begin{equation}
+    \Heritability = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
+\end{equation}
+Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($\Heritability$) as:
+\begin{equation}
+    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRatePheno \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
+\end{equation}
 
-    The boundary between the two models is given by \textcite{turelli_heritable_1984} as:
-    \begin{align}
-        20 \MutationRate \VarSelection_S &< \E \left[ a^2 \right] \text{ for the House of card model}, \\
-        20 \MutationRate \VarSelection_S &> \E \left[ a^2 \right] \text{ for the geometric model}.
-    \end{align}
+\newpage
+\section{Bayesian estimate}\label{sec:bayesian-estimate}
 
-    \subsection{Selection at the phylogenetic scale}
-    If the trait is under directional selection, but the strength and direction of selection varies randomly from one generation to the next (variance $\VarSelection_D$), then the trait still follows a Brownian process:
-    \begin{align}
-        \VarPhy &= \Var \left( \Trait_z \right), \\
-        &= \left(\frac{\VarGenetic}{\Ne} + \VarGenetic \VarSelection_D \VarGenetic \right) t. \label{eq-selection-vary-direct}
-    \end{align}
+\subsection{Multivariate Brownian process}\label{subsec:multivariate-brownian-process}
+Here we model $\Ntrait$ traits evolving along the phylogenetic tree that are correlated between them.
+Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology.
+The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
+Along branch $\Branch$ with length $d_{\Branch}$, the Brownian process start at the ancestral node $\mathcal{A}(\Branch)$ with value $\Brownian(\mathcal{A}(\Branch))$, and ends at node $\mathcal{R}(\Branch)$  with value $\Brownian(\mathcal{R}(\Branch))$.
+The independent contrast $\contrast_{\Branch}$ defined as change in trait along the branch normalized by $\sqrt {d_{\Branch}}$ is a multivariate Gaussian:
+\begin{equation}
+    \label{eq:DistribBrownian}
+    \contrast_{\Branch} = \frac{\Brownian (\mathcal{R}(\Branch)) - \Brownian (\mathcal{A}(\Branch)) }{\sqrt {d_{\Branch}}} \sim \mathcal{N}\left(\VecZero, \CovarianceMatrix \right).
+\end{equation}
 
-    If the trait is under stabilizing selection, but the strength and direction of the optimum varies randomly from one generation to the next (variance $\VarSelection_{\theta}$), then the trait still follows a Brownian process:
-    \begin{align}
-        \VarPhy &= \VarSelection_{\theta} t. \label{eq-selection-vary-stabi}
-    \end{align}
+\subsection{Sampling the covariance matrix}\label{subsec:sampling-the-covariance-matrix}
+From the independent contrast at each branch of the tree ($\contrast_{\Branch}$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$, as:
+\begin{equation}
+    \Scattermatrix = \sum\limits_{\Branch=1}^{\Nbranch} \contrast_{\Branch} \MultiplyMatrix \contrast_{\Branch}\tr\label{eq:bayes-scatter},
+\end{equation}
+where $\Nbranch$ is the number of branches in the tree and $\NbrTaxa$ the number of taxa.
 
-    Altogether, traits that are found to be under a long term Brownian process are not necessarily neutral.
+The {prior} on the covariance matrix is an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
+\begin{equation}
+    \label{eq:Distribcovariance}
+    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
+\end{equation}
 
-    \printbibliography
+By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\Brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
+\begin{equation}
+    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)\label{eq:bayes-posterior}
+\end{equation}
+This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate_{\WishartIDD}$ defined by
+\begin{equation}
+    \Multivariate_{\WishartIDD} \sim \mathcal{N} \left( \VecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).\label{eq:bayes-multivariate}
+\end{equation}
+And from these multivariate samples, $\CovarianceMatrix$ is Gibbs sampled as:
+\begin{equation}
+    \CovarianceMatrix = \left( \sum\limits_{\WishartIDD=1}^{\WishartPostDf} \Multivariate_{\WishartIDD} \MultiplyMatrix  \left[\Multivariate_{\WishartIDD} \right] \tr \right)\inv \label{eq:bayes-gibbs}
+\end{equation}
 
+\newpage
+\section{Bayesian and Maximum-likelihood implementation}\label{sec:implementation}
+
+Implementation is included within the \textit{BayesCode} software, available at \url{https://github.com/ThibaultLatrille/bayescode}.
+
+\subsection{Data formatting}\label{subsec:data-formatting}
+
+Running the analysis on your dataset and compute posterior probabilities requires three files:
+\begin{enumerate}
+    \item A phylogenetic tree in newick format, with branch lengths in number of substitutions per site (neutral markers), from which the values of nucleotide divergence ($d$) used in denominator of eq.~26 are used.
+    \item A file containing the mean trait values for each species.
+    \item A file containing the variation within-species for each trait and the genetic variation within-species (neutral markers).
+\end{enumerate}
+
+\subsubsection{Phylogenetic tree}
+
+The phylogenetic tree must be in newick format, with branch lengths in substitutions per site (neutral markers).
+
+\subsubsection{Mean trait for each species}
+
+The file containing mean trait values for each species must be in a tab-delimited file with the following format:
+\begin{center}
+    \begin{adjustbox}{width = 0.35\textwidth}
+        \begin{tabular}{|l|c|c|}
+            \hline
+            TaxonName            & Body\_mass & Brain\_mass \\
+            \hline
+            Panthera\_tigris     & 12.26      & 5.676       \\
+            Pithecia\_pithecia   & 7.256      & 3.436       \\
+            Colobus\_angolensis  & 9.176      & 4.284       \\
+            Saimiri\_boliviensis & 6.845      & 3.279       \\
+            $\vdots$             & $\vdots$   & $\vdots$    \\
+            \hline
+        \end{tabular}\label{tab:trait-mean}
+    \end{adjustbox}
+\end{center}
+
+The columns are:
+\begin{itemize}
+    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
+    \item As many columns as traits, without spaces or special characters in the trait.
+    \item The values can be \texttt{NaN} to indicate that the trait is not available for that taxon.
+\end{itemize}
+
+\newpage
+\subsubsection{Trait variation for each species}
+
+The file containing trait variation for each species must be in a tab-delimited file with the following format:
+\begin{center}
+    \begin{adjustbox}{width = 1.0\textwidth}
+        \begin{tabular}{|l|c|c|c|c|c|c|}
+            \hline
+            TaxonName            & Nucleotide\_diversity & Body\_mass\_variance & Body\_mass\_heritability & Brain\_mass\_variance & Brain\_mass\_heritability \\
+            \hline
+            Pithecia\_pithecia   & 0.0016                & 0.22871              & 0.2                      & 0.00737               & 0.2                       \\
+            Colobus\_angolensis  & 0.0017                & 0.00393              & 0.2                      & 0.00416               & 0.2                       \\
+            Saimiri\_boliviensis & 0.0013                & 0.00022              & 0.2                      & 0.00045               & 0.2                       \\
+            Pygathrix\_nemaeus   & 0.0016                & 0.00347              & 0.2                      & 0.00097               & 0.2                       \\
+            $\vdots$             & $\vdots$              & $\vdots$             & $\vdots$                 & $\vdots$              & $\vdots$                  \\
+            \hline
+        \end{tabular}
+        \label{tab:trait-variance}
+    \end{adjustbox}
+\end{center}
+
+\begin{itemize}
+    \item \emph{TaxonName}: the name of the taxon matching the name in the alignment and the tree.
+    \item \emph{Nucleotide\_diversity}: the nucleotide diversity within-species (neutral markers), cannot be \texttt{NaN}.
+    \item As many columns as traits, without spaces or special characters in the trait.
+    \item \emph{TraitName\_variance}: the phenotypic variance of the trait within-species, can be \texttt{NaN} to indicate that the trait variance is not available for that taxon.
+    \item \emph{TraitName\_heritability} (optional): the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
+    \item The columns with the suffix \texttt{\_variance} and \texttt{\_heritability} are repeated for each trait.
+    \item \emph{TraitName\_heritability\_lower} (optional): the lower bound of the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
+    \item \emph{TraitName\_heritability\_upper} (optional): the upper bound of the heritability of the trait within-species, between 0 and 1, cannot be \texttt{NaN}.
+    \item If the columns with the suffix \texttt{\_heritability\_lower} and \texttt{\_heritability\_upper} are present, the heritability is randomly drawn from a uniform distribution between the lower and upper bounds.
+    \item If the columns with the suffix \texttt{\_heritability} is present, it is taken as is.
+    \item If the additive genetic variance (instead of phenotypic variance) is available for a trait, the heritability can be omitted and will automatically be set to 1.0.
+\end{itemize}
+
+\newpage
+\subsection{Bayesian estimation}\label{subsec:running-nodetraitsand-readnodetraits}
+
+The executable \texttt{nodetraits} from \textit{BayesCode} is used to run the Bayesian estimation of the model, and the executable \texttt{readnodetraits} is used to read the results.
+
+Assuming that the file \texttt{data/body\_size/mammals.male.tsv} contains the mean trait values for each species, the file \texttt{data/body\_size/mammals.male.var\_trait.tsv} contains the variation within-species for each trait and the genetic variation within-species (neutral markers), and the file \texttt{data/body\_size/mammals.male.tree} contains the phylogenetic tree, the following commands are used to run the model and read the results.
+
+\subsubsection{Running the model}
+\texttt{nodetraits} is run with the following command:
+\begin{lstlisting}[language = sh,label={lst:nodetraits-run}]
+nodetraits  --until 2000
+            --tree data/body_size/mammals.male.tree
+            --traitsfile data/body_size/mammals.male.tsv
+            run_mammals_male
+\end{lstlisting}
+
+\subsubsection{Reading the results}
+Once the model has run, the chain \texttt{run\_mammals\_male} is used to compute the posterior distribution of the ratio of between-species variation over within-species variation with \texttt{readnodetraits}:
+\begin{lstlisting}[language = sh,label={lst:readnodetraits-rho}]
+readnodetraits --burnin 1000
+               --var_within data/body_size/mammals.male.var_trait.tsv
+               --output results_mammals_male.tsv
+               run_mammals_male
+\end{lstlisting}
+The file \texttt{data\_empirical/chain\_name.ratio.tsv} then contains the posterior mean of the ratio of between-species variation over within-species variation, the 95\% and 99\% credible interval, and the posterior probability that the ratio is greater than 1.
+
+\subsection{Maximum likelihood estimation}\label{subsec:maximum-likelihood-estimation}
+
+To obtain the ratio (without the posterior credible interval and probability) using maximum likelihood computation, the following python script can be used:
+\begin{lstlisting}[language = sh, label={lst:neutrality_index}]
+python3 utils/neutrality_index.py --tree data/body_size/mammals.male.tree
+                                  --traitsfile data/body_size/mammals.male.tsv
+                                  --var_within data/body_size/mammals.male.var_trait.tsv
+                                  --output results_ML_mammals_male.tsv
+\end{lstlisting}
+
+\newpage
+\section{Saturation of phenotypic divergence}\label{sec:supp-distance}
+
+In our simulation setting, the genetic architecture underlying the phenotype is not changing along the phylogeny.
+As a result, the phenotype may be bounded by what such genetic architecture can produce, and this could cause a slowdown of phenotypic divergence over time, ultimately resulting in a decreased $\RateBetween$.
+
+\begin{center}
+    \captionof{figure}{Saturation of phenotypic divergence.}
+    \includegraphics[width=1.0\textwidth, page=1] {figureS2}
+    \label{fig:supp-distance}
+\end{center}
+\begin{itemize}
+    \item Left column: $1,000$ simulations with low divergence between species (half that of mammals).
+    \item Middle column: $1,000$ simulations with high divergence between species (4 times that of mammals).
+    \item Right column: $200$ simulations with very high divergence between species (20 times that of mammals).
+    \item Top row: Simulation of neutral trait, phenotypic divergence between species as a function of the nucleotide divergence.
+Phenotypic divergence is computed between two species as the covariance between the trait ($\VarPhy$), and the nucleotide divergence is computed as the number of substitutions per site shared between the two species ($d$).
+Each point is a pair of species, and the bounds of the intervals in grey are the 2.5\% and 97.5\% quantiles across the replicates.
+The blue line is the linear regression and the red line is the saturation model ($y = \alpha \Multiply x / (\beta + x)$).
+    \item Bottom row: Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
+Histogram of probabilities of $\NI$ being greater than $1$.
+\end{itemize}
+Under a model of neutral trait evolution, when mutation rate increases, or equivalently the divergence between species increases, the phenotypic divergence between species saturates faster than the nucleotide divergence.
+This saturation effect can result in a spurious signal of stabilizing selection ($\NI < 1$) for deeper phylogeny when the trait is evolving neutrally.
 \end{document}
+%TC:endignore

### 2024-07-05 data_simulated/RevBayesMerge.py

@@ -1,56 +1,0 @@
-#!python3
-import os
-from glob import glob
-
-
-def minipage(size, file, name=""):
-    out = "\\begin{minipage}{" + str(size) + "\\linewidth} \n"
-    out += "\\centering " + name + "\n"
-    out += "\\includegraphics[width=\\linewidth, page=1]{" + file + "} \n"
-    out += "\\end{minipage}\n"
-    return out
-
-
-def write(exp, o):
-    o.write("\\begin{center}\n")
-
-    for method in ["neutral", "moving_optimum", "stabilizing"]:
-        distance_path = f"{exp}/results/inference_ML.distance.{method}.pdf"
-        if not os.path.exists(distance_path):
-            continue
-        o.write(minipage(0.32, distance_path, method.replace("_", " ").title()))
-
-    for param in ["is_OU", "t_half", "sigma2"]:
-        hist_path = f"{exp}/results/inference_RevBayes.simple_OU.{param}.pdf"
-        if not os.path.exists(hist_path):
-            continue
-        o.write(minipage(0.32, hist_path, param.replace("_", " ").title()))
-
-    for param in ["num_theta_changes", "t_half", "sigma2"]:
-        hist_path = f"{exp}/results/inference_RevBayes.relaxed_OU.{param}.pdf"
-        if not os.path.exists(hist_path):
-            continue
-        o.write(minipage(0.32, hist_path, param.replace("_", " ").title()))
-
-    o.write("\\end{center}\n")
-
-
-def main():
-    tex_include = "docs/include-figures.tex"
-    tex_target = "docs/supp-mat.tex"
-    experiments = glob("experiments/*_u")
-    o = open(tex_include, 'w')
-    o.write("\\section{Experiments} \n")
-    for exp in experiments:
-        subsection = os.path.basename(exp).replace("_", " ").title()
-        o.write("\\subsection{" + subsection + "}\n")
-        write(exp, o)
-    o.close()
-    output_dir = os.path.dirname(tex_target)
-    tex_to_pdf = f"pdflatex -synctex=1 -interaction=nonstopmode -output-directory={output_dir} {tex_target}"
-    os.system(tex_to_pdf)
-    os.system(tex_to_pdf)
-
-
-if __name__ == '__main__':
-    main()

### 2024-07-05 data_simulated/run.sh

@@ -1,4 +1,4 @@
 CPU=8
-for EXPERIMENT in ./config/constant_pop_size_many_sites*.yaml; do
+for EXPERIMENT in ./config/*.yaml; do
 python3 simulated_experiment.py -c $(basename "${EXPERIMENT}") -j ${CPU}
 done

### 2024-07-05 data_simulated/update_snake.sh

@@ -1,11 +1,0 @@
-# shellcheck disable=SC2164
-# shellcheck disable=SC2046
-FOLDER=$(pwd)
-for EXPERIMENT in $(pwd)/experiments/*_u; do
-  echo "Updating snakemake in: ${EXPERIMENT}"
-  rm -rf "${EXPERIMENT}"/Snakefile
-  ln $(pwd)/Snakefile "${EXPERIMENT}"/Snakefile
-  cd "${EXPERIMENT}"
-  snakemake -j 8 -k
-  cd "${FOLDER}"
-done

### 2024-07-05 utils/BayesCode/src/AAMutSelDM5Model.hpp
Include BayesCode source code

@@ -1,0 +1,1351 @@
+#include "AAMutSelCodonMatrixBidimArray.hpp"
+#include "Chrono.hpp"
+#include "CodonSequenceAlignment.hpp"
+#include "CodonSuffStat.hpp"
+#include "DGamma.hpp"
+#include "GTRSubMatrix.hpp"
+#include "GammaSuffStat.hpp"
+#include "IIDDirichlet.hpp"
+#include "IIDGamma.hpp"
+#include "Move.hpp"
+#include "MultinomialAllocationVector.hpp"
+#include "Permutation.hpp"
+#include "PhyloProcess.hpp"
+#include "StickBreakingProcess.hpp"
+#include "Weights.hpp"
+#include "components/ChainComponent.hpp"
+#include "components/Tracer.hpp"
+#include "components/model_decl_utils.hpp"
+
+/**
+ * \brief The mutation-selection model with constant fitness landscape over the
+ * tree -- double Dirichlet process version.
+ *
+ * The model is parameterized by
+ * - an fixed unrooted phylogenetic tree tau, with branch lengths l = (l_j), for
+ * j running over branches
+ * - a GTR nucleotide matrix Q = rho * pi, specifying the mutation process
+ * (assumed homogeneous across sites and lineages)
+ * - an array of site-specific amino-acid fitness profiles F_ia, for site i and
+ * amino-acid a
+ * - an omega multiplier, capturing deviations of the non-syn rate from the
+ * model (see Rodrigue and Lartillot, 2107); this parameter is fixed to 1 by
+ * default.
+ *
+ * Site-specific amino-acid fitness profiles are drawn from a Dirichlet process,
+ * implemented using a truncated stick-breaking process, of concentration
+ * parameter kappa, and truncated at Ncat.
+ *
+ * The base distribution of this Dirichlet process, G_0, is itself a (truncated
+ * stick breaking) mixture of Dirichlet distributions, parameterized by
+ * basekappa, and truncated at baseNcat. Each component of this mixture
+ * Dirichlet is parameterized by a center (a 20-freqeuncy vector) and a
+ * concentration.
+ *
+ * Concerning the first-level stick-breaking process, by default, Ncat == 100
+ * (can be changed using the  -ncat option). As for baseNcat, it is equal to 1,
+ * in which case the base distribution G_0 is just a simple Dirichlet (as in
+ * Rodrigue et al, 2010). This simple model is probably the one that should be
[…]

### 2024-07-05 utils/BayesCode/src/CodonM2aModel.cpp

@@ -1,0 +1,541 @@
+#include "CodonM2aModel.hpp"
+#include "Move.hpp"
+
+// constuction allocation
+//
+
+void CodonM2aModel::init() {
+    blmode = 0;
+    nucmode = 0;
+    data = new FileSequenceAlignment(datafile);
+    if (data->GetNsite() % 3) {
+        cerr << "error : not a correctly formatted codon-alignment: " << datafile << '\n';
+        exit(1);
+    }
+    codondata = new CodonSequenceAlignment(data, true);
+
+    Nsite = codondata->GetNsite();  // # columns
+    Ntaxa = codondata->GetNtaxa();
+
+    taxonset = codondata->GetTaxonSet();
+
+    // get tree from file (newick format)
+    std::ifstream tree_stream{treefile};
+    NHXParser parser{tree_stream};
+    tree = make_from_parser(parser);
+
+    Nbranch = tree->nb_branches();
+
+    Allocate();
+    tracer = unique_ptr<Tracer>(new Tracer(*this));
+}
+
+void CodonM2aModel::Allocate() {
+    // Branch lengths
+    blhypermean = 0.1;
+    blhyperinvshape = 1.0;
+    blhypermeanarray = new SimpleBranchArray<double>(*tree, blhypermean);
+    branchlength = new GammaWhiteNoise(*tree, *blhypermeanarray, blhyperinvshape);
+
+    purom = 0.5;
+    puromhypermean = 0.5;
+    puromhyperinvconc = 0.5;
+
+    dposom = 1.0;
+    dposomhypermean = 0.5;
+    dposomhyperinvshape = 0.5;
+
+    purw = 0.1;
+    purwhypermean = 0.5;
+    purwhyperinvconc = 0.5;
+
+    if (!pi) {
+        posw = 0;
+        poswhypermean = 0;
+        poswhyperinvconc = 0;
+    } else {
+        posw = 0.1;
+        poswhypermean = 0.5;
+        poswhyperinvconc = 0.5;
+    }
+
+    componentomegaarray = new M2aMix(purom, dposom + 1, purw, posw);
+    sitealloc = new MultinomialAllocationVector(GetNsite(), componentomegaarray->GetWeights());
+    sitepostprobarray.assign(GetNsite(), vector<double>(3, 0));
+
+    nucrelratehypercenter.assign(Nrr, 1.0 / Nrr);
[…]

### 2024-07-05 utils/BayesCode/src/DatedNodeMutSel.cpp

@@ -1,0 +1,110 @@
+#include <cmath>
+#include <fstream>
+#include "DatedNodeMutSelModel.hpp"
+#include "components/ChainCheckpoint.hpp"
+#include "components/ChainDriver.hpp"
+#include "components/ConsoleLogger.hpp"
+#include "components/InferenceAppArgParse.hpp"
+#include "components/StandardTracer.hpp"
+#include "components/restart_check.hpp"
+
+using namespace std;
+
+class DatedNodeMutselArgParse : public BaseArgParse {
+  public:
+    explicit DatedNodeMutselArgParse(ChainCmdLine &cmd) : BaseArgParse(cmd) {}
+
+    ValueArg<int> ncat{"", "ncat",
+        "Number of components for the amino-acid fitness profiles "
+        "(truncation of the stick-breaking process).",
+        false, 30, "int", cmd};
+    /*
+    SwitchArg condition_aware{"b", "condition_aware",
+        "One Ne per condition, if the tree doesn't have condition, then one Ne per branch
+    (experimental).", cmd, false};
+    */
+    ValueArg<std::string> traitsfile{
+        "", "traitsfile", "File path to the life-history trait (in log-space) in tsv format. "
+        "The First column is `TaxonName` (taxon matching the name in the alignment) and the next columns are traits."
+        "", false, "Null", "string", cmd};
+    ValueArg<std::string> profiles{"", "profiles",
+        "File path the fitness profiles (tsv or csv), thus considered fixed. "
+        "Each line must contains the fitness of each of the 20 amino-acid, thus summing to one. "
+        "If same number of profiles as the codon alignment, site allocations are considered fixed. "
+        "If smaller than the alignment size, site allocations are computed and `ncat` is given by "
+        "the number of profiles in the file.",
+        false, "Null", "string", cmd};
+    SwitchArg move_root_pop_size{"", "move_root_pop_size",
+        "Move Ne at the root, for the equilibrium frequencies (experimental)", cmd, false};
+    SwitchArg clamp_pop_sizes{
+        "", "clamp_pop_sizes", "Clamp the branch population size (experimental).", cmd, false};
+    SwitchArg clamp_nuc_matrix{
+        "", "clamp_nuc_matrix", "Clamp the nucleotide matrix (experimental).", cmd, false};
+    SwitchArg clamp_corr_matrix{
[…]

### 2024-07-05 utils/BayesCode/src/DatedNodeMutSelModel.hpp

@@ -1,0 +1,2006 @@
+#pragma once
+
+#include <cassert>
+#include "AAMutSelNeCodonMatrixBidimArray.hpp"
+#include "BranchComponentMatrixSelector.hpp"
+#include "BranchProduct.hpp"
+#include "Chrono.hpp"
+#include "Chronogram.hpp"
+#include "CodonSequenceAlignment.hpp"
+#include "CodonSuffStat.hpp"
+#include "GTRSubMatrix.hpp"
+#include "GammaSuffStat.hpp"
+#include "IIDDirichlet.hpp"
+#include "IIDGamma.hpp"
+#include "Move.hpp"
+#include "MultinomialAllocationVector.hpp"
+#include "MultivariateProcess.hpp"
+#include "Permutation.hpp"
+#include "PhyloProcess.hpp"
+#include "PolyProcess.hpp"
+#include "PolySuffStat.hpp"
+#include "ScaledMutationRateCompound.hpp"
+#include "ScatterSuffStat.hpp"
+#include "StickBreakingProcess.hpp"
+#include "TaxonTraits.hpp"
+#include "components/ChainComponent.hpp"
+#include "components/Tracer.hpp"
+
+/**
+ * \brief The mutation-selection model with constant fitness landscape over the
+ * tree -- double Dirichlet process version.
+ *
+ * The model is parameterized by
+ * - an fixed unrooted phylogenetic tree tau, with branch lengths l = (l_j), for
+ * j running over branches
+ * - a GTR nucleotide matrix Q = rho * pi, specifying the mutation process
+ * (assumed homogeneous across sites and lineages)
+ * - an array of site-specific amino-acid fitness profiles F_ia, for site i and
+ * amino-acid a
+ * - A branch Ne
+ * - A branch mutation rate (correlated to Ne)
+ *
+ * Site-specific amino-acid fitness profiles are drawn from a Dirichlet process,
+ * implemented using a truncated stick-breaking process, of concentration
+ * parameter kappa, and truncated at Ncat.
+ *
+ * The base distribution of this Dirichlet process, G_0, is itself a (truncated
+ * stick breaking) mixture of Dirichlet distributions, parameterized by
+ * basekappa, and truncated at baseNcat. Each component of this mixture
+ * Dirichlet is parameterized by a center (a 20-freqeuncy vector) and a
+ * concentration.
+ *
+ * Concerning the first-level stick-breaking process, by default, Ncat == 100
+ * (can be changed using the  -ncat option). As for baseNcat, it is equal to 1,
[…]

### 2024-07-05 utils/BayesCode/src/ReadAAMutSelMultipleOmega.cpp

@@ -1,0 +1,300 @@
+#include <cmath>
+#include <fstream>
+#include "AAMutSelMultipleOmegaModel.hpp"
+#include "components/ChainDriver.hpp"
+#include "components/ChainReader.hpp"
+#include "components/ReadArgParse.hpp"
+#include "components/stats_posterior.hpp"
+#include "tclap/CmdLine.h"
+
+using namespace std;
+using namespace TCLAP;
+
+class ReadAAMutSelDSBDPOmegaArgParse : public ReadArgParse {
+  public:
+    explicit ReadAAMutSelDSBDPOmegaArgParse(CmdLine &cmd) : ReadArgParse(cmd) {}
+
+    SwitchArg nuc{"n", "nuc",
+        "Mean posterior 4x4 nucleotide matrix."
+        "Results are written in {chain_name}.nucmatrix.tsv by default (optionally use the --output argument "
+        " to specify a different output path).",
+        cmd};
+    ValueArg<string> chain_omega{"", "chain_omega",
+        "A second chain ran with the option --freeomega and --flatfitness to obtain the classical "
+        "ω-based codon model (Muse & Gaut). "
+        "These two chains allow to compute posterior of ω, ω₀, ωᴬ=ω-ω₀ and p(ωᴬ>0) for each site "
+        "and at the gene level. "
+        "Results are written in {chain_name}.omegaA.tsv by default (optionally use the --output argument "
+        " to specify a different output path).",
+        false, "", "string", cmd};
+    SwitchArg omega{"", "omega",
+        "Compute posterior credible interval for ω for each site and at the gene level. "
+        "Can be combined with the option `confidence_interval` to change the default value (0.025 "
+        "at each side of the distribution). "
+        "Results are written in {chain_name}.ci{confidence_interval}.tsv by default (optionally "
+        "use the --output argument to specify a different output path).",
+        cmd};
+    SwitchArg omega_knot{"", "omega_0",
+        "Compute posterior credible interval for ω₀ predicted at the mutation-selection "
+        "equilibrium from the fitness profiles, for each site and at the gene level. "
+        "Can be combined with the option `confidence_interval` to change the default value (0.025 "
+        "at each side of the distribution). "
+        "Results are written in {chain_name}.ci{confidence_interval}.tsv by default (optionally "
[…]

### 2024-07-05 utils/BayesCode/src/ReadDatedNode.cpp

@@ -1,0 +1,399 @@
+#include <cmath>
+#include <fstream>
+#include "DatedNodeModel.hpp"
+#include "components/ChainDriver.hpp"
+#include "components/ChainReader.hpp"
+#include "components/ReadArgParse.hpp"
+#include "components/csv_parser.hpp"
+#include "components/stats_posterior.hpp"
+#include "tclap/CmdLine.h"
+#include "tree/export.hpp"
+
+using namespace std;
+using namespace TCLAP;
+
+
+class ReadNodeArgParse : public ReadArgParse {
+  public:
+    explicit ReadNodeArgParse(CmdLine& cmd) : ReadArgParse(cmd) {}
+
+    SwitchArg newick{"t", "newick",
+        "Computes the mean posterior node-specific entries of the multivariate Brownian process. "
+        "Each entry of the multivariate Brownian process is written in a newick extended (.nhx) "
+        "format file."
+        "For each trait, results are written in {chain_name}.{trait}.nhx by default (optionally "
+        "use the --output argument to specify a different output path).",
+        cmd};
+    SwitchArg wAIC{"w", "wAIC",
+        "Computes the wAIC for the model."
+        "Results are written in {chain_name}.wAIC.txt by default (optionally "
+        "use the --output argument to specify a different output path).",
+        cmd};
+    SwitchArg cov{"c", "cov",
+        "Computes the mean posterior covariance matrix, precision matrix and correlation matrix. "
+        "Results are written in {chain_name}.cov by default (optionally use the --output argument "
+        "to specify a different output path).",
+        cmd};
+    ValueArg<string> var_within{"", "var_within",
+        "An input .tsv file containing the within-population variance of each trait (see "
+        "documentation for file format).",
+        false, "", "string", cmd};
+};
+
+class VarWithinSample {
+  public:
+    VarWithinSample() = default;
+
+    VarWithinSample(const CSVParser& var_within, const string& trait_name) {
+        size_t taxa_col_index = var_within.find_column_index("TaxonName");
+        if (taxa_col_index == string::npos) {
+            cerr << "Error: Could not find column 'TaxonName'" << endl;
+            exit(1);
+        }
+        size_t ps_index = var_within.find_column_index("Nucleotide_diversity");
+        if (ps_index == string::npos) {
+            cerr << "Error: Could not find column 'Nucleotide_diversity'" << endl;
+            exit(1);
+        }
[…]

### 2024-07-05 utils/BayesCode/src/ReadDatedNodeMutSel.cpp

@@ -1,0 +1,191 @@
+#include <cmath>
+#include <fstream>
+#include "DatedNodeMutSelModel.hpp"
+#include "components/ChainDriver.hpp"
+#include "components/ChainReader.hpp"
+#include "components/ReadArgParse.hpp"
+#include "components/stats_posterior.hpp"
+#include "tclap/CmdLine.h"
+#include "tree/export.hpp"
+
+using namespace std;
+using namespace TCLAP;
+
+
+class ReadNodeMutSelArgParse : public ReadArgParse {
+  public:
+    explicit ReadNodeMutSelArgParse(CmdLine &cmd) : ReadArgParse(cmd) {}
+
+    SwitchArg ss{"s", "ss",
+        "Computes the mean posterior site-specific amino-acid equilibrium frequencies"
+        "(amino-acid fitness profiles). "
+        "Results are written in {chain_name}.siteprofiles by default (optionally use the --output argument "
+        " to specify a different output path).",
+        cmd};
+    SwitchArg newick{"t", "newick",
+        "Computes the mean posterior node-specific entries of the multivariate Brownian process. "
+        "Each entry of the multivariate Brownian process is written in a newick extended (.nhx) "
+        "format file."
+        "For each trait, results are written in {chain_name}.{trait}.nhx by default (optionally "
+        "use the --output argument to specify a different output path).",
+        cmd};
+    SwitchArg cov{"c", "cov",
+        "Computes the mean posterior covariance matrix, precision matrix and correlation matrix. "
+        "Results are written in {chain_name}.cov by default (optionally use the --output argument "
+        "to specify a different output path).",
+        cmd};
+};
+
+
+int main(int argc, char *argv[]) {
+    CmdLine cmd{"DatedMutSel", ' ', "0.1"};
+    ReadNodeMutSelArgParse read_args(cmd);
+    cmd.parse(argc, argv);
+
+    string chain_name = read_args.GetChainName();
+    int burnin = read_args.GetBurnIn();
+    int every = read_args.GetEvery();
+    int size = read_args.GetSize();
+
+    ifstream is{chain_name + ".param"};
+    unique_ptr<DatedNodeMutSelModel> model = nullptr;
+    new ChainDriver(is);
+    is >> model;
+    ChainReader cr(*model, chain_name + ".chain");
+
+    cr.skip(burnin);
+    cerr << size << " points to read\n";
+
+    if (read_args.GetPpred()) {
+        for (int i = 0; i < size; i++) {
+            cerr << '.';
[…]

### 2024-07-05 utils/BayesCode/src/SingleOmegaModel.hpp

@@ -1,0 +1,554 @@
+#pragma once
+
+#include "CodonSequenceAlignment.hpp"
+#include "CodonSubMatrix.hpp"
+#include "CodonSuffStat.hpp"
+#include "GTRSubMatrix.hpp"
+#include "GammaSuffStat.hpp"
+#include "IIDGamma.hpp"
+#include "Move.hpp"
+#include "PhyloProcess.hpp"
+#include "components/ChainComponent.hpp"
+#include "components/Tracer.hpp"
+
+/**
+ * \brief A standard site- and branch-homogeneous Muse and Gaut omega-codon
+ * model
+ *
+ * The model has the following structure:
+ * - branch lengths iid Exponential of mean blhypermean
+ * - nucleotide relative exchangeabilities and stationaries are uniform
+ * Dirichlet
+ * - there is one single omega=dN/dS for all sites and across all branches
+ * - prior over omega ~ Gamma(omegahypermean,omegahyperinvshape).
+ *
+ * The 2 hyperparameters omegahypermean and hyperinvshape are fixed when this
+ * model is used in isolation. On the other hand, this model can be used in a
+ * multigene context (see MultiGeneSingeOmegaModel), in which case
+ * omegahypermean and hyperinvshape are estimated across genes.
+ */
+
+// mode shared:      global
+// mode shrunken:    gene specific, with hyperparameters estimated across genes
+// mode independent: gene-specific, with fixed hyperparameters
+enum param_mode_t { shared, shrunken, independent };
+
+class SingleOmegaModel : public ChainComponent {
+    // tree and data
+    std::string datafile, treefile;
+    std::unique_ptr<const Tree> tree;
+    FileSequenceAlignment *data;
+    const TaxonSet *taxonset;
+    CodonSequenceAlignment *codondata;
+
+    int Nsite;
+    int Ntaxa;
+    int Nbranch;
+
+    param_mode_t blmode;
+    param_mode_t nucmode;
+
+    // Branch lengths
+    double blhypermean;
+    double blhyperinvshape;
+    SimpleBranchArray<double> *blhypermeanarray;
+    GammaWhiteNoise *branchlength;
+
+    // Poisson suffstats for substitution histories, as a function of branch
+    // lengths
+    PoissonSuffStatBranchArray *lengthpathsuffstatarray;
+
+    // Nucleotide rates
+    std::vector<double> nucrelratehypercenter;
+    double nucrelratehyperinvconc;
[…]

### 2024-07-05 utils/BayesCode/src/all_tests.cpp

@@ -1,0 +1,9 @@
+#define DOCTEST_CONFIG_IMPLEMENT_WITH_MAIN
+
+#include "components/test.cpp"
+
+#include "operations/test.cpp"
+
+#include "tags/test.cpp"
+
+#include "mpi_components/test/seq_test.cpp"

### 2024-07-05 utils/BayesCode/src/components/ChainReader.hpp

@@ -1,0 +1,19 @@
+#pragma once
+
+#include "Tracer.hpp"
+
+class ChainReader {
+    Tracer tracer;
+    std::ifstream is;
+
+  public:
+    template <class M>
+    ChainReader(M& model, std::string filename) : tracer(model), is(filename) {
+        tracer.ignore_header(is);
+    }
+
+    void next() { tracer.read_line(is); }
+    void skip(int n) {
+        for (int i = 0; i < n; i++) next();
+    }
+};

### 2024-07-05 utils/BayesCode/src/components/ReadArgParse.hpp

@@ -1,0 +1,81 @@
+#pragma once
+
+#include <cassert>
+#include <fstream>
+#include "tclap/CmdLine.h"
+
+class ReadArgParse {
+  protected:
+    TCLAP::CmdLine& cmd;
+    int until{-1};
+    int burnin{-1};
+
+  public:
+    explicit ReadArgParse(TCLAP::CmdLine& cmd) : cmd{cmd} {}
+
+    TCLAP::ValueArg<int> every{"e", "every",
+        "Number of MCMC iterations between two saved point in the trace.", false, 1, "int", cmd};
+    TCLAP::ValueArg<int> until_input{"u", "until",
+        "Maximum number of (saved) iterations (-1 means unlimited).", false, -1, "int", cmd};
+    TCLAP::ValueArg<int> burnin_input{
+        "b", "burnin", "Number of MCMC iterations for the burn-in.", false, 0, "int", cmd};
+    TCLAP::SwitchArg ppred{"p", "ppred",
+        "For each point of the chain (after burn-in), produces a data replicate simulated "
+        "from the posterior predictive distribution",
+        cmd};
+    TCLAP::SwitchArg trace{"", "trace",
+        "Recompute the trace."
+        "Trace is written in {chain_name}.trace.tsv by default (optionally "
+        "use the --output argument to specify a different output path).",
+        cmd};
+    TCLAP::UnlabeledValueArg<std::string> chain_name{
+        "chain_name", "Chain name (output file prefix).", true, "chain", "string", cmd};
+    TCLAP::ValueArg<string> output{
+        "o", "output", "Output file path (optional)", false, "", "string", cmd};
+
+    std::string GetChainName() { return chain_name.getValue(); }
+
+    bool GetPpred() { return ppred.getValue(); }
+
+    int GetBurnIn() {
+        if (burnin == -1) {
+            if (int((GetUntil() - burnin_input.getValue()) / GetEvery()) < 1) {
+                std::cerr << "Burn-in was set to " << burnin_input.getValue()
+                          << " but there is only " << GetUntil() << "  points to read.\n";
+                burnin = std::max(GetUntil() - 10 * GetEvery(), 0);
+                std::cerr << "Setting the burn-in to " << burnin << std::endl;
+            } else {
+                burnin = burnin_input.getValue();
+            }
+        }
+        return burnin;
+    }
+
+    int GetEvery() { return every.getValue(); }
+
[…]

### 2024-07-05 utils/BayesCode/src/components/StandardTracer.hpp

@@ -1,0 +1,31 @@
+#pragma once
+#include <fstream>
+#include "ChainComponent.hpp"
+#include "Tracer.hpp"
+
+class StandardTracer : public ChainComponent {
+    Tracer model_tracer;
+    Tracer stats_tracer;
+    std::string chain_name;
+
+  public:
+    template <class M>
+    StandardTracer(M& m, std::string chain_name)
+        : model_tracer(m), stats_tracer(m, processing::HasTag<Stat>()), chain_name(chain_name) {}
+
+    void start() final {
+        std::ofstream model_os{chain_file(chain_name), std::ios_base::trunc};
+        model_tracer.write_header(model_os);
+        std::ofstream stats_os{chain_name + ".trace", std::ios_base::trunc};
+        stats_tracer.write_header(stats_os);
+    }
+
+    void savepoint(int) final {
+        std::ofstream model_os{chain_file(chain_name), std::ios_base::app};
+        model_tracer.write_line(model_os);
+        std::ofstream trace_os{chain_name + ".trace", std::ios_base::app};
+        stats_tracer.write_line(trace_os);
+    }
+
+    static std::string chain_file(std::string chain_name) { return chain_name + ".chain"; }
+};

### 2024-07-05 utils/BayesCode/src/components/test.cpp

@@ -1,0 +1,116 @@
+#include "doctest.h"
+
+#include "BaseArgParse.hpp"
+#include "Tracer.hpp"
+
+using namespace std;
+
+struct TracerDummyModel {
+    double a, b;
+    vector<double> v;
+
+    template <class Info>
+    void declare_interface(Info info) {
+        declare(info, "a", a);
+        declare(info, "v", v);
+        declare(info, "b", b);
+    }
+};
+
+TEST_CASE("Tracer reading test") {
+    stringstream ss("1.1 2.2 3.3 4.4 5.5");
+    TracerDummyModel m = {-1, -1, {-1, -1, -1}};
+    Tracer t(m, processing::True());
+
+    t.read_line(ss);
+
+    CHECK(m.a == 1.1);
+    CHECK(m.b == 5.5);
+    vector<double> expected{2.2, 3.3, 4.4};
+    CHECK(m.v == expected);
+}
+
+TEST_CASE("Tracer writing test") {
+    stringstream ss;
+    TracerDummyModel m = {1.1, 2.2, {3.3, 4.4, 5.5}};
+    Tracer t(m, processing::True());
+
+    t.write_header(ss);
+    t.write_line(ss);
+    auto old_v_addr = &(m.v[0]);
+    m.a = 1.11;
+    m.b = 2.21;
+    m.v[0] = 3.31;
+    m.v[1] = 4.41;
+    m.v[2] = 5.51;
+    CHECK(old_v_addr == &(m.v[0]));
+    CHECK(m.v[0] == 3.31);
+    CHECK(m.v[1] == 4.41);
+    CHECK(m.v[2] == 5.51);
+    t.write_line(ss);
+
+    CHECK(ss.str() ==
+          "a\tv[0]\tv[1]\tv[2]\tb\n1.1\t3.3\t4.4\t5.5\t2.2\n1.11\t3.31\t4.41\t5.51\t2.21");
+}
+
+struct MyTracerData {
+    int x, y;
+    template <class Info>
+    void declare_interface(Info info) {
+        declare(info, "x", x);
+        declare(info, "y", y);
+    }
+};
+
+struct MyTracerStruct {
+    int a{17};
+    vector<int> b{2, 3, 4, 5};
+    vector<MyTracerData> c{{1, 2}, {2, 3}};
+    template <class Info>
+    void declare_interface(Info info) {
+        declare(info, "a", a);
+        declare(info, "b", b);
[…]

### 2024-07-05 utils/BayesCode/src/global/logging.hpp

@@ -1,0 +1,60 @@
+#pragma once
+
+#include <iostream>
+#include "spdlog/spdlog.h"
+#include "spdlog/fmt/fmt.h"
+#include "spdlog/sinks/basic_file_sink.h"
+#include "spdlog/sinks/stdout_color_sinks.h"
+
+using std::string;
+
+// Type for logger pointers
+using logger_t = std::shared_ptr<spdlog::logger>;
+
+// Global variables to simplify formatting
+const std::vector<int> color_codes{33, 34, 35, 36, 91, 92, 93, 94, 95, 96};
+const string dim_code{"\e[0m\e[2m"}, bold_code{"\e[0m\e[1m"}, normal_code{"\e[0m"};
+const string colored_pattern_prefix{"[" + dim_code + "%Y-%m-%d %H:%M:%S.%e" + normal_code + "] [" +
+                                    dim_code + "%P" + normal_code + "] [" + dim_code + "%n" +
+                                    normal_code + "] [" + dim_code + "%^%l%$" + normal_code + "] "};
+
+// Factory functions to create loggers
+inline logger_t stdout_logger(const string& name) {
+    auto result = spdlog::get(name);
+    if (result == nullptr) {
+        result = spdlog::stdout_color_mt(name);
+        result->set_pattern(colored_pattern_prefix + "%v");
+    }
+    return result;
+}
+
+inline logger_t file_logger(const string& name, const string& filename) {
+    auto result = spdlog::get(name);
+    if (result == nullptr) {
+        result = spdlog::basic_logger_mt(name, filename);
+        result->set_pattern("[%Y-%m-%d %H:%M:%S.%e] [%P] [%n] [%^%l%$] %v");
+    }
+    return result;
+}
+
+inline logger_t global_logger() {
+    auto result = spdlog::get("global");
+    if (result == nullptr) {
+        result = spdlog::stdout_color_mt("global");
+        result->set_pattern(colored_pattern_prefix + "[" + dim_code + "%@" + normal_code + "] " +
+                            normal_code + "%v");
+    }
+    return result;
+}
+
+// Macros for global logging
+#define INFO(...) SPDLOG_LOGGER_INFO(global_logger(), __VA_ARGS__)
+#define WARNING(...) SPDLOG_LOGGER_WARN(global_logger(), __VA_ARGS__)
[…]

### 2024-07-05 utils/BayesCode/src/lib/AADiffSelCodonMatrixBidimArray.hpp

@@ -1,0 +1,101 @@
+
+#ifndef AADIFFSELCODONMATRIXARRAY_H
+#define AADIFFSELCODONMATRIXARRAY_H
+
+#include "AAMutSelOmegaCodonSubMatrix.hpp"
+#include "BidimArray.hpp"
+
+/**
+ * \brief A BidimArray of AAMutSelOmegaCodonSubMatrix (used in DiffSelModel and
+ * DiffSelSparseModel)
+ *
+ * The constructor takes as arguments a single nucleotide matrix and a bidim
+ * array of amino-acid fitness profiles (across sites and conditions). It then
+ * constructs an array of AAMutSelOmegaCodonSubMatrix of same size as the array
+ * of fitness profiles.
+ */
+
+class AADiffSelCodonMatrixBidimArray : public BidimArray<SubMatrix>,
+                                       public BidimArray<AAMutSelOmegaCodonSubMatrix> {
+  public:
+    // two possible ways to do it:
+    // AAMutSelNeCodonMatrixBidimArray(const Selector<vector<double> >&
+    // infitnessarray, const Selector<double>& inNe, const CodonStateSpace&
+    // incodonstatespace, const SubMatrix& innucmatrix);
+    // AAMutSelNeCodonMatrixBidimArray(const Selector<vector<double> >&
+    // infitnessarray, const vector<double>& inNe, const CodonStateSpace&
+    // incodonstatespace, const SubMatrix& innucmatrix);
+
+    //! constructor parameterized by a bidim array of fitness profiles, a codon
+    //! state space and a single nucleotide matrix.
+    AADiffSelCodonMatrixBidimArray(const BidimSelector<std::vector<double>> &infitnessarray,
+        const CodonStateSpace &incodonstatespace, const SubMatrix &innucmatrix)
+        : fitnessarray(infitnessarray),
+          codonstatespace(incodonstatespace),
+          nucmatrix(innucmatrix),
+          matrixarray(infitnessarray.GetNrow(),
+              std::vector<AAMutSelOmegaCodonSubMatrix *>(
+                  infitnessarray.GetNcol(), (AAMutSelOmegaCodonSubMatrix *)0)) {
+        Create();
+    }
+
+    ~AADiffSelCodonMatrixBidimArray() { Delete(); }
+
+    //! return the number of rows (number of conditions)
+    virtual int GetNrow() const override { return fitnessarray.GetNrow(); }
+    //! return the number of columns (number of sites)
[…]

### 2024-07-05 utils/BayesCode/src/lib/AAMutSelOmegaCodonSubMatrix.hpp

@@ -1,0 +1,79 @@
+#pragma once
+
+#include "CodonSubMatrix.hpp"
+#include <cassert>
+
+/**
+ * \brief A mutation-selection codon substitution process.
+ *
+ * This codon substitution process describes the evolution of a coding position,
+ * under a constant fitness landscape over the 20 amino-acids. The model is
+ * parameterized by a nucleotide matrix, specifying the mutation process, and a
+ * vector of 20 scaled fitness parameters (summing to 1), for the 20
+ * amino-acids. The process also takes a real parameter, omega, which acts as a
+ * multiplier in front of all non-synonymous substitutions. The standard
+ * mutation-selection process is obtained by setting omega=1. Letting omega be
+ * different from 1 was explored in Rodrigue and Lartillot, 2017, MBE (detecting
+ * deviations from expected non-syn rate under the standard mut-sel model).
+ */
+
+class AAMutSelOmegaCodonSubMatrix : public virtual NucCodonSubMatrix,
+                                    public virtual OmegaCodonSubMatrix {
+  public:
+    //! constructor, parameterized by a codon state space (genetic code), a
+    //! nucleotide mutation matrix, a 20-vector of amino-acid fitnesss, and a
+    //! positive real parameter omega (=1 in the standard model).
+    AAMutSelOmegaCodonSubMatrix(const CodonStateSpace *instatespace, const SubMatrix *inNucMatrix,
+        const std::vector<double> &inaa, double inomega, double inNe, bool innormalise = false)
+        : SubMatrix(instatespace->GetNstate(), innormalise),
+          CodonSubMatrix(instatespace, innormalise),
+          NucCodonSubMatrix(instatespace, inNucMatrix, innormalise),
+          OmegaCodonSubMatrix(instatespace, inomega, innormalise),
+          fitnesses(inaa.size(), 0.0),
+          logfitnesses(inaa.size(), 1.0 / inaa.size()),
+          aa(inaa),
+          Ne(inNe) {}
+
+    //! \brief access by copy to fitness of a given amino-acid
+    //!
+    //! Note: to avoid numerical errors, this function adds 1e-8.
+    double GetFitness(int a) const {
+        assert(std::abs((exp(Ne * log(aa[a])) + 1e-8) - fitnesses[a]) < 1e-6);
+        return fitnesses[a];
+    }
+
+    double GetLogFitness(int a) const {
[…]

### 2024-07-05 utils/BayesCode/src/lib/BranchOmegaModel.hpp

@@ -1,0 +1,512 @@
+#include "BranchArray.hpp"
+#include "BranchProduct.hpp"
+#include "BranchSpecificMeanGammaTree.hpp"
+#include "CodonSequenceAlignment.hpp"
+#include "CodonSubMatrixBranchArray.hpp"
+#include "CodonSuffStat.hpp"
+#include "GTRSubMatrix.hpp"
+#include "GammaSuffStat.hpp"
+#include "IIDGamma.hpp"
+#include "PhyloProcess.hpp"
+#include "ProbModel.hpp"
+#include "Tree.hpp"
+
+/**
+ * \brief A site-homogeneous and branch-heterogeneous Muse and Gaut omega-codon
+ * model
+ *
+ * The model has the following structure:
+ * - branch lengths iid Exponential of rate lambda
+ * - nucleotide relative exchangeabilities and stationaries are uniform
+ * Dirichlet
+ * - over branch j, omega_j = w * v_j, with v_j ~ Gamma(branchmean,
+ * branchinvshape)
+ * - w ~ Gamma(sitemean,siteinvshape)
+ *
+ * when model used in isolation: w = 1, and the v_j's are estimated, along with
+ * their hyperparameters branchmean and branchinvshape. On the other hand, in a
+ * multi gene context, the v_j's are shared across genes, and w_i for gene i is
+ * allowed to vary (and then all 4 hyperparameters are estimated across genes).
+ */
+
+class BranchOmegaModel : public ProbModel {
+    // tree and data
+    Tree *tree;
+    FileSequenceAlignment *data;
+    const TaxonSet *taxonset;
+    CodonSequenceAlignment *codondata;
+
+    int Nsite;
+    int Ntaxa;
+    int Nbranch;
+
+    // branch lengths iid expo (gamma of shape 1 and scale lambda)
+    // where lambda is a hyperparameter
+    double lambda;
+    BranchIIDGamma *branchlength;
+
+    // nucleotide exchange rates and equilibrium frequencies (stationary
+    // probabilities)
+    std::vector<double> nucrelrate;
+    std::vector<double> nucstat;
+
+    // all this is fixed, being controlled at the inter-gene level
+    double branchvhypermean;
+    double branchvhyperinvshape;
+    BranchIIDGamma *branchv;
+    double genew;
+    BranchProduct *meanomegatree;
+    double omegainvshape;
+
+    // branch-specific omega's
[…]

### 2024-07-05 utils/BayesCode/src/lib/BranchSitePath.cpp

@@ -1,0 +1,256 @@
+#include <cmath>
+#include <sstream>
+
+#include "BranchSitePath.hpp"
+
+#include "PathSuffStat.hpp"
+#include "PoissonSuffStat.hpp"
+#include "SubMatrix.hpp"
+#include "SuffStat.hpp"
+
+using namespace std;
+
+//-------------------------------------------------------------------------
+//	* BranchSitePath
+//-------------------------------------------------------------------------
+
+BranchSitePath::BranchSitePath() {
+    init = last = new Plink;
+    // bkinit = bklast = new Plink;
+    nsub = 0;
+}
+
+BranchSitePath::BranchSitePath(int state) {
+    init = last = new Plink;
+    // bkinit = bklast = new Plink;
+    nsub = 0;
+    init->SetState(state);
+    init->SetRelativeTime(0);
+}
+
+BranchSitePath::~BranchSitePath() {
+    Reset(0);
+    delete init;
+    // delete bkinit;
+}
+
+void BranchSitePath::Reset(int state) {
+    Plink *link = last;
+    while (link != init) {
+        Plink *prev = link->Prev();
+        delete link;
+        link = prev;
+    }
+    nsub = 0;
+    init->SetState(state);
+    init->SetRelativeTime(0);
+    last = init;
+}
+
+/*
+void BranchSitePath::BKReset(int state) {
+    Plink *link = bklast;
+    while (link != bkinit) {
+        Plink *prev = link->Prev();
+        delete link;
+        link = prev;
+    }
+    bknsub = 0;
+    bkinit->SetState(state);
+    bkinit->SetRelativeTime(0);
+    bklast = bkinit;
+}
+
+void BranchSitePath::BackupPath() {
+    BKReset(init->GetState());
+    Plink *link = init;
+    while (link != last) {
+        BKAppend(link->Next()->GetState(), link->GetRelativeTime());
+        link = link->Next();
+    }
+    bklast->SetRelativeTime(last->GetRelativeTime());
+    bknsub = nsub;
+}
+
+void BranchSitePath::RestorePath() {
+    Reset(bkinit->GetState());
+    Plink *link = bkinit;
+    while (link != bklast) {
+        Append(link->Next()->GetState(), link->GetRelativeTime());
+        link = link->Next();
+    }
+    last->SetRelativeTime(bklast->GetRelativeTime());
+    nsub = bknsub;
+}
+
+string BranchSitePath::ToString(bool redundant) {
[…]

### 2024-07-05 utils/BayesCode/src/lib/BranchSitePath.hpp

@@ -1,0 +1,243 @@
+#pragma once
+
+#include <map>
+#include <string>
+
+class SuffStat;
+class PoissonSuffStat;
+class PathSuffStat;
+class PolySuffStat;
+class PathSuffStatBidimArray;
+class SubMatrix;
+
+/**
+ * \brief The building block for substitution histories, or paths
+ * (BranchSitePath class)
+ *
+ * A substitution history over some period of time, and with n substitution
+ * events in total, is encoded as doubly-linked chains of n+1 Plink objects (see
+ * BranchSitePath for an example). Each Plink encodes the current state and the
+ * relative waiting time (relative to total time t) until either the next event
+ * or the endpoint. Each Plink points to the previous Plink (Plink* prev) or to
+ * the next (Plink* next). The first Plink has prev=null and the last Pling has
+ * next=null;
+ *
+ * Thus, for instance
+ * the following history, over a total time of t=10 units: A---C--T-----
+ * (starting from A, waiting 3 time units, then making a substitution toward C,
+ * then waiting 2 time units, making a substitution toward T, then waiting 10
+ * time units and then stopping) would be encoded by a chain of three Plink
+ * objects: (A,0.3) -> (C,0.2) -> (T,0.5).
+ */
+
+class Plink {
+    friend class BranchSitePath;
+
+  public:
+    //! default constructor (empty)
+    Plink();
+
+    //! constructor specifying the current state and the relative time until next
+    //! event
+    Plink(int instate, double inrel_time);
+    ~Plink();
+
+    Plink *Prev();
+    Plink *Next();
+
+    const Plink *Prev() const;
+    const Plink *Next() const;
+
+    bool IsFirst() const;
+    bool IsLast() const;
+
+    void Splice();
+    void Insert(Plink *link);
+
+    void SetState(int instate);
+    int GetState() const;
+
+    void SetRelativeTime(double inrel_time);
+    double GetRelativeTime() const;
+
+  private:
+    Plink *next;
+    Plink *prev;
+
+    int state;
+    double rel_time;
+};
+
+/**
+ * \brief A doubly-linked structure specifying the detailed substitution history
+ * over a branch, for a given site
+ *
+ * A substitution history over some period of time, and with n substitution
+ * events in total, is encoded as doubly-linked chains of n+1 Plink objects.
+ * Thus, for instance
+ * the following history, over a total time of t=10 units: A---C--T-----
[…]

### 2024-07-05 utils/BayesCode/src/lib/Chi2.cpp

@@ -1,0 +1,38 @@
+
+#include "IncompleteGamma.hpp"
+#include "Random.hpp"
+
+int main(int argc, char *argv[]) {
+    if (argc == 1) {
+        cerr << "chi2 infile df outfile\n";
+        cerr << "\tdf = 1 or 2: number of degrees of freedom for simple chi2\n";
+        cerr << "\tdf = -1: mixture of 50% point mass 50% chi2 df = 1\n";
+        cerr << "\tdf = -2: mixture of 25% point mass, 50% chi2 df=1, 25% chi2 "
+                "df=2\n";
+        exit(1);
+    }
+
+    // 1,2 : 1 or 2 degrees of freedom
+    // -1 : mix of 50% 0 50% 1
+    // -2: mix of 25% 0 50% 1 25% 2
+    ifstream is(argv[1]);
+    int df = atoi(argv[2]);
+    ofstream os(argv[3]);
+
+    int Ngene;
+    is >> Ngene;
+    for (int i = 0; i < Ngene; i++) {
+        string name;
+        double x;
+        is >> name >> x;
+        x *= 2;
+        if (x <= 0) {
+            os << name << '\t' << x << '\t' << 1 << '\t' << "100" << '\n';
+        } else {
+            double p = 0.25 * Chi2Pval(x, df);
+            double q = (int)(100 * Ngene * p / (i + 1));
+            if (q > 100) { q = 100; }
+            os << name << '\t' << x << '\t' << p << '\t' << q << '\n';
+        }
+    }
+}

### 2024-07-05 utils/BayesCode/src/lib/CodonStateSpace.cpp

@@ -1,0 +1,234 @@
+#include "CodonStateSpace.hpp"
+#include <cstdlib>
+#include <iostream>
+#include <sstream>
+#include "Exception.hpp"
+using namespace std;
+
+CodonStateSpace::CodonStateSpace(GeneticCodeType type) {
+    nucstatespace = new DNAStateSpace;
+    protstatespace = new ProteinStateSpace;
+
+    code = type;
+    if (code == Universal) {
+        Nstate = Ncodon - UniNStopCodons;
+
+        CodonCodeWithStops = new int[Ncodon];  // stops included
+        CodonCode = new int[Nstate];           // stops excluded
+
+        CodonPos = new int *[Npos];
+        for (int pos = 0; pos < Npos; pos++) {
+            CodonPos[pos] = new int[Nstate];  // stops excluded
+        }
+
+        int k = 0;
+        for (int i = 0; i < Ncodon; i++) {
+            CodonCodeWithStops[i] = UniCodonCode[i];
+            if (CodonCodeWithStops[i] != -1) {
+                CodonCode[k] = CodonCodeWithStops[i];
+                for (int pos = 0; pos < Npos; pos++) { CodonPos[pos][k] = codonpos[pos][i]; }
+                k++;
+            }
+        }
+
+        Nstop = UniNStopCodons;
+        StopCodons = new int[Nstop];
+        StopPos1 = new int[Nstop];
+        StopPos2 = new int[Nstop];
+        StopPos3 = new int[Nstop];
+        for (int i = 0; i < Nstop; i++) {
+            StopCodons[i] = UniStopCodons[i];
+            StopPos1[i] = UniStopPos1[i];
+            StopPos2[i] = UniStopPos2[i];
+            StopPos3[i] = UniStopPos3[i];
+        }
+
+    } else if (code == MtInv) {
+        Nstate = Ncodon - MtInvNStopCodons;
+
+        CodonCodeWithStops = new int[Ncodon];  // stops included
+        CodonCode = new int[Nstate];           // stops excluded
+
+        CodonPos = new int *[Npos];
+        for (int pos = 0; pos < Npos; pos++) {
+            CodonPos[pos] = new int[Nstate];  // stops excluded
+        }
+
+        int k = 0;
+        for (int i = 0; i < Ncodon; i++) {
+            CodonCodeWithStops[i] = MtInvCodonCode[i];
[…]

### 2024-07-05 utils/BayesCode/src/lib/CodonSubMatrix.cpp

@@ -1,0 +1,48 @@
+#include "CodonSubMatrix.hpp"
+using namespace std;
+
+void MGCodonSubMatrix::ComputeArray(int i) const {
+    double total = 0;
+    for (auto j : statespace->GetNeighbors(i)) {
+        int pos = GetDifferingPosition(i, j);
+        int a = GetCodonPosition(pos, i);
+        int b = GetCodonPosition(pos, j);
+
+        assert(a != b);
+        Q(i, j) = (*NucMatrix)(a, b);
+        total += Q(i, j);
+    }
+    Q(i, i) = -total;
+    assert(total >= 0);
+}
+
+void MGCodonSubMatrix::ComputeStationary() const {
+    // compute stationary probabilities
+    double total = 0;
+    for (int i = 0; i < GetNstate(); i++) {
+        mStationary[i] = NucMatrix->Stationary(GetCodonPosition(0, i)) *
+                         NucMatrix->Stationary(GetCodonPosition(1, i)) *
+                         NucMatrix->Stationary(GetCodonPosition(2, i));
+        total += mStationary[i];
+    }
+
+    // renormalize stationary probabilities
+    for (int i = 0; i < GetNstate(); i++) { mStationary[i] /= total; }
+}
+
+void MGOmegaCodonSubMatrix::ComputeArray(int i) const {
+    double total = 0;
+    for (auto j : statespace->GetNeighbors(i)) {
+        int pos = GetDifferingPosition(i, j);
+        int a = GetCodonPosition(pos, i);
+        int b = GetCodonPosition(pos, j);
+
+        assert(a != b);
+        Q(i, j) = (*NucMatrix)(a, b);
+        if (!Synonymous(i, j)) { Q(i, j) *= GetOmega(); }
+
+        total += Q(i, j);
+    }
+    Q(i, i) = -total;
+    assert(total >= 0);
+}

### 2024-07-05 utils/BayesCode/src/lib/CodonSubMatrix.hpp

@@ -1,0 +1,134 @@
+#pragma once
+
+#include "CodonStateSpace.hpp"
+#include "SubMatrix.hpp"
+
+const double omegamin = 1e-5;
+
+/**
+ * \brief A general class representing all codon matrices
+ *
+ * Provides a few generic methods for dealing with codon-specific issues (syn,
+ * non syn, etc). However, still needs to be specialized into actual classes
+ * (the two most important being MGOmegaCodonSubMatrix and
+ * AAMutSelOmegaCodonSubMatrix)
+ */
+
+class CodonSubMatrix : public virtual SubMatrix {
+  public:
+    //! constructor parameterized by codon state space (itself specifying the
+    //! genetic code)
+    CodonSubMatrix(const CodonStateSpace *instatespace, bool innormalise)
+        : SubMatrix(instatespace->GetNstate(), innormalise), statespace(instatespace) {}
+
+    const CodonStateSpace *GetCodonStateSpace() const { return statespace; }
+
+    //! see CodonStateSpace::Synonymous
+    bool Synonymous(int codon1, int codon2) const { return statespace->Synonymous(codon1, codon2); }
+    //! see CodonStateSpace::GetCodonPosition
+    int GetCodonPosition(int pos, int codon) const {
+        return statespace->GetCodonPosition(pos, codon);
+    }
+    //! see CodonStateSpace::GetDifferingPosition
+    int GetDifferingPosition(int codon1, int codon2) const {
+        return statespace->GetDifferingPosition(codon1, codon2);
+    }
+
+  protected:
+    const CodonStateSpace *statespace;
+};
+
+/**
+ * \brief A generic class for codon matrices based on an underlying nucleotide
+ * rate matrix
+ *
+ * Most codon matrices rely on a mutation process at the level of nucleotides.
+ * NucCodonSubMatrix takes a const pointer to a nucleotide substitution matrix
+ * representing the nucleotide 4x4 mutation process.
+ * Still an abstract class.
+ */
+class NucCodonSubMatrix : public virtual CodonSubMatrix {
+  public:
+    NucCodonSubMatrix(
+        const CodonStateSpace *instatespace, const SubMatrix *inNucMatrix, bool innormalise)
+        : SubMatrix(instatespace->GetNstate(), innormalise),
+          CodonSubMatrix(instatespace, innormalise) {
+        SetNucMatrix(inNucMatrix);
+    }
+
[…]

### 2024-07-05 utils/BayesCode/src/lib/CodonSubMatrixArray.hpp

@@ -1,0 +1,213 @@
+#pragma once
+
+#include "AAMutSelOmegaCodonSubMatrix.hpp"
+#include "Array.hpp"
+#include "CodonSubMatrix.hpp"
+#include "SubMatrix.hpp"
+
+/**
+ * \brief An Array of MGOmegaCodonSubMatrix
+ *
+ * The constructor takes as arguments a single nucleotide matrix and a
+ * Selector<double> which returns the value of omega for each site. It then
+ * constructs an array of MGOmegaCodonSubMatrix of same size as the omega array,
+ * and such that the ith matrix takes the ith value of the omega array as its
+ * dN/dS.
+ *
+ * This array derives from Array and not SimpleArray, because it is implemented
+ * as a vector<MGOmegaCodonSubMatrix*>.
+ */
+
+class MGOmegaCodonSubMatrixArray : public Array<SubMatrix>, public Array<MGOmegaCodonSubMatrix> {
+  public:
+    //! constructor parameterized by a codon state space, a single nucleotide
+    //! matrix and an array (in fact, a Selector) of omega's
+    MGOmegaCodonSubMatrixArray(const CodonStateSpace *incodonstatespace,
+        const SubMatrix *innucmatrix, const Selector<double> *inomegaarray)
+        : codonstatespace(incodonstatespace),
+          nucmatrix(innucmatrix),
+          omegaarray(inomegaarray),
+          matrixarray(inomegaarray->GetSize()) {
+        Create();
+    }
+
+    ~MGOmegaCodonSubMatrixArray() { Delete(); }
+
+    //! return array size
+    int GetSize() const { return omegaarray->GetSize(); }
+    //! const access to matrix i
+    const MGOmegaCodonSubMatrix &GetVal(int i) const { return *matrixarray[i]; }
+    //! non-const access to matrix i
+    MGOmegaCodonSubMatrix &operator[](int i) { return *matrixarray[i]; }
+
+    //! const access to underlying nucleotide matrix
+    const SubMatrix &GetNucMatrix() const { return *nucmatrix; }
+
+    //! update all matrices
+    void UpdateCodonMatrices() {
+        for (int i = 0; i < GetSize(); i++) {
+            (*this)[i].SetOmega(omegaarray->GetVal(i));
+            (*this)[i].CorruptMatrix();
+        }
+    }
+
+    //! update only those matrices for which occupancy[i] != 0
+    void UpdateCodonMatrices(const Selector<int> &occupancy) {
+        if (occupancy.GetSize() != GetSize()) {
[…]

### 2024-07-05 utils/BayesCode/src/lib/GTRSuffStat.hpp

@@ -1,0 +1,239 @@
+
+#ifndef GTRSUFFSTAT_H
+#define GTRSUFFSTAT_H
+
+#include "AASubSelSubMatrix.hpp"
+#include "AASubSelSubMatrixArray.hpp"
+#include "GTRSubMatrix.hpp"
+#include "GTRSubMatrixArray.hpp"
+#include "PathSuffStat.hpp"
+
+class RelRateSuffStat : public SuffStat {
+  public:
+    RelRateSuffStat(int innstate) : nstate(innstate) {
+        nrr = innstate * (innstate - 1) / 2;
+        rrcount.assign(nrr, 0);
+        rrbeta.assign(nrr, 0);
+    }
+
+    ~RelRateSuffStat() {}
+
+    int GetCount(int i) const { return rrcount[i]; }
+
+    double GetBeta(int i) const { return rrbeta[i]; }
+
+    void Clear() {
+        for (int i = 0; i < nrr; i++) {
+            rrcount[i] = 0;
+            rrbeta[i] = 0;
+        }
+    }
+
+    void AddSuffStat(
+        const GTRSubMatrix &matrix, const PathSuffStat &pathsuffstat, double rate = 1) {
+        const std::map<int, double> &waitingtime = pathsuffstat.GetWaitingTimeMap();
+        for (std::map<int, double>::const_iterator i = waitingtime.begin(); i != waitingtime.end();
+             i++) {
+            int a = i->first;
+            for (int b = 0; b < nstate; b++) {
+                if (b != a) { rrbeta[rrindex(a, b)] += matrix.Stationary(b) * i->second * rate; }
+            }
+        }
+
+        const std::map<pair<int, int>, int> &paircount = pathsuffstat.GetPairCountMap();
+        for (std::map<pair<int, int>, int>::const_iterator i = paircount.begin();
+             i != paircount.end(); i++) {
+            rrcount[rrindex(i->first.first, i->first.second)] += i->second;
+        }
+    }
+
+    void AddSuffStat(
+        const GTRSubMatrixArray &matrixarray, const PathSuffStatArray &pathsuffstatarray) {
+        for (int i = 0; i < matrixarray.GetSize(); i++) {
+            AddSuffStat(matrixarray.GetVal(i), pathsuffstatarray.GetVal(i));
+        }
+    }
+
+    void AddSuffStat(const GTRSubMatrixArray &matrixarray,
[…]

### 2024-07-05 utils/BayesCode/src/lib/IIDDirichlet.hpp

@@ -1,0 +1,387 @@
+#pragma once
+
+#include "Array.hpp"
+#include "Random.hpp"
+#include "SuffStat.hpp"
+
+/**
+ * \brief A sufficient statistic for a collection of Dirichlet random variables,
+ * as a function of the parameters of the Dirichlet distribution
+ *
+ * Consider a Dirichlet distribution, over vectors of dimension K, and with
+ * parameter alpha = (alpha_k)_k=1..K. Consider N Dirichlet random variables
+ * (x_i)_i=1..N ~ Dirichlet(alpha). Then, up to an additive constant (not
+ * depending on alpha):
+ *
+ * sum_i log p(x_i | alpha) = N (log Gamma(|alpha|) - sum_k log Gamma(alpha_k))
+ * + sum_k S_k ^ (alpha_k-1)
+ *
+ * where S_k = sum_i log(x_ik).
+ */
+
+class DirichletSuffStat : public SuffStat {
+  public:
+    DirichletSuffStat(int indim) : sumlog(indim, 0), n(0) {}
+    ~DirichletSuffStat() {}
+
+    //! set suff stats to 0
+    void Clear() {
+        for (unsigned int i = 0; i < sumlog.size(); i++) { sumlog[i] = 0; }
+        n = 0;
+    }
+
+    //! get dimension of underlying Dirichlet distribution
+    int GetDim() { return (int)sumlog.size(); }
+
+    //! add the contribution of one variate (x) to this suffstat
+    void AddSuffStat(const std::vector<double> &pi) {
+        for (unsigned int i = 0; i < sumlog.size(); i++) {
+            if (pi[i] <= 0) {
+                std::cerr << "error: negative pi in DirichletSuffStat: " << pi[i] << '\n';
+                for (unsigned int j = 0; j < sumlog.size(); j++) { std::cerr << pi[j] << '\t'; }
+                std::cerr << '\n';
+                exit(1);
+            }
+            sumlog[i] += log(pi[i]);
+        }
+        n++;
+    }
+
+    //! add the contribution of one variate (x) to this suffstat
+    void AddSuffStat(const double *insumlog, int d) {
+        for (unsigned int i = 0; i < sumlog.size(); i++) { sumlog[i] += insumlog[i]; }
+        n += d;
+    }
+
+    //! (*this) += from
+    void Add(const DirichletSuffStat &from) {
[…]

### 2024-07-05 utils/BayesCode/src/lib/IIDProfileMask.hpp

@@ -1,0 +1,126 @@
+#pragma once
+
+#include <cmath>
+#include "Array.hpp"
+
+/**
+ * \brief An array of IID 0/1 masks of a fixed dimension
+ *
+ * This class is used in AAMutSelSparseOmegaModel and DiffSelDoublySparseModel.
+ * In those two cases, the masks are over the 20 amino-acids, and the array if
+ * of size Nsite (i.e. the array implements site-specific masks over the
+ * alignment). The masks are meant to define a low/high fitness distribution
+ * over the 20 amino-acids.
+ *
+ * Each mask is made of dim iid Bernoulli(pi), conditional on at least one entry
+ * of the mask being equal to 1.
+ */
+
+class IIDProfileMask : public SimpleArray<std::vector<int>> {
+  public:
+    //! constructor, parameterized by array size, mask dimension and Bernoulli
+    //! probability parameter
+    IIDProfileMask(int size, int indim, double pi)
+        : SimpleArray(size, std::vector<int>(indim, 1)), dim(indim), pi(pi) {}
+
+    //! return dimension of the masks
+    int GetDim() const { return dim; }
+
+    //! set probability parameter of the Bernoulli to a new value
+    void SetPi(double inpi) { pi = inpi; }
+
+    //! return total log probability over entire array
+    double GetLogProb() const {
+        double total = 0;
+        for (int i = 0; i < GetSize(); i++) { total += GetLogProb(i); }
+        return total;
+    }
+
+    //! return log probability for entry i
+    double GetLogProb(int i) const {
+        int naa = 0;
+        const std::vector<int> &x = GetVal(i);
+        for (int k = 0; k < GetDim(); k++) { naa += x[k]; }
+        if (!naa) {
+            std::cerr << "error in IIDProfileMask: all entries are null\n";
+            exit(1);
+        }
+        // probability is conditional on at least one entry being 1
+        return naa * log(pi) + (GetDim() - naa) * log(1.0 - pi) -
+               log(1.0 - exp(GetDim() * log(1.0 - pi)));
+    }
+
+    //! return mean width (i.e. mean number of entries equal to 1) across the
+    //! array
+    double GetMeanWidth() const {
+        double mean = 0;
+        for (int i = 0; i < GetSize(); i++) {
+            const std::vector<int> &x = GetVal(i);
+            for (int k = 0; k < GetDim(); k++) { mean += x[k]; }
+        }
+        mean /= GetSize();
+        return mean;
+    }
+
+  private:
[…]

### 2024-07-05 utils/BayesCode/src/lib/MultinomialAllocationVector.hpp

@@ -1,0 +1,75 @@
+#ifndef MULTINOMALLOC_H
+#define MULTINOMALLOC_H
+
+#include "Array.hpp"
+#include "OccupancySuffStat.hpp"
+#include "Random.hpp"
+
+/**
+ * \brief A multinomial allocation vector
+ *
+ * MultinomialAllocationVector specifies the probability distribution of the
+ * multinomial allocation of N items to the K components of the finite mixture.
+ * The Distribution is parameterized by a vector of K weight.
+ */
+
+class MultinomialAllocationVector : public SimpleArray<int> {
+  public:
+    //! Constructor (parameterized by number of items (N) and mixture weights
+    //! (vector of dimension K)
+    MultinomialAllocationVector(int insize, const std::vector<double> &inweight)
+        : SimpleArray<int>(insize), weight(inweight) {
+        SampleAlloc();
+    }
+
+    ~MultinomialAllocationVector() {}
+
+    //! sample allocations based on current weights (i.e. from the prior)
+    void SampleAlloc() {
+        for (int i = 0; i < GetSize(); i++) {
+            (*this)[i] = Random::DrawFromDiscreteDistribution(weight);
+        }
+    }
+
+    //! resample allocation of item i based on an externally given vector of
+    //! posterior probabilities
+    void GibbsResample(int i, const std::vector<double> &postprob) {
+        (*this)[i] = Random::DrawFromDiscreteDistribution(postprob);
+    }
+
+    //! resample all allocations based on an externally given array of
+    //! item-specific posterior probabilities
+    void GibbsResample(const std::vector<std::vector<double>> &postprobarray) {
+        for (int i = 0; i < GetSize(); i++) {
+            (*this)[i] = Random::DrawFromDiscreteDistribution(postprobarray[i]);
+        }
+    }
+
+    //! apply permutation to allocation vector
+    void Permute(const Selector<int> &permut) override {
+        if (permut.GetSize() != int(weight.size())) {
+            std::cerr << "error in MultinomialAllocationVector::Permute: non matching "
+                         "array size\n";
+            exit(1);
+        }
+        std::vector<int> invpermut(permut.GetSize(), 0);
+        for (int k = 0; k < permut.GetSize(); k++) { invpermut[permut.GetVal(k)] = k; }
+        for (int i = 0; i < GetSize(); i++) { (*this)[i] = invpermut[(*this)[i]]; }
+    }
+
[…]

### 2024-07-05 utils/BayesCode/src/lib/MultivariateProcess.cpp

@@ -1,0 +1,114 @@
+#include <cassert>
+#include <cmath>
+
+#include "MultivariateProcess.hpp"
+
+NodeMultivariateProcess::NodeMultivariateProcess(
+    const Chronogram &inchrono, const PrecisionMatrix &inprecision_matrix, int indimensions)
+    : SimpleNodeArray<EVector>(inchrono.GetTree()),
+      chronogram(inchrono),
+      dimensions(indimensions),
+      precision_matrix(inprecision_matrix) {
+    for (Tree::NodeIndex node = 0; node < Tree::NodeIndex(GetTree().nb_nodes()); node++) {
+        (*this)[node] = EVector::Zero(dimensions);
+    }
+}
+
+double NodeMultivariateProcess::GetSigma(int dimension) const {
+    return sqrt(1.0 / precision_matrix(dimension, dimension));
+}
+
+double NodeMultivariateProcess::GetLogProb(Tree::NodeIndex node) const {
+    if (GetTree().is_root(node)) {
+        return 0.0;
+    } else {
+        return Random::logNormalDensity(GetContrast(node), precision_matrix);
+    }
+}
+
+EVector NodeMultivariateProcess::GetContrast(Tree::NodeIndex node) const {
+    assert(!GetTree().is_root(node));
+    return (this->GetVal(node) - this->GetVal(GetTree().parent(node))) /
+           sqrt(chronogram.GetVal(GetTree().branch_index(node)));
+}
+
+double NodeMultivariateProcess::GetLocalLogProb(Tree::NodeIndex node) const {
+    double tot = GetLogProb(node);
+    for (auto const &child : GetTree().children(node)) { tot += GetLogProb(child); }
+    return tot;
+}
+
+double NodeMultivariateProcess::GetLogProb() const {
+    double tot = 0;
+    for (Tree::NodeIndex node = 0; node < Tree::NodeIndex(GetTree().nb_nodes()); node++) {
+        if (!GetTree().is_root(node)) { tot += GetLogProb(node); }
+    }
+    return tot;
+}
+
+NodeProcess::NodeProcess(NodeMultivariateProcess &innode_multivariate, int indimension)
+    : dimension(indimension), node_multivariate(innode_multivariate) {}
+
+double NodeProcess::GetSigma() const { return node_multivariate.GetSigma(dimension); }
+
+const Tree &NodeProcess::GetTree() const { return node_multivariate.GetTree(); }
+
[…]

### 2024-07-05 utils/BayesCode/src/lib/Permutation.hpp

@@ -1,0 +1,23 @@
+#ifndef PERM_H
+#define PERM_H
+
+#include "Array.hpp"
+
+/**
+ * \brief A permutation of 1..N
+ */
+
+class Permutation : public SimpleArray<int> {
+  public:
+    //! constructor (parameterized by N)
+    Permutation(int size) : SimpleArray<int>(size) { Reset(); }
+
+    ~Permutation() {}
+
+    //! set equal to identity permuation
+    void Reset() {
+        for (int i = 0; i < GetSize(); i++) { (*this)[i] = i; }
+    }
+};
+
+#endif

### 2024-07-05 utils/BayesCode/src/lib/PoissonSuffStat.hpp

@@ -1,0 +1,280 @@
+#pragma once
+
+#include <cmath>
+#include "Array.hpp"
+#include "BranchArray.hpp"
+#include "PhyloProcess.hpp"
+#include "SuffStat.hpp"
+
+/**
+ * \brief A Poisson-like sufficient statistic
+ *
+ * This sufficient statistic deals with all cases where the probability of the
+ * variable(s) of interest (say X), as a function of some rate parameter r
+ * (positive real number), can be written as P(X | r) \propto r^count
+ * exp(-beta*r), for some suff stats count (integer) and beta (positive real
+ * number). When several independent variables are entailed by the generic X
+ * mentioned above, the suff stats just have to be summed over all items
+ * (separately for count and beta).
+ *
+ * As an example, the probability of the substitution histories over all sites,
+ * for a given branch of the tree, as a function of the branch length l, can be
+ * written as follows: p(S | l) = K l^count exp(-beta*l), where count is the
+ * total number of substitution events along the branch (across all sites), beta
+ * is the average rate away from the current state (averaged over all paths) and
+ * K is some normalization constant (that may depend on other parameters than
+ * l). Thus, conditional on the current substitution histories, the sufficient
+ * statistics count and beta can be first computed, and then MCMC moves on
+ * branch lengths can be done based on the knowledge of these two numbers. Other
+ * examples include the probability of substitution histories as a function of
+ * omega = dN/dS (in that case, the count suff stat is the number of
+ * non-synonymous substitutions only). In these two cases, the suffstats have to
+ * be calculated based on the substitution histories (see
+ * BranchSitePath::AddLengthSuffStat) and then summed over branches, sites, or
+ * any other more subtle pattern, depending on the exact structure of the model.
+ *
+ * PoissonSuffStat implements this general idea:
+ * collecting count and beta suff stats across all relevant items
+ * and providing methods for calculating the probability, as a function of the
+ * rate parameter.
+ */
+
+class PoissonSuffStat : public SuffStat {
+  public:
+    PoissonSuffStat() { Clear(); }
+    ~PoissonSuffStat() {}
+
+    //! set count and beta to 0
+    void Clear() { count = beta = 0; }
+
[…]

### 2024-07-05 utils/BayesCode/src/lib/PolySuffStat.cpp

@@ -1,0 +1,82 @@
+#include "PolySuffStat.hpp"
+
+double PolySuffStat::GetLogProb(PoissonRandomField &poissonrandomfield,
+    const std::vector<double> &aafitnessarray, const GTRSubMatrix &nucmatrix,
+    const double &theta) const {
+    double total = 0;
+    for (auto const &i : polycount) {
+        double proba = poissonrandomfield.GetProb(std::get<0>(i.first), std::get<1>(i.first),
+            std::get<2>(i.first), std::get<3>(i.first), aafitnessarray, nucmatrix, theta);
+        if (proba > 0) {
+            total += i.second * log(proba);
+        } else {
+            total = -std::numeric_limits<double>::infinity();
+        }
+    }
+    return total;
+}
+
+void PolySuffStat::Add(const PolySuffStat &suffstat) {
+    for (auto const &i : suffstat.GetPairCountMap()) { AddPairCount(i.first, i.second); }
+}
+
+int PolySuffStat::GetPairCount(std::tuple<int, int, unsigned, unsigned> poly_tuple) const {
+    auto const i = polycount.find(poly_tuple);
+    if (i == polycount.end()) { return 0; }
+    return i->second;
+}
+
+
+void PolySuffStatArray::Clear() {
+    for (int i = 0; i < GetSize(); i++) { (*this)[i].Clear(); }
+}
+
+double PolySuffStatArray::GetLogProb(PoissonRandomField &poissonrandomfield,
+    const Selector<std::vector<double>> &siteaafitnessarray, const GTRSubMatrix &nucmatrix,
+    const double &theta) const {
+    double total = 0;
+    for (int i = 0; i < GetSize(); i++) {
+        total += GetVal(i).GetLogProb(
+            poissonrandomfield, siteaafitnessarray.GetVal(i), nucmatrix, theta);
+    }
+    return total;
+}
+
+void PolySuffStatArray::Add(
+    const Selector<PolySuffStat> &suffstatarray, const Selector<int> &alloc) {
+    assert(alloc.GetSize() == suffstatarray.GetSize());
+    for (int i = 0; i < suffstatarray.GetSize(); i++) {
+        (*this)[alloc.GetVal(i)] += suffstatarray.GetVal(i);
+    }
+}
+
+void PolySuffStatBidimArray::Clear() {
+    for (int row = 0; row < GetNrow(); row++) {
[…]

### 2024-07-05 utils/BayesCode/src/lib/Random.cpp

@@ -1,0 +1,628 @@
+#include "Random.hpp"
+#include <sys/time.h>
+#include <cmath>
+#include <iostream>
+#include "global/logging.hpp"
+
+/* =======================================================
+   (VL) Magical constants, to be used only in this file.
+   ======================================================= */
+#define MT_IA 397
+#define MT_IB (MT_LEN - MT_IA)
+#define UPPER_MASK 0x80000000
+#define LOWER_MASK 0x7FFFFFFF
+#define MATRIX_A 0x9908B0DF
+#define TWIST(b, i, j) ((b)[i] & UPPER_MASK) | ((b)[j] & LOWER_MASK)
+#define MAGIC(s) (((s)&1) * MATRIX_A)
+
+// #define SAFE_EXP(x) ((x)<-200.0 ? 0.0 : exp(x))
+#define SAFE_EXP(x) exp(x)
+
+using namespace std;
+
+const double gammacoefs[] = {0.9999999999995183, 676.5203681218835, -1259.139216722289,
+    771.3234287757674, -176.6150291498386, 12.50734324009056, -0.1385710331296526,
+    0.9934937113930748e-05, 0.1659470187408462e-06};
+const double Logroot2pi = 0.918938533204673;
+
+// -------------------------------------------------
+// just a trick for random number initialisation
+// function to be called before entering main()
+class random_init {
+  public:
+    random_init() {
+        Random::InitRandom();
+        INFO("Random seed is {}", Random::GetSeed());
+    }
+};
+
+static random_init init;
+
+int Random::Seed = 0;
+int Random::mt_index = 0;
+unsigned long long Random::mt_buffer[MT_LEN];
+
+const double Random::INFPROB = -250;
+
+// ---------------------------------------------------------------------------------
+//		Random()
+// ---------------------------------------------------------------------------------
+void Random::InitRandom(int seed) {
+    if (seed == -1) {
+        struct timeval tod;
+        gettimeofday(&tod, nullptr);
+        seed = tod.tv_usec;
+    }
+    Seed = seed;
+    srand(seed);
+    int i;
+
+    if (RAND_MAX == 32767) {
+        for (i = 0; i < MT_LEN; i++) { mt_buffer[i] = rand() * 32768 + rand(); }
+    } else {
+        for (i = 0; i < MT_LEN; i++) { mt_buffer[i] = rand(); }
+    }
+    mt_index = 0;
+}
+
+Random::Random(int seed) { InitRandom(seed); }
+
[…]

### 2024-07-05 utils/BayesCode/src/lib/ScaledMutationRateCompound.hpp

@@ -1,0 +1,41 @@
+#pragma once
+
+#include "MultivariateProcess.hpp"
+#include "ScaledMutationRate.hpp"
+#include "TaxonMapping.hpp"
+
+/**
+ * \brief A ScaledMutationRate that returns the scaled mutation rate (theta=4*Ne*u) as a compound
+ * parameter given Ne (population size) and u (mutation rate per generation) of a given taxon.
+ *
+ * Useful for implementing models assuming each taxon has its own Ne (population size) and u
+ * (mutation rate per generation). Moreover Ne and u are assumed to follow a brownian process
+ * (NodeProcess).
+ */
+class NodeProcessScaledMutationRate : public ScaledMutationRate {
+  public:
+    //! \brief Constructor, taking as arguments the Nbr of taxa, the scaling factor, the mutation
+    //! rate (a NodeProcess) and the population size (a NodeProcess).
+    NodeProcessScaledMutationRate(double const &intheta_scale, NodeProcess const *innode_popsize,
+        NodeProcess const *innode_rates, int Ntaxa);
+
+    ~NodeProcessScaledMutationRate() override = default;
+
+    void SetTaxonMap(TaxonMap const *intaxon_map) { taxon_map = intaxon_map; }
+
+    double GetTheta(int taxon) const override;
+
+    double GetNodeTheta(Tree::NodeIndex node) const;
+
+    double &operator[](int taxon) { return theta[taxon]; }
+
+    void Update();
+
+  private:
+    double const &theta_scale;
+    NodeProcess const *node_popsize;
+    NodeProcess const *node_rates;
+    std::vector<double> theta;
+
+    TaxonMap const *taxon_map{nullptr};
+};

### 2024-07-05 utils/BayesCode/src/lib/ScatterSuffStat.hpp

@@ -1,0 +1,24 @@
+#pragma once
+
+#include "MultivariateProcess.hpp"
+#include "SuffStat.hpp"
+
+class ScatterSuffStat : public SuffStat {
+  public:
+    //! Constructor (with only the tree given as argument)
+    explicit ScatterSuffStat(const Tree &intree);
+
+    ~ScatterSuffStat() = default;
+
+    void Clear();
+
+    void AddSuffStat(NodeMultivariateProcess const &nodeprocessses);
+
+    void SamplePrecisionMatrix(
+        PrecisionMatrix &precision_matrix, PriorCovariance const &prior_matrix) const;
+
+  protected:
+    const Tree &tree;
+    int dimensions;
+    EMatrix scattermatrix;
+};

### 2024-07-05 utils/BayesCode/src/lib/SequenceAlignment.cpp

@@ -1,0 +1,752 @@
+#include "SequenceAlignment.hpp"
+#include <fstream>
+#include "BiologicalSequences.hpp"
+#include "Random.hpp"
+#include "StringStreamUtils.hpp"
+using namespace std;
+
+// ---------------------------------------------------------------------------
+// ---------------------------------------------------------------------------
+//     SequenceAlignment
+// ---------------------------------------------------------------------------
+// ---------------------------------------------------------------------------
+
+int Int(string s) { return atoi(s.c_str()); }
+
+double Double(string s) { return atof(s.c_str()); }
+
+vector<double> SequenceAlignment::GetEmpiricalFreq() const {
+    vector<double> in(GetNstate(), 0);
+    int n = 0;
+    for (int i = 0; i < GetNtaxa(); i++) {
+        for (int j = 0; j < GetNsite(); j++) {
+            if (GetState(i, j) != unknown) {
+                in[GetState(i, j)]++;
+                n++;
+            }
+        }
+    }
+    for (int i = 0; i < GetNstate(); i++) { in[i] /= n; }
+    return in;
+}
+
+void SequenceAlignment::ToStream(ostream &os) const {
+    os << Ntaxa << '\t' << GetPrintNsite() << '\n';
+    for (int i = 0; i < Ntaxa; i++) {
+        os << taxset->GetTaxon(i) << '\t';
+        for (int j = 0; j < Nsite; j++) { os << statespace->GetState(GetState(i, j)); }
+        os << '\n';
+    }
+}
+
+void SequenceAlignment::ToFasta(ostream &os) const {
+    for (int i = 0; i < Ntaxa; i++) {
+        os << '>' << taxset->GetTaxon(i) << '\n';
+        for (int j = 0; j < GetPrintNsite(); j++) { os << statespace->GetState(GetState(i, j)); }
+        os << '\n';
+    }
+}
+
+int SequenceAlignment::GetNonMissingTriplet() const {
+    int nsite = 0;
+    for (int j = 2; j < Nsite - 3; j += 3) {
+        if (NoMissingColumn(j - 1) && NoMissingColumn(j) && NoMissingColumn(j + 1)) { nsite++; }
+    }
+    return nsite;
+}
+
+void SequenceAlignment::ToStreamTriplet(ostream &os) const {
+    int nsite = 0;
[…]

### 2024-07-05 utils/BayesCode/src/lib/StateSpace.hpp

@@ -1,0 +1,132 @@
+#pragma once
+
+#include <string>
+#include "BiologicalSequences.hpp"  //FIXME only here because constant unknown
+
+/**
+ * \brief Generic interface for a state space (nucleotide, codons, etc)
+ *
+ * In practice, states are encoded as integers ranging from 0 to Nstate-1, where
+ * Nstate=4 for nucleotides, 61 for codons (universal code), etc. This is true
+ * in particular for matrix calculation (SubMatrix), for sequence alignments
+ * (SequenceAlignment). However, the StateSpace class provides the general
+ * interface for converting integer states into strings or conversely.
+ * Specialized versions, such as DNAStateSpace or CodonStateSpace provide
+ * context-specific additional methods.
+ */
+
+class StateSpace {
+  public:
+    virtual ~StateSpace() = default;
+
+    //! return size of state space
+    virtual int GetNstate() const = 0;
+
+    //! return integer for a given string-formatted state
+    virtual int GetState(std::string from) const = 0;
+
+    //! return a formatted string output for given integer-encoded state
+    virtual std::string GetState(int state) const = 0;
+
+    //! whether the two states are compatible (for the moment: just whether or not
+    //! the two states are equal)
+    virtual bool isCompatible(int state1, int state2) const {
+        return ((state1 == unknown) || (state2 == unknown) || (state1 == state2));
+    }
+
+    //! return length of symbol used when printing state (normally, 1 for
+    //! nucleotides or amino-acids, 3 for codons)
+    virtual int GetSymbolLength() const { return 1; }
+};
+
+/**
+ * \brief A StateSpace class built from an array of characters
+ *
+ * This class provides the implementation for all cases where states are
+ * canonically referred to using a one-letter code (thus, nucleotides,
+ * amino-acids, RNA). This excludes the case of codons.
+ */
+
+class SimpleStateSpace : public StateSpace {
+  public:
+    int GetState(std::string from) const override;
+
+    int GetNstate() const override { return Nstate; }
+
+    std::string GetState(int state) const override;
+
+  protected:
+    int Nstate;
+    char *Alphabet;
+    int NAlphabetSet;
+    char *AlphabetSet;
[…]

### 2024-07-05 utils/BayesCode/src/lib/StickBreakingProcess.hpp

@@ -1,0 +1,202 @@
+#ifndef SBP_H
+#define SBP_H
+
+#include "Array.hpp"
+#include "OccupancySuffStat.hpp"
+#include "Permutation.hpp"
+#include "Random.hpp"
+
+/**
+ * \brief A truncated stick-breaking weight vector
+ */
+
+class StickBreakingProcess : public SimpleArray<double> {
+  public:
+    //! constructor, parameterized by array size (truncation upper limit) and
+    //! concentration parameter kappa
+    StickBreakingProcess(int inncat, const double &inkappa)
+        : SimpleArray<double>(inncat), V(inncat), kappa(inkappa) {
+        Sample();
+    }
+
+    ~StickBreakingProcess() {}
+
+    const double &GetKappa() const { return kappa; }
+
+    //! get underlying array of Beta variates
+    std::vector<double> &GetBetaVariates() { return V; }
+    const std::vector<double> &GetBetaVariates() const { return V; }
+
+    //! swap the two components
+    void SwapComponents(int cat1, int cat2) {
+        Swap(cat1, cat2);
+        double tmp = V[cat1];
+        V[cat1] = V[cat2];
+        V[cat2] = tmp;
+    }
+
+    //! Sample (from prior distribution)
+    void Sample() {
+        double cumulProduct = 1.0;
+        double totweight = 0;
+        for (int k = 0; k < GetSize(); k++) {
+            double x = Random::sGamma(1.0);
+            double y = Random::sGamma(kappa);
+            double v = x / (x + y);
+            V[k] = v;
+            if (k == GetSize() - 1) {
+                V[k] = 1;
+                v = 1;
+            }
+            (*this)[k] = v * cumulProduct;
+            cumulProduct *= (1 - v);
+            totweight += (*this)[k];
+        }
+    }
+
+    //! Sample from posterior distribution (given occupany sufficient statistics)
+    void GibbsResample(const OccupancySuffStat &occupancy) {
+        int remainingOcc = 0;
+        for (int i = 0; i < occupancy.GetSize(); i++) { remainingOcc += occupancy.GetVal(i); }
+
+        double cumulProduct = 1.0;
+        double totweight = 0;
+        for (int k = 0; k < GetSize(); k++) {
+            remainingOcc -= occupancy.GetVal(k);
+            double x = Random::sGamma(1 + occupancy.GetVal(k));
+            double y = Random::sGamma(kappa + remainingOcc);
[…]

### 2024-07-05 utils/BayesCode/src/lib/SubMatrix.cpp

@@ -1,0 +1,332 @@
+#include "SubMatrix.hpp"
+#include <cmath>
+#include <cstdlib>
+#include <fstream>
+#include <iostream>
+using namespace std;
+
+int SubMatrix::nuni = 0;
+int SubMatrix::nunimax = 0;
+int SubMatrix::nunisubcount = 0;
+int SubMatrix::diagcount = 0;
+double SubMatrix::diagerr = 0;
+
+double SubMatrix::nz = 0;
+double SubMatrix::meanz = 0;
+double SubMatrix::maxz = 0;
+
+const int witheigen = 1;
+
+// ---------------------------------------------------------------------------
+//     SubMatrix()
+// ---------------------------------------------------------------------------
+SubMatrix::SubMatrix(int inNstate, bool innormalise) : Nstate(inNstate), normalise(innormalise) {
+    ndiagfailed = 0;
+    Create();
+}
+
+void SubMatrix::Create() {
+    Q = EMatrix::Zero(Nstate, Nstate);
+    u = EMatrix(Nstate, Nstate);
+    invu = EMatrix(Nstate, Nstate);
+    v = EVector(Nstate);
+    vi = EVector(Nstate);
+    mStationary = EVector(Nstate);
+
+    ptrQ = nullptr;
+    ptru = nullptr;
+    ptrinvu = nullptr;
+    ptrv = nullptr;
+    ptrStationary = nullptr;
+
+    if (!witheigen) {
+        ptrQ = new double *[Nstate];
+        for (int i = 0; i < Nstate; i++) { ptrQ[i] = new double[Nstate]; }
+
+        ptru = new double *[Nstate];
+        for (int i = 0; i < Nstate; i++) { ptru[i] = new double[Nstate]; }
+
+        ptrinvu = new double *[Nstate];
+        for (int i = 0; i < Nstate; i++) { ptrinvu[i] = new double[Nstate]; }
+
+        ptrv = new double[Nstate];
+        ptrStationary = new double[Nstate];
+    }
+
+    UniMu = 1;
+    mPow = new double **[UniSubNmax];
+    for (int n = 0; n < UniSubNmax; n++) { mPow[n] = nullptr; }
+
+    flagarray = new bool[Nstate];
+    diagflag = false;
+    statflag = false;
+    for (int i = 0; i < Nstate; i++) { flagarray[i] = false; }
+    powflag = false;
+}
+
+// ---------------------------------------------------------------------------
+//     ~SubMatrix()
[…]

### 2024-07-05 utils/BayesCode/src/lib/TaxonSet.cpp

@@ -1,0 +1,69 @@
+#include "TaxonSet.hpp"
+#include <iostream>
+#include "BiologicalSequences.hpp"
+#include "global/logging.hpp"
+#include "tree/implem.hpp"
+
+using namespace std;
+
+// ---------------------------------------------------------------------------
+// ---------------------------------------------------------------------------
+//     TaxonSet
+// ---------------------------------------------------------------------------
+// ---------------------------------------------------------------------------
+
+TaxonSet::TaxonSet(const std::vector<string> &names) : Ntaxa(names.size()), taxlist(names) {
+    for (int i = 0; i < Ntaxa; i++) {
+        if (taxmap[names[i]] != 0) { FAIL("Found several taxa with same name: {}", names[i]); }
+        taxmap[names[i]] = i + 1;
+    }
+}
+
+std::vector<Tree::NodeIndex> TaxonSet::get_index_table(const Tree *tree) const {
+    // The taxon given the node
+    std::vector<Tree::NodeIndex> ret(tree->nb_nodes(), -1);
+    for (size_t node = 0; node < tree->nb_nodes(); node++) {
+        if (tree->is_leaf(node)) {
+            if (tree->node_name(node).empty()) { FAIL("Leaf has no name") }
+            ret[node] = GetTaxonIndex(tree->node_name(node));
+        }
+    }
+    INFO("Get index table ok");
+    return ret;
+}
+
+std::vector<int> TaxonSet::get_reverse_index_table(const Tree *tree) const {
+    // The node given the taxon
+    std::vector<int> ret(Ntaxa, -1);
+    for (Tree::NodeIndex node = 0; node < Tree::NodeIndex(tree->nb_nodes()); node++) {
+        if (tree->is_leaf(node)) { ret.at(GetTaxonIndex(tree->node_name(node))) = node; }
+    }
+    return ret;
+}
+
+TaxonSet::TaxonSet(const TaxonSet &from)
+    : Ntaxa(from.GetNtaxa()), taxmap(from.taxmap), taxlist(from.taxlist) {}
+
+void TaxonSet::ToStream(ostream &os) const {
+    os << Ntaxa << '\n';
+    for (int i = 0; i < Ntaxa; i++) { os << taxlist[i] << '\n'; }
+}
+
+int TaxonSet::GetTaxonIndexWithIncompleteName(string const &taxname) const {
+    int found = -1;
+    for (int i = 0; i < Ntaxa; i++) {
[…]

### 2024-07-05 utils/BayesCode/src/lib/TaxonSet.hpp

@@ -1,0 +1,79 @@
+#pragma once
+
+#include <iostream>
+#include <map>
+#include <string>
+#include <vector>
+
+class Tree;
+
+/**
+ * \brief A set of taxa (simply, a bi-directional correspondance between taxon
+ * names and integers between 0 and Ntaxa-1)
+ *
+ * TaxonSet implements a globally consistent correspondance of taxon names and
+ * indices between trees and sequence alignments. Each Tree and
+ * SequenceAlignment object has a pointer to a TaxonSet. Typically, a
+ * SequenceAlignment is created by reading from a file; the TaxonSet is created
+ * at that step, by the SequenceAlignment constructor itself. Then, a Tree is
+ * read from file and is registered with the taxon set given by the sequence
+ * alignment. Once this is done, both objects have mutually consistent indexing
+ * systems and can work together (e.g. to calculate a likelihood).
+ *
+ * Of note: some work to do here to ensure proper destruction of TaxonSet (smart
+ * pointers?)
+ */
+
+class TaxonSet {
+  public:
+    //! constructor, based on a vector of taxon names
+    explicit TaxonSet(const std::vector<std::string> &names);
+
+    //! constructor, based on a vector of taxon names
+    TaxonSet(const TaxonSet &from);
+
+    //! default constructor
+    ~TaxonSet() = default;
+
+    //! return number of taxa
+    int GetNtaxa() const { return Ntaxa; }
+
+    //! return taxon name, given the index
+    std::string GetTaxon(int index) const { return taxlist[index]; }
+
+    //! return true if taxon is present, given the name
+    bool TaxonPresent(std::string const &intaxon) const;
+
+    //! return taxon index, given the name
+    int GetTaxonIndex(std::string const &intaxon) const;
+
+    //! return taxon index, given incomplete name (first part)
+    int GetTaxonIndexWithIncompleteName(std::string const &taxname) const;
+
+    //! formatted output to stream
+    void ToStream(std::ostream &os) const;
+
+    //! The taxon given the node
+    std::vector<int> get_index_table(const Tree *tree) const;
+
+    //! The node given the taxon
+    std::vector<int> get_reverse_index_table(const Tree *tree) const;
+
+  private:
+    int Ntaxa;
+    std::map<std::string, int> taxmap;
+    std::vector<std::string> taxlist;
+};
+
[…]

### 2024-07-05 utils/BayesCode/src/lib/jph.hpp

@@ -1,0 +1,20 @@
+#undef NO_ERROR
+#undef ERROR
+#define NO_ERROR 0
+#define ERROR 1
+
+#undef FALSE
+#undef TRUE
+#define FALSE 0
+#define TRUE 1
+
+#define NO 0
+#define YES 1
+
+#define UP 0
+#define DOWN 1
+
+#define ORDERED 0
+#define UNORDERED 1
+
+#define pos(i, j, n) ((i) * (n) + (j))

### 2024-07-05 utils/BayesCode/src/mpi_components/gather.hpp

@@ -1,0 +1,70 @@
+#pragma once
+
+#include "PartitionedBufferManager.hpp"
+#include "partition.hpp"
+
+/*==================================================================================================
+  GatherMaster
+==================================================================================================*/
+class GatherMaster : public Proxy {
+    Partition partition;
+    PartitionedBufferManager manager;
+
+  public:
+    GatherMaster(Partition partition) : partition(partition), manager(partition) {}
+
+    template <class... Variables>
+    void add(Variables&&... vars) {
+        manager.add(std::forward<Variables>(vars)...);
+    }
+
+    void acquire() final {
+        assert(manager.buffer_size() > 0);
+        auto buf = manager.receive_buffer();
+        MPI_Gatherv(NULL, 0, MPI_PACKED, buf, manager.revcounts(), manager.displs(), MPI_PACKED, 0,
+            MPI_COMM_WORLD);
+        manager.receive();
+    }
+};
+
+/*==================================================================================================
+  GatherSlave
+==================================================================================================*/
+class GatherSlave : public Proxy {
+    BufferManager manager;
+    Partition partition;
+
+  public:
+    GatherSlave(Partition partition) : partition(partition) {}
+
+    template <class... Variables>
+    void add(Variables&&... vars) {
+        manager.add(std::forward<Variables>(vars)...);
+    }
+
+    void release() final {
+        assert(manager.buffer_size() > 0);
+        auto buf = manager.send_buffer();
+        MPI_Gatherv(buf, manager.buffer_size(), MPI_PACKED, NULL, NULL, NULL, MPI_PACKED, 0,
+            MPI_COMM_WORLD);
+    }
+};
+
+
+/*==================================================================================================
+  Gather functions
+  Functions that are meant to be called globally and that will create either a master or slave
+  component depending on the process
+==================================================================================================*/
+template <class... Variables>
+std::unique_ptr<Proxy> gather(Partition partition, Variables&&... vars) {
+    if (!MPI::p->rank) {  // master
+        auto gatherer = new GatherMaster(partition);
+        gatherer->add(std::forward<Variables>(vars)...);
+        return std::unique_ptr<Proxy>(dynamic_cast<Proxy*>(gatherer));
+    } else {  // slave
+        auto gatherer = new GatherSlave(partition);
+        gatherer->add(std::forward<Variables>(vars)...);
+        return std::unique_ptr<Proxy>(dynamic_cast<Proxy*>(gatherer));
+    }
+}

### 2024-07-05 utils/BayesCode/src/tree/test.cpp

@@ -1,0 +1,63 @@
+#include <fstream>
+#include <map>
+#include "implem.hpp"
+
+int tree_size(const Tree* tree, Tree::NodeIndex from) {
+    if (tree->is_leaf(from)) { return 1; }
+    int tot = 0;
+    for (auto c : tree->children(from)) { tot += tree_size(tree, c); }
+    return tot;
+}
+
+int main() {
+    // parsing file
+    std::ifstream file("data/besnard/cyp_coding.Chrysithr_root.nhx");
+    NHXParser parser{file};
+
+    // creating topology
+    auto tree = make_from_parser(parser);
+    std::cout << "Nb nodes = " << tree->nb_nodes() << std::endl;
+    std::cout << "root index = " << tree->root() << '\n';
+
+    const Tree* tree2 = tree.get();
+
+    std::cout << "tree size : " << tree_size(tree2, tree2->root()) << '\n';
+
+    for (size_t i = 0; i < tree->nb_nodes(); i++) {
+        std::cout << "[" << i << ": " << tree->node_name(i) << "] ";
+    }
+    std::cout << "\n\n";
+
+    // creating vector of node names
+    auto node_names = node_container_from_parser<std::string>(
+        parser, [](int i, const AnnotatedTree& t) { return t.tag(i, "name"); });
+    for (auto e : node_names) {
+        if (e != "") std::cout << e << " ";
+    }
+    std::cout << "\n\n";
+
+    // slightly shuffled taxa list for example tree
+    std::vector<std::string> taxa = {"Ele.bald", "Ele.bal2", "Ele.bal4", "Ele.vivi", "Ele.vivA",
+        "Ele.bal3", "Ele.viv2", "Ele.fici", "Ele.grac", "Ele.lim2", "Ele.rost", "Ele.limo",
+        "Ele.pal2", "Ele.acut", "Ele.palu", "Ele.gra2", "Ele.lim3", "Ele.geni", "Ele.quan",
+        "Abildgaar", "Bulbostyl", "Bolboscho", "Fuir.abn", "Fuir.umb", "Scho.lac", "Scho.val",
+        "Scho.muc", "Hellmut1", "Actinosch", "Fimb.lit", "Fimb.dic", "Fimb.fe2", "Fimb.li2",
[…]

### 2024-07-05 manuscript/main.tex
Changes suggested by editor in pp of diversifying selection

@@
 Combined with nucleotide diversity and divergence, we estimated $\NI$ and posterior probabilities of diversifying selection under different assumptions for trait heritability as shown in the Table~\ref{table:empirical}.
-Assuming complete heritability, body mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females.
+For body mass, assuming complete heritability led to zero posterior probabilities of diversifying selection for both males and females ($\proba [\NI > 1 ] = 0.0$).
 If we assumed that heritability ($\Heritability$) of body mass was uniformly distributed between 20\% and 40\%~\citep{hu_bringing_2022}, posterior probabilities of diversifying selection became $0.635$ for males and $0.324$ for females.
@@
 We also analyzed a similar dataset for body mass focusing this time only at Primates (Table~\ref{table:empirical}).
-For primates body mass, assuming complete heritability, body mass was found to be under diversifying selection with posterior probabilities of $0.0$ for both males and females, exactly as in the mammal dataset.
+For primates body mass, assuming complete heritability led to zero posterior probabilities of diversifying selection for both males and females, exactly as in the mammal dataset.
 However, we found posterior probabilities of diversifying selection of $1.0$ for males and $0.914$ for females when assuming a uniform distribution for the heritability of body mass between 20\% and 40\%.
